---
title: "Formatting data for use in mvgam"
author: "Nicholas J Clark"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: yes
vignette: >
  %\VignetteIndexEntry{Formatting data for use in mvgam}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
params:
  EVAL: !r identical(tolower(Sys.getenv("NOT_CRAN")), "true")
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  eval = if (isTRUE(exists("params"))) params$EVAL else FALSE
)
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  dpi = 100,
  fig.asp = 0.7,
  fig.width = 6,
  out.width = "70%",
  fig.align = "center"
)
library(mvgam)
library(ggplot2)
theme_set(theme_bw(base_size = 12, base_family = "serif"))
```

`mvgam` expects data in a specific shape. The most common cause
of a failed fit is a data frame that does not quite meet that
shape: a `series` column that is character instead of a factor,
gaps in the `time` index, missing values where the model cannot
accept them, or a response that violates the chosen family's
support. This vignette walks through the requirements and shows
how `check_mvgam_data()` (an alias for `mvgam_data()`) raises
the same friendly errors that `mvgam()` does at fit time,
without paying for a Stan compile to find out. By the end, you
should be able to take an arbitrary long-format data frame and
know whether `mvgam()` will accept it.

## The long-format requirement

Every observation needs its own row, indexed by a time column
and (when multiple series are present) by a series column. The
defaults are `time` and `series`, but every trend constructor
takes bare-name `time = ` and `series = ` arguments that override
these defaults, and `jsdgam()` exposes `unit = ` and `species = `
for the same purpose. So if your data carry the timestamp under
`epiWeek` and the series identifier under `plotID`, calling
`AR(time = epiWeek, series = plotID)` (or
`jsdgam(unit = epiWeek, species = plotID, ...)`) keeps your own
column names intact. The package's in-built `birdsong` dataset
follows the defaults: four species counted weekly across 80
weeks.

```{r birdsong-head}
data(birdsong)
str(birdsong)
head(birdsong, 4)
```

There is one row per `(series, time)` cell, with `series` as a
`factor` and `time` as an integer index of week. The response
`y` carries the weekly count. A multi-series fit needs the four
columns: a series identifier (`series` by default), a time
index (`time` by default), the response (whatever the formula
names), and any covariates the formula references. The
`species`, `week` and `week_in_year` columns in `birdsong` are
not required by `mvgam` and would be ignored, but they would
work as the series and time columns if a trend constructor were
called with `RW(time = week, series = species)`.

The same long format is what `sim_mvgam()` returns by default
when generating fixtures:

```{r sim-mvgam-head}
set.seed(1)
simdat <- sim_mvgam(n_series = 3L, n_timepoints = 24L,
                     prop_missing = 0.2)
head(simdat$data_train, 6)
```

`sim_mvgam()` injects 20% missing-at-random `NA`s into the
response to mirror the realistic case where surveys occasionally
miss observations. Missing values in the response are fine;
`mvgam()` skips them in the likelihood and still infers the
latent trajectory at those timepoints.

## The `series` factor

Whatever column carries the series identifier (`series` by
default, or any column named through `AR(series = ...)` /
`jsdgam(species = ...)`) must be a `factor`, and its levels
must match the unique values present in the data. The two
failure modes are *character instead of factor* and *extra
levels that no row references*. The first is easy to catch by
class; the second is sneaky because the data look fine until
`mvgam()` tries to index a series that isn't there.

```{r series-class}
class(birdsong$series)
levels(birdsong$series)
setdiff(levels(birdsong$series), unique(birdsong$series))
```

An empty `setdiff()` confirms the levels and the realised values
agree. When they disagree, `droplevels()` is the fix:

```{r dropping-levels}
bad_levels <- birdsong
bad_levels$series <- factor(
  bad_levels$series,
  levels = c(levels(bad_levels$series), "phantom_species")
)
setdiff(levels(bad_levels$series), unique(bad_levels$series))

# Fix by dropping unused levels.
good_levels <- bad_levels
good_levels$series <- droplevels(good_levels$series)
levels(good_levels$series)
```

You can technically supply data with no `series` column, in
which case `mvgam()` assumes a single time series and assigns
the default name. For models with shared dynamics across
non-temporal groupings, `ZMVN()` and `jsdgam()` accept the
`series` factor as a grouping variable rather than a time index.
See `?ZMVN` and `?jsdgam` for those entry points.

## The `time` variable

The time column must be numeric or integer (the default name is
`time`; any other column name passed through `AR(time = ...)`,
`jsdgam(unit = ...)`, etc. is honoured). Trend constructors that
assume regular sampling (`RW()`, `AR()`, `VAR()`, `PW()`) further
require the spacing between consecutive time values to be
constant within each series, although the response can be
missing at any timepoint. Smaller time values are interpreted as
*earlier* observations.

```{r time-range}
range(birdsong$time)
tapply(birdsong$time, birdsong$series, function(t) diff(range(t)) + 1L)
```

Each species in `birdsong` covers the same 80 timepoints. The
expansion-by-`expand.grid()` pattern is the canonical fix when
some `(series, time)` combinations are missing entirely:

```{r expand-grid-fix}
sparse <- birdsong[!(birdsong$series == "warbler" &
                       birdsong$time %in% c(5L, 6L, 7L)), ]
nrow(sparse)
filled <- merge(
  expand.grid(series = levels(sparse$series),
              time   = sort(unique(sparse$time))),
  sparse, by = c("series", "time"), all.x = TRUE
)
nrow(filled)
sum(is.na(filled$y))
```

The expansion restores the missing `(series, time)` rows with
`NA` responses, which `mvgam()` handles correctly.

### Irregular sampling

If the spacing between timepoints is not constant and cannot be
sensibly regularised, the AR-family constructors will reject the
data. The escape hatch is `CAR()`, the continuous-time AR(1)
process, which carries the time gap into the lag coefficient
itself. See `?CAR` for the formulation and a worked example. For
non-temporal data where the goal is purely residual covariance
among series, `ZMVN()` ignores the `time` axis entirely.

## The response and family compatibility

Every observation family in `mvgam` carries a support that the
response must respect: `poisson()` and `negbinomial()` need
non-negative integers; `Beta()` (and `Beta_binomial()`'s second
arm) need values strictly in $(0, 1)$; `Gamma()` and
`lognormal()` need strictly positive reals; `bernoulli()` needs
0/1 values. Most regression packages will silently fit something
when the response doesn't match and leave the user to puzzle
over a degenerate posterior. `mvgam()` raises an error early
instead.

```{r bad-family, error = TRUE}
bad_family <- data.frame(
  y      = rnorm(20),
  time   = 1:20,
  series = factor("series_1")
)
mvgam(y ~ 1, data = bad_family, family = Beta(),
      chains = 1L, samples = 100L, silent = 2L)
```

The error states the actual problem (Beta needs $(0, 1)$, the
response has negative values) rather than a numerical collapse
mid-sampling. The same check runs inside the data inspector
below, which is the recommended pre-flight call.

## `check_mvgam_data()`: pre-flight validation

`check_mvgam_data()` (the same function as `mvgam_data()`,
exposed under a more discoverable name) runs every validator
`mvgam()` runs at fit time. It checks required columns, factor
level alignment, regular time spacing for trends that need it,
response-family compatibility, and any covariates the user
plans to reference. When `plot = TRUE` it also draws the
standard EDA view (time-series, ACF, histogram, ECDF) so the
shape of the data is visible at a glance.

```{r check-good, fig.asp = 0.85, out.width = "85%"}
check_mvgam_data(
  data    = birdsong,
  family  = poisson(),
  series  = 1L
)
```

The print line confirms the family, the number of series, the
time range and the row count. The 4-panel patchwork shows the
time-series, ACF, ECDF, and histogram for the first species,
which is the right level of detail when checking a single
series; passing `series = "all"` instead returns one faceted
time-series plot covering every species. The same call returns
a structured `mvgam_data` object invisibly so it can be assigned
and re-used if desired.

When the data are *not* valid, the same function raises an
informative error. Three common failure modes:

```{r check-missing-cov, error = TRUE}
miss_cov <- birdsong
miss_cov$temp <- rnorm(nrow(miss_cov))
miss_cov$temp[5L] <- NA
check_mvgam_data(miss_cov, family = poisson(), plot = FALSE)
```

The check passes because no formula referenced the missing
column. The same call with `mvgam(y ~ temp, ...)` would catch
it; covariate-completeness checks fire when the formula is
parsed.

```{r check-missing-times, error = TRUE}
gap <- birdsong[birdsong$time != 5L, ]
check_mvgam_data(gap, family = poisson(), trend_model = AR(),
                  plot = FALSE)
```

The `trend_model` argument switches on the regular-spacing
check, which catches missing timepoints that an AR fit cannot
absorb. The error message tells the user which series is broken
and suggests either `expand.grid()` for filling gaps or `CAR()`
for accepting irregular spacing as a modelling choice.

```{r check-bad-response, error = TRUE}
bad_resp <- birdsong
bad_resp$y <- bad_resp$y + 0.5  # introduce non-integers
check_mvgam_data(bad_resp, family = poisson(), plot = FALSE)
```

The response-shape check rejects the data because Poisson
requires non-negative integers and `+ 0.5` makes them all
non-integer. Switching to `Gamma()` or `lognormal()` would
require the response to be strictly positive instead. Each
family's support is documented under `?mvgam_families`.

## Newdata for forecasts: `validate_newdata()`

Forecasting requires a `newdata` frame in the same shape as the
training data, with the same factor levels and covariate
columns. `validate_newdata()` runs the structural check before
`predict()` / `forecast()` are called, which is faster than
finding out at the back end of an expensive prediction
pipeline.

```{r validate-newdata, error = TRUE}
future_df <- data.frame(
  series = factor(
    rep(levels(birdsong$series), each = 4L),
    levels = levels(birdsong$series)
  ),
  time   = rep(81L:84L, times = 4L)
)
validate_newdata(future_df, training_data = birdsong)
```

A clean newdata frame returns invisibly; a misaligned one (wrong
factor levels, missing time column, missing covariate) raises
the same error the prediction call would, with the same hint
about how to fix it. This is the right call to make in any
forecasting pipeline before the first `predict()`.

## Closure-unit data shape

Closure-unit families (`occ()`, `nmix()` and its variants)
require an extra `visit` column and a per-unit `cap` column on
top of the long format above. Each row is a single visit to a
sampling unit, not a single time observation. The
`sim_closure_unit_data()` helper builds this layout directly:

```{r closure-shape}
occ_dat <- sim_closure_unit_data(
  family    = occ(),
  type      = 1L,
  n_species = 2L,
  n_sites   = 12L,
  n_visits  = 3L,
  seed      = 1L
)
str(occ_dat$data_train)
head(occ_dat$data_train, 6)
```

`series` carries the species identifier, `time` carries the
closure-unit identifier (the site index for a single-season
design; the season index when `multi_season = TRUE`), `visit`
indexes the within-unit repeated observations, `y` is the
binary detection record (for occupancy) or count (for N-mixture),
and `cap` is the per-unit upper bound used by the
marginalisation. The state-side and detection-side covariates
come along as additional columns referenced by the
`bf(y ~ ..., p ~ ...)` formula.

`check_mvgam_data()` validates closure-unit shape when an `occ()`
or `nmix()` family is passed:

```{r check-closure}
check_mvgam_data(occ_dat$data_train, family = occ(),
                  plot = FALSE)
```

The same family-aware validator that runs inside `mvgam()` and
`jsdgam()` confirms the data are ready: closure-unit rows count
correctly, the `cap` column is consistent across visits within
a unit and the factor levels line up.

The closure-unit grain is covered in more detail in the
[N-mixture vignette](https://nicholasjclark.github.io/mvgam/articles/nmixtures.html);
this section exists so you can recognise the layout when you
encounter it and not confuse it with the per-time-step long
format.

## Custom column names

The default `series` and `time` column names work for the
majority of fits, but they are not required. Every trend
constructor accepts bare-name `time = ` and `series = ` arguments
that point at any column in the data; `jsdgam()` exposes the same
through `unit = ` and `species = `. A `birdsong`-style fit can
run directly off the `week` and `species` columns without
renaming:

```{r custom-names, results = "hide", eval = FALSE}
# Equivalent to a `time` / `series` setup, no renaming required:
mvgam(
  y ~ s(week_in_year, bs = "cc", k = 10),
  trend_formula = ~ AR(time = week, series = species),
  data    = birdsong,
  family  = poisson(),
  chains  = 2L, silent = 2L
)
```

The trend constructor's `time` / `series` arguments override the
defaults consistently across `RW()`, `AR()`, `VAR()`, `CAR()`,
`PW()` and `ZMVN()`. For `jsdgam()` fits the equivalent is
`unit = ` on the top-level call. The closure-unit families
(`occ()`, `nmix()`) take a different route. They inherit the
unit identifier from the package validator and pass it through
the detection arm of `bf()`.

## A pre-fit checklist

Before any `mvgam()` call, the checklist that catches the most
common failures is short:

1. The series column is a `factor` with no orphan levels
   (`setdiff(levels(data$series), unique(data$series))` is
   empty). If you point a trend constructor at a different
   column with `AR(series = my_col)`, the same rule applies to
   `my_col`.
2. The time column is `numeric` or `integer` and (for AR-family
   trends) the spacing within each series is constant.
3. The response respects the chosen family's support
   (`?mvgam_families`).
4. Every covariate the formula references has no `NA`s.
5. `check_mvgam_data(data, family = ..., trend_model = ...)`
   runs without error.

The fifth item subsumes the first four; the others remain useful
when triaging an error message before reaching for the
inspector.

## Where to next

- `?mvgam_data` / `?check_mvgam_data` for the inspector's full
  argument list.
- `?validate_newdata` for the newdata pre-check used in
  forecasting pipelines.
- `?mvgam_families` for the per-family response-shape rules.
- `?CAR` for the irregular-sampling escape hatch.
- The
  [overview vignette](https://nicholasjclark.github.io/mvgam/articles/mvgam_overview.html)
  for the model surface that consumes the data validated here.
- The
  [N-mixture vignette](https://nicholasjclark.github.io/mvgam/articles/nmixtures.html)
  for the closure-unit grain in worked detail.
