---
title: "Formatting data for mvgam"
author: "Nicholas J Clark"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: yes
vignette: >
  %\VignetteIndexEntry{Formatting data for mvgam}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
params:
  EVAL: !r identical(tolower(Sys.getenv("NOT_CRAN")), "true")
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  eval = if (isTRUE(exists("params"))) params$EVAL else FALSE
)
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  dpi = 100,
  fig.asp = 0.7,
  fig.width = 6,
  out.width = "70%",
  fig.align = "center"
)
library(mvgam)
library(ggplot2)
theme_set(mvgam_theme(base_size = 12, base_family = "serif"))
```

`mvgam()` takes data in long format, one row per series and time.
Most failed fits come from a frame that misses that shape: a
`series` column stored as character, gaps in the `time` index, a
column the formula names but the data lack, missing covariate
values or a response outside the family's support. This vignette
sets out each requirement and shows how `mvgam_data()` checks a
frame against them before any Stan code is compiled.

## The long-format requirement

Each observation takes its own row, indexed by a time column and,
for a model with a latent trend, a series column. The default
names are `time` and `series`. Every trend constructor takes
`time` and `series` arguments naming other columns, as in
`AR(time = epiWeek, series = plotID)`, and `jsdgam()` takes `unit`
and `species`. The `birdsong` data that ship with mvgam use the
defaults, with weekly counts of four species over 80 weeks:

```{r birdsong-head}
data(birdsong)
str(birdsong)
head(birdsong, 4)
```

Each `(series, time)` cell has one row, `series` is a factor and
`time` an integer week. A fit uses the series and time columns,
the response its formula names and any covariates the formulas
refer to. The `species` and `week` columns duplicate `series` and
`time`, and `RW(time = week, series = species)` would use them
instead.

`sim_mvgam()` returns data in the same format:

```{r sim-mvgam-head}
set.seed(1)
simdat <- sim_mvgam(n_series = 3L, n_timepoints = 24L,
                     prop_missing = 0.2)
head(simdat$data_train, 6)
```

`prop_missing = 0.2` sets a fifth of the responses to `NA`. A
missing response is allowed: `mvgam()` leaves it out of the
likelihood and still estimates the latent state at that time.

## The `series` factor

A model with a latent trend places every row on one series. Its data
needs a series column: `series` by default, or any column named
through `AR(series = ...)` or `jsdgam(species = ...)`. That column
must be a `factor`, and its levels fix the order of the series in
every summary, plot and forecast. A `character` column looks the
same in `head()` and fails the check:

```{r series-class}
class(birdsong$series)
levels(birdsong$series)
```

```{r series-character-fail, error = TRUE}
char_series <- birdsong
char_series$series <- as.character(char_series$series)
mvgam_data(char_series, family = poisson(),
                 trend_formula = ~ AR(), plot = FALSE)
```

`factor()` converts the column. A single time series needs the
column too, holding one level:

```{r single-series}
one <- birdsong[birdsong$series == levels(birdsong$series)[1L], ]
one$series <- factor("series_1")
levels(one$series)
```

A model without a trend has no series axis and needs no series
column.

`mvgam()` drops a level that no row holds, and `droplevels()` does
the same before fitting:

```{r dropping-levels}
extra_level <- birdsong
extra_level$series <- factor(
  extra_level$series,
  levels = c(levels(extra_level$series), "phantom_species")
)
setdiff(levels(extra_level$series), unique(extra_level$series))
extra_level$series <- droplevels(extra_level$series)
levels(extra_level$series)
```

## The `time` variable

The time column must be numeric. `RW()`, `AR()`, `VAR()` and
`PW()` advance the latent state one time step at a time. They need
evenly spaced times within each series, although the response may
be missing at any time.

```{r time-range}
range(birdsong$time)
tapply(birdsong$time, birdsong$series, function(t) diff(range(t)) + 1L)
```

Each species covers the same 80 weeks. Where some
`(series, time)` rows are absent altogether, `expand.grid()`
builds the full grid and a merge restores the rows:

```{r expand-grid-fix}
sparse <- birdsong[!(birdsong$series == "warbler" &
                       birdsong$time %in% c(5L, 6L, 7L)), ]
nrow(sparse)
filled <- merge(
  expand.grid(series = levels(sparse$series),
              time   = sort(unique(sparse$time))),
  sparse, by = c("series", "time"), all.x = TRUE
)
nrow(filled)
sum(is.na(filled$y))
```

The merge gives each restored row an `NA` response.

### Irregular sampling

Where observations are irregular by design, `CAR()` fits a
continuous-time AR(1) process whose autocorrelation decays with
the time elapsed between observations. `ZMVN()` has no time
dynamics and accepts any spacing. `?CAR` gives a worked example.

## The response and family compatibility

Each family has a support the response must lie in: non-negative
integers for `poisson()` and `negbinomial()`, $(0, 1)$ for
`Beta()`, positive reals for `Gamma()` and `lognormal()` and 0
or 1 for `bernoulli()`. `mvgam()` checks the response against the
support before compiling:

```{r bad-family, error = TRUE}
bad_family <- data.frame(
  y      = rnorm(20),
  time   = 1:20,
  series = factor("series_1")
)
mvgam(y ~ 1, data = bad_family, family = Beta(),
      chains = 1L, iter = 1100L, warmup = 1000L, silent = 2L)
```

The error names the support, the observed range and the first
offending row.

## Columns the formulas use

A formula names more columns than its covariates. brms builds the
model frame from the variable inside an `offset()`, the grouping
factor of a random effect and the columns of addition terms such
as `trials()`, `weights()` and `cens()`, and each of them must be
in the data. To show the checks, the chunk below adds survey
minutes as an exposure and an observer as a grouping factor to
`birdsong`, and then drops the observer column:

```{r check-absent-column, error = TRUE}
surveys <- birdsong
surveys$minutes <- rep(c(30, 45, 60, 45), length.out = nrow(surveys))
surveys$observer <- factor(rep(c("a", "b", "c"),
                               length.out = nrow(surveys)))
no_observer <- surveys[, names(surveys) != "observer"]
mvgam_data(no_observer,
           formula = y ~ offset(log(minutes)) + (1 | observer),
           family = poisson(), plot = FALSE)
```

`mvgam()` and `stancode()` refuse the same frame before brms
builds a model. A name held in `data2`, such as a matrix
covariate, counts as present when `data2` is passed to
`mvgam_data()`.

A binomial response counts successes out of the number of trials
that `trials()` names. A count above its trials lies outside the
family's support. Here each week records how many of 12 listening
points detected the species, and one row claims 14:

```{r check-trials, error = TRUE}
set.seed(2)
surveys$points <- 12L
surveys$detected <- rbinom(nrow(surveys), size = 12L, prob = 0.3)
surveys$detected[10L] <- 14L
mvgam_data(surveys, formula = detected | trials(points) ~ 1,
           family = binomial(), plot = FALSE)
```

## Grouped trends

A hierarchical trend shares its dynamics among the series of a
group. The constructor's `gr` argument names the column holding
each series' group. Its `subgr` argument names the column holding
the series within that group. The two columns together identify each
series, and the data need no `series` column. Here the four
`birdsong` species are split into two habitats:

```{r check-grouped, message = TRUE}
grouped <- birdsong[, names(birdsong) != "series"]
grouped$habitat <- factor(
  ifelse(grouped$species %in% c("warbler", "thrush"),
         "forest", "wetland")
)
mvgam_data(grouped, family = poisson(),
           trend_formula = ~ AR(gr = habitat, subgr = species),
           plot = FALSE)
```

Every group must hold the same number of series, and each series
must stay in one group. The check refuses an unbalanced design:

```{r check-grouped-uneven, error = TRUE}
uneven <- grouped
uneven$habitat[uneven$species == "wren"] <- "forest"
mvgam_data(uneven, family = poisson(),
           trend_formula = ~ AR(gr = habitat, subgr = species),
           plot = FALSE)
```

## `mvgam_data()`: pre-flight validation

`mvgam_data()` runs the checks `mvgam()` runs at fit time: the
columns the formulas and the trend constructor name, the series
factor, even spacing for a trend that needs it, the response
support and missing values in the model's columns. With
`plot = TRUE` it also draws the series, its histogram, its
autocorrelation and its empirical distribution:

```{r check-good, fig.asp = 0.85, out.width = "85%", message = TRUE}
mvgam_data(
  data    = birdsong,
  family  = poisson(),
  series  = 1L
)
```

The check reports the family, the number of series, the time
range and the number of rows. `series = 1L` draws the first
species, and `series = "all"` draws every species in one faceted
plot. The call returns an `mvgam_data` object invisibly.

Data that fail a check raise an error. Three common cases:

```{r check-bad-family, error = TRUE}
bad_y <- birdsong
bad_y$y <- bad_y$y - 0.5  # introduce negative non-integers
mvgam_data(bad_y, family = poisson(), plot = FALSE)
```

Subtracting 0.5 left the counts fractional and some negative.
`?brmsfamily` gives the support of each brms family, and
`?tweedie`, `?beta_nb`, `?com_binomial`, `?nmix` and `?occ` give
those of the families mvgam adds.

```{r check-missing-times, error = TRUE}
gap <- birdsong[birdsong$time != 5L, ]
mvgam_data(gap, family = poisson(), trend_formula = ~ AR(),
                  plot = FALSE)
```

`trend_formula = ~ AR()` turns on the spacing check, which finds
the removed week. Without a `trend_formula`, `mvgam_data()` checks
the data for a model with no trend, which needs neither even
spacing nor a series column.

A missing response is allowed, but a missing covariate value is
not: every row needs its covariates, including a row whose
response is missing. `mvgam_data()` checks the columns that the
`formula` and `trend_formula` it is given refer to:

```{r check-covariate-na, error = TRUE}
miss_cov <- birdsong
miss_cov$temp <- rnorm(nrow(miss_cov))
miss_cov$temp[5L] <- NA
mvgam_data(miss_cov, formula = y ~ temp,
                  family = poisson(), plot = FALSE)
```

`mvgam()` runs the same check before compiling. Without a
`formula`, `mvgam_data()` cannot tell which columns the model
uses, and it does not check `temp`.

## Newdata for forecasts: `check_newdata()`

`forecast()` and `predict()` take a `newdata` frame holding the
future times and every covariate the model uses.
`check_newdata(newdata, data, trend_formula)` compares its time and
series columns with the training data before any prediction runs.
Like `mvgam()`, it takes the column names from the trend
constructor. It also coerces the series column to the training
levels:

```{r check-newdata-ok}
future_df <- data.frame(
  series = factor(rep(levels(birdsong$series), each = 4L)),
  time   = rep(81L:84L, times = 4L)
)
str(check_newdata(future_df, birdsong, trend_formula = ~ AR()))
```

A series the model never saw has no fitted trend to forecast from,
and the check refuses it:

```{r check-newdata-bad, error = TRUE}
bad_future <- data.frame(
  series = factor("phantom_species"),
  time   = 81L
)
check_newdata(bad_future, birdsong, trend_formula = ~ AR())
```

`mvgam(newdata = )` runs the same check on held-out data, and also
refuses a frame missing a column the formulas use.

## Closure-unit data shape

Closure-unit families (`occ()`, `nmix()` and its variants) add a
`visit` column to the long format above. Each row records one
visit to a sampling unit. The `sim_closure_unit_data()` helper
builds this layout:

```{r closure-shape}
occ_dat <- sim_closure_unit_data(
  family    = occ(),
  type      = 1L,
  n_species = 2L,
  n_sites   = 12L,
  n_visits  = 3L,
  seed      = 1L
)
head(occ_dat$data_train, 6)
```

`series` holds the species and `time` the closure unit, which is
the site here and the season when `multi_season = TRUE`. `visit`
numbers the repeat visits within a unit and `y` records the
detection. The covariates of the state and detection formulas,
`bf(y ~ ..., p ~ ...)`, are further columns. `mvgam_data()` checks
this layout when given `occ()` or `nmix()`:

```{r check-closure, message = TRUE}
mvgam_data(occ_dat$data_train, family = occ(),
           formula = bf(y ~ env, p ~ tod_c), plot = FALSE)
```

`nmix()` models counts, and its `cap` column sets the upper bound
on the latent abundance of each unit. The bound must be constant
within a unit and at least as large as every count observed there:

```{r check-nmix-cap, error = TRUE}
nmix_dat <- sim_closure_unit_data(
  family = nmix(), n_species = 2L, n_sites = 12L, n_visits = 3L,
  seed = 1L
)$data_train
nmix_dat$cap <- 2L
mvgam_data(nmix_dat, family = nmix(),
           formula = bf(y ~ env, p ~ tod_c), plot = FALSE)
```

The [N-mixture article](https://nicholasjclark.github.io/mvgam/articles/nmix.html)
fits these data.

## Joint species distribution models

`jsdgam()` takes the same long format, with one row per sampling
unit and species. Its `unit` argument names the column of units.
Its `species` argument names the column of species, a factor. The units
carry no time order. `jsdgam()` fits a `ZMVN()` factor model over
them. Given that trend, `mvgam_data()` runs the checks `jsdgam()`
runs:

```{r check-jsdm, message = TRUE}
set.seed(3)
jsdm_dat <- expand.grid(
  site = 1:20,
  species = factor(c("sp_a", "sp_b", "sp_c"))
)
jsdm_dat$elev <- rnorm(20)[jsdm_dat$site]
jsdm_dat$present <- rbinom(nrow(jsdm_dat), size = 1L, prob = 0.4)
mvgam_data(jsdm_dat, y = "present", formula = present ~ elev,
           family = bernoulli(),
           trend_formula = ~ ZMVN(time = site, series = species),
           plot = FALSE)
```

## Custom column names

The `time` and `series` arguments of a trend constructor name other
columns, and a fit to `birdsong` can use `week` and `species`
directly:

```{r custom-names, results = "hide"}
mvgam(
  y ~ s(week_in_year, bs = "cc", k = 10),
  trend_formula = ~ AR(time = week, series = species),
  data    = birdsong,
  family  = poisson(),
  chains  = 2L, silent = 2L
)
```

`RW()`, `AR()`, `VAR()`, `CAR()`, `PW()` and `ZMVN()` all take
these arguments.

## A pre-fit checklist

Before calling `mvgam()`, check that:

1. The series column is a factor. A hierarchical trend takes its
   series from the `gr` and `subgr` columns instead.
2. The time column is numeric, and evenly spaced within each
   series for `RW()`, `AR()`, `VAR()` and `PW()`.
3. Every column the formulas use is in the data: covariates,
   offsets, random-effect grouping factors and addition terms.
4. No such column has missing values.
5. The response lies in the family's support.

`mvgam_data(data, family = ..., formula = ..., trend_formula = ...)`
checks all five.

## Where to next

- `?mvgam_data` for the inspector's full
  argument list.
- `?check_newdata` for the check on forecast frames.
- `?brmsfamily` for the brms family library, plus `?tweedie`,
  `?beta_nb`, `?com_binomial`, `?nmix` and `?occ` for the
  response-shape rules of the families mvgam adds.
- `?CAR` for irregularly spaced times.
- The
  [overview vignette](https://nicholasjclark.github.io/mvgam/articles/mvgam_overview.html)
  for the model surface that consumes the data validated here.
- The
  [N-mixture article](https://nicholasjclark.github.io/mvgam/articles/nmix.html)
  for the closure-unit grain in worked detail.
- The
  [dynamic factor vignette](https://nicholasjclark.github.io/mvgam/articles/dfm.html)
  for the `trend_map` structures that decide which series share a
  latent path once the data are in shape.
