---
title: "Dynamic factor models"
author: "Nicholas J Clark"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: yes
vignette: >
  %\VignetteIndexEntry{Dynamic factor models}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
params:
  EVAL: !r identical(tolower(Sys.getenv("NOT_CRAN")), "true")
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  eval = if (isTRUE(exists("params"))) params$EVAL else FALSE
)
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  dpi = 100,
  fig.asp = 0.7,
  fig.width = 6,
  out.width = "70%",
  fig.align = "center"
)
library(mvgam)
library(ggplot2)
theme_set(theme_bw(base_size = 12, base_family = "serif"))
```

Multivariate time series often move together for reasons the
covariates do not record. Sites in the same habitat share weather
signals; species in the same guild ride shared resource pulses.
Dynamic factor models compress that common variation into a small
number of latent processes and pin each observed series to a
weighted sum of those processes (Zuur et al.
[2003](https://doi.org/10.1002/env.611); Holmes et al.
[2012](https://doi.org/10.32614/RJ-2012-002)). mvgam's `trend_map`
argument is how you supply the fixed sharing structure for a
state-space fit: you say which series belong to which trend, and the
posterior latent path is shared verbatim across the group. This
vignette recovers a small AR(1) dynamic factor model against a
simulated truth, hands the same problem to a model that learns the
loadings, and hands it to a deliberately misspecified model.
mvgam's score functions quantify how much the sharing choice
matters for forecasts on held-out data.

## Dynamic factor models

For $N_{\text{series}}$ observed series indexed by $j$ and $N_{lv}$
latent trends indexed by $k$, the dynamic factor model writes each
series' link-scale mean as a linear combination of the trends'
current states:

\begin{align*}
y_{j, t}      & \sim F(\mu_{j, t}, \boldsymbol{\theta}) \\
g(\mu_{j, t}) & = \beta_0 + \beta^\top \mathbf{x}_{j, t}
                       + \sum_{k = 1}^{N_{lv}} Z_{j, k} \, z_{k, t} \\
z_{k, t}      & \sim \text{AR}(1) \quad\text{or any other trend model.}
\end{align*}

$Z$ is an $N_{\text{series}} \times N_{lv}$ loading matrix. Row $j$
of $Z$ says which trends drive series $j$ and with what weight; the
posterior latent path $z_{k, 1:T}$ is shared verbatim across every
series whose row of $Z$ places non-zero weight on trend $k$. When
$N_{lv} < N_{\text{series}}$ the model is a genuine dimension
reduction: many series ride a small number of latent paths, and the
posterior on those paths pools information across every contributing
series.

`trend_map` is how you supply $Z$. mvgam accepts three input shapes:

- A character shorthand: `"identity"` fits one independent trend per
  series ($Z = I$); `"shared"` gives every series the same latent
  path ($Z$ is a column of ones).
- A `data.frame` with columns `series` and `trend` for sparse binary
  sharing. Each row says "series X uses trend Y"; two series pointing
  at the same trend integer are fixed to share.
- A numeric matrix of dimension $N_{\text{series}} \times N_{lv}$.
  Finite entries are treated as fixed loadings and preserved exactly
  in the posterior; `NA` entries mark loadings that should be sampled
  under a partial-Z parameterisation.

Supplying any fixed entry via `trend_map` bypasses the post-hoc QR
identification mvgam applies to free factor models (Heaps & Jermyn
[2024](https://doi.org/10.1007/s11222-024-10454-0)), so the encoded
structure is preserved verbatim in the posterior.

## A simulated example

We simulate six series driven by three latent AR(1) trends, with
one lone series that does not share its dynamics with any other.
Trend 1 has long memory ($\phi_1 = 0.85$) and drives series 1-3;
trend 2 has short memory ($\phi_2 = 0.15$) and drives series 4-5.
Trend 3 sits between them ($\phi_3 = 0.55$) with a larger
innovation SD ($\sigma_3 = 0.8$) and drives series 6 alone. The
last 15 timepoints are held out for forecast scoring.

```{r simulate}
set.seed(3)

# Dimensions, true AR(1) parameters and time split
n_time <- 80
n_series <- 6
n_lv <- 3
n_train <- 65
phi_true <- c(0.85, 0.15, 0.55)
sig_true <- c(0.5, 0.35, 0.8)

# Simulate the three latent trends; first state drawn from the
# AR(1) stationary marginal so the process starts in equilibrium
lv <- matrix(NA, n_time, n_lv)
lv[1, ] <- rnorm(n_lv, 0, sig_true / sqrt(1 - phi_true^2))
for (t in 2:n_time) {
  lv[t, ] <- phi_true * lv[t - 1L, ] + rnorm(n_lv, 0, sig_true)
}

# Fixed loadings: series 1-3 -> trend 1, series 4-5 -> trend 2,
# series 6 alone on trend 3
Z_true <- matrix(0, nrow = n_series, ncol = n_lv)
Z_true[1, 1] <- 1
Z_true[2, 1] <- 1
Z_true[3, 1] <- 1
Z_true[4, 2] <- 1
Z_true[5, 2] <- 1
Z_true[6, 3] <- 1

# Observations: series-level mean plus Gaussian noise (sigma = 0.2)
mu <- lv %*% t(Z_true)
y_mat <- mu + matrix(rnorm(n_time * n_series, 0, 0.2), n_time, n_series)

# Long-format data frame + train / test split
sim_data <- data.frame(
  time = rep(seq_len(n_time), n_series),
  series = factor(
    rep(paste0("s", seq_len(n_series)), each = n_time),
    levels = paste0("s", seq_len(n_series))
  ),
  y = as.numeric(y_mat)
)
train <- sim_data[sim_data$time <= n_train, ]
test  <- sim_data[sim_data$time >  n_train, ]
```

`mvgam_data()` renders the observed series with the training slice
in colour and any newdata slice overlaid in the same panel, so we
can see the sharing structure and the held-out horizon at once.

```{r plot-observed, fig.cap = "Six simulated series with the last 15 timepoints held out for forecast scoring. Series 1-3 track a long-memory trend, series 4-5 track a short-memory trend, and series 6 rides its own medium-memory trend with a larger amplitude."}
mvgam_data(train, y = "y", family = gaussian(), newdata = test)
```

## Fitting with a `trend_map`

We pass the sharing structure as a two-column `data.frame`. The
`series` column lines up with the observed factor's levels and the
`trend` column carries the integer trend index; two rows pointing at
the same integer share a latent path.

```{r trend-map}
tm_correct <- data.frame(
  series = paste0("s", seq_len(n_series)),
  trend = c(1L, 1L, 1L, 2L, 2L, 3L)
)
tm_correct
```

The observation side is a single population intercept (`y ~ 1`);
the trend side is an AR(1) process, applied to each of the three
trends the `trend_map` implies. Passing `newdata = test` at fit time
stores the test slice with the fitted object so later `forecast()`
calls can use it without re-supplying.

```{r fit-correct, results = "hide"}
fit_correct <- mvgam(
  formula = y ~ 1,
  trend_formula = ~ AR(p = 1),
  trend_map = tm_correct,
  data = train,
  newdata = test,
  family = gaussian(),
  chains = 2L,
  iter = 1000L,
  warmup = 500L,
  silent = 2L,
  refresh = 0
)
```

```{r recovery-values, echo = FALSE}
post_sum <- posterior::summarise_draws(as_draws_df(fit_correct), "mean")
grab <- function(pat) {
  round(post_sum$mean[grepl(pat, post_sum$variable)], 2)
}
ar1_est <- grab("^ar1_trend\\[")
sig_est <- grab("^sigma_trend\\[")
obs_sd <- grab("^sigma$")
```

`summary()` reports the AR coefficient and innovation SD for each of
the three trends alongside the observation intercept and residual
SD. Every simulated value falls inside its 95% credible interval:
`ar1_trend` recovers to (`r paste(ar1_est, collapse = ", ")`) against
a truth of (0.85, 0.15, 0.55), and `sigma_trend` to
(`r paste(sig_est, collapse = ", ")`) against (0.5, 0.35, 0.8). The
residual observation SD lands at `r obs_sd`, the value used to
generate the data.

```{r summary}
summary(fit_correct)
```

## Recovering the latent trends

`hindcast(fit, type = "trend")` returns the latent-state posterior
on its own scale, and `plot()` on that object renders one facet per
series. Because we fixed series 1-3 to share trend 1, series 4-5 to
share trend 2, and series 6 to ride trend 3, the panels within each
sharing group carry identical draws:

```{r plot-trend-hindcast, fig.cap = "Trend-scale hindcasts. Series 1-3 (top row) share one latent path; series 4-5 (bottom left, centre) share another; series 6 (bottom right) rides its own trend."}
plot(hindcast(fit_correct, type = "trend"))
```

`plot(fit, type = "factors")` shows the latent processes directly,
one panel per trend, annotated with the fraction of series-level
variance each factor explains. Trend 1 dominates because three of
the six series load on it; trend 2 contributes least; trend 3 sits
in between with the outlier series' larger-amplitude dynamics.

```{r plot-factors-correct, fig.cap = "The three latent AR(1) factors recovered by the correct-loadings fit, with the percentage of series-level variance each explains. Trend 1 (three series) contributes most, trend 3 (the outlier series) contributes a substantial share, and trend 2 (two short-memory series) contributes least."}
plot(fit_correct, type = "factors")
```

## Forecasting on shared trends

The same projection carries into forecasts. The AR(1) recursion
runs in latent space per posterior draw and the fixed $Z$ matrix
maps each draw's LV trajectory back to series scale. Series that
share a trend receive the same latent forecast paths cell for cell,
so the ribbons within each sharing group also match.

```{r plot-forecast-correct, fig.cap = "Forecasts under the correct trend_map. Series 1-3 reproduce the same latent forecast; series 4-5 do the same for the short-memory trend; series 6 forecasts its own dynamics. The long-memory ribbon fans out gradually; the short-memory ribbon returns to zero within a few timesteps."}
plot(forecast(fit_correct, newdata = test))
```

## Letting the model learn the loadings

Fixing the loadings is only the right call when domain knowledge
tells you which series belong together. When the sharing structure
is uncertain, hand the loadings to mvgam and let the sampler infer
them. Passing an all-`NA` numeric matrix of the target size to
`trend_map` tells mvgam to sample $Z$ under the default Student-t
prior with the QR identification of Heaps & Jermyn
([2024](https://doi.org/10.1007/s11222-024-10454-0)).

```{r free-z-fit, results = "hide"}
Z_free <- matrix(NA_real_, nrow = n_series, ncol = n_lv)
fit_free <- mvgam(
  formula = y ~ 1,
  trend_formula = ~ AR(p = 1),
  trend_map = Z_free,
  data = train,
  newdata = test,
  family = gaussian(),
  chains = 2L,
  iter = 1000L,
  warmup = 500L,
  silent = 2L,
  refresh = 0
)
```

The factor plot from the free-loadings fit recovers the same
dominant latent pattern as the correct fit. The leading factor
tracks the same slow oscillation, and the remaining two capture the
faster dynamics; the variance shares reallocate a little because the
QR identification distributes them differently across columns, but
the underlying signals are the same.

```{r plot-factors-free, fig.cap = "The three latent factors recovered by the free-loadings fit. The leading factor is the same long-memory pattern the correct-loadings fit found; the remaining two capture the same faster dynamics, with a slightly different variance allocation because the QR identification does not know which sharing structure was used to generate the data."}
plot(fit_free, type = "factors")
```

## When `trend_map` is wrong

`trend_map` is a hard prior on the sharing structure. Getting it
wrong forces the model to explain distinct dynamics with a single
latent path. To see how much that hurts, we fit a deliberately
misspecified model that pools every series onto one shared trend
(`trend_map = "shared"`) and compare its held-out forecast scores
against the correct fit.

```{r shared-fit, results = "hide"}
fit_shared <- mvgam(
  formula = y ~ 1,
  trend_formula = ~ AR(p = 1),
  trend_map = "shared",
  data = train,
  newdata = test,
  family = gaussian(),
  chains = 2L,
  iter = 1000L,
  warmup = 500L,
  silent = 2L,
  refresh = 0
)
```

The forecast plot from the misspecified fit shows the cost visually.
Every series' ribbon widens because a single latent path cannot
match the three distinct dynamical regimes at once, and the
hindcast bands during training are similarly wide.

```{r plot-forecast-shared, fig.cap = "Forecasts under the misspecified fit that pools every series onto one shared trend. Compared with the correct-loadings forecasts above, both the hindcast and forecast ribbons are visibly wider on every series because one latent process is trying to carry three distinct dynamical regimes."}
plot(forecast(fit_shared, newdata = test))
```

`score()` quantifies the cost using strictly proper scoring rules.
CRPS and the log score both penalise sharper forecasts less when
they are close to the observation and more when they are far, so
they measure both the mean and the calibration of the predictive
distribution. Lower is better for both.

```{r scores}
score_totals <- function(fit) {
  fc <- forecast(fit, newdata = test)
  c(
    crps = sum(sapply(
      score(fc, score = "crps")[levels(train$series)],
      function(x) sum(x$score, na.rm = TRUE)
    )),
    logs = sum(sapply(
      score(fc, score = "logs")[levels(train$series)],
      function(x) sum(x$score, na.rm = TRUE)
    )),
    energy = sum(score(fc, score = "energy")$all_series$score,
                  na.rm = TRUE)
  )
}
scores_tbl <- rbind(
  correct = score_totals(fit_correct),
  shared  = score_totals(fit_shared),
  free    = score_totals(fit_free)
)
round(scores_tbl, 2)
```

```{r score-summary, echo = FALSE}
logs_gap <- round(100 * (scores_tbl["shared", "logs"] /
                           scores_tbl["correct", "logs"] - 1))
free_rank <- vapply(colnames(scores_tbl), function(m) {
  unname(rank(scores_tbl[, m])["free"])
}, numeric(1))
free_worst <- names(free_rank)[free_rank == 3]
```

The correct-loadings fit wins on all three metrics. The log score
gap is the sharpest, `r round(scores_tbl["correct", "logs"])` for the
correct fit against `r round(scores_tbl["shared", "logs"])` for the
misspecified `"shared"` fit, worse by `r logs_gap`%. That comes from
series 4-6, whose dynamics are washed out when their observations
are pooled onto the dominant long-memory trend.

The free-loadings fit does not simply sit between the other two. It
beats `"shared"` on the log score but comes last on
`r paste(free_worst, collapse = " and ")`. It recovers the right
latent shapes, but it pays for identifying $Z$ jointly with the
trend parameters, and the QR rotation spreads variance across
factors in a way that does not match the sharing structure used to
generate the data.

The practical takeaway is that a held-out forecast score is a
useful diagnostic on the sharing structure itself. When two
plausible `trend_map` specifications disagree, fitting both and
comparing `score()` outputs on a real test slice tells you which
sharing pattern the data actually support.

## Take-aways

A dynamic factor model in mvgam is a state-space fit where fewer
latent processes drive more observed series. `trend_map` supplies
the sharing structure. The accepted input shapes range from the
`"identity"` and `"shared"` character shorthands through a
`data.frame` of named sharing groups to a numeric matrix that
handles the general case with optional partial-Z entries. Downstream
methods work unchanged. The parameter catalogue only grows by the
per-trend scalars belonging to whatever trend model was supplied.

When the sharing structure is known, fixing it via `trend_map` gives
the tightest and best-calibrated forecasts. When the structure is
uncertain, an all-`NA` matrix lets mvgam learn the loadings jointly
with the trend parameters, recovering the same underlying latent
shapes at a moderate score cost from the extra identification
uncertainty. When two candidate structures disagree, fitting both
and comparing `score()` outputs on a held-out slice is a practical
way to let the data decide.

Reach for `how_to_cite(fit)` when you want a citation-ready methods
paragraph naming the trend, family and sampler configuration used on
the fit. For the same content rendered as Markdown plus LaTeX for a
manuscript methods section, `methods_md(fit)` produces a paste-ready
block.

## Further reading

The natural next functions to explore on the same fit are:

- `?mvgam` for the full fitter signature, including the sibling
  `loadings_prior` argument that adds a structured prior to the
  sampled entries of $Z$ when using the numeric matrix form.
- `?hindcast.mvgam` and `?forecast.mvgam` for the prediction surface,
  including the `type = "trend"` view used above.
- `?score` for the strictly proper scoring rules used to compare
  forecast performance in this vignette.
- `?ensemble` and `?lfo_cv` for the two forecast-comparison
  workhorses built on top of `score`.
- `?jsdgam` for the joint species distribution wrapper, which uses
  the same partial-Z path to identify a factor model on multispecies
  count or occurrence data.
- `?mcmc_plot` for parameter posterior visualisation on any variable
  reported by `variables(fit)`.

Sibling articles in this series carry the workflow further:

- The
  [forecast evaluation article](https://nicholasjclark.github.io/mvgam/articles/forecast_evaluation.html)
  walks through `forecast()`, `score()`, `lfo_cv()` and `ensemble()`
  on a worked example with count observations.
- The
  [multi-response article](https://nicholasjclark.github.io/mvgam/articles/mvbf.html)
  covers the different case where several outcomes on different
  observation families share a single latent trend via a
  multivariate observation formula.
- The
  [JSDM article](https://nicholasjclark.github.io/mvgam/articles/jsdgam.html)
  applies the factor-model architecture to community ecology data
  with informed loadings, including the multiplicative gamma
  process prior that lets the data choose the factor count.
- The
  [VAR article](https://nicholasjclark.github.io/mvgam/articles/var.html)
  covers the opposite extreme, where every series keeps its own
  latent process and the interest is in how those processes
  drive one another.
- The
  [overview vignette](https://nicholasjclark.github.io/mvgam/articles/mvgam_overview.html)
  places `trend_map` alongside the other trend-side switches
  (`cor`, `gr`, `n_lv`) it composes with.

## Acknowledgements

The dynamic factor model on which `trend_map` is built follows Zuur
et al. ([2003](https://doi.org/10.1002/env.611)); the MARSS R
package (Holmes et al.
[2012](https://doi.org/10.32614/RJ-2012-002)) is the reference
implementation the syntax deliberately echoes. The identification
approach for factor models follows Heaps & Jermyn
([2024](https://doi.org/10.1007/s11222-024-10454-0)); mvgam applies
a QR rotation in generated quantities to identify factor models with
free loadings, and skips this rotation whenever a user supplies a
`trend_map`. Formulas rely on mgcv smooths (Wood
[2017](https://doi.org/10.1201/9781315370279)) and brms's multilevel
apparatus (Bürkner
[2017](https://doi.org/10.18637/jss.v080.i01)); the dynamic
generalised additive model framing follows Clark & Wells
([2023](https://doi.org/10.1111/2041-210X.13974)); sampling is
performed with Stan (Carpenter et al.
[2017](https://doi.org/10.18637/jss.v076.i01)).

## References

Bürkner, P.-C. (2017). brms: An R package for Bayesian
multilevel models using Stan. *Journal of Statistical Software*
80(1), 1-28.
[doi:10.18637/jss.v080.i01](https://doi.org/10.18637/jss.v080.i01)

Carpenter, B., Gelman, A., Hoffman, M. D., Lee, D., Goodrich, B.,
Betancourt, M., Brubaker, M., Guo, J., Li, P. and Riddell, A.
(2017). Stan: A probabilistic programming language. *Journal of
Statistical Software* 76(1), 1-32.
[doi:10.18637/jss.v076.i01](https://doi.org/10.18637/jss.v076.i01)

Clark, N. J. and Wells, K. (2023). Dynamic Generalized Additive
Models (DGAMs) for forecasting discrete ecological time series.
*Methods in Ecology and Evolution* 14, 771-784.
[doi:10.1111/2041-210X.13974](https://doi.org/10.1111/2041-210X.13974)

Heaps, S. E. and Jermyn, I. H. (2024). Structured prior
distributions for the covariance matrix in latent factor models.
*Statistics and Computing* 34, 143.
[doi:10.1007/s11222-024-10454-0](https://doi.org/10.1007/s11222-024-10454-0)

Holmes, E. E., Ward, E. J. and Wills, K. (2012). MARSS: Multivariate
Autoregressive State-space Models for Analyzing Time-series Data.
*The R Journal* 4(1), 11-19.
[doi:10.32614/RJ-2012-002](https://doi.org/10.32614/RJ-2012-002)

Wood, S. N. (2017). *Generalized Additive Models: An Introduction
with R* (2nd edition). Chapman and Hall/CRC.
[doi:10.1201/9781315370279](https://doi.org/10.1201/9781315370279)

Zuur, A. F., Fryer, R. J., Jolliffe, I. T., Dekker, R. and Beukema,
J. J. (2003). Estimating common trends in multivariate time series
using dynamic factor analysis. *Environmetrics* 14(7), 665-685.
[doi:10.1002/env.611](https://doi.org/10.1002/env.611)
