---
title: "Dynamic factor models"
author: "Nicholas J Clark"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: yes
vignette: >
  %\VignetteIndexEntry{Dynamic factor models}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
params:
  EVAL: !r identical(tolower(Sys.getenv("NOT_CRAN")), "true")
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  eval = if (isTRUE(exists("params"))) params$EVAL else FALSE
)
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  dpi = 100,
  fig.asp = 0.7,
  fig.width = 6,
  out.width = "70%",
  fig.align = "center"
)
library(mvgam)
library(ggplot2)
theme_set(mvgam_theme(base_size = 12, base_family = "serif"))
```

Series measured together often move together. Sites in one habitat
share weather, and species in one guild share resource pulses. A
dynamic factor model describes that common variation with a small
number of latent trends, and writes each series as a weighted sum of
them (Zuur et al. [2003](https://doi.org/10.1002/env.611); Holmes et
al. [2012](https://doi.org/10.32614/RJ-2012-002)). In mvgam the
`trend_map` argument sets which series load on which trend. This
vignette simulates six series from three trends and fits three
models: one given the true loadings, one that estimates them and one
that pools every series onto a single trend. Forecast scores on
held-out data then compare the three.

## The model

For series $j$ and latent trend $k$, the mean of each series on the
link scale is a linear combination of the trends:

\begin{align*}
y_{j, t}      & \sim F(\mu_{j, t}, \boldsymbol{\theta}) \\
g(\mu_{j, t}) & = \beta_0 + \beta^\top \mathbf{x}_{j, t}
                       + \sum_{k = 1}^{N_{lv}} Z_{j, k} \, z_{k, t} \\
z_{k, t}      & \sim \text{AR}(1) \text{ or any other trend model}
\end{align*}

$Z$ is the loading matrix, with one row per series and one column per
trend. Row $j$ gives the weight of each trend in series $j$. Every
series with a non-zero weight on trend $k$ uses the same latent path
$z_{k, 1:T}$, and all of those series inform its posterior. With fewer
trends than series, the model reduces the dimension of the problem.

## Three ways to write `trend_map`

`trend_map` supplies $Z$ in one of three forms:

- A keyword. `"identity"` gives each series its own trend, and
  `"shared"` gives every series the same trend.
- A `data.frame` with columns `series` and `trend`. Series given the
  same trend number share that trend with a loading of one.
- A numeric matrix with one row per series and one column per trend.
  mvgam holds each number fixed and estimates each `NA`.

The matrix form covers partial knowledge. Here five series have known
loadings and the fifth series is free to load on any of three trends:

```{r partial-map}
Z_partial <- rbind(
  s1 = c(1, 0, 0),
  s2 = c(1, 0, 0),
  s3 = c(1, 0, 0),
  s4 = c(0, 1, 0),
  s5 = c(NA, NA, NA),
  s6 = c(0, 0, 1)
)
```

Passed as `trend_map = Z_partial`, the fit estimates the three `NA`
entries, and `summary()` lists them under `Factor Loadings`. A matrix
of all `NA` estimates every loading, as shown later in this vignette.

## Simulated data

To compare the three models with known loadings, we simulated six
series from three AR(1) trends. Trend 1 is strongly autocorrelated
($\phi_1 = 0.85$) and drives series 1 to 3. Trend 2 is weakly
autocorrelated ($\phi_2 = 0.15$) and drives series 4 and 5. Trend 3
($\phi_3 = 0.55$) has the largest innovation SD and drives series 6
alone. The last 15 of the 80 time points are held out.

```{r simulate}
set.seed(3)

n_time <- 80
n_series <- 6
n_lv <- 3
n_train <- 65
phi_true <- c(0.85, 0.15, 0.55)
sig_true <- c(0.5, 0.35, 0.8)

# Each trend starts at its stationary distribution
lv <- matrix(NA, n_time, n_lv)
lv[1, ] <- rnorm(n_lv, 0, sig_true / sqrt(1 - phi_true^2))
for (t in 2:n_time) {
  lv[t, ] <- phi_true * lv[t - 1L, ] + rnorm(n_lv, 0, sig_true)
}

# Series 1-3 load on trend 1, series 4-5 on trend 2, series 6 on trend 3
Z_true <- matrix(0, nrow = n_series, ncol = n_lv)
Z_true[cbind(1:6, c(1, 1, 1, 2, 2, 3))] <- 1

# Observations add Gaussian noise with SD 0.2
y_mat <- lv %*% t(Z_true) +
  matrix(rnorm(n_time * n_series, 0, 0.2), n_time, n_series)

sim_data <- data.frame(
  time = rep(seq_len(n_time), n_series),
  series = factor(
    rep(paste0("s", seq_len(n_series)), each = n_time),
    levels = paste0("s", seq_len(n_series))
  ),
  y = as.numeric(y_mat)
)
train <- sim_data[sim_data$time <= n_train, ]
test  <- sim_data[sim_data$time >  n_train, ]
```

`mvgam_data()` checks the training data and plots each series, with
the held-out points in black:

```{r plot-observed, fig.cap = "The six simulated series. Series 1 to 3 follow one trend, series 4 and 5 a second, and series 6 a third. Black points after the dashed line are held out."}
mvgam_data(train, family = gaussian(), trend_formula = ~ AR(),
           newdata = test)
```

## Fitting known loadings

The first model is given the true grouping as a `data.frame`:

```{r trend-map}
tm_known <- data.frame(
  series = paste0("s", seq_len(n_series)),
  trend = c(1L, 1L, 1L, 2L, 2L, 3L)
)
tm_known
```

The observation formula holds an intercept alone, and `AR(p = 1)`
gives each of the three trends its own AR(1) process. Passing
`newdata = test` stores the held-out data on the fit, and `forecast()`
uses it later.

```{r fit-known, results = "hide"}
fit_known <- mvgam(
  formula = y ~ 1,
  trend_formula = ~ AR(p = 1),
  trend_map = tm_known,
  data = train,
  newdata = test,
  family = gaussian(),
  chains = 2L,
  iter = 1000L,
  warmup = 500L,
  silent = 2L
)
```

```{r summary-known}
summary(fit_known)
```

`summary()` reports one AR coefficient and one innovation SD per
trend. The 95% intervals of all six parameters contain the simulated
values, and the observation SD is estimated at its true value of 0.2.

`plot(type = "factors")` draws each latent trend with the share of
the series-level variance it explains. Because three of the six
series load on it, trend 1 explains the most. Trend 3 drives one
series and comes second, as its innovations are the largest.

```{r plot-factors-known, fig.cap = "The three latent trends of the model given the true loadings, each labelled with the share of series-level variance it explains."}
plot(fit_known, type = "factors")
```

`forecast()` extends each trend over the held-out times and maps the
draws back to the series through $Z$. Series on the same trend receive
the same latent forecast. The forecast for the weakly autocorrelated
trend returns to its mean within a few steps, and the forecast for the
strongly autocorrelated trend returns slowly.

```{r plot-forecast-known, fig.cap = "Forecasts from the model given the true loadings. Points are observations, and those after the dashed line were held out."}
fc_known <- forecast(fit_known)
plot(fc_known)
```

## Estimating the loadings

When the grouping is unknown, a matrix of `NA` asks mvgam to estimate
every loading. The loadings take a Student-t prior. A factor model
with free loadings gives the same fit under any rotation of its
factors. mvgam reports the loadings and factors in the rotation of
Heaps & Jermyn
([2024](https://doi.org/10.1007/s11222-024-10454-0)), which orders the
factors and fixes their signs.

```{r fit-free, results = "hide"}
fit_free <- mvgam(
  formula = y ~ 1,
  trend_formula = ~ AR(p = 1),
  trend_map = matrix(NA_real_, nrow = n_series, ncol = n_lv),
  data = train,
  newdata = test,
  family = gaussian(),
  chains = 2L,
  iter = 1000L,
  warmup = 500L,
  silent = 2L
)
```

```{r plot-factors-free, fig.cap = "The three latent factors of the model that estimates its loadings. The first factor matches trend 1 of the simulation."}
plot(fit_free, type = "factors")
```

The first factor reproduces trend 1 and explains a similar share of
the variance as it does in the model given the true loadings. Because
the rotation orders factors by the variance they explain, with no
knowledge of the simulated grouping, the second and third factors are
mixtures of trends 2 and 3. A single factor of a model with free
loadings seldom maps onto one process.

The dependence among series is the same under every rotation, and it
is often the target of a factor model. `residual_cor()` computes the
correlations that the loadings imply between each pair of series:

```{r plot-residcor-free, fig.cap = "Posterior median correlations among the six series, implied by the estimated loadings."}
plot(residual_cor(fit_free))
```

The estimated correlations recover the simulated grouping, which this
model was never given. Series 1 to 3 are strongly correlated with one
another, as are series 4 and 5, and series 6 is weakly correlated
with the rest.

`ordinate()` shows the same structure as a biplot. Each arrow is the
loading of one series on the first two factors, and each numbered
point is a time. Series that share a trend point in the same
direction.

```{r plot-ordinate-free, fig.cap = "Ordination of the model that estimates its loadings. Arrows are series loadings on the first two factors, and numbered points are times."}
ordinate(fit_free)
```

## Pooling every series

To see what a wrong grouping costs, the third model puts all six
series on one trend with `trend_map = "shared"`:

```{r fit-pooled, results = "hide"}
fit_pooled <- mvgam(
  formula = y ~ 1,
  trend_formula = ~ AR(p = 1),
  trend_map = "shared",
  data = train,
  newdata = test,
  family = gaussian(),
  chains = 2L,
  iter = 1000L,
  warmup = 500L,
  silent = 2L
)
```

```{r plot-forecast-pooled, fig.cap = "Forecasts from the model that pools all six series onto one trend. The intervals over the training period are much wider than those of the model given the true loadings."}
fc_pooled <- forecast(fit_pooled)
plot(fc_pooled)
```

One trend cannot follow three different sets of dynamics. The model
assigns the remaining variation to observation noise, and its
intervals over the training period are several times wider than those
of the first model.

## Comparing forecasts

`compare_scores()` scores the three forecasts against the held-out
data. The continuous ranked probability score (CRPS) and the log
score are computed for each series and horizon, and the energy score
is computed across all series at each horizon. Lower values are
better for all three. The chunk below sums each score over the
held-out period.

```{r scores}
fc_free <- forecast(fit_free)
total_score <- function(score) {
  by_horizon <- summary(compare_scores(
    fc_known, fc_free, fc_pooled,
    score = score, model_names = c("known", "free", "pooled")
  ))
  colSums(by_horizon[, c("known", "free", "pooled")])
}
totals <- sapply(c("crps", "logs", "energy"), total_score)
round(totals, 1)
```

```{r score-text, include = FALSE}
txt <- function(model, score) {
  formatC(totals[model, score], format = "f", digits = 1)
}
```

The pooled model has the worst total on all three scores. Its log
score of `r txt("pooled", "logs")` compares with
`r txt("known", "logs")` for the model given the true loadings. The
model that estimates its loadings scores `r txt("free", "crps")` on
CRPS and `r txt("free", "logs")` on the log score, close to the
`r txt("known", "crps")` and `r txt("known", "logs")` of the model
given the true loadings. In this simulation, estimating the loadings
cost little forecast accuracy, and a wrong grouping cost much more.

A held-out forecast score is a practical check on a proposed
grouping. Where two candidate `trend_map` structures are plausible,
fit both and compare their scores.

## Where to next

- `?mvgam` describes `trend_map` and `loadings_prior`, which sets a
  structured prior on estimated loadings.
- `?compare_scores` and `?score` list the available scoring rules.
- `?residual_cor` and `?shared_variation` give the correlations and
  covariances among series, with their intervals.
- `?ordinate` describes the biplot, and `?active_factors` helps
  choose the number of factors.
- The
  [forecast evaluation article](https://nicholasjclark.github.io/mvgam/articles/forecast_evaluation.html)
  covers `lfo_cv()` and `ensemble()` for comparing and combining
  models.
- The
  [JSDM article](https://nicholasjclark.github.io/mvgam/articles/jsdgam.html)
  applies the same factor structure to community data with
  `jsdgam()`.
- The
  [VAR article](https://nicholasjclark.github.io/mvgam/articles/var.html)
  covers models where each series keeps its own trend and the trends
  affect one another.
- The
  [overview vignette](https://nicholasjclark.github.io/mvgam/articles/mvgam_overview.html)
  introduces the other trend models and their arguments.

## References

Heaps, S. E. and Jermyn, I. H. (2024). Structured prior
distributions for the covariance matrix in latent factor models.
*Statistics and Computing* 34, 143.
[doi:10.1007/s11222-024-10454-0](https://doi.org/10.1007/s11222-024-10454-0)

Holmes, E. E., Ward, E. J. and Wills, K. (2012). MARSS: Multivariate
Autoregressive State-space Models for Analyzing Time-series Data.
*The R Journal* 4(1), 11-19.
[doi:10.32614/RJ-2012-002](https://doi.org/10.32614/RJ-2012-002)

Zuur, A. F., Fryer, R. J., Jolliffe, I. T., Dekker, R. and Beukema,
J. J. (2003). Estimating common trends in multivariate time series
using dynamic factor analysis. *Environmetrics* 14(7), 665-685.
[doi:10.1002/env.611](https://doi.org/10.1002/env.611)
