---
title: "Dynamic factor models with trend_map"
author: "Nicholas J Clark"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: yes
vignette: >
  %\VignetteIndexEntry{Dynamic factor models with trend_map}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
params:
  EVAL: !r identical(tolower(Sys.getenv("NOT_CRAN")), "true")
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  eval = if (isTRUE(exists("params"))) params$EVAL else FALSE
)
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  dpi = 100,
  fig.asp = 0.7,
  fig.width = 6,
  out.width = "70%",
  fig.align = "center"
)
library(mvgam)
library(ggplot2)
theme_set(theme_bw(base_size = 12, base_family = "serif"))
```

Multivariate time series often move together for reasons the
covariates do not record. Sites in the same habitat share weather
signals; species in the same guild ride shared resource pulses.
Dynamic factor models compress that common variation into a small
number of latent processes and pin each observed series to a
weighted sum of those processes (Zuur et al.
[2003](https://doi.org/10.1002/env.611); Holmes et al.
[2012](https://doi.org/10.32614/RJ-2012-002)). mvgam's `trend_map`
argument is how you supply the fixed sharing structure for a
state-space fit: you say which series belong to which trend, and the
posterior latent path is shared verbatim across the group. This
vignette introduces the three input shapes `trend_map` accepts and
recovers a small AR(1) dynamic factor model against a simulated
truth.

## Dynamic factor models

For $N_{\text{series}}$ observed series indexed by $j$ and $N_{lv}$
latent trends indexed by $k$, the dynamic factor model writes each
series' link-scale mean as a linear combination of the trends'
current states:

\begin{align*}
y_{j, t}      & \sim F(\mu_{j, t}, \boldsymbol{\theta}) \\
g(\mu_{j, t}) & = \beta_0 + \beta^\top \mathbf{x}_{j, t}
                       + \sum_{k = 1}^{N_{lv}} Z_{j, k} \, z_{k, t} \\
z_{k, t}      & \sim \text{AR}(1) \quad\text{or any other trend model.}
\end{align*}

$Z$ is an $N_{\text{series}} \times N_{lv}$ loading matrix. Row $j$
of $Z$ says which trends drive series $j$ and with what weight; the
posterior latent path $z_{k, 1:T}$ is shared verbatim across every
series whose row of $Z$ places non-zero weight on trend $k$. When
$N_{lv} < N_{\text{series}}$ the model is a genuine dimension
reduction: many series ride a small number of latent paths, and the
posterior on those paths pools information across every contributing
series.

`trend_map` is how you supply $Z$. mvgam accepts three input shapes:

- A character shorthand: `"identity"` fits one independent trend per
  series ($Z = I$); `"shared"` gives every series the same latent
  path ($Z$ is a column of ones).
- A `data.frame` with columns `series` and `trend` for sparse binary
  sharing. Each row says "series X uses trend Y"; two series pointing
  at the same trend integer are fixed to share.
- A numeric matrix of dimension $N_{\text{series}} \times N_{lv}$.
  Finite entries are treated as fixed loadings and preserved exactly
  in the posterior; `NA` entries mark loadings that should be sampled
  under a partial-Z parameterisation.

Supplying any fixed entry via `trend_map` bypasses the post-hoc QR
identification mvgam applies to free factor models (Heaps & Jermyn
[2024](https://doi.org/10.1007/s11222-024-10454-0)), so the encoded
structure is preserved verbatim in the posterior.

## A simulated example

We simulate four series driven by two latent AR(1) trends over 80
timepoints. Trend 1 has a long memory ($\phi_1 = 0.85$) and a
moderate innovation SD; trend 2 has short memory ($\phi_2 = 0.15$)
and a smaller innovation SD. Series 1 and 2 are driven by trend 1;
series 3 and 4 by trend 2. Observations are Gaussian with residual
SD 0.2.

```{r simulate}
set.seed(2)

# Dimensions and true parameters for the two latent AR(1) processes
n_time <- 80
n_series <- 4
n_lv <- 2
phi_true <- c(0.85, 0.15)
sig_true <- c(0.5, 0.35)

# Simulate the two latent trends; first state is drawn from the AR(1)
# stationary marginal so the process starts in equilibrium
lv <- matrix(NA, n_time, n_lv)
lv[1, ] <- rnorm(n_lv, 0, sig_true / sqrt(1 - phi_true^2))
for (t in 2:n_time) {
  lv[t, ] <- phi_true * lv[t - 1L, ] + rnorm(n_lv, 0, sig_true)
}

# Fixed loadings: series 1-2 driven by trend 1, series 3-4 by trend 2
Z_true <- matrix(0, nrow = n_series, ncol = n_lv)
Z_true[1, 1] <- 1
Z_true[2, 1] <- 1
Z_true[3, 2] <- 1
Z_true[4, 2] <- 1

# Assemble observations: series-level mean plus Gaussian noise
mu <- lv %*% t(Z_true)
y_mat <- mu + matrix(rnorm(n_time * n_series, 0, 0.2), n_time, n_series)

# mvgam long-format data frame: one row per series-time observation
sim_data <- data.frame(
  time = rep(seq_len(n_time), n_series),
  series = factor(
    rep(paste0("s", seq_len(n_series)), each = n_time),
    levels = paste0("s", seq_len(n_series))
  ),
  y = as.numeric(y_mat)
)
```

The sharing structure is already visible in the raw observations:
series 1 and 2 track a trend with long memory (slow swings), while
series 3 and 4 track a noisier trend with short memory.

```{r plot-observed, fig.cap = "Simulated observations rendered via the package's own inspection surface. Series 1-2 track one latent AR(1) trend with long memory; series 3-4 track a second trend with short memory."}
mvgam_data(sim_data, y = "y", family = gaussian())
```

## Fitting with a `trend_map`

We pass the sharing structure as a two-column `data.frame`. The
`series` column lines up with the observed factor's levels and the
`trend` column carries the integer trend index; two rows pointing at
the same integer share a latent path.

```{r trend-map}
tm <- data.frame(
  series = paste0("s", seq_len(n_series)),
  trend = c(1L, 1L, 2L, 2L)
)
tm
```

The rest of the call is a standard `mvgam()` fit. The observation
side is a single population intercept (`y ~ 1`); the trend side is an
AR(1) process, applied to each of the two trends the `trend_map`
implies.

```{r fit, results = "hide"}
mod <- mvgam(
  formula = y ~ 1,
  trend_formula = ~ AR(p = 1),
  trend_map = tm,
  data = sim_data,
  family = gaussian(),
  chains = 2L,
  samples = 500L,
  burnin = 500L,
  silent = 2L,
  refresh = 0
)
```

`summary()` reports the AR coefficient and innovation SD for each of
the two trends alongside the observation intercept and residual SD.
The posterior means sit close to the simulated truth: `ar1_trend[1]`
recovers to 0.90 (95% CI 0.81-0.97; truth 0.85) and `ar1_trend[2]`
to 0.18 (95% CI -0.08-0.45; truth 0.15), and the innovation SDs
recover to 0.54 and 0.37 against truth 0.5 and 0.35. Residual
observation SD lands at 0.20 (95% CI 0.18-0.22), which is the value
used to generate the data.

```{r summary}
summary(mod)
```

## Recovering the latent trends

`hindcast(mod, type = "trend")` returns the latent-state posterior on
its own scale, and `plot()` on that object renders one facet per
series. Because we fixed series 1 and 2 to share trend 1, and series
3 and 4 to share trend 2, the paired ribbons carry identical draws:
the top row is the posterior for the trend with long memory, the
bottom row for the trend with short memory.

```{r plot-trend-hindcast, fig.cap = "Trend-scale hindcasts. The top row (series 1-2) plots one latent path; the bottom row (series 3-4) plots the other. The paired ribbons are identical because trend_map pinned the sharing."}
plot(hindcast(mod, type = "trend"))
```

Overlaying the simulated truth as crosses confirms that both latent
paths are recovered inside the credible ribbon. We join the truth
for each series to the ribbon plot returned by
`plot(hindcast())` via an added `geom_point()` layer.

```{r plot-trend-truth, fig.cap = "Trend hindcasts with the simulated truth overlaid as crosses. The ribbon covers the truth throughout for both latent paths."}
truth_long <- data.frame(
  time = rep(seq_len(n_time), n_series),
  series = factor(
    rep(paste0("s", seq_len(n_series)), each = n_time),
    levels = paste0("s", seq_len(n_series))
  ),
  truth = as.numeric(lv %*% t(Z_true))
)

plot(hindcast(mod, type = "trend")) +
  geom_point(
    data = truth_long,
    aes(x = time, y = truth),
    shape = 4,
    inherit.aes = FALSE,
    size = 1.2
  )
```

Series-level hindcasts stack the latent state and the observation
intercept, so series that share a trend also share the same shape at
the response scale, differing only by the observation noise the
ribbon absorbs.

```{r plot-series-hindcast, fig.cap = "Response-scale hindcasts for each series. Series 1 and 2 track together; series 3 and 4 track together. The ribbons are slightly wider than the trend-scale hindcasts because they include the observation noise."}
plot(hindcast(mod))
```

## Forecasting on shared trends

The same projection carries into forecasts. When we ask for 15
steps ahead, the AR(1) recursion runs in latent space per
posterior draw and the fixed `Z` matrix maps each draw's LV
trajectory back to series scale. Series that share a trend
receive the same latent forecast paths cell for cell.

```{r forecast}
n_forecast <- 15L
fc_newdata <- data.frame(
  time = rep((n_time + 1L):(n_time + n_forecast), n_series),
  series = factor(
    rep(paste0("s", seq_len(n_series)), each = n_forecast),
    levels = paste0("s", seq_len(n_series))
  ),
  y = NA_real_
)
fc <- forecast(mod, newdata = fc_newdata)
```

```{r plot-forecast, fig.cap = "Forecasts for the four series. Series 1 and 2 (top row) reproduce the same latent AR(1) forecast draw for draw; series 3 and 4 do the same for the other trend. The long-memory trend fans out gradually across the horizon; the short-memory trend collapses back to its mean within a few timesteps."}
plot(fc)
```

## Letting the model learn the loadings

Fixing the loadings is only the right call when domain
knowledge tells you which series belong together. When the
sharing structure is uncertain, hand the loadings to mvgam and
let the sampler infer them. Passing an all-`NA` numeric matrix
of the target size to `trend_map` tells mvgam to sample `Z`
under the default Student-t prior with the QR identification of
Heaps & Jermyn ([2024](https://doi.org/10.1007/s11222-024-10454-0)).

```{r free-z-fit, results = "hide"}
Z_free <- matrix(NA_real_, nrow = n_series, ncol = n_lv)
mod_free <- mvgam(
  formula = y ~ 1,
  trend_formula = ~ AR(p = 1),
  trend_map = Z_free,
  data = sim_data,
  family = gaussian(),
  chains = 2L,
  samples = 500L,
  burnin = 500L,
  silent = 2L,
  refresh = 0
)
```

Compared with the fit that fixed `Z`, the free-loadings model
has to jointly identify the two latent trajectories and the
four loadings that map them onto the series. The trend
hindcasts carry more uncertainty in return:

```{r plot-free-z, fig.cap = "Trend hindcasts under the free-loadings fit. The ribbons are visibly wider than the fixed-Z hindcasts because the fit now samples the loadings jointly with the latent paths."}
plot(hindcast(mod_free, type = "trend"))
```

This is the usual identification versus flexibility trade.
Fixing `Z` when you know the sharing structure gives you
tighter posteriors on the trends. Letting `Z` sample lets the
data tell you which series belong together at the cost of
somewhat wider bands.

## Take-aways

A dynamic factor model in mvgam is a state-space fit where fewer
latent processes drive more observed series. `trend_map` supplies
the sharing structure. The accepted input shapes range from the
`"identity"` and `"shared"` character shorthands through a
`data.frame` of named sharing groups to a numeric matrix that
handles the general case with optional partial-Z entries. Downstream
methods work unchanged. The parameter catalogue only grows by the
per-trend scalars belonging to whatever trend model was supplied.

Reach for `how_to_cite(mod)` when you want a citation-ready methods
paragraph naming the trend, family and sampler configuration used on
the fit. For the same content rendered as Markdown plus LaTeX for a
manuscript methods section, `methods_md(mod)` produces a paste-ready
block.

## Further reading

The natural next functions to explore on the same fit are:

- `?mvgam` for the full fitter signature, including the sibling
  `loadings_prior` argument that adds a structured prior to the
  sampled entries of $Z$ when using the numeric matrix form.
- `?hindcast.mvgam` and `?forecast.mvgam` for the prediction surface,
  including the `type = "trend"` view used above.
- `?jsdgam` for the joint species distribution wrapper, which uses
  the same partial-Z path to identify a factor model on multispecies
  count or occurrence data.
- `?mcmc_plot` for parameter posterior visualisation on any variable
  reported by `variables(mod)`.

Sibling articles in this series carry the workflow further:

- The
  [multi-response article](https://nicholasjclark.github.io/mvgam/articles/multivariate.html)
  covers the different case where several outcomes on different
  observation families share a single latent trend via a
  multivariate observation formula.
- The
  [JSDM article](https://nicholasjclark.github.io/mvgam/articles/jsdgam.html)
  applies the factor-model architecture to community ecology data
  with informed loadings.

## Acknowledgements

The dynamic factor model on which `trend_map` is built follows Zuur
et al. ([2003](https://doi.org/10.1002/env.611)); the MARSS R
package (Holmes et al.
[2012](https://doi.org/10.32614/RJ-2012-002)) is the reference
implementation the syntax deliberately echoes. The identification approach for factor models
follows Heaps & Jermyn
([2024](https://doi.org/10.1007/s11222-024-10454-0)); mvgam applies
a QR rotation in generated quantities to identify factor models with
free loadings, and skips this rotation whenever a user supplies a
`trend_map`. Formulas rely on mgcv smooths (Wood
[2017](https://doi.org/10.1201/9781315370279)) and brms's multilevel
apparatus (Bürkner
[2017](https://doi.org/10.18637/jss.v080.i01)); the dynamic
generalised additive model framing follows Clark & Wells
([2023](https://doi.org/10.1111/2041-210X.13974)); sampling is
performed with Stan (Carpenter et al.
[2017](https://doi.org/10.18637/jss.v076.i01)).

## References

Bürkner, P.-C. (2017). brms: An R package for Bayesian
multilevel models using Stan. *Journal of Statistical Software*
80(1), 1-28.
[doi:10.18637/jss.v080.i01](https://doi.org/10.18637/jss.v080.i01)

Carpenter, B., Gelman, A., Hoffman, M. D., Lee, D., Goodrich, B.,
Betancourt, M., Brubaker, M., Guo, J., Li, P. and Riddell, A.
(2017). Stan: A probabilistic programming language. *Journal of
Statistical Software* 76(1), 1-32.
[doi:10.18637/jss.v076.i01](https://doi.org/10.18637/jss.v076.i01)

Clark, N. J. and Wells, K. (2023). Dynamic Generalized Additive
Models (DGAMs) for forecasting discrete ecological time series.
*Methods in Ecology and Evolution* 14, 771-784.
[doi:10.1111/2041-210X.13974](https://doi.org/10.1111/2041-210X.13974)

Heaps, S. E. and Jermyn, I. H. (2024). Structured prior
distributions for the covariance matrix in latent factor models.
*Statistics and Computing* 34, 143.
[doi:10.1007/s11222-024-10454-0](https://doi.org/10.1007/s11222-024-10454-0)

Holmes, E. E., Ward, E. J. and Wills, K. (2012). MARSS: Multivariate
Autoregressive State-space Models for Analyzing Time-series Data.
*The R Journal* 4(1), 11-19.
[doi:10.32614/RJ-2012-002](https://doi.org/10.32614/RJ-2012-002)

Wood, S. N. (2017). *Generalized Additive Models: An Introduction
with R* (2nd edition). Chapman and Hall/CRC.
[doi:10.1201/9781315370279](https://doi.org/10.1201/9781315370279)

Zuur, A. F., Fryer, R. J., Jolliffe, I. T., Dekker, R. and Beukema,
J. J. (2003). Estimating common trends in multivariate time series
using dynamic factor analysis. *Environmetrics* 14(7), 665-685.
[doi:10.1002/env.611](https://doi.org/10.1002/env.611)
