---
title: "Overview of the mvgam package"
author: "Nicholas J Clark"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: yes
vignette: >
  %\VignetteIndexEntry{Overview of the mvgam package}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
params:
  EVAL: !r identical(tolower(Sys.getenv("NOT_CRAN")), "true")
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  eval = if (isTRUE(exists("params"))) params$EVAL else FALSE
)
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  dpi = 100,
  fig.asp = 0.8,
  fig.width = 6,
  out.width = "60%",
  fig.align = "center"
)
library(mvgam)
library(ggplot2)
theme_set(theme_bw(base_size = 12, base_family = "serif"))
```

`mvgam` fits, plots and forecasts Bayesian dynamic generalized
additive models. Version 2 is built on top of `brms`: every
observation model is constructed as a `brms` formula and compiled
through the `brms` Stan code generator, and `mvgam` then extends
that pipeline with a latent trend layer so the fitted model is a
state-space model rather than a regression. An observation
process maps a latent state to the response through a chosen
likelihood, and a trend process governs how the latent state
evolves over time.

The `brms` backend brings the full `brms` modelling surface to
state-space fits:

- **Families.** Every `brms` observation family is supported
  natively (`gaussian()`, `student()`, `poisson()`,
  `negbinomial()`, `binomial()`, `beta_binomial()`,
  `bernoulli()`, `Beta()`, `Gamma()`, `lognormal()`, the
  ZI and hurdle variants, ordinal families). `mvgam` adds a
  small library of families for time series and ecological
  use cases that `brms` does not ship (`tweedie`,
  `com_binomial`, `nmix`, `occ`, `diri`, `multi`, `categ`,
  `mvn`, `mvt`); see the table below.
- **Effect types.** Linear terms, smooths (`s()`, `t2()`,
  `te()`), Gaussian process smooths (`gp()`), monotonic
  effects (`mo()`), random effects through `(... | ...)`,
  smooths split by a factor (`by = `), and nonlinear formulas
  via `bf(..., nl = TRUE)`. Anything `brms` accepts as a
  predictor is available on both `formula` and
  `trend_formula`.
- **Multiple response variables.** Joint-response fits compose
  through `bf()` arms, including distributional regression on
  any auxiliary parameter (e.g. `bf(y ~ ..., sigma ~ ...)`).
  Closure-unit families pass a detection model as a second
  `bf()` arm; `nmix` and `occ` use this to separate occupancy
  or abundance from the detection process.
- **Post-fit tooling.** `marginaleffects::predictions()`,
  `comparisons()`, `hypotheses()` and `slopes()` all dispatch
  on `mvgam` objects through the standard `posterior_epred()`,
  `posterior_predict()` and `posterior_linpred()` methods.
  `bayesplot`, `posterior` and `loo` work without
  special-casing. `tidy()`, `augment()` and `glance()` return
  broom-standard data frames.

The state-space layer is what's new on top of `brms`. The latent
trend pipeline accepts random walks, autoregressions (with
arbitrary sparse lag sets), vector autoregressions, continuous
time AR for irregular sampling, piecewise growth, hierarchical
correlations, and low-rank factor structure for joint species
distribution models. Smooths and covariate effects can sit on
the latent state alongside the response, so the same brms
modelling vocabulary applies in both layers.

This vignette walks through the package on a simulated dataset
generated by `sim_mvgam()`. The simulator returns the true smooth
functions and latent trajectory it drew from, so the recovery of
every component can be checked directly. The pedagogical question
is sharp: when the data has both a deterministic seasonal cycle
and a stochastic dynamic process beyond that cycle, what does a
fixed-seasonality smooth miss, and what does adding a latent state
buy?

## State-space formulation

For an observation $y_{i, t}$ on series $i$ at time $t$ with link
function $g$ and observation family $\mathcal{F}$:

\begin{align*}
y_{i, t} & \sim \mathcal{F}(\tilde{y}_{i, t}, \phi) \\
g(\tilde{y}_{i, t}) & = \alpha_i +
  \sum_{j=1}^{J} s_{i, j}(x_{j, t}) +
  Z_{i, \cdot} \, z_{\cdot, t} \, .
\end{align*}

The intercept $\alpha_i$ and smooths $s_{i, j}$ form the
observation-side linear predictor. The latent state $z_{k, t}$
evolves through a trend process (AR(1), VAR, RW, CAR and so on);
$Z$ is a loadings matrix mapping the $K$-dimensional latent state
onto the $N$ series. With $K = N$ and $Z = I$ each series gets
its own latent state; with $K < N$ the latent states are shared
factors. Smooths and covariate effects passed through
`trend_formula` sit on the latent state; everything in the
top-level `formula` sits on the response.

## Supported observation families

Every observation family in `brms` is supported through the brms
integration layer (`gaussian()`, `student()`, `poisson()`,
`negbinomial()`, `binomial()`, `beta_binomial()`,
`bernoulli()`, `Beta()`, `Gamma()`, `lognormal()`, and the rest).
`mvgam` adds a small library of families that brms does not ship
natively:

| Family | Use case | Trend dpars |
|:------:|:---------|:-----------:|
| `tweedie()` | Compound Poisson-Gamma counts with zero inflation | `phi`, `power` |
| `com_binomial()` | Conway-Maxwell-Binomial for over-, under-, super-dispersed bounded counts | `nu` |
| `nmix()` | N-mixture detection-error models (Poisson-Binomial, Royle-Nichols, Poisson-Poisson variants) | family-specific |
| `occ()` | Single-season occupancy with imperfect detection | `psi` |
| `diri()` | Dirichlet compositional responses | `phi` |
| `multi()`, `categ()` | Multinomial and categorical responses on the closure-unit grain | -- |
| `mvn()`, `mvt()` | Multivariate normal / Student-t residual covariance via gllvm-style factor structure | `Psi` |

These families (`nmix`, `occ`, `diri`, `multi`, `categ`, `mvn`,
`mvt`) typically go through `jsdgam()` rather than `mvgam()`.
Each is documented on its own help page with a worked recovery
example.

### Closure-unit and joint-species families

`jsdgam()` is the entry point for joint species distribution
modelling and closure-unit observation processes. Data arrive
in long format with one row per `(unit, species)` cell:
sites by species for `diri` / `multi` / `categ` / `mvn` / `mvt`,
or sites by visit by species for `nmix` / `occ` where the
detection sub-formula models imperfect observation. The latent
trend pipeline composes the same way as for `mvgam()`: factor
structure via `n_lv`, structured `loadings_prior`, hierarchical
correlations through `gr` / `subgr`. Three concrete cases:

- **`occ()` and `nmix()`** marginalise over the latent
  occupancy or abundance state inside the Stan likelihood. The
  chain only samples the detection sub-formula and the
  state process linear predictor. `nmix()` supports three
  observation processes (`"poisson_binomial"`, `"royle_nichols"`,
  `"poisson_poisson"`) chosen at the constructor. The detection
  sub-formula is a second `bf()` arm in the model formula.
- **`diri()` / `multi()` / `categ()`** model compositional
  responses on the simplex. A reference category fixes one slot
  in the length K `mu` vector and a `sum_to_zero_vector[K]`
  constraint is applied to each column of `Z`, so the joint
  identification works without user constraints.
  `posterior_epred()` returns one probability per
  `(site, species)` row summing to 1 within each site.
- **`mvn()` and `mvt()`** model continuous multivariate
  responses with the gllvm covariance
  $\Sigma = Z Z^\top + \mathrm{diag}(\Psi^2)$.
  `residual_cor()` extracts the implied cross-species
  correlations after Heaps QR identification.

See `?jsdgam` for the full surface, including phylogenetic
distance matrices through `phylo`, trait-environment
interactions through `trait_formula`, and the
`?ordinate.mvgam` / `?ordinate.jsdgam` biplot helpers.

## Latent trend processes

The trend constructors live in the `trend_formula` argument and
fall into six families. Each is documented in its own help page;
the formulas below state the per-step transition the constructor
emits.

**Random walk**, `RW()`. Each series has its own non-stationary
state; the moving average term ($\theta_i$) engages when
`ma = TRUE`:

\begin{align*}
z_{i, t} & = z_{i, t - 1} + \theta_i \, \epsilon_{i, t - 1}
                + \epsilon_{i, t} \\
\epsilon_{i, t} & \sim \text{Normal}(0, \sigma_i)
\end{align*}

**Autoregression**, `AR(p = ...)`. The lag set
$\mathcal{L} = \{1, \dots, p\}$ when `p` is a scalar, or
$\mathcal{L} = p$ when `p` is a vector (so `AR(p = c(1, 12))`
sets $\mathcal{L} = \{1, 12\}$). Stationary by construction
through a Heaps (2023) prior on the lag coefficients:

\begin{align*}
z_{i, t} & \sim \text{Normal}\!\Bigl(
  \textstyle \sum_{l \in \mathcal{L}} \rho_{i, l} \, z_{i, t - l}
  + \theta_i \, \epsilon_{i, t - 1},\;
  \sigma_i \Bigr)
\end{align*}

**Vector autoregression**, `VAR(p = ...)`. Each $A_l$ is an
$N_{\text{series}} \times N_{\text{series}}$ transition matrix.
Innovations are correlated across series through $\Sigma$, with
an LKJ prior on the Cholesky factor:

\begin{align*}
z_t & \sim \text{Normal}\!\Bigl(
  \textstyle \sum_{l=1}^{p} A_l \, z_{t - l},\; \Sigma \Bigr)
\end{align*}

**Zero-mean multivariate normal**, `ZMVN()`. Cross-series
covariance with no temporal recursion; useful for joint species
distributions and similar grouped responses:

\begin{align*}
z_t & \sim \text{Normal}(0, \Sigma)
\end{align*}

**Continuous-time AR(1)**, `CAR()`. The lag coefficient decays
geometrically with the time gap $\Delta_t = t - t_{-}$ between
the current observation and the most recent one. Lets the AR
recursion run on irregularly-spaced timestamps:

\begin{align*}
z_{i, t} & \sim \text{Normal}\!\bigl(
  \rho_i^{\Delta_t} \, z_{i, t_{-}},\;
  \sigma_i \, \sqrt{1 - \rho_i^{2 \Delta_t}} \bigr)
\end{align*}

**Piecewise growth**, `PW()`. The Prophet kernel with $S$
changepoints at known times $\{s_1, \dots, s_S\}$. The lag-zero
rate is $k$ at $t = 0$, the offset is $m$ and the magnitude
shift at changepoint $s_j$ is $\delta_j$. Let $a(t)$ be the
$\mathbb{R}^S$ indicator with $a_j(t) = \mathbb{1}\{t \geq s_j\}$
and $\gamma$ the offset-correction vector that keeps the
trajectory continuous. Linear growth:

\begin{align*}
z_t & = \bigl(k + a(t)^\top \delta\bigr) t
       + \bigl(m + a(t)^\top \gamma\bigr)
\end{align*}

Logistic growth uses the same $(k, m, \delta)$ but maps through
a sigmoid with a (possibly time-varying) carrying capacity
$C(t)$:

\begin{align*}
z_t & = \frac{C(t)}{1 + \exp\!\bigl(
  -(k + a(t)^\top \delta)\,(t - (m + a(t)^\top \gamma))
\bigr)}
\end{align*}

### Composing options

Three orthogonal switches layer on top of any per-series trend.
The first is `cor = TRUE`, which switches to correlated
innovations through an LKJ prior on the Cholesky factor of
$\Sigma$ (default: independent per-series).

The second is `gr = ` / `subgr = `, which sets up hierarchical
residual correlation matrices. The per-group covariance is
shrunk toward a population matrix through a weight
$\alpha_{\text{cor}} \in (0, 1)$ on the population draw:

\begin{align*}
\Omega_g & = \alpha_{\text{cor}} \, \Omega_{\text{global}}
           + (1 - \alpha_{\text{cor}}) \, \Omega_{g, \text{local}}
\end{align*}

The third is `n_lv < n_series`. A $K$-dimensional latent process
$\tilde{z}_t$ drives the $N_{\text{series}}$-dimensional state
through a loadings matrix
$Z \in \mathbb{R}^{N_{\text{series}} \times K}$:

\begin{align*}
z_t & = Z \, \tilde{z}_t, \qquad
\tilde{z}_t \in \mathbb{R}^K
\end{align*}

Factor structure delivers three concrete advantages over running
$N_{\text{series}}$ independent latent processes.

The first is dimensionality reduction. Communities with dozens
of correlated outcomes (species, financial assets, regions) can
share dynamics through a small set of latent axes, cutting
parameter count and identifying the joint structure that
fully-independent fits miss.

The second is rotation invariance. The marginal covariance
$Z Z^\top$ identifies what the data can resolve about $Z$, and
mvgam follows the Heaps and Jermyn (2024) prior. That prior
samples $Z$ unconstrained and applies a thin QR rotation in
generated quantities to identify $Z$ post hoc. The diagonal of
the resulting rotated loadings is non-negative by construction.
This removes the $2^K$ sign mode equivalence that plagues
constrained factor models.

The third is structured priors. `loadings_prior` accepts either
a Matrix Normal prior built from per-series features and
distance matrices, or the multiplicative gamma process (MGP)
prior of Bhattacharya and Dunson (2011). MGP shrinks later
columns of $Z$ toward zero and lets `n_lv` act as a truncation
ceiling rather than the exact factor count. Under MGP,
`active_factors()` reports the posterior distribution of the
number of active columns.

For VAR trend factor models, the lag coefficient cube also
rotates. The rotated dynamics matrix is
$\tilde{A}_l = Q^\top A_l Q$ for the orthogonal $Q$ delivered by
the QR factorisation. Per-factor scalar parameters
($\rho_l$, $\sigma$, MA coefficients, $L_{\Omega}$) stay in the
unrotated latent basis. See `?loadings_prior` for the structured
priors and `?ordinate.mvgam` for biplot helpers on the rotated
loadings.

The example below uses `AR(p = c(1, 12))` to fit lag-1 momentum
together with annual recurrence at lag 12 from monthly data. The
full surface is documented at `?mvgam_trends`; constructor
arguments live at `?AR`, `?VAR`, `?RW`, `?ZMVN`, `?CAR`, `?PW`.
Smooth Gaussian process effects on covariates remain available
through `gp()` inside the formula or `trend_formula`. The
standalone GP trend constructor was retired during v2
development.

## A simulated example

`sim_mvgam(type = 7)` generates a single Gaussian series with a
deterministic seasonal cycle on the obs side and a sparse-lag
`AR(p = c(1, 12))` latent state. The seasonal cycle is
identifiable from the smooth alone; the latent state carries
short-term momentum (lag 1) and year-on-year recurrence (lag 12)
beyond the fixed cycle. The simulator returns 20 years of
monthly data and a 12-month held-out tail so the question "does
adding the latent state buy us anything on the out-of-sample
forecast?" has somewhere to land.

```{r sim-data}
set.seed(7)
sim <- sim_mvgam(
  type               = 7L,
  family             = gaussian(),
  n_series           = 1L,
  n_timepoints       = 240L,
  proportional_train = 12 / 13
)
train <- sim$data_train
test  <- sim$data_test
nrow(train)
nrow(test)
head(train)
```

`sim$true_smooths[["s(season)"]]` holds the generative seasonal
function on a fine grid; `sim$true_trend` is the realised
AR(1, 12) trajectory on the link scale. The two together drive
the response:

```{r sim-plot, fig.asp = 0.55, out.width = "80%"}
trend_df <- data.frame(
  time = seq_along(sim$true_trend[, 1L]),
  z    = sim$true_trend[, 1L]
)
ggplot() +
  geom_line(data = train, aes(time, y), colour = "grey40") +
  geom_point(data = train, aes(time, y), size = 0.6,
             colour = "grey20") +
  geom_line(data = trend_df, aes(time, z), colour = "steelblue",
            linewidth = 0.8) +
  geom_vline(xintercept = max(train$time), linetype = "dashed",
             colour = "grey50") +
  labs(x = "time (months)", y = "y",
       caption = paste0(
         "Black: response. Blue: true AR(1, 12) latent state ",
         "(link scale). Vertical line: end of training."
       )) +
  theme(plot.caption = element_text(hjust = 0))
```

The seasonal cycle is visible in the annual repetition of the
response. The latent state (blue line) drifts substantially
across the 240-month window: $z$ ranges from about $-9$ to $+8$,
much wider than the deterministic cycle itself. A fixed seasonal
fit will pick up the average annual pattern but cannot follow
that drift into the held out year.

## Model 1: seasonal smooth, no latent state

The first model fits the cyclic seasonal smooth and treats
everything else as observation noise:

\begin{align*}
y_t & \sim \text{Normal}(\mu_t, \sigma) \\
\mu_t & = \alpha + s_{\text{season}}(\text{season}_t) \\
\beta_{\text{smooth}} & \sim \text{Normal}\!\bigl(0, (S \lambda)^{-1}\bigr) \\
\sigma & \sim \text{half-Student-t}(3, 0, 2.5)
\end{align*}

The cyclic smooth wraps January (1) onto December (12) through
the `bs = "cc"` basis; the `knots` argument places the cycle
endpoints just outside the observed range so the basis treats
month 1 and month 12 as adjacent.

```{r model-step1, results = "hide"}
step1 <- mvgam(
  y ~ s(season, bs = "cc", k = 12),
  knots   = list(season = c(0.5, 12.5)),
  family  = gaussian(),
  data    = train,
  newdata = test,
  chains  = 2,
  samples = 400,
  burnin  = 400,
  silent  = 2
)
```

```{r}
summary(step1, include_betas = FALSE)
```

The headline number is `sigma`: roughly $3.2$ on the response
scale. The cyclic smooth has been pulled toward zero
(`sds_1` posterior median around $0.09$, smooth basis
coefficients all close to zero) because most of the temporal
variation in the response is unstructured to a fit that ignores
time. With nothing else to explain that variation, the obs noise
absorbs it.

```{r step1-cond-effects, fig.asp = 0.5, out.width = "70%"}
conditional_effects(step1)
```

The marginal smooth is essentially flat. The Dunn-Smyth residuals
show why this fit is misspecified:

```{r step1-resid, fig.asp = 0.5, out.width = "80%"}
plot(step1, type = "residuals")
```

Lag-1 and lag-12 autocorrelation are both visible in the residual
ACF. Model 1 discards this structure entirely, forcing the lag-1
and lag-12 signal into residual noise, a sign of fundamental
misspecification when a state-space model would absorb that
signal into a latent state.

## Model 2: add an AR(1, 12) latent state

The second model keeps the seasonal smooth and lets the latent
state carry the structure model 1 missed:

\begin{align*}
y_t & \sim \text{Normal}(\mu_t + z_t, \sigma) \\
\mu_t & = \alpha + s_{\text{season}}(\text{season}_t) \\
z_t & \sim \text{Normal}\!\bigl(\rho_1 \, z_{t - 1} + \rho_{12} \, z_{t - 12},
                               \sigma_{\text{trend}}\bigr) \\
\rho_1, \rho_{12} & \sim \text{Normal}(0, 1)[-1, 1] \\
\sigma_{\text{trend}} & \sim \text{half-Student-t}(3, 0, 2.5)
\end{align*}

The obs-side formula is unchanged; the latent state enters
through `trend_formula`:

```{r model-step2, results = "hide"}
step2 <- mvgam(
  y ~ s(season, bs = "cc", k = 12),
  knots         = list(season = c(0.5, 12.5)),
  trend_formula = ~ AR(p = c(1, 12)),
  family        = gaussian(),
  data          = train,
  newdata       = test,
  chains        = 2,
  samples       = 400,
  burnin        = 400,
  silent        = 2
)
```

```{r}
summary(step2, include_betas = FALSE)
```

Model 1 misattributed the dynamic structure to observation noise
and reported $\sigma \approx 3.2$; model 2 correctly partitions
that signal into the latent state and brings $\sigma$ down to
$\approx 1.1$. The trend
block reports `ar1_trend` and `ar12_trend`:

```{r step2-ar-recovery}
ar1  <- as.numeric(as.array(step2, variable = "ar1_trend[1]"))
ar12 <- as.numeric(as.array(step2, variable = "ar12_trend[1]"))
data.frame(
  parameter = c("ar1_trend", "ar12_trend"),
  truth     = c(0.55, 0.40),
  median    = c(median(ar1),  median(ar12)),
  lower_90  = c(quantile(ar1,  0.05), quantile(ar12, 0.05)),
  upper_90  = c(quantile(ar1,  0.95), quantile(ar12, 0.95)),
  prob_gt_0 = c(mean(ar1 > 0), mean(ar12 > 0))
)
```

Both lag coefficients are clearly above zero, with posterior
probability $1.0$ for each. The 90% intervals cover the simulator
truths of $0.55$ and $0.40$. `mcmc_plot()` returns the same
posteriors as densities:

```{r step2-mcmc, fig.asp = 0.45, out.width = "70%"}
mcmc_plot(step2, variable = "ar[0-9]+_trend",
          regex = TRUE, type = "areas")
```

The Dunn-Smyth residuals no longer show the lag-1 and lag-12
autocorrelation that model 1's residuals carried:

```{r step2-resid, fig.asp = 0.5, out.width = "80%"}
plot(step2, type = "residuals")
```

The cyclic smooth, freed from the job of explaining
time-varying structure, now identifies the seasonal cycle
through the smoother penalty:

```{r step2-cond-effects, fig.asp = 0.5, out.width = "70%"}
conditional_effects(step2)
```

## Hindcast versus forecast

`hindcast()` returns the posterior trajectory at the training
times. `forecast()` propagates the latent state forward to held
out times. Both return a single `mvgam_forecast` object with the
same structure; the forecast arm is populated only when
`newdata` covers times beyond the training window:

```{r hindcast-forecast}
fc <- forecast(step2, newdata = test)
dim(fc$hindcasts[[1]])
dim(fc$forecasts[[1]])
```

```{r forecast-plot, fig.asp = 0.55, out.width = "80%"}
plot(fc, series = 1)
```

The hindcast trajectory reads the Stan-fitted latent state
directly (deterministic, exact). The forecast runs the AR(1, 12)
recursion forward from the last fitted state, accumulating
uncertainty from `sigma_trend` and the obs noise.

## Forecast scoring across rules

PSIS-LOO is unreliable for comparing these two fits. Leaving an
observation out while keeping the latent state estimated from it
raises the in-sample density and rewards the wrong model. The
honest yardstick is the held-out forecast score. `score()`
computes proper scoring rules per timepoint; the sum across the
held out set ranks models on out-of-sample predictive
performance. The Continuous Ranked Probability Score is the
default, but the same call works for the log score (`logs`),
the Dawid-Sebastiani score (`dss`), and the Discrete Ranked
Probability Score (`drps`):

```{r forecast-score}
fc1 <- forecast(step1, newdata = test)
fc2 <- forecast(step2, newdata = test)
score_rules <- c("crps", "logs", "dss", "drps")
scores <- vapply(score_rules, function(rule) {
  s1 <- score(fc1, score = rule)
  s2 <- score(fc2, score = rule)
  c(no_trend = sum(s1[[1]]$score, na.rm = TRUE),
    ar_1_12  = sum(s2[[1]]$score, na.rm = TRUE))
}, numeric(2))
round(scores, 2)
```

A lower score is better. The state-space fit wins on every rule.
CRPS drops by roughly $3.4$, the log score by $1.3$, the
Dawid-Sebastiani score by $4.0$, and the DRPS by $1.9$. DSS shows
the biggest margin because that rule penalises poorly calibrated
forecast variance most aggressively, and the trendless fit stays
overconfident around its (flat) cycle.

`score()` covers additional rules through the `score` argument:
`sis` (interval score), `qs` (quantile score) and a weighted
CRPS variant (`twcrps`) for emphasising tail behaviour. For
multivariate forecasts from VAR fits or factor models, `score()`
also supports `energy` (energy score) and `variogram` (variogram
score). Both penalise dependence patterns that per-series rules
miss. The `interval_width` and `per_series` arguments control
nominal coverage and aggregation.

### Leave-future-out cross-validation

A single train / test split rewards models that happen to win on
one specific horizon. `lfo_cv.mvgam()` automates the alternative:
it scores forecasts at every refit point along the series, then
aggregates the per-step scores across many forecast origins.

```{r lfo-example, eval = FALSE}
lfo_step2 <- lfo_cv(step2, min_t = 180, fc_horizon = 12)
plot(lfo_step2)
```

`min_t` sets the first refit point, `fc_horizon` sets the
prediction window per fold, and the chain is reseated through
`update()` between folds so the Stan program does not recompile.
The default score is the expected log predictive density (ELPD),
returned per fold and per step ahead. The companion
`kfold.mvgam()` runs block CV for grouped designs through
cluster-level holdout. It supports user-defined fold maps
through the `folds` argument.

`loo()` and `loo_compare()` both dispatch on `mvgam` objects.
Both inherit the same pathology. PSIS-LOO approximates leaving
$y_t$ out of the likelihood, but the chain still conditioned on
$y_t$ when it estimated the latent state $z_t$. The approximate
density is therefore evaluated against a posterior that has
already seen the observation through the latent path. Pareto-k
diagnostics warn about exactly this. Within-sample time points
raise warnings because the importance weights diverge.
Differencing two contaminated ELPDs in `loo_compare()` does not
remove the contamination. The honest comparison for time series
fits is either a held-out forecast score (as above) or
`lfo_cv()`. `lfo_cv()` refits the chain on shorter prefixes, so
the latent state at the scoring time was never conditioned on
its own observation.

## How to cite

`how_to_cite()` returns a methods-section paragraph adapted to
the model's actual structure (likelihood, link, trend type,
sampler configuration, backend):

```{r how-to-cite}
how_to_cite(step2)
```

For a math-only model statement rendered as Markdown plus LaTeX,
use `methods_md(step2)`.

## Where to next

- `?mvgam` for the full fitter signature, including
  `loadings_prior`, `trend_map`, and threading options.
- `?jsdgam` for joint species distribution and closure-unit
  family fits.
- `?forecast.mvgam` and `?hindcast.mvgam` for the prediction
  surface, including `type = "link" / "expected" / "trend"`.
- `?conditional_effects.mvgam` for the `marginaleffects`-backed
  effect-summary path; `predictions()`, `comparisons()`, and
  `hypotheses()` from the `marginaleffects` package also dispatch
  on `mvgam` objects.
- `?kfold.mvgam` and `?lfo_cv.mvgam` for k-fold and
  leave-future-out cross-validation.
- `?irf`, `?fevd`, `?stability` for impulse-response, forecast
  error variance decomposition, and stability summaries for VAR
  trends.

## Further reading

Clark, N. J. and Wells, K. (2023). Dynamic Generalized Additive
Models for forecasting discrete ecological time series.
*Methods in Ecology and Evolution* 14, 771-784.

Clark, N. J., Ernest, S. K. M., Senyondo, H., Simonis, J.,
White, E. P., Yenni, G. M. and Karunarathna, K. A. N. K. (2025).
Beyond single-species models: multispecies forecasts for
ecological predictability. *PeerJ* 13:e18929.

Heaps, S. E. and Jermyn, I. H. (2024). Structured prior
distributions for the covariance matrix in latent factor models.
*Statistics and Computing* 34:143.

Burkner, P.-C. (2017). brms: An R package for Bayesian multilevel
models using Stan. *Journal of Statistical Software* 80(1):1-28.

Wood, S. N. (2017). *Generalized Additive Models: An Introduction
with R* (2nd edition). Chapman and Hall/CRC.
