---
title: "Overview of the mvgam package"
author: "Nicholas J Clark"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: yes
vignette: >
  %\VignetteIndexEntry{Overview of the mvgam package}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
params:
  EVAL: !r identical(tolower(Sys.getenv("NOT_CRAN")), "true")
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  eval = if (isTRUE(exists("params"))) params$EVAL else FALSE
)
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  dpi = 100,
  fig.asp = 0.8,
  fig.width = 6,
  out.width = "60%",
  fig.align = "center"
)
library(mvgam)
library(ggplot2)
theme_set(mvgam_theme(base_size = 12, base_family = "serif"))
```

`mvgam` fits Bayesian state-space models to time series, extending
the dynamic generalised additive models of Clark and Wells (2023).
The observation model is a `brms` formula (Bürkner 2017), and a
second formula, `trend_formula`, adds a latent state that evolves
through time. Temporal structure that the covariates do not explain
is estimated as part of the latent state, and forecasts carry that
structure forward. Stan fits the model, and the fitted object works
with the `brms`, `marginaleffects`, `loo` and `bayesplot` tooling.

For an observation $y_{i,t}$ on series $i$ at time $t$, with link
function $g$ and a response distribution $\mathcal{F}$:

\begin{align*}
y_{i,t} & \sim \mathcal{F}(\mu_{i,t}, \theta) \\
g(\mu_{i,t}) & = \alpha_i + \textstyle\sum_j f_j(x_{j,i,t})
  + Z_{i,\cdot} \, \eta_{\cdot,t} \\
\eta_{\cdot,t} & \sim \text{trend process}(\eta_{\cdot,t-1}, \dots)
\end{align*}

The $f_j$ are any terms a `brms` formula accepts: linear effects,
smooths (Wood 2017), Gaussian processes, monotonic effects and random
effects. The latent state $\eta$ follows one of the trend processes
below, and the loadings $Z$ map it onto the series. Each series has
its own state when $Z$ is the identity, and a few shared factors
drive many series when $Z$ has fewer columns than there are series.

## What mvgam can model

Every `brms` family works as the observation model, and `mvgam` adds
families for ecological and count data:

| Family | Use case |
|:-------|:---------|
| `tweedie()` | Non-negative data with many exact zeros |
| `beta_nb()` | Counts with heavier tails than a negative binomial |
| `com_binomial()` | Bounded counts that are over- or under-dispersed |
| `occ()` | Occupancy from repeated detection surveys |
| `nmix()` | Abundance from repeated counts with imperfect detection |
| `diri()`, `multi()`, `categ()` | Compositions and categorical outcomes |
| `mvn()`, `mvt()` | Correlated continuous responses |

`trend_formula` names the latent state with one constructor, and it
can hold covariates and smooths that act on the state:

| Constructor | Latent state | Typical use |
|:------------|:-------------|:------------|
| `RW()` | Random walk | Series with no fixed mean |
| `AR(p = )` | Autoregression, including sparse lags such as `p = c(1, 12)` | Momentum and seasonal recurrence |
| `VAR(p = )` | Vector autoregression | Series that drive one another |
| `CAR()` | Continuous-time AR(1) | Irregular times, and series observed at their own times |
| `ZMVN()` | Correlated residuals with no time recursion | Sites, regions or species |
| `PW()` | Piecewise linear or logistic growth | Trends with changepoints |

The constructors share a set of options. `cor = TRUE` correlates the
innovations across series, and `gr` with `subgr` pools correlation
matrices across groups. `n_lv` sets a number of shared latent factors
below the number of series, and `df` gives the innovations
Student-t tails. `ma = TRUE` adds a moving-average term to `RW()`,
`AR()` and `VAR()`. The help page of each constructor lists the
options it takes.

`mvgam()` fits data in long format, one row per series and time.
`jsdgam()` fits joint species distribution models to one row per site
and species, with phylogenetic and trait-informed loadings.

## Portal rodent captures

`portal_data` holds monthly captures of four desert rodent species
from the Portal Project, a long-term field study in Arizona (Ernest
et al. 2018). Trapping follows the lunar cycle, and some months
went unsampled, which leaves missing counts. Two covariates come
with the counts: minimum temperature (`mintemp`) and a 12-month
moving average of vegetation greenness (`ndvi_ma12`). We held out
the last 12 months of each series to score forecasts.

```{r data}
data(portal_data)
train <- subset(portal_data, time <= 68)
test  <- subset(portal_data, time >  68)
```

Fitting any model below needs a working Stan installation, through
either `cmdstanr` or `rstan`; the
[cmdstanr guide](https://mc-stan.org/cmdstanr/articles/cmdstanr.html)
covers the setup.

### One species

The Desert Pocket Mouse (PP) is dormant through the colder months,
and its captures run from long stretches of zeros to peaks above
20. `mvgam_data()` checks the counts against the family and plots
the series, its distribution and its autocorrelation, with the
held-out months in black:

```{r pp-data, fig.asp = 0.85, out.width = "85%"}
pp_train <- droplevels(subset(train, series == "PP"))
pp_test  <- droplevels(subset(test, series == "PP"))
mvgam_data(pp_train, family = poisson(), y = "captures",
           newdata = pp_test)
```

The first model relates captures to smooth functions of the two
covariates, with no latent state:

```{r fit-gam, results = "hide"}
fit_gam <- mvgam(
  captures ~ s(mintemp, k = 6) + s(ndvi_ma12, k = 6),
  family  = poisson(),
  data    = pp_train,
  newdata = pp_test,
  silent  = 2
)
```

Captures in one month carry over into the next beyond anything
the covariates explain, and the randomised quantile residuals of
this model remain strongly autocorrelated at lag 1:

```{r gam-resid, fig.asp = 0.5, out.width = "80%"}
plot(fit_gam, type = "residuals", series = "PP")
```

The second model moves both smooths into `trend_formula`, where
they act on a latent state that follows an AR(1) process. The
observation formula keeps the intercept:

```{r fit-ar, results = "hide"}
fit_ar <- mvgam(
  captures ~ 1,
  trend_formula = ~ s(mintemp, k = 6) + s(ndvi_ma12, k = 6) + AR(),
  family        = poisson(),
  data          = pp_train,
  newdata       = pp_test,
  silent        = 2
)
```

```{r ar-summary}
summary(fit_ar)
```

The latent state absorbs the persistence from one month to the
next, and the residuals of this fit show no autocorrelation left:

```{r ar-resid, fig.asp = 0.5, out.width = "80%"}
plot(fit_ar, type = "residuals", series = "PP")
```

`pp_check()` compares the observed counts with counts drawn from
the fitted model. The rootogram shows the model reproducing the
distribution of captures, zeros included:

```{r ar-ppc, fig.asp = 0.5, out.width = "70%"}
pp_check(fit_ar, type = "rootogram")
```

`conditional_effects()` plots each covariate's effect on the
expected count, and the `marginaleffects` functions
`predictions()`, `comparisons()` and `slopes()` work on the fit
directly:

```{r ar-effects, fig.asp = 0.5, out.width = "70%"}
plot(conditional_effects(fit_ar, type = "expected"), ask = FALSE)
```

`forecast()` continues the latent state from its last estimated
values into the held-out months. Without a latent state, the
first model forecasts from the covariates alone, and its
intervals miss most of the high counts:

```{r forecasts, fig.asp = 0.5, out.width = "80%"}
fc_gam <- forecast(fit_gam)
fc_ar  <- forecast(fit_ar)
plot(fc_gam)
plot(fc_ar)
```

`compare_scores()` scores both forecasts with the continuous
ranked probability score (CRPS), where lower is better, and
`summary()` sets the scores out by forecast horizon:

```{r scores, fig.asp = 0.5, out.width = "70%"}
scores <- compare_scores(fc_gam, fc_ar, score = "crps")
summary(scores)
plot(scores, relative = "fc_gam")
```

```{r score-numbers, include = FALSE}
by_horizon <- summary(scores)
totals <- round(colSums(by_horizon[c("fc_gam", "fc_ar")], na.rm = TRUE), 1)
n_scored <- sum(!is.na(by_horizon$fc_ar))
ar_wins <- sum(by_horizon$fc_ar < by_horizon$fc_gam, na.rm = TRUE)
pp_rows <- subset(scores, series != "all_series")
covered <- tapply(pp_rows$in_interval, pp_rows$model, sum, na.rm = TRUE)
```

The latent state improves the forecasts. The AR model has the lower
CRPS in `r ar_wins` of the `r n_scored` scored months, and its CRPS
sums to `r totals[["fc_ar"]]` against `r totals[["fc_gam"]]` for the
smooth-only model. Its 90% intervals contain `r covered[["fc_ar"]]`
of the `r n_scored` held-out counts. Those of the model without a
latent state contain `r covered[["fc_gam"]]`.

`score()` also returns the log score, interval scores and, for
several series, energy and variogram scores. `lfo_cv()` repeats
the comparison over many forecast origins by refitting on
successively longer windows.

### Four species

The four species share the covariates but respond to them
differently. The observation formula gives each species its own
mean through `series`. In `trend_formula`, `by = lv_axis()` gives
each species its own smooths of temperature and vegetation, and
`AR(cor = TRUE)` gives each species its own latent AR(1) state,
with innovations correlated across species:

```{r fit-cor, results = "hide"}
fit_cor <- mvgam(
  captures ~ series,
  trend_formula = ~ s(mintemp, by = lv_axis(), k = 6) +
    s(ndvi_ma12, by = lv_axis(), k = 6) + AR(cor = TRUE),
  family        = poisson(),
  data          = train,
  newdata       = test,
  silent        = 2
)
```

`conditional_effects()` with `series = "all"` draws each species'
response to temperature in a panel of its own. The responses differ
clearly among species. Captures of PP rise with minimum temperature
and those of PB rise at the warmest temperatures, while captures of
the two kangaroo rats, DM and DO, fall slightly:

```{r cor-effects, fig.asp = 0.8, out.width = "85%"}
plot(
  conditional_effects(fit_cor, effects = "mintemp", series = "all",
                      type = "expected"),
  ask = FALSE
)
```

The forecasts come from one joint posterior, and each species
keeps its own scale and persistence:

```{r cor-forecast, fig.asp = 0.75, out.width = "90%"}
fc_cor <- forecast(fit_cor)
plot(fc_cor)
```

```{r cor-numbers, include = FALSE}
cor_scores <- score(fc_cor, score = "crps")
in_interval <- unlist(lapply(
  cor_scores[names(cor_scores) != "all_series"], `[[`, "in_interval"
))
```

The forecasts track the held-out counts of all four species. Their
90% intervals contain `r sum(in_interval, na.rm = TRUE)` of the
`r sum(!is.na(in_interval))` held-out counts.

`residual_cor()` returns the correlations among the species'
innovations, which measure how strongly the species rise and fall
together once temperature and vegetation are accounted for:

```{r cor-resid, fig.asp = 0.8, out.width = "60%"}
rc <- residual_cor(fit_cor)
plot(rc)
```

```{r cor-resid-numbers, include = FALSE}
pb_pp <- sprintf("%.2f (95%% interval %.2f to %.2f)",
                 rc$cor["PB", "PP"], rc$cor_lower["PB", "PP"],
                 rc$cor_upper["PB", "PP"])
```

The two pocket mouse species, PB and PP, share the strongest
correlation, `r pb_pp`. `n_lv` would instead estimate a few shared
latent factors for many species, and `VAR()` would estimate how
each species' state affects the others at the next time step.

### Reporting

`methods_md()` writes the fitted model as Markdown equations with
its priors, and `how_to_cite()` returns a methods paragraph and the
references for the model's components:

```{r methods, results = "asis"}
methods_md(fit_ar, heading_level = 4)
```

```{r how-to-cite}
how_to_cite(fit_ar)
```

## Where to go next

- The
  [data formatting vignette](https://nicholasjclark.github.io/mvgam/articles/data.html)
  prepares data for `mvgam()`, including missing values and
  irregular times.
- The
  [dynamic factor vignette](https://nicholasjclark.github.io/mvgam/articles/dfm.html)
  shares latent states across many series.
- The
  [VAR article](https://nicholasjclark.github.io/mvgam/articles/var.html)
  estimates how series drive one another, with impulse responses and
  variance decompositions.
- The
  [JSDM article](https://nicholasjclark.github.io/mvgam/articles/jsdgam.html)
  fits multispecies data with `jsdgam()`, and the
  [N-mixture article](https://nicholasjclark.github.io/mvgam/articles/nmix.html)
  covers detection-error families.
- The
  [forecast evaluation article](https://nicholasjclark.github.io/mvgam/articles/forecast_evaluation.html)
  compares models with `score()`, `lfo_cv()` and `ensemble()`.

## References

Bürkner, P.-C. (2017). brms: An R package for Bayesian
multilevel models using Stan. *Journal of Statistical Software*
80(1), 1-28.
[doi:10.18637/jss.v080.i01](https://doi.org/10.18637/jss.v080.i01)

Clark, N. J. and Wells, K. (2023). Dynamic Generalized Additive
Models (DGAMs) for forecasting discrete ecological time series.
*Methods in Ecology and Evolution* 14, 771-784.
[doi:10.1111/2041-210X.13974](https://doi.org/10.1111/2041-210X.13974)

Ernest, S. K. M. et al. (2018). The Portal Project: a long-term
study of a Chihuahuan desert ecosystem. *bioRxiv* 332783.
[doi:10.1101/332783](https://doi.org/10.1101/332783)

Wood, S. N. (2017). *Generalized Additive Models: An Introduction
with R* (2nd edition). Chapman and Hall/CRC.
[doi:10.1201/9781315370279](https://doi.org/10.1201/9781315370279)
