---
title: "Overview of the mvgam package"
author: "Nicholas J Clark"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: yes
vignette: >
  %\VignetteIndexEntry{Overview of the mvgam package}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
params:
  EVAL: !r identical(tolower(Sys.getenv("NOT_CRAN")), "true")
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  eval = if (isTRUE(exists("params"))) params$EVAL else FALSE
)
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  dpi = 100,
  fig.asp = 0.8,
  fig.width = 6,
  out.width = "60%",
  fig.align = "center"
)
library(mvgam)
library(ggplot2)
theme_set(theme_bw(base_size = 12, base_family = "serif"))
```

`mvgam` fits Bayesian dynamic generalized additive models and
helps with plotting and forecasting them. The observation model is written as a `brms`
formula and compiled through the `brms` Stan code generator;
`mvgam` extends that pipeline with a latent state layer so the
fitted model is a state-space model rather than a regression. An
observation process maps the latent state to the response through
a chosen likelihood, and a trend process governs how the latent
state evolves over time.

Because the observation side is compiled by `brms`, the full
`brms` modelling vocabulary carries through into state-space fits.
Linear terms, smooths (`s()`, `t2()`, `te()`), Gaussian process
smooths (`gp()`), monotonic effects (`mo()`), random effects
through `(... | ...)`, smooths split by a factor (`by = `), and
nonlinear formulas through `bf(..., nl = TRUE)` all work on
either side of the model. Joint responses compose through `bf()`
arms with distributional regression on auxiliary parameters
(e.g. `bf(y ~ ..., sigma ~ ...)`).

`mvgam` then adds the latent state layer. Inside `trend_formula`
sit random walks, autoregressions with arbitrary sparse lag sets,
vector autoregressions, piecewise growth and hierarchical
residual correlations, alongside a continuous time AR for
irregular sampling and a low-rank factor structure for joint
species distribution models. Smooths and covariate effects can
run on the latent state alongside the response, so the same
modelling vocabulary applies in both layers.

Once fitted, the resulting object plugs into the wider Bayesian
tooling: `marginaleffects::predictions()`, `comparisons()`,
`hypotheses()` and `slopes()` dispatch on `mvgam` objects through
the standard `posterior_epred()`, `posterior_predict()` and
`posterior_linpred()` methods. `bayesplot` and `loo` plug in
directly. `as_draws()` from the `posterior` package returns the
draws as an array. `tidy()` and `glance()` return broom-format
data frames, with `augment()` for adding posterior summaries back
to the original data.

This vignette walks through the package on a simulated dataset
generated by `sim_mvgam()`. The simulator returns the true smooth
functions and latent trajectory it drew from, so every component's
recovery can be checked directly. The pedagogical question is
sharp: when the data carries both a deterministic seasonal cycle
and a stochastic dynamic process beyond that cycle, what does a
fixed-seasonality smooth miss, and what does adding a latent state
buy?

## State-space formulation

For an observation $y_{i, t}$ on series $i$ at time $t$ with link
function $g$ and observation family $\mathcal{F}$:

\begin{align*}
y_{i, t} & \sim \mathcal{F}(\tilde{y}_{i, t}, \phi) \\
g(\tilde{y}_{i, t}) & = \alpha_i +
  \sum_{j=1}^{J} s_{i, j}(x_{j, t}) +
  Z_{i, \cdot} \, z_{\cdot, t} \, .
\end{align*}

The intercept $\alpha_i$ and smooths $s_{i, j}$ form the
observation-side linear predictor. The latent state $z_{k, t}$
evolves through a trend process (AR(1), VAR, RW, CAR and so on);
$Z$ is a loadings matrix mapping the $K$-dimensional latent state
onto the $N$ series. With $K = N$ and $Z = I$ each series gets
its own latent state; with $K < N$ the latent states are shared
factors. Smooths and covariate effects passed through
`trend_formula` sit on the latent state; everything in the
top-level `formula` sits on the response.

## Modelling APIs

Two fitters at the top of the package expose the same state-space
machinery. They differ only in how the data are laid out and
what the response represents.

`mvgam()` takes a long-format data frame indexed by a time
column and (optionally) by a series column. The defaults are
`time` and `series`, but every trend constructor accepts
`time = ` / `series = ` arguments that point at any column in
the data, so existing column names need not be renamed. The
series dimension drives the multivariate trend constructors:
`VAR()`, the `cor = TRUE` switches and hierarchical
correlations through `gr`. The response is a single column on
the `brms` side. Use this fitter to forecast one or more time
series, including the closure-unit families when each unit
produces a scalar response.

`jsdgam()` is the entry point for joint species distribution and
closure-unit fits where the grouping is biological rather than
temporal. Data arrive as one row per `(unit, species)` cell.
Joint multivariate response families work on sites by species;
occupancy and N-mixture models work on sites by visit by
species, with the detection process attached as a second `bf()`
arm. `jsdgam()` composes the same trend pipeline as `mvgam()`,
with factor structure via `n_lv` and `loadings_prior`, and adds
phylogenetic distance matrices through `phylo` and trait by
environment interactions through `trait_formula`.

The family table below applies to both fitters. The closure-unit
and joint multivariate response families work in either, and the
choice between `mvgam()` and `jsdgam()` is driven by what the
rows of the data represent rather than by the likelihood itself.

## Supported observation families

The complete `brms` family list is available through the
integration layer. `mvgam` adds a small library of families for
time-series and ecological use cases that `brms` does not ship
natively:

| Family | Use case | Family parameters |
|:------:|:---------|:-----------------:|
| `tweedie()` | Compound Poisson-Gamma counts that include many zeros | `phi`, `power` |
| `com_binomial()` | Conway-Maxwell-Binomial for over-, under-, super-dispersed bounded counts | `nu` |
| `nmix()` | N-mixture detection-error models (Poisson-Binomial, Royle-Nichols, Poisson-Poisson variants) | family-specific |
| `occ()` | Single-season occupancy with imperfect detection | `psi` |
| `diri()` | Dirichlet compositional responses | `phi` |
| `multi()`, `categ()` | Multinomial and categorical responses on the closure-unit grain | -- |
| `mvn()`, `mvt()` | Multivariate normal / Student-t responses with low-rank residual covariance | `Psi` |

Each family is documented on its own help page with a worked
recovery example. The two groups that need extra explanation are
the closure-unit families and the joint multivariate response
families, both summarised below.

### Closure-unit families: `occ()` and `nmix()`

Closure-unit data hold repeated observations of the same sampling
unit (a site, a visit, a season) under a process that may detect
the underlying state imperfectly. `occ()` and `nmix()` marginalise
the latent occupancy or abundance state inside the Stan
likelihood, so the chain samples only the detection model and
the linear predictor for the state. Both enter through a second
`bf()` arm.

`occ()` implements the single-season occupancy model of
[MacKenzie et al. (2002)](https://doi.org/10.1890/0012-9658\(2002\)083[2248:ESORWD]2.0.CO;2),
which separates occupancy probability $\psi$ from per-visit
detection probability. `nmix()` supports three observation
processes chosen at the constructor: `"poisson_binomial"`
follows [Royle (2004)](https://doi.org/10.1111/j.0006-341X.2004.00142.x),
`"royle_nichols"` follows
[Royle and Nichols (2003)](https://doi.org/10.1890/0012-9658\(2003\)084[0777:EAFRPA]2.0.CO;2),
and `"poisson_poisson"` treats each observed count as a noisy
Poisson realisation of an underlying Poisson abundance.

### Joint multivariate response families

These families model a multivariate response on the closure-unit
grain. The likelihood is evaluated at the unit level (one row
per `(unit, species)` cell with `K` species), and the trend
pipeline composes through the latent state as for any other
family. Two complementary identification mechanisms are needed
because both compositional and continuous multivariate
likelihoods are non-identified without them.

`diri()`, `multi()` and `categ()` model compositional or
categorical responses on the simplex. The likelihood subtracts
the reference category's linear predictor from every entry of
the length-$K$ `mu` vector (a softmax shift equivalent to the
$K - 1$ reference parameterisation), and a `sum_to_zero_vector[K]`
constraint is applied to each column of $Z$. Together these
identify the joint model without user-supplied constraints.
`posterior_epred()` returns one probability per
`(unit, species)` row summing to 1 within each unit.

`mvn()` and `mvt()` model continuous multivariate responses.
The residual covariance is the sum of a loadings outer product
and a diagonal of species-specific variances $\Psi^2$:

\begin{align*}
\Sigma & = Z Z^\top + \mathrm{diag}(\Psi^2)
\end{align*}

`residual_cor()` converts this $\Sigma$ to a correlation matrix
between species on the response scale. The loadings are
identified by the same rotation described in the trend factor
model section below.

See `?jsdgam` for the full surface, including the phylogenetic
and trait-environment hooks, and `?ordinate.mvgam` /
`?ordinate.jsdgam` for biplot helpers built on the rotated
loadings. The online article
[Joint species distribution GAMs with informed loadings](https://nicholasjclark.github.io/mvgam/articles/jsdgam.html)
walks through a complete `jsdgam()` workflow on count data
with a negative binomial observation model, trait-informed
loadings, ordination rotations and MGP shrinkage.

## Latent trend processes

Six families of trend constructor live inside `trend_formula`.
Each is documented on its own help page; the formulas below give
the per-step transition that the constructor compiles into Stan.
The presentation starts with the simplest case (random walks),
then covers stationary autoregressions, an analogue for irregular
sampling, and a deterministic piecewise kernel.

### Random walks: `RW()`

The simplest dynamic process. Each series carries a
non-stationary state that drifts by a Normal innovation at every
step. Setting `ma = TRUE` adds a moving-average error term
$\theta_i$ that lets the chain absorb brief shocks without
forcing them into the next state:

\begin{align*}
z_{i, t} & = z_{i, t - 1} + \theta_i \, \epsilon_{i, t - 1}
                + \epsilon_{i, t} \\
\epsilon_{i, t} & \sim \text{Normal}(0, \sigma_i)
\end{align*}

### Autoregressions: `AR(p = ...)`

`AR()` scales the previous state by lag coefficients $\rho_l$.
When `p` is a scalar, the lag set is
$\mathcal{L} = \{1, \dots, p\}$; when `p` is a vector, the lag
set is $\mathcal{L} = p$ and the user can skip intermediate
lags. `AR(p = c(1, 12))` evaluates only the lag 1 and lag 12
coefficients, with no $\rho_2$ through $\rho_{11}$ in the model.
Sparse lag sets are the right shape for periodic seasonal
recurrence (lag 1 momentum and lag 12 annual cycle for monthly
data; lag 1 and lag 7 for weekly data) without paying for the
intermediate coefficients. Default priors bound each $\rho_l$
in $(-1, 1)$. For AR(1), the initial state is rescaled to the
stationary marginal variance, so the prior at $t = 1$ matches
the stationary distribution implied by $\rho_1$ and $\sigma_i$:

\begin{align*}
z_{i, t} & \sim \text{Normal}\!\Bigl(
  \textstyle \sum_{l \in \mathcal{L}} \rho_{i, l} \, z_{i, t - l}
  + \theta_i \, \epsilon_{i, t - 1},\;
  \sigma_i \Bigr)
\end{align*}

### Vector autoregressions: `VAR(p = ...)`

`VAR()` generalises `AR()` to multiple correlated series. Each
$A_l$ is an $N_{\text{series}} \times N_{\text{series}}$
transition matrix. Entries off the diagonal carry the lagged
effect of one series on another. Unlike `AR()`, `VAR()` requires
`p` to be a scalar (`VAR(p = 2)` includes lags 1 and 2): the
Heaps (2023) joint stationary prior used here needs consecutive
companion-form structure, so sparse lag vectors are not
supported. Innovations
are correlated through $\Sigma$, with an LKJ prior on the
Cholesky factor:

\begin{align*}
z_t & \sim \text{Normal}\!\Bigl(
  \textstyle \sum_{l=1}^{p} A_l \, z_{t - l},\; \Sigma \Bigr)
\end{align*}

### Zero-mean multivariate normal: `ZMVN()`

`ZMVN()` is `VAR()` stripped of its temporal recursion. It is the
right kernel when the data are organised by a non-temporal
grouping (sites or regions, for example) and the question is
residual covariance rather than dynamics, which is the typical setup for
joint species distribution models that have no time dimension:

\begin{align*}
z_t & \sim \text{Normal}(0, \Sigma)
\end{align*}

### Continuous-time AR(1): `CAR()`

For data sampled at irregular intervals, where the discrete AR
recursion no longer makes sense, `CAR()` runs an AR(1) recursion
whose lag coefficient decays geometrically with the time gap
$\Delta_t = t - t_{-}$ between the current observation and the
most recent one. This is the Ornstein-Uhlenbeck process
specialised to discrete observations:

\begin{align*}
z_{i, t} & \sim \text{Normal}\!\bigl(
  \rho_i^{\Delta_t} \, z_{i, t_{-}},\;
  \sigma_i \bigr)
\end{align*}

### Piecewise growth: `PW()`

Where the trends above are stochastic, `PW()` is deterministic.
The constructor compiles the Prophet kernel of
[Taylor and Letham (2018)](https://doi.org/10.1080/00031305.2017.1380080).
It fits a piecewise linear or piecewise logistic trajectory with
growth rate changes at a set of changepoints
$\{s_1, \dots, s_S\}$. The lag-zero rate is $k$, the offset is
$m$, and the magnitude shift at changepoint $s_j$ is $\delta_j$.
Let $a(t)$ be the $\mathbb{R}^S$ indicator with
$a_j(t) = \mathbb{1}\{t \geq s_j\}$ and $\gamma$ the
offset-correction vector that keeps the trajectory continuous.
Linear growth:

\begin{align*}
z_t & = \bigl(k + a(t)^\top \delta\bigr) t
       + \bigl(m + a(t)^\top \gamma\bigr)
\end{align*}

Logistic growth uses the same $(k, m, \delta)$ but maps through
a sigmoid with a (possibly time-varying) carrying capacity
$C(t)$:

\begin{align*}
z_t & = \frac{C(t)}{1 + \exp\!\bigl(
  -(k + a(t)^\top \delta)\,(t - (m + a(t)^\top \gamma))
\bigr)}
\end{align*}

### Composing options

Three orthogonal switches layer on top of any per-series trend,
so the six constructors above cover many configurations without
new kernels.

The first is `cor = TRUE`, which switches independent innovations
to correlated innovations through an LKJ prior on the Cholesky
factor of $\Sigma$.

The second is `gr = ` / `subgr = `, which sets up hierarchical
residual correlation matrices. The per-group covariance is
shrunk toward a population matrix through a weight
$\alpha_{\text{cor}} \in (0, 1)$ on the population draw:

\begin{align*}
\Omega_g & = \alpha_{\text{cor}} \, \Omega_{\text{global}}
           + (1 - \alpha_{\text{cor}}) \, \Omega_{g, \text{local}}
\end{align*}

The third is `n_lv < n_series`, which moves the trend into a
factor model. A $K$-dimensional latent process $\tilde{z}_t$
drives the $N_{\text{series}}$-dimensional state through a
loadings matrix
$Z \in \mathbb{R}^{N_{\text{series}} \times K}$:

\begin{align*}
z_t & = Z \, \tilde{z}_t, \qquad
\tilde{z}_t \in \mathbb{R}^K
\end{align*}

Factor structure delivers three advantages over running
$N_{\text{series}}$ independent latent processes. The first is
dimensionality reduction: communities with dozens of correlated
outcomes (species, financial assets, regions) can share dynamics
through a small set of latent axes, cutting parameter count and
identifying joint structure that fully-independent fits miss.

The second is rotation invariance. The marginal covariance
$Z Z^\top$ identifies what the data can resolve about $Z$, and
mvgam follows the
[Heaps and Jermyn (2024)](https://doi.org/10.1007/s11222-024-10454-0)
prior: sample $Z$ unconstrained, then apply a thin QR rotation
in generated quantities to identify $Z$ post hoc. The diagonal
of the rotated loadings is non-negative by construction. This
removes the $2^K$ sign equivalence that constrained factor
models suffer from. For VAR factor models the lag-coefficient
cube rotates the same way:
$\tilde{A}_l = Q \, A_l \, Q^\top$ for the orthogonal $Q$
delivered by the QR factorisation. Per-factor scalar parameters
($\rho_l$, $\sigma$, MA coefficients, $L_{\Omega}$) stay in the
unrotated latent basis.

The third is structured priors. `loadings_prior` accepts either
a Matrix Normal prior built from per-series features and
distance matrices, or the multiplicative gamma process (MGP)
prior of
[Bhattacharya and Dunson (2011)](https://doi.org/10.1093/biomet/asr013).
MGP shrinks later columns of $Z$ toward zero, so `n_lv` becomes a
truncation ceiling on the factor count. Under MGP,
`active_factors()` reports the posterior distribution of the
number of active columns.

See `?mvgam_trends` for the full surface, `?AR`, `?VAR`, `?RW`,
`?ZMVN`, `?CAR`, `?PW` for the per-constructor arguments,
`?loadings_prior` for the structured priors, and
`?ordinate.mvgam` for biplot helpers on the rotated loadings.
Smooth Gaussian process effects on covariates remain available
through `gp()` inside either `formula` or `trend_formula`.

The example below uses `AR(p = c(1, 12))` to fit lag-1 momentum
together with annual recurrence at lag 12 from monthly data,
which is the dynamic structure the simulator was set up to
generate.

## A simulated example

`sim_mvgam(type = 7)` generates a single Gaussian series with a
deterministic seasonal cycle on the obs side and a sparse-lag
`AR(p = c(1, 12))` latent state. The seasonal cycle is
identifiable from the smooth alone. The latent state carries
momentum at lag 1 and recurrence at lag 12 on top of that fixed
cycle. The simulator returns 240 monthly observations and holds
the final 12 back as a test set, so the question "does adding
the latent state help on a forecast the model has never seen?"
has somewhere to land.

```{r sim-data}
set.seed(7)
sim <- sim_mvgam(
  type               = 7L,
  family             = gaussian(),
  n_series           = 1L,
  n_timepoints       = 240L,
  proportional_train = 12 / 13
)
train <- sim$data_train
test  <- sim$data_test
nrow(train)
nrow(test)
head(train)
```

`sim$true_smooths[["s(season)"]]` holds the generative seasonal
function on a fine grid; `sim$true_trend` is the realised
AR(1, 12) trajectory on the link scale. `mvgam_data()` runs the
same validators `mvgam()` would run at fit time and draws the
package's standard EDA view of the response (time series,
histogram, ACF, empirical CDF) with the held-out window overlaid
in black:

```{r sim-plot, fig.asp = 0.85, out.width = "85%"}
mvgam_data(train, family = gaussian(), newdata = test)
```

The annual repetition in the time-series panel is the seasonal
cycle the smooth will fit. The ACF panel shows lag 1 and lag 12
spikes that the smooth alone cannot absorb. The realised latent
state in `sim$true_trend` drifts much wider than the
deterministic cycle itself: `range(sim$true_trend)` runs from
roughly $-9$ to $+8$, where the cycle alone spans only a couple
of units, so a fixed seasonal fit will pick up the average annual
pattern but cannot follow that drift into the held-out year. The
next two sections fit exactly that fixed-seasonal model and then
add the latent state back in to recover what the smooth cannot
see.

## Model 1: seasonal smooth, no latent state

The first model fits the cyclic seasonal smooth and treats
everything else as observation noise:

\begin{align*}
y_t & \sim \text{Normal}(\mu_t, \sigma) \\
\mu_t & = \alpha + s_{\text{season}}(\text{season}_t) \\
\beta_{\text{smooth}} & \sim \text{Normal}\!\bigl(0, (S \lambda)^{-1}\bigr) \\
\sigma & \sim \text{half-}t_3(0, 2.5)
\end{align*}

The cyclic smooth wraps January (1) onto December (12) through
the `bs = "cc"` basis; the `knots` argument places the cycle
endpoints just outside the observed range so the basis treats
month 1 and month 12 as adjacent.

```{r model-step1, results = "hide"}
step1 <- mvgam(
  y ~ s(season, bs = "cc", k = 12),
  knots   = list(season = c(0.5, 12.5)),
  family  = gaussian(),
  data    = train,
  newdata = test,
  chains  = 2,
  samples = 400,
  burnin  = 400,
  silent  = 2
)
```

```{r}
summary(step1, include_betas = FALSE)
```

The headline number is `sigma`: roughly $3.2$ on the response
scale. The cyclic smooth has been pulled toward zero
(`sds_1` posterior median around $0.09$, smooth basis
coefficients all close to zero) because most of the temporal
variation in the response is unstructured to a fit that ignores
time. With nothing else to explain that variation, the obs noise
absorbs it.

```{r step1-cond-effects, fig.asp = 0.5, out.width = "70%"}
conditional_effects(step1)
```

The marginal smooth is essentially flat. The Dunn-Smyth residuals
show why this fit is misspecified:

```{r step1-resid, fig.asp = 0.5, out.width = "80%"}
plot(step1, type = "residuals")
```

Lag-1 and lag-12 autocorrelation are both visible in the residual
ACF. The dynamic structure that the simulator built into the data
has nowhere to go in this fit, so it leaks into the residuals
and pushes `sigma` up. A state-space model would absorb that
signal into a latent state and leave the residuals clean.

## Model 2: add an AR(1, 12) latent state

The second model keeps the seasonal smooth and lets the latent
state carry the structure model 1 missed:

\begin{align*}
y_t & \sim \text{Normal}(\mu_t + z_t, \sigma) \\
\mu_t & = \alpha + s_{\text{season}}(\text{season}_t) \\
z_t & \sim \text{Normal}\!\bigl(\rho_1 \, z_{t - 1} + \rho_{12} \, z_{t - 12},
                               \sigma_{\text{trend}}\bigr) \\
\rho_1, \rho_{12} & \sim \text{Normal}(0, 1)[-1, 1] \\
\sigma_{\text{trend}} & \sim \text{half-}t_3(0, 2.5)
\end{align*}

The obs-side formula is unchanged; the latent state enters
through `trend_formula`:

```{r model-step2, results = "hide"}
step2 <- mvgam(
  y ~ s(season, bs = "cc", k = 12),
  knots         = list(season = c(0.5, 12.5)),
  trend_formula = ~ AR(p = c(1, 12)),
  family        = gaussian(),
  data          = train,
  newdata       = test,
  chains        = 2,
  samples       = 400,
  burnin        = 400,
  silent        = 2
)
```

```{r}
summary(step2, include_betas = FALSE)
```

`sigma` drops from $\approx 3.2$ under model 1 to $\approx 1.1$
here, which is what the partition was supposed to deliver: the
dynamic signal moves out of observation noise and into the
latent state. The trend block reports `ar1_trend` and
`ar12_trend`, which can be checked against the simulator truths:

```{r step2-ar-recovery}
ar1  <- as.numeric(as.array(step2, variable = "ar1_trend[1]"))
ar12 <- as.numeric(as.array(step2, variable = "ar12_trend[1]"))
data.frame(
  parameter = c("ar1_trend", "ar12_trend"),
  truth     = c(0.55, 0.40),
  median    = c(median(ar1),  median(ar12)),
  lower_90  = c(quantile(ar1,  0.05), quantile(ar12, 0.05)),
  upper_90  = c(quantile(ar1,  0.95), quantile(ar12, 0.95)),
  prob_gt_0 = c(mean(ar1 > 0), mean(ar12 > 0))
)
```

Both lag coefficients are clearly above zero, with posterior
probability $1.0$ for each. The 90% intervals cover the
simulator truths of $0.55$ and $0.40$. `mcmc_plot()` returns
the same posteriors as densities:

```{r step2-mcmc, fig.asp = 0.45, out.width = "70%"}
mcmc_plot(step2, variable = "ar[0-9]+_trend",
          regex = TRUE, type = "areas")
```

With the dynamic structure now carried by the latent state, the
Dunn-Smyth residuals no longer show the lag-1 and lag-12
autocorrelation that model 1's residuals carried:

```{r step2-resid, fig.asp = 0.5, out.width = "80%"}
plot(step2, type = "residuals")
```

And the cyclic smooth, freed from the job of explaining
time-varying structure that wasn't seasonal, identifies the
underlying annual cycle:

```{r step2-cond-effects, fig.asp = 0.5, out.width = "70%"}
conditional_effects(step2)
```

## Hindcast versus forecast

The fit produces two posterior trajectories that get separate
names. `hindcast()` returns the trajectory at the training times,
read directly from the Stan-fitted latent state and therefore
deterministic at every draw. `forecast()` propagates the AR(1, 12)
recursion forward from the last fitted state to held-out times,
accumulating uncertainty from `sigma_trend` and the obs noise.
Both return a single `mvgam_forecast` object with the same
structure; the forecast arm is populated only when `newdata`
covers times beyond the training window:

```{r hindcast-forecast}
fc <- forecast(step2, newdata = test)
dim(fc$hindcasts[[1]])
dim(fc$forecasts[[1]])
```

```{r forecast-plot, fig.asp = 0.55, out.width = "80%"}
plot(fc, series = 1)
```

The forecast trajectory is the held-out test the rest of the
analysis hinges on, and the next two sections grade it.

## Forecast scoring across rules

Proper scoring rules grade probabilistic forecasts on calibration
and sharpness in one number. `score()` returns one score per
timepoint from an `mvgam_forecast` object. Summing across the
held-out window gives a single number per model that can be
compared across fits. The Continuous Ranked Probability Score (CRPS) is
the default, but the same call supports the log score (`logs`),
the Dawid-Sebastiani score (`dss`), and the Discrete Ranked
Probability Score (`drps`):

```{r forecast-score}
fc1 <- forecast(step1, newdata = test)
fc2 <- forecast(step2, newdata = test)
score_rules <- c("crps", "logs", "dss", "drps")
scores <- vapply(score_rules, function(rule) {
  s1 <- score(fc1, score = rule)
  s2 <- score(fc2, score = rule)
  c(no_trend = sum(s1[[1]]$score, na.rm = TRUE),
    ar_1_12  = sum(s2[[1]]$score, na.rm = TRUE))
}, numeric(2))
round(scores, 2)
```

A lower score is better. The state-space fit wins on every rule.
CRPS drops by roughly $3.4$, the log score by $1.3$, the
Dawid-Sebastiani score by $4.0$, and the DRPS by $1.9$. DSS shows
the biggest margin because that rule penalises poorly calibrated
forecast variance most aggressively, and the trendless fit stays
overconfident around its (flat) cycle.

`score()` covers additional rules through the `score` argument:
the interval score (`sis`) and the quantile score (`qs`) for
calibration tasks; `twcrps` weights CRPS toward tail behaviour. For
multivariate forecasts from VAR fits or factor models, `score()`
also supports `energy` (energy score) and `variogram` (variogram
score); both penalise dependence patterns that per-series rules
miss. The `interval_width` argument sets the nominal coverage
band used by the interval-based rules.

### Leave-future-out cross-validation

A single train/test split rewards models that happen to win on
one specific horizon. `lfo_cv.mvgam()` automates the alternative:
it scores forecasts at every refit point along the series, then
aggregates the per-step scores across many forecast origins.

```{r lfo-example, fig.asp = 0.55, out.width = "80%", results = "hide"}
lfo_step2 <- lfo_cv(step2, min_t = 200, fc_horizon = 12,
                    silent = 2)
plot(lfo_step2)
```

`min_t` sets the first refit point, `fc_horizon` sets the
prediction window per fold, and the chain is reseated through
`update()` between folds so the Stan program does not recompile.
The default score is the expected log predictive density (ELPD),
returned per fold and per step ahead. The companion
`kfold.mvgam()` runs block CV for grouped designs through
cluster-level holdout. It supports user-defined fold maps
through the `folds` argument.

`loo()` and `loo_compare()` also dispatch on `mvgam` objects,
but both inherit the same pathology when applied to state-space
fits. PSIS-LOO approximates leaving $y_t$ out of the likelihood,
but the chain still conditioned on $y_t$ when it estimated the
latent state $z_t$. The approximate density is therefore
evaluated against a posterior that has already seen the
observation through the latent path; Pareto-$k$ diagnostics
warn about exactly this, and differencing two contaminated
ELPDs in `loo_compare()` does not remove the contamination. The
honest comparison uses either the forecast score above on the
held-out window or `lfo_cv()` below. `lfo_cv()` refits the chain
on shorter prefixes, so the latent state at the scoring time was
never conditioned on its own observation.

## How to cite

`how_to_cite()` returns a methods-section paragraph adapted to
the model's actual structure (likelihood, link, trend type,
sampler configuration, backend):

```{r how-to-cite}
how_to_cite(step2)
```

For a math-only model statement rendered as Markdown plus LaTeX,
use `methods_md(step2)`.

## Where to next

- `?mvgam` for the full fitter signature, including
  `loadings_prior` and `trend_map`.
- `?jsdgam` for joint species distribution and closure-unit
  family fits.
- `?forecast.mvgam` and `?hindcast.mvgam` for the prediction
  surface, including `type = "link" / "expected" / "trend"`.
- `?conditional_effects.mvgam` for the `marginaleffects`-backed
  effect-summary path; `predictions()`, `comparisons()`, and
  `hypotheses()` from the `marginaleffects` package also dispatch
  on `mvgam` objects.
- `?kfold.mvgam` for k-fold CV.
- `?lfo_cv.mvgam` for CV that holds out future timepoints.
- `?irf` and `?fevd` for impulse responses and forecast error
  variance decomposition; `?stability` for eigenvalue checks on
  VAR trends.

## Further reading

Bhattacharya, A. and Dunson, D. B. (2011). Sparse Bayesian
infinite factor models. *Biometrika* 98, 291-306.
[doi:10.1093/biomet/asr013](https://doi.org/10.1093/biomet/asr013)

Bürkner, P.-C. (2017). brms: An R package for Bayesian
multilevel models using Stan. *Journal of Statistical Software*
80(1), 1-28.
[doi:10.18637/jss.v080.i01](https://doi.org/10.18637/jss.v080.i01)

Clark, N. J. and Wells, K. (2023). Dynamic Generalized Additive
Models (DGAMs) for forecasting discrete ecological time series.
*Methods in Ecology and Evolution* 14, 771-784.
[doi:10.1111/2041-210X.13974](https://doi.org/10.1111/2041-210X.13974)

Clark, N. J., Ernest, S. K. M., Senyondo, H., Simonis, J.,
White, E. P., Yenni, G. M. and Karunarathna, K. A. N. K. (2025).
Beyond single-species models: multispecies forecasts for
ecological predictability. *PeerJ* 13, e18929.
[doi:10.7717/peerj.18929](https://doi.org/10.7717/peerj.18929)

Heaps, S. E. (2023). Enforcing stationarity through the prior in
vector autoregressions. *Journal of Computational and Graphical
Statistics* 32, 74-83.
[doi:10.1080/10618600.2022.2079648](https://doi.org/10.1080/10618600.2022.2079648)

Heaps, S. E. and Jermyn, I. H. (2024). Structured prior
distributions for the covariance matrix in latent factor models.
*Statistics and Computing* 34, 143.
[doi:10.1007/s11222-024-10454-0](https://doi.org/10.1007/s11222-024-10454-0)

MacKenzie, D. I., Nichols, J. D., Lachman, G. B., Droege, S.,
Royle, J. A. and Langtimm, C. A. (2002). Estimating site
occupancy rates when detection probabilities are less than one.
*Ecology* 83(8), 2248-2255. DOI:
<https://doi.org/10.1890/0012-9658(2002)083[2248:ESORWD]2.0.CO;2>

Royle, J. A. (2004). N-mixture models for estimating population
size from spatially replicated counts. *Biometrics* 60, 108-115.
[doi:10.1111/j.0006-341X.2004.00142.x](https://doi.org/10.1111/j.0006-341X.2004.00142.x)

Royle, J. A. and Nichols, J. D. (2003). Estimating abundance
from repeated presence-absence data or point counts. *Ecology*
84(3), 777-790. DOI:
<https://doi.org/10.1890/0012-9658(2003)084[0777:EAFRPA]2.0.CO;2>

Taylor, S. and Letham, B. (2018). Forecasting at scale. *The
American Statistician* 72(1), 37-45.
[doi:10.1080/00031305.2017.1380080](https://doi.org/10.1080/00031305.2017.1380080)

Wood, S. N. (2017). *Generalized Additive Models: An Introduction
with R* (2nd edition). Chapman and Hall/CRC.
[doi:10.1201/9781315370279](https://doi.org/10.1201/9781315370279)
