---
title: "Overview of the mvgam package"
author: "Nicholas J Clark"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: yes
vignette: >
  %\VignetteIndexEntry{Overview of the mvgam package}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
params:
  EVAL: !r identical(tolower(Sys.getenv("NOT_CRAN")), "true")
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  eval = if (isTRUE(exists("params"))) params$EVAL else FALSE
)
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  dpi = 100,
  fig.asp = 0.8,
  fig.width = 6,
  out.width = "60%",
  fig.align = "center"
)
library(mvgam)
library(ggplot2)
theme_set(theme_bw(base_size = 12, base_family = "serif"))
```

`mvgam` fits, plots and forecasts Bayesian dynamic generalized
additive models. The package combines `mgcv`-style smooth terms
with `brms`-style observation families and a library of latent
trend specifications (random walks, autoregressions, vector
autoregressions, continuous-time AR, piecewise growth, Gaussian
processes). The fitted models are state-space: an observation
process maps a latent state to the response through a chosen
likelihood, and a trend process governs how the latent state
evolves over time.

This vignette walks through the package on a simulated dataset
generated by `sim_mvgam()`. The simulator returns the true smooth
functions and latent trajectory it drew from, so the recovery of
every component can be checked directly. The pedagogical question
is sharp: when the data has both a deterministic seasonal cycle
and a stochastic dynamic process beyond that cycle, what does a
fixed-seasonality smooth miss, and what does adding a latent state
buy?

## State-space formulation

For an observation $y_{i, t}$ on series $i$ at time $t$ with link
function $g$ and observation family $\mathcal{F}$:

\begin{align*}
y_{i, t} & \sim \mathcal{F}(\tilde{y}_{i, t}, \phi) \\
g(\tilde{y}_{i, t}) & = \alpha_i +
  \sum_{j=1}^{J} s_{i, j}(x_{j, t}) +
  Z_{i, \cdot} \, z_{\cdot, t} \, .
\end{align*}

The intercept $\alpha_i$ and smooths $s_{i, j}$ form the
observation-side linear predictor. The latent state $z_{k, t}$
evolves through a trend process (AR(1), VAR, RW, CAR, GP and so
on); $Z$ is a loadings matrix mapping the $K$-dimensional latent
state onto the $N$ series. With $K = N$ and $Z = I$ each series
gets its own latent state; with $K < N$ the latent states are
shared factors. Smooths and covariate effects passed through
`trend_formula` sit on the latent state; everything in the
top-level `formula` sits on the response.

## Supported observation families

Every observation family in `brms` is supported through the brms
integration layer (`gaussian()`, `student()`, `poisson()`,
`negbinomial()`, `binomial()`, `beta_binomial()`,
`bernoulli()`, `Beta()`, `Gamma()`, `lognormal()`, and the rest).
`mvgam` adds a small library of families that brms does not ship
natively:

| Family | Use case | Trend dpars |
|:------:|:---------|:-----------:|
| `tweedie()` | Compound Poisson-Gamma counts with zero inflation | `phi`, `power` |
| `com_binomial()` | Conway-Maxwell-Binomial for over-, under-, super-dispersed bounded counts | `nu` |
| `nmix()` | N-mixture detection-error models (Poisson-Binomial, Royle-Nichols, Poisson-Poisson variants) | family-specific |
| `occ()` | Single-season occupancy with imperfect detection | `psi` |
| `diri()` | Dirichlet compositional responses | `phi` |
| `multi()`, `categ()` | Multinomial and categorical responses on the closure-unit grain | -- |
| `mvn()`, `mvt()` | Multivariate normal / Student-t residual covariance via gllvm-style factor structure | `Psi` |

The closure-unit and multi-response families
(`nmix`, `occ`, `diri`, `multi`, `categ`, `mvn`, `mvt`) typically
go through `jsdgam()` rather than `mvgam()`; see the
per-constructor help pages for worked examples.

## Latent trend processes

The trend constructors live in the `trend_formula` argument:

\begin{align*}
\text{RW:} \quad & z_{i, t} \sim \text{Normal}(z_{i, t - 1}, \sigma_i) \\
\text{AR(p):} \quad & z_{i, t} \sim \text{Normal}\!\Bigl(\sum_{l \in \mathcal{L}} \rho_{i, l} \, z_{i, t - l}, \sigma_i\Bigr) \\
\text{VAR(p):} \quad & z_t \sim \text{Normal}\!\Bigl(\sum_{l=1}^{p} A_l \, z_{t - l}, \Sigma\Bigr) \\
\text{ZMVN:} \quad & z_t \sim \text{Normal}(0, \Sigma) \\
\text{CAR(1):} \quad & z_{i, t} \sim \text{Normal}\!\Bigl(\rho_i^{\Delta t} \, z_{i, t - 1}, \sigma_i\Bigr) \\
\text{GP:} \quad & z \sim \text{Normal}(0, K_{\alpha, \rho})
\end{align*}

The lag set $\mathcal{L}$ for `AR(p)` is `1:p` when `p` is a
scalar and `p` itself when `p` is a vector. The example below
uses `AR(p = c(1, 12))` to fit lag-1 momentum together with
year-on-year recurrence at lag 12 from monthly data. Multivariate
options compose through `cor = TRUE` (correlated innovations),
`gr = ` and `subgr = ` (hierarchical partial pooling across
groups), and factor structure (`n_lv < n_series`, with post-hoc
QR identification of the loadings matrix following Heaps and
Jermyn, 2024). Piecewise linear and logistic growth are available
through `PW()` for Prophet-style changepoint trends. The full set
is documented at `?mvgam_trends`; constructor arguments live at
`?AR`, `?VAR`, `?RW`, `?ZMVN`, `?CAR`, `?PW`, `?GP`.

## A simulated example

`sim_mvgam(type = 7)` generates a single Gaussian series with a
deterministic seasonal cycle on the obs side and a sparse-lag
`AR(p = c(1, 12))` latent state. The seasonal cycle is
identifiable from the smooth alone; the latent state carries
short-term momentum (lag 1) and year-on-year recurrence (lag 12)
beyond the fixed cycle. The simulator returns 20 years of
monthly data and a 12-month held-out tail so the question "does
adding the latent state buy us anything on the out-of-sample
forecast?" has somewhere to land.

```{r sim-data}
set.seed(7)
sim <- sim_mvgam(
  type               = 7L,
  family             = gaussian(),
  n_series           = 1L,
  n_timepoints       = 240L,
  proportional_train = 12 / 13
)
train <- sim$data_train
test  <- sim$data_test
nrow(train)
nrow(test)
head(train)
```

`sim$true_smooths[["s(season)"]]` holds the generative seasonal
function on a fine grid; `sim$true_trend` is the realised
AR(1, 12) trajectory on the link scale. The two together drive
the response:

```{r sim-plot, fig.asp = 0.55, out.width = "80%"}
trend_df <- data.frame(
  time = seq_along(sim$true_trend[, 1L]),
  z    = sim$true_trend[, 1L]
)
ggplot() +
  geom_line(data = train, aes(time, y), colour = "grey40") +
  geom_point(data = train, aes(time, y), size = 0.6,
             colour = "grey20") +
  geom_line(data = trend_df, aes(time, z), colour = "steelblue",
            linewidth = 0.8) +
  geom_vline(xintercept = max(train$time), linetype = "dashed",
             colour = "grey50") +
  labs(x = "time (months)", y = "y",
       caption = paste0(
         "Black: response. Blue: true AR(1, 12) latent state ",
         "(link scale). Vertical line: end of training."
       )) +
  theme(plot.caption = element_text(hjust = 0))
```

The seasonal cycle is visible in the annual repetition of the
response. The latent state (blue line) drifts substantially
across the 240-month window: $z$ ranges from about $-9$ to $+8$,
much wider than the deterministic cycle itself. A fixed seasonal
fit will pick up the average annual pattern but cannot follow
that drift into the held out year.

## Model 1: seasonal smooth, no latent state

The first model fits the cyclic seasonal smooth and treats
everything else as observation noise:

\begin{align*}
y_t & \sim \text{Normal}(\mu_t, \sigma) \\
\mu_t & = \alpha + s_{\text{season}}(\text{season}_t) \\
\beta_{\text{smooth}} & \sim \text{Normal}\!\bigl(0, (S \lambda)^{-1}\bigr) \\
\sigma & \sim \text{half-Student-t}(3, 0, 2.5)
\end{align*}

The cyclic smooth wraps January (1) onto December (12) through
the `bs = "cc"` basis; the `knots` argument places the cycle
endpoints just outside the observed range so the basis treats
month 1 and month 12 as adjacent.

```{r model-step1, results = "hide"}
step1 <- mvgam(
  y ~ s(season, bs = "cc", k = 12),
  knots   = list(season = c(0.5, 12.5)),
  family  = gaussian(),
  data    = train,
  newdata = test,
  chains  = 2,
  samples = 400,
  burnin  = 400,
  silent  = 2
)
```

```{r}
summary(step1, include_betas = FALSE)
```

The headline number is `sigma`: roughly $3.2$ on the response
scale. The cyclic smooth has been pulled toward zero
(`sds_1` posterior median around $0.09$, smooth basis
coefficients all close to zero) because most of the temporal
variation in the response is unstructured to a fit that ignores
time. With nothing else to explain that variation, the obs noise
absorbs it.

```{r step1-cond-effects, fig.asp = 0.5, out.width = "70%"}
conditional_effects(step1)
```

The marginal smooth is essentially flat. The Dunn-Smyth residuals
show why this fit is misspecified:

```{r step1-resid, fig.asp = 0.5, out.width = "80%"}
plot(step1, type = "residuals")
```

Lag-1 and lag-12 autocorrelation are both visible in the residual
ACF. Model 1 discards this structure entirely, forcing the lag-1
and lag-12 signal into residual noise, a sign of fundamental
misspecification when a state-space model would absorb that
signal into a latent state.

## Model 2: add an AR(1, 12) latent state

The second model keeps the seasonal smooth and lets the latent
state carry the structure model 1 missed:

\begin{align*}
y_t & \sim \text{Normal}(\mu_t + z_t, \sigma) \\
\mu_t & = \alpha + s_{\text{season}}(\text{season}_t) \\
z_t & \sim \text{Normal}\!\bigl(\rho_1 \, z_{t - 1} + \rho_{12} \, z_{t - 12},
                               \sigma_{\text{trend}}\bigr) \\
\rho_1, \rho_{12} & \sim \text{Normal}(0, 1)[-1, 1] \\
\sigma_{\text{trend}} & \sim \text{half-Student-t}(3, 0, 2.5)
\end{align*}

The obs-side formula is unchanged; the latent state enters
through `trend_formula`:

```{r model-step2, results = "hide"}
step2 <- mvgam(
  y ~ s(season, bs = "cc", k = 12),
  knots         = list(season = c(0.5, 12.5)),
  trend_formula = ~ AR(p = c(1, 12)),
  family        = gaussian(),
  data          = train,
  newdata       = test,
  chains        = 2,
  samples       = 400,
  burnin        = 400,
  silent        = 2
)
```

```{r}
summary(step2, include_betas = FALSE)
```

Model 1 misattributed the dynamic structure to observation noise
and reported $\sigma \approx 3.2$; model 2 correctly partitions
that signal into the latent state and brings $\sigma$ down to
$\approx 1.1$. The trend
block reports `ar1_trend` and `ar12_trend`:

```{r step2-ar-recovery}
ar1  <- as.numeric(as.array(step2, variable = "ar1_trend[1]"))
ar12 <- as.numeric(as.array(step2, variable = "ar12_trend[1]"))
data.frame(
  parameter = c("ar1_trend", "ar12_trend"),
  truth     = c(0.55, 0.40),
  median    = c(median(ar1),  median(ar12)),
  lower_90  = c(quantile(ar1,  0.05), quantile(ar12, 0.05)),
  upper_90  = c(quantile(ar1,  0.95), quantile(ar12, 0.95)),
  prob_gt_0 = c(mean(ar1 > 0), mean(ar12 > 0))
)
```

Both lag coefficients are clearly above zero, with posterior
probability $1.0$ for each. The 90% intervals cover the simulator
truths of $0.55$ and $0.40$. `mcmc_plot()` returns the same
posteriors as densities:

```{r step2-mcmc, fig.asp = 0.45, out.width = "70%"}
mcmc_plot(step2, variable = "ar[0-9]+_trend",
          regex = TRUE, type = "areas")
```

The Dunn-Smyth residuals no longer show the lag-1 and lag-12
autocorrelation that model 1's residuals carried:

```{r step2-resid, fig.asp = 0.5, out.width = "80%"}
plot(step2, type = "residuals")
```

The cyclic smooth, freed from the job of explaining
time-varying structure, now identifies the seasonal cycle
through the smoother penalty:

```{r step2-cond-effects, fig.asp = 0.5, out.width = "70%"}
conditional_effects(step2)
```

## Hindcast versus forecast

`hindcast()` returns the posterior trajectory at the training
times. `forecast()` propagates the latent state forward to held
out times. Both return a single `mvgam_forecast` object with the
same structure; the forecast arm is populated only when
`newdata` covers times beyond the training window:

```{r hindcast-forecast}
fc <- forecast(step2, newdata = test)
dim(fc$hindcasts[[1]])
dim(fc$forecasts[[1]])
```

```{r forecast-plot, fig.asp = 0.55, out.width = "80%"}
plot(fc, series = 1)
```

The hindcast trajectory reads the Stan-fitted latent state
directly (deterministic, exact). The forecast runs the AR(1, 12)
recursion forward from the last fitted state, accumulating
uncertainty from `sigma_trend` and the obs noise.

## Forecast scoring across rules

PSIS-LOO is unreliable for comparing these two fits. Leaving an
observation out while keeping the latent state estimated from it
raises the in-sample density and rewards the wrong model. The
honest yardstick is the held-out forecast score. `score()`
computes proper scoring rules per timepoint; the sum across the
held out set ranks models on out-of-sample predictive
performance. The Continuous Ranked Probability Score is the
default, but the same call works for the log score (`logs`),
the Dawid-Sebastiani score (`dss`), and the Discrete Ranked
Probability Score (`drps`):

```{r forecast-score}
fc1 <- forecast(step1, newdata = test)
fc2 <- forecast(step2, newdata = test)
score_rules <- c("crps", "logs", "dss", "drps")
scores <- vapply(score_rules, function(rule) {
  s1 <- score(fc1, score = rule)
  s2 <- score(fc2, score = rule)
  c(no_trend = sum(s1[[1]]$score, na.rm = TRUE),
    ar_1_12  = sum(s2[[1]]$score, na.rm = TRUE))
}, numeric(2))
round(scores, 2)
```

A lower score is better. The state-space fit wins on every rule.
CRPS drops by roughly $3.4$, the log score by $1.3$, the
Dawid-Sebastiani score by $4.0$, and the DRPS by $1.9$. DSS shows
the biggest margin because that rule penalises poorly calibrated
forecast variance most aggressively, and the trendless fit stays
overconfident around its (flat) cycle.

## How to cite

`how_to_cite()` returns a methods-section paragraph adapted to
the model's actual structure (likelihood, link, trend type,
sampler configuration, backend):

```{r how-to-cite}
how_to_cite(step2)
```

For a math-only model statement rendered as Markdown plus LaTeX,
use `methods_md(step2)`.

## Where to next

- `?mvgam` for the full fitter signature, including
  `loadings_prior`, `trend_map`, and threading options.
- `?jsdgam` for joint species distribution and closure-unit
  family fits.
- `?forecast.mvgam` and `?hindcast.mvgam` for the prediction
  surface, including `type = "link" / "expected" / "trend"`.
- `?conditional_effects.mvgam` for the `marginaleffects`-backed
  effect-summary path; `predictions()`, `comparisons()`, and
  `hypotheses()` from the `marginaleffects` package also dispatch
  on `mvgam` objects.
- `?kfold.mvgam` and `?lfo_cv.mvgam` for k-fold and
  leave-future-out cross-validation.
- `?irf`, `?fevd`, `?stability` for impulse-response, forecast
  error variance decomposition, and stability summaries for VAR
  trends.

## Further reading

Clark, N. J. and Wells, K. (2023). Dynamic Generalized Additive
Models for forecasting discrete ecological time series.
*Methods in Ecology and Evolution* 14, 771-784.

Clark, N. J., Ernest, S. K. M., Senyondo, H., Simonis, J.,
White, E. P., Yenni, G. M. and Karunarathna, K. A. N. K. (2025).
Beyond single-species models: multispecies forecasts for
ecological predictability. *PeerJ* 13:e18929.

Heaps, S. E. and Jermyn, I. H. (2024). Structured prior
distributions for the covariance matrix in latent factor models.
*Statistics and Computing* 34:143.

Burkner, P.-C. (2017). brms: An R package for Bayesian multilevel
models using Stan. *Journal of Statistical Software* 80(1):1-28.

Wood, S. N. (2017). *Generalized Additive Models: An Introduction
with R* (2nd edition). Chapman and Hall/CRC.
