---
title: "N-mixture models with mvgam"
author: "Nicholas J Clark"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: yes
vignette: >
  %\VignetteIndexEntry{N-mixture models with mvgam}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
params:
  EVAL: !r identical(tolower(Sys.getenv("NOT_CRAN")), "true")
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  eval = if (isTRUE(exists("params"))) params$EVAL else FALSE
)
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  dpi = 100,
  fig.asp = 0.7,
  fig.width = 6,
  out.width = "70%",
  fig.align = "center"
)
library(mvgam)
library(ggplot2)
theme_set(theme_bw(base_size = 12, base_family = "serif"))
```

N-mixture models infer the true abundance of a closed population
from repeated counts that are recorded with imperfect detection.
The model partitions an observed count $y_{i, r}$ into a latent
state (the unobserved abundance $N_i$ at site $i$) and a per-visit
detection process that records each individual independently with
probability $p_r$. This vignette walks through fitting the
single-season Poisson-binomial N-mixture in `mvgam`, recovering
the latent abundance against a known truth, and checking model
fit with closure-unit-aware goodness-of-fit statistics. A
two-species worked example carries the discussion; the same
`bf(y ~ ..., p ~ ...)` formula extends to any number of species.

## N-mixture models

For an observation $y_{i, r}$ on site $i$ at visit $r$, the
single-season Poisson-binomial N-mixture of
[Royle (2004)](https://doi.org/10.1111/j.0006-341X.2004.00142.x)
factors the data-generating process into a state layer and an
observation layer:

\begin{align*}
y_{i, r} & \sim \text{Binomial}(N_i, p_r) \\
N_i      & \sim \text{Poisson}(\lambda_i) \, .
\end{align*}

Covariates enter through the linear predictors for the abundance
mean $\lambda_i$ (log link) and the detection probability $p_r$
(logit link):

\begin{align*}
\log \lambda_i  & = \beta_0 + \beta^\top \mathbf{x}_i \\
\text{logit}\, p_r & = \gamma_0 + \gamma^\top \mathbf{q}_r \, .
\end{align*}

State and observation are conditionally independent given $N_i$.
A single visit cannot identify $\lambda$ and $p$ separately, but
replicated visits at the same site under the closure assumption
let the binomial mean $N_i p_r$ and the binomial variance
$N_i p_r (1 - p_r)$ jointly inform both parameters.

The latent $N_i$ is integer-valued, which leaves Hamiltonian Monte
Carlo nothing to differentiate. `mvgam` handles this the same way
the Stan documentation handles any discrete latent variable: the
likelihood is marginalised analytically, summing over a finite
range $N_i \in \{y_{i, \max}, \dots, K_{\max}\}$ where $K_{\max}$
is a per-site upper bound the user supplies (via the `cap`
column). The chain only ever samples the continuous parameters
on the right-hand side of the linear predictors. For a friendly
walk-through of marginalising discrete parameters in Stan, see
the chapter on
[Latent Discrete Parameters in the Stan User's Guide](https://mc-stan.org/docs/stan-users-guide/latent-discrete.html)
and McElreath's
[Algebra and the Missing Oxen](https://elevanth.org/blog/2018/01/29/algebra-and-missingness/)
walkthrough.

## Closure-unit data shape

Closure-unit families need the data in a long format that
preserves the closure assumption: one row per `(species, site,
visit)` cell, with the species identifier in `series`, the
closure-unit identifier in `time`, the visit index in `visit`,
the count in `y`, and the per-unit upper truncation for the
marginalisation in `cap`. The columns are required because
[`mvgam()`](https://nicholasjclark.github.io/mvgam/reference/mvgam.html)
needs to align observations into the marginalised likelihood
without the user pivoting to and from a multidimensional array.

The `sim_closure_unit_data()` helper generates a data frame in
this shape together with the true generative parameters, so
recovery checks read directly off `truth$lambda`, `truth$N`,
`truth$p`, `truth$state_coefs`, and `truth$detection_coefs`. Each
recipe carries a per-site state-level covariate set and a
per-visit detection-level covariate set; recipe 1 uses a single
covariate on each side (`env` on the state side, `tod_c` on the
detection side) and is the cleanest entry point for an
N-mixture example.

## A two-species worked example

Two species, 60 sites, four visits per site. Recipe 1 carries
the site-level covariate `env ~ N(0, 1)` on the state side and
the visit-level covariate `tod_c ~ U(-6, 6)` on the detection
side. The simulator seeds the coefficient draws and the visit
covariate values, returning the generative truth on `simdat$truth`.

```{r sim-data}
set.seed(0)
simdat <- sim_closure_unit_data(
  family    = nmix(),
  type      = 1L,
  n_species = 2L,
  n_sites   = 60L,
  n_visits  = 4L,
  seed      = 130L,
  K_max     = 30L
)
str(simdat$data_train)
head(simdat$truth$state_coefs)
head(simdat$truth$detection_coefs)
```

The data frame is ready for [`mvgam()`](https://nicholasjclark.github.io/mvgam/reference/mvgam.html)
without further manipulation. The truth bundle records the true
per-species intercepts and slopes on the state side
(`state_coefs`) and the detection side (`detection_coefs`), the
true per-site expected abundance `lambda` and realised abundance
`N`, and the true per-visit detection probability `p`. Recovery
checks throughout the rest of the vignette point back to these
slots.

### Fit

The fit combines two `brms`-style formulas inside a `bf()` call.
The first arm gives the state-side linear predictor for $\log
\lambda$; the second arm gives the detection-side linear
predictor for $\text{logit}\, p$. Each arm includes a `series`
main effect to estimate per-species intercepts plus an
interaction with the covariate (`env:series`, `tod_c:series`) so
each species gets its own slope:

```{r fit, results = "hide"}
mod <- mvgam(
  formula = bf(y ~ series + env:series,
                p ~ series + tod_c:series),
  data    = simdat$data_train,
  family  = nmix(),
  chains  = 2L,
  samples = 500L,
  burnin  = 500L,
  silent  = 2L
)
```

```{r}
summary(mod, include_betas = FALSE)
```

The chain mixed cleanly: all R-hat values sit at 1.00 and bulk
effective sample sizes reach into the thousands across the
linear-predictor coefficients, so the posterior carries enough
independent draws to support reliable inference about each
species' abundance and detection process. The next section
checks how well the recovered coefficients match the simulator's
true parameters.

### Recovery against truth

`as.matrix(mod$fit)` exposes the posterior draws under the
brms parameter names, which is what the recovery check uses. The
`b[k]` indexing follows the column order of the design matrix
that `mvgam()` built from the formula (`mod$standata$X` for the
state side, `mod$standata$X_p` for the detection side):

```{r recovery}
post <- as.matrix(mod$fit)
truth_state <- simdat$truth$state_coefs
truth_det   <- simdat$truth$detection_coefs

recovery <- data.frame(
  parameter = c("sp_1 log-lambda intercept",
                "sp_2 - sp_1 intercept",
                "sp_1 env slope",
                "sp_2 env slope",
                "sp_1 logit-p intercept",
                "sp_2 - sp_1 detection intercept",
                "sp_1 tod_c slope",
                "sp_2 tod_c slope"),
  truth     = c(truth_state[1, "intercept"],
                truth_state[2, "intercept"] -
                  truth_state[1, "intercept"],
                truth_state[1, "env"],
                truth_state[2, "env"],
                truth_det[1, "intercept"],
                truth_det[2, "intercept"] -
                  truth_det[1, "intercept"],
                truth_det[1, "tod_c"],
                truth_det[2, "tod_c"]),
  median    = c(median(post[, "Intercept"]),
                median(post[, "b[1]"]),
                median(post[, "b[2]"]),
                median(post[, "b[3]"]),
                median(post[, "Intercept_p"]),
                median(post[, "b_p[1]"]),
                median(post[, "b_p[2]"]),
                median(post[, "b_p[3]"]))
)
recovery$bias <- recovery$median - recovery$truth
recovery[, -1L] <- lapply(recovery[, -1L],
                           function(x) round(x, 2L))
recovery
```

Detection slopes (`tod_c:series`) recover precisely, with bias
near zero on both species. The state-side intercepts and one of
the env slopes carry a small bias on the order of a few tenths.
The bias is the classic N-mixture identifiability footprint:
detection probability and abundance compensate for each other in
the binomial mean
$\mathbb{E}[Y_{i, r}] = N_i p_r$, so a low $p$ shifts the
posterior toward a higher $\lambda$ at the same data. With four
visits per site this confound rarely resolves to zero bias even
with otherwise well-specified data;
[Knape and Korner-Nievergelt (2015)](https://doi.org/10.1111/2041-210X.12377)
discuss this and the conditions that mitigate it (more visits,
informative priors, abundance covariates with strong effects on
the log-lambda scale). The latent abundance posterior remains
well-calibrated under this bias, which is shown next.

### Latent abundance: hindcast vs truth

The point of fitting an N-mixture is the latent abundance $N$,
not the regression coefficients per se. `hindcast(mod, type =
"latent_state")` returns the posterior at every closure unit as
an `mvgam_latent_state` object that prints a short summary and
hands back a tidy data frame through `summary()` (or its alias
`as.data.frame()`):

```{r latent-N-hindcast}
hc <- hindcast(mod, type = "latent_state")
print(hc)
head(as.data.frame(hc), 6)
```

`plot()` on the same object returns a ggplot ribbon (step-style
for integer-valued `N`) faceted by species. Because the output
is a plain ggplot, the simulator truth can be overlaid as a
points layer with `+ geom_point()` without writing any plotting
code by hand:

```{r latent-N-plot, fig.asp = 0.5, out.width = "85%"}
truth_long <- data.frame(
  series = factor(rep(c("sp_1", "sp_2"), each = 60L),
                  levels = c("sp_1", "sp_2")),
  time   = rep(seq_len(60L), 2L),
  true_N = c(simdat$truth$N[1L, ], simdat$truth$N[2L, ])
)
plot(hc) +
  geom_point(data = truth_long,
              aes(x = time, y = true_N),
              colour = "black", size = 1.3) +
  labs(x = "Closure unit (site)",
        y = expression("Latent abundance " * N),
        title = "Posterior latent N (ribbons) vs simulator truth (points)")
```

Almost every truth point sits inside the 95% credible ribbon and
most sit inside the 50% ribbon. Two diagnostics quantify what
the figure shows:

```{r latent-N-coverage}
smry <- as.data.frame(hc)
merged <- merge(smry, truth_long, by = c("series", "time"))
mean(merged$true_N >= merged$lower_50 &
       merged$true_N <= merged$upper_50)
mean(merged$true_N >= merged$lower_95 &
       merged$true_N <= merged$upper_95)
```

The 95% credible interval contains the simulator's true $N$ at
roughly the nominal rate. The 50% interval is slightly over-wide,
which is the expected face of the coefficient confound discussed
above: when the model spreads uncertainty between $\lambda$ and
$p$ rather than collapsing onto the true pair, the joint
posterior on $N$ widens to match. The headline is that the
recovered latent abundance lines up with truth even when
individual regression coefficients carry small biases.

### Covariate effects on abundance and detection

`conditional_effects()` returns a partial-effect plot per
covariate-by-species combination on the response scale. Because
`y` is a count from the N-mixture, the response scale is the
expected count $\mathbb{E}[Y] = \lambda \cdot p$ per visit;
covariate slopes on $\lambda$ and on $p$ both contribute. To
read each side separately, the `marginaleffects` interface
exposes the closure-unit `latent_state` and `detection` types
directly:

```{r ce-link, fig.asp = 0.55, out.width = "80%"}
conditional_effects(mod)
```

The state-side panel shows expected abundance rising or falling
with `env` differently across the two species, matching the
sign of the simulator's true `env` slopes (sp_1 positive, sp_2
also positive but flatter). The detection-side panel mirrors the
recovered `tod_c` effect: detection probability decreases as
`tod_c` rises, with similar slopes on both species.

`conditional_effects()` puts both species on the same panel by
mixing their marginal expected counts. To inspect detection
probability on its native probability scale, with each species
on its own curve, `marginaleffects::plot_predictions()` with
`type = "detection"` reads the per-visit detection process out
of the fit and plots it directly:

```{r me-detection, fig.asp = 0.55, out.width = "80%"}
marginaleffects::plot_predictions(
  mod,
  condition = c("tod_c", "series"),
  type      = "detection"
) +
  ylab("Pr(detection per visit)") +
  ylim(c(0, 1))
```

Both species' detection curves slope downward in `tod_c` and stay
inside the unit interval across the range of the visit-time
covariate. The curves separate cleanly. Sp_2's curve sits higher
than sp_1's at every value of `tod_c`, which means sp_2 is the
more readily detected species whatever the time of day. The gap
between the two intercepts is what the `series` term on the
detection sub-formula picked up, and it matches the positive
`seriessp_2` detection contrast that the recovery table
reported.

### Goodness-of-fit: closure-unit Bayesian p-values

`pp_check(type = "fit_stat")` runs the
[MacKenzie and Bailey (2004)](https://doi.org/10.1198/108571104X3361)
closure-unit goodness-of-fit test. The statistic collapses the
visit-level counts within each closure unit to a sufficient
discrepancy (chi-squared by default; Freeman-Tukey as an
alternative that handles small counts better), then compares the
observed
discrepancy to its posterior predictive distribution. A Bayesian
p-value near 0.5 means the observed discrepancy is typical under
the model; values near 0 or 1 indicate that the model
under-disperses or over-disperses the observed counts.

```{r ppc-chi}
pp_check(mod, type = "fit_stat", stat = "chi_squared",
          ndraws = 200L)
```

The chi-squared Bayesian p-value lands in the centre of the null
distribution, well away from 0 or 1. The model's per-unit
predicted counts span the same range of dispersion as the
observed counts, so there is no sign that the
Binomial-Poisson mixture is over- or under-dispersing the data.
Chi-squared weights every cell equally and can react harshly to
the small counts that closure-unit data typically carry, so the
Freeman-Tukey discrepancy is the natural cross-check:

```{r ppc-ft}
pp_check(mod, type = "fit_stat", stat = "freeman_tukey",
          ndraws = 200L)
```

The Freeman-Tukey p-value also lands away from the boundaries.
Both discrepancies agree. The closure-unit likelihood is
describing both the hidden abundance process and the per-visit
detection layer well, even with the modest intercept bias the
recovery table documented, and the latent abundance coverage
shown earlier confirms it.

## Alternative observation processes

The Poisson-binomial assumption that detection is an independent
Bernoulli trial for each of $N$ individuals is the most common
N-mixture formulation, but it is not the only one. `nmix()`
exposes two further variants chosen by the `type` argument; the
Stan implementation marginalises latent $N$ for each of them and
the post-fit surface (latent-state hindcast, detection
predictions, GOF) carries through unchanged.

| Variant | Constructor | Detection scale | When to prefer |
|---------|-------------|-----------------|----------------|
| Poisson-binomial | `nmix()` | logit $p$ per visit | Default. Direct Royle (2004) parameterisation; cleanest interpretation when individuals are independently detectable. |
| Royle-Nichols | `nmix("royle_nichols")` | log $r$ per individual | Repeated presence-absence rather than counts. Heterogeneous detection across individuals collapses into a per-individual hazard $r$; the binary observation arises from $1 - (1 - r)^N$. |
| Poisson-Poisson | `nmix("poisson_poisson")` | log $\rho$ per visit | Encounter-rate data where the observed count is a noisy Poisson realisation of an underlying Poisson encounter rate, with no a-priori detection ceiling. |

Switching variants is a one-line change in the `family`
argument; the data shape and formula stay the same.

```{r variant-switch, eval = FALSE}
mod_rn  <- mvgam(formula = bf(y ~ series + env:series,
                                 p ~ series + tod_c:series),
                 data    = simdat$data_train,
                 family  = nmix("royle_nichols"),
                 chains  = 2L, samples = 500L,
                 burnin  = 500L, silent = 2L)

mod_ppm <- mvgam(formula = bf(y ~ series + env:series,
                                 p ~ series + tod_c:series),
                 data    = simdat$data_train,
                 family  = nmix("poisson_poisson"),
                 chains  = 2L, samples = 500L,
                 burnin  = 500L, silent = 2L)
```

The Royle-Nichols variant follows
[Royle and Nichols (2003)](https://doi.org/10.1890/0012-9658\(2003\)084%5B0777:EAFRPA%5D2.0.CO;2);
the Poisson-Poisson variant is the encounter-rate parameterisation
adopted for `mvgam`. See `?nmix` for the constructor's full
argument list, including the `multi_season` opt-in for fits with
a real seasonal axis.

## Where to next

- `?nmix` for the constructor's family-specific dpars,
  identification caveats, and the per-variant kernel references.
- `?occ` for the single-season occupancy family that shares the
  closure-unit data shape with `nmix`.
- `?sim_closure_unit_data` for the simulator's recipe catalogue
  and the `truth$` bundle layout.
- `?jsdgam` for joint species distribution fits with a latent
  factor structure (`n_lv > 0`) on the abundance side, useful
  when residual cross-species correlation needs explicit
  modelling.
- The
  [overview vignette](https://nicholasjclark.github.io/mvgam/articles/mvgam_overview.html)
  for the state-space architecture that the closure-unit families
  fit into.

## Further reading

Knape, J. and Korner-Nievergelt, F. (2015). Estimates from
non-replicated population surveys rely on critical assumptions.
*Methods in Ecology and Evolution* 6, 902-906.
[doi:10.1111/2041-210X.12377](https://doi.org/10.1111/2041-210X.12377)

MacKenzie, D. I. and Bailey, L. L. (2004). Assessing the fit of
site-occupancy models. *Journal of Agricultural, Biological, and
Environmental Statistics* 9, 300-318.
[doi:10.1198/108571104X3361](https://doi.org/10.1198/108571104X3361)

McElreath, R. (2018).
[Algebra and the Missing Oxen](https://elevanth.org/blog/2018/01/29/algebra-and-missingness/).
*Elements of Evolutionary Anthropology* (blog).

Royle, J. A. (2004). N-mixture models for estimating population
size from spatially replicated counts. *Biometrics* 60, 108-115.
[doi:10.1111/j.0006-341X.2004.00142.x](https://doi.org/10.1111/j.0006-341X.2004.00142.x)

Royle, J. A. and Nichols, J. D. (2003). Estimating abundance
from repeated presence-absence data or point counts. *Ecology*
84(3), 777-790. DOI:
<https://doi.org/10.1890/0012-9658(2003)084[0777:EAFRPA]2.0.CO;2>

Stan Development Team (2024).
[Latent Discrete Parameters](https://mc-stan.org/docs/stan-users-guide/latent-discrete.html).
*Stan User's Guide*.
