---
title: "Forecast evaluation in mvgam"
author: "Nicholas J Clark"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
    toc: yes
    toc_float: yes
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  dpi = 100,
  fig.asp = 0.7,
  fig.width = 6,
  out.width = "70%",
  fig.align = "center"
)
library(mvgam)
library(ggplot2)
theme_set(mvgam_theme(base_size = 12, base_family = "serif"))

# Cached fits + LFO objects live under pkgdown/, which is
# excluded from the CRAN tarball by .Rbuildignore. pkgdown runs
# articles from the project root, so the relative path resolves
# at build time. The visible fit chunks remain as documentation
# of what produced each object.
cache_dir <- file.path("pkgdown", "forecast_eval_cache")
read_cache <- function(name) {
  readRDS(file.path(cache_dir, paste0(name, ".rds")))
}
```

A forecast is only as useful as the score that judges it. The
choice of scoring rule shapes which model wins a comparison,
which forecast horizons get the credit, and which corner of the
posterior gets blamed for misses. This article walks through the
mvgam forecast-evaluation surface on a real ecological dataset.
We cover the three workflow stages every applied user runs into.
`loo()` ranks the training fit. `forecast()` plus
`score()` gives a fixed-window evaluation against a holdout.
`lfo_cv()` runs rolling-origin evaluation that simulates a
production forecasting workflow. We then build an ensemble that combines the
strongest models and score it against its components.

All scoring rules used below are *proper* in the sense of
Gneiting & Raftery
([2007](https://doi.org/10.1198/016214506000001437)): they are
minimised in expectation by the true predictive distribution, so
ranking models by score is well-defined.

## The data

`portal_data` ships with the package and carries four desert
rodent species (DM, DO, PB, PP) trapped monthly at the Portal
LTREB site. Two covariates come pre-centred: a 12-month moving
average of NDVI (`ndvi_ma12`) and the previous month's minimum
temperature (`mintemp`). Averaging over twelve months damps the
NDVI series considerably, so it varies on a much narrower scale
than `mintemp` does. The response `captures` is an integer
count, and roughly a fifth of the cells are missing because the
field team could not always reach the plot.

```{r data-head}
data(portal_data)
str(portal_data)
head(portal_data, 4)
```

We split the 80 monthly timepoints into a 68-month training window
and a 12-month holdout for forecast evaluation.

```{r train-test-split}
train <- subset(portal_data, time <= 68L)
test  <- subset(portal_data, time >  68L)
c(train_rows = nrow(train), test_rows = nrow(test))
```

The four series have markedly different abundance levels and
temporal patterns. `mvgam_data()` gives the four-panel EDA view
for one series and the faceted multi-series view when
`series = "all"`:

```{r plot-series, fig.asp = 0.55, out.width = "85%"}
mvgam_data(train, family = poisson(), y = "captures",
            series = "all")
```

```{r series-shape, include = FALSE}
zero_run <- function(y) {
  y[is.na(y)] <- -1
  r <- rle(y == 0)
  if (any(r$values)) max(r$lengths[r$values]) else 0L
}
fe_pb_zero <- round(100 * mean(
  train$captures[train$series == "PB"] == 0, na.rm = TRUE
))
fe_pp_zero <- round(100 * mean(
  train$captures[train$series == "PP"] == 0, na.rm = TRUE
))
fe_pp_run <- zero_run(train$captures[train$series == "PP"])
```

DM dominates the captures and never hits zero. The other three
are an order of magnitude smaller and each is sparse in its own
way: PB records the most zeros at `r fe_pb_zero`% of its
sampled months, while PP records fewer at `r fe_pp_zero`% but
strings them into the longest run, `r fe_pp_run` months without
a capture. A useful forecast model has to accommodate both the
scale gap and that heterogeneity.

## Three candidate models

We fit three models in increasing structural complexity. Each is
a Bayesian Poisson regression on the captures, but they differ
in how they handle the temporal signal beyond the smooths on
NDVI and minimum temperature.

Model 1, smooths only (`mod_spline`), is a baseline GAM with
per-species temporal smooths. No state-space dynamics; the
forecast extrapolates the spline trajectories, which often misses
on count data because splines were not designed for forecasting.

```{r mod_spline_show, eval = FALSE}
mod_spline <- mvgam(
  captures ~ s(mintemp, k = 6) + s(ndvi_ma12, k = 6) +
    s(time, by = series, k = 8) + series,
  family = poisson(),
  data   = train,
  chains = 2L, iter = 1600L, warmup = 1000L, silent = 2L
)
```

```{r mod_spline_load, include = FALSE}
mod_spline <- read_cache("mod_spline")
```

Model 2, per-species AR(1) (`mod_ar`), replaces the temporal
spline with an independent AR(1) state-space process for each
species. Each species' latent state evolves with its own
auto-correlation; the obs-side smooths handle the covariate
response.

```{r mod_ar_show, eval = FALSE}
mod_ar <- mvgam(
  captures ~ s(mintemp, k = 6) + s(ndvi_ma12, k = 6) + series,
  trend_formula = ~ AR(),
  family = poisson(),
  data   = train,
  chains = 2L, iter = 1600L, warmup = 1000L, silent = 2L
)
```

```{r mod_ar_load, include = FALSE}
mod_ar <- read_cache("mod_ar")
```

Model 3, correlated VAR(1) (`mod_var`), is a vector
autoregression with `cor = TRUE`, allowing latent states for the
four species to be cross-correlated. If one species' deviation
predicts another's, this model can use that information to
sharpen multi-species forecasts.

```{r mod_var_show, eval = FALSE}
mod_var <- mvgam(
  captures ~ s(mintemp, k = 6) + s(ndvi_ma12, k = 6) + series,
  trend_formula = ~ VAR(cor = TRUE),
  family = poisson(),
  data   = train,
  chains = 2L, iter = 1600L, warmup = 1000L, silent = 2L
)
```

```{r mod_var_load, include = FALSE}
mod_var <- read_cache("mod_var")
```

The VAR fit's `summary()` shows the joint parameters that
distinguish it from the simpler models:

```{r mod_var_summary}
summary(mod_var, include_betas = FALSE)
```

The `A_trend` matrix carries the lag-1 cross-effects between
species, and `Sigma_trend` (with `cor = TRUE`) is unstructured
rather than diagonal.

Read the `Rhat` and `Bulk_ESS` columns before anything else. The
entries of `A_trend` sample poorly here, and raising the
iterations does not rescue them: sixteen transition coefficients
are a great deal to ask of four series over 68 months with a
fifth of the cells missing. Carry that into every comparison
below, because a model whose parameters are this uncertain can
still score well for reasons that have nothing to do with
forecasting. The innovation covariance fares better than the
transition matrix does, which matters later.

## In-sample comparison: PSIS-LOO

PSIS-LOO (Vehtari et al.
[2017](https://doi.org/10.1007/s11222-016-9696-4)) approximates
leave-one-observation-out predictive density on the training set
using Pareto-smoothed importance sampling, avoiding a full refit
per fold.

```{r loo-each}
loo_spline <- loo(mod_spline)
loo_ar     <- loo(mod_ar)
loo_var    <- loo(mod_var)
loo_compare(loo_spline, loo_ar, loo_var)
```

```{r loo-numbers, include = FALSE}
elpd_of <- function(l) l$estimates["elpd_loo", "Estimate"]
fe_spline_gap <- abs(round(elpd_of(loo_var) - elpd_of(loo_spline)))
k_ar <- loo_ar$diagnostics$pareto_k
k_sp <- loo_spline$diagnostics$pareto_k
fe_k_ar_pct <- round(100 * mean(k_ar > 0.7))
fe_k_sp_pct <- round(100 * mean(k_sp > 0.7), 1)
fe_k_ar_max <- round(max(k_ar), 2)
```

The ranking puts the two state-space models on top, with the
spline GAM `r fe_spline_gap` ELPD units behind the VAR.
`loo_model_weights()` makes that look decisive:

```{r model-weights}
mw <- loo_model_weights(
  list(spline = loo_spline, ar = loo_ar, var = loo_var),
  method = "stacking"
)
mw
```

Nearly all the weight goes to the VAR and none to the spline. It
would be easy to read that as a recommendation about forecasting.
It is not one, and the diagnostics say so before the scores do:

```{r pareto-k}
sapply(list(spline = loo_spline, ar = loo_ar, var = loo_var),
       function(l) round(mean(l$diagnostics$pareto_k > 0.7), 3))
```

The spline's importance weights are almost all well behaved, at
`r fe_k_sp_pct`% of folds past the 0.7 threshold. The AR model
reaches `r fe_k_ar_pct`%, with a worst value of
`r fe_k_ar_max`. A Pareto-k above 0.7 means the
importance-sampling estimate for that fold cannot be trusted, so
part of the ranking above rests on numbers PSIS itself flags.

The reason is structural, and it also explains which way the
ranking fell. PSIS-LOO estimates the density of observation $t$
under a posterior fitted without it, by reweighting the posterior
that did include it. That works while dropping one point barely
moves the posterior. A latent trend breaks the assumption twice.

The state at time $t$ sits between the states at $t-1$ and
$t+1$, and Markov dynamics interpolate across the gap. A model
carrying a latent trend therefore still holds a well-informed
state at $t$ when $y_t$ is dropped, and `log_lik()` evaluates
the density at that state. The quantity that comes back is
closer to an in-sample density than an out-of-sample one, and it
flatters whichever model carries the most latent structure. The
spline has no latent state to flatter, which is why it ranks
last rather than first.

The same dependence ruins the weights. Because $y_t$ largely
determines the state at $t$, removing it moves the posterior in
a direction that one observation's importance ratio cannot
cover. That is what the Pareto-k values are reporting.

None of this makes the ranking wrong on its own terms. PSIS-LOO
asks how well each model describes the points it trained on, and
the state-space models do describe them better. It is the wrong
question for a forecaster, and the approximation answering it
has partly broken.

For forecast comparison the question is to leave the *future*
out.
Every evaluation point is then unseen relative to its own
training window, so nothing interpolates through neighbours and
no importance ratio has to stretch. The `lfo_cv()` section below
asks it, and gets a different answer.

## Out-of-sample forecasts

`forecast()` produces an `mvgam_forecast` object carrying the
posterior predictive distribution at each held-out timepoint.

```{r forecasts}
fc_spline <- forecast(mod_spline, newdata = test)
fc_ar     <- forecast(mod_ar,     newdata = test)
fc_var    <- forecast(mod_var,    newdata = test)
```

The default `plot()` method shows the median forecast with 50%
and 90% credible bands, with held-out observations overlaid:

```{r plot-fc-var, fig.asp = 0.55, out.width = "85%"}
plot(fc_var, series = 1)
```

The VAR forecast tracks the observed DM captures with bands that
widen over the first half of the horizon and then level off.
That is the signature of a stationary process: forecast variance
climbs towards the marginal variance and stops there, rather
than growing without bound the way a random walk's would.

For likelihood-based scoring (ELPD) we also need a link-scale
forecast object:

```{r link-forecasts}
fc_spline_link <- forecast(mod_spline, newdata = test, type = "link")
fc_ar_link     <- forecast(mod_ar,     newdata = test, type = "link")
fc_var_link    <- forecast(mod_var,    newdata = test, type = "link")
```

## Univariate scoring

The CRPS (Continuous Ranked Probability Score) is the
distributional analogue of mean absolute error. Lower is better.
`score()` returns one row per evaluated timepoint per series:

```{r crps}
crps_var <- score(fc_var, score = "crps")
str(crps_var, max.level = 1)
head(crps_var$DM, 4)
```

The `in_interval` column carries a 0/1 indicator of whether the
held-out observation fell inside the 90% predictive interval.
Average it within a series for an empirical coverage estimate,
which should sit near 0.9 for a calibrated model:

```{r coverage}
sapply(crps_var[setdiff(names(crps_var), "all_series")],
       function(d) mean(d$in_interval, na.rm = TRUE))
```

Only DM lands there. DO falls short and both sparse series cover
every held-out point, which means their intervals are wider than
they need to be rather than better calibrated. Twelve months per
species is far too short a window to read much into any of
these, and that is the honest limitation of a single fixed
holdout: it gives one draw of the evaluation, with no sense of
how much it would move on a different window. The rolling-origin
section below is the answer to that.

`compare_scores()` bundles scores from several `mvgam_forecast`
objects into one tidy frame. It returns one row per model x
series x horizon, suitable for ggplot facets and summaries:

```{r crps-compare}
crps_all <- compare_scores(fc_spline, fc_ar, fc_var, score = "crps")
head(crps_all)
```

`plot()` dispatches on the `mvgam_compare_scores` class, faceting by series for a
univariate score and taking `relative = "<model_name>"` to
switch to differences against a baseline:

```{r crps-by-horizon, fig.asp = 0.55, out.width = "85%"}
plot(crps_all)
```

```{r crps-means, include = FALSE}
cm <- aggregate(
  score ~ model + series,
  data = crps_all[crps_all$series != "all_series", ],
  FUN = function(z) mean(z, na.rm = TRUE)
)
crps_mean <- function(m, sp) {
  round(cm$score[cm$model == m & cm$series == sp], 1)
}
```

The picture is not the same for every species. On DO the spline
is not competitive at all, averaging a CRPS of
`r crps_mean("fc_spline", "DO")` against
`r crps_mean("fc_ar", "DO")` for the AR model, because its
per-species smooth extrapolates a trajectory the observed series
never follows. DM shows a milder version of the same thing. On
the two sparse series the ordering reverses, and the spline is
the better forecast.

To see which horizons each state-space model beats the spline
at, pivot to differences:

```{r crps-vs-spline, fig.asp = 0.55, out.width = "85%"}
plot(crps_all, relative = "fc_spline")
```

Negative values mean the AR or VAR model scored lower than the
spline at that horizon. The gains are concentrated on DO and DM,
the two series with enough signal for a latent state to track.
On PB and PP the state-space models are modestly worse: those
series spend much of the holdout at or near zero, and a latent
AR state that keeps drifting adds variance a flat spline
prediction avoids. Averaging CRPS over all four species would
hide that entirely, which is the reason to score per series
before scoring jointly.

The log score (`score = "logs"`) evaluates the predictive
log-density at each held out observation. The forecast object
must use `type = "link"`:

```{r logs}
logs_var <- score(fc_var_link, score = "logs")
head(logs_var$DM, 4)
```

For integer counts, the discrete ranked probability score is
the count-aware sibling of CRPS:

```{r drps}
drps_var <- score(fc_var, score = "drps")
head(drps_var$DM, 4)
```

DRPS and CRPS will rank models similarly, but DRPS is the
principled choice when the response is bounded to integers
because its units are the natural ones.

## Multivariate scoring

Univariate scoring sums per-series performance; it cannot reward
a model for capturing the *joint* structure across species. Two
proper scoring rules for joint forecasts are the energy score
(Gneiting et al. [2008](https://doi.org/10.1007/s11749-008-0114-x)),
a multivariate generalisation of CRPS, and the variogram
score of order 0.5 (Scheuerer & Hamill
[2015](https://doi.org/10.1175/MWR-D-14-00269.1)), which targets
cross-series correlation structure.

```{r mv-scores}
energy_var <- score(fc_var, score = "energy")
vario_var  <- score(fc_var, score = "variogram")
head(energy_var$all_series, 4)
head(vario_var$all_series, 4)
```

Per-series rows still carry the coverage indicator but the joint
score lives in `all_series`. Lower energy means better joint
calibration; lower variogram means better recovery of pairwise
dependencies.

```{r mv-compare, fig.asp = 0.5, out.width = "85%"}
energy_all <- compare_scores(fc_spline, fc_ar, fc_var,
                              score = "energy")
plot(energy_all)
```

```{r energy-numbers, include = FALSE}
ej <- as.data.frame(energy_all)
ej <- ej[ej$series == "all_series", ]
eband <- function(m) {
  z <- ej$score[ej$model == m]
  round(range(z, na.rm = TRUE), 1)
}
esp <- eband("fc_spline")
ear <- eband("fc_ar")
evr <- eband("fc_var")
# Forecast draws are stochastic, so count the horizons rather
# than asserting a margin that could flip between renders.
epair <- merge(
  ej[ej$model == "fc_ar", c("eval_horizon", "score")],
  ej[ej$model == "fc_var", c("eval_horizon", "score")],
  by = "eval_horizon", suffixes = c("_ar", "_var")
)
epair <- epair[stats::complete.cases(epair), ]
e_ar_wins <- sum(epair$score_ar < epair$score_var)
e_n_h <- nrow(epair)
```

Read the spline column first. Its energy score climbs from
`r esp[1]` at horizon one to `r esp[2]` by month 12, because the
linear extrapolation drifts further from the joint trajectory of
the four series with every step ahead. The AR and VAR models
stay flat over the same window, spanning `r ear[1]` to
`r ear[2]` and `r evr[1]` to `r evr[2]`, with the latent
dynamics holding the joint forecast in place. AR and VAR run
close, AR scoring lower at `r e_ar_wins` of the `r e_n_h`
horizons. On these data and over this window the modelled
cross-species dynamics add no detectable benefit to the joint
forecast, and the simpler AR is the better buy if forecast
accuracy is the only thing being bought.

That ordering disagrees with the per-series CRPS above, where
the spline won on PB and PP. The two scores ask different
things. CRPS takes each species on its own, and the spline is
adequate on the two that sit near zero. The energy score takes
all four at once, where a forecast that drifts on DM and DO is
wrong about the system whatever it does on the quiet series.

Dropping the VAR on that score alone would be a mistake, though.
Its transition matrix and its innovation covariance answer
different questions, and the covariance is the better identified
of the two:

```{r resid-cor}
rc <- residual_cor(mod_var)
round(rc$cor, 2)
```

```{r cor-numbers, include = FALSE}
cor_ci <- function(a, b) {
  sprintf("%.2f (%.2f to %.2f)", rc$cor[a, b],
          rc$cor_lower[a, b], rc$cor_upper[a, b])
}
fe_pbpp <- cor_ci("PB", "PP")
fe_dmdo <- cor_ci("DM", "DO")
```

One pair clears zero. PB and PP correlate at `r fe_pbpp`, and
those two are congeners, both *Chaetodipus* pocket mice. After
the covariates and each species' own dynamics are accounted for,
what is left of their latent states still rises and falls
together. The two *Dipodomys* kangaroo rats lean the same way at
`r fe_dmdo`, but that interval covers zero, so 68 months will
not settle it.

The AR model cannot produce this at any horizon, because it
gives every species an independent latent state by construction.
So the VAR earns nothing on forecast accuracy over this window
and still answers a question the simpler model cannot be asked.
Which of the two matters is a question about the study rather
than about the scores, and it is worth deciding before the
scores are consulted rather than after. The
[VAR article](https://nicholasjclark.github.io/mvgam/articles/var.html)
takes this structure apart properly, with impulse responses and
variance decompositions.

## Rolling-origin evaluation with `lfo_cv()`

Approximate leave-future-out cross-validation (Bürkner, Gabry &
Vehtari [2020](https://link.springer.com/article/10.1007/s00180-020-01045-4))
is the time-series-honest cousin of PSIS-LOO. Starting at a
minimum training length `min_t`, the procedure expands the
training window one timepoint at a time and forecasts `fc_horizon`
steps ahead. At each step, PSIS importance weights are computed
against the existing fit; only when Pareto-k exceeds a threshold
does mvgam refit the model. Leaving `pareto_k_threshold = NULL`
takes the adaptive rule `min(1 - 1 / log10(S), 0.7)` on the
number of posterior draws `S`, which tightens the gate for fits
with few draws and clamps at 0.7 once PSIS is on firm ground.
That hybrid scheme gives near-exact LFO ELPD at a fraction of
the cost of refitting everywhere.

```{r lfo_show, eval = FALSE}
lfo_spline <- lfo_cv(mod_spline, min_t = 60L, fc_horizon = 4L,
                      silent = 2L)
lfo_ar     <- lfo_cv(mod_ar,     min_t = 60L, fc_horizon = 4L,
                      silent = 2L)
lfo_var    <- lfo_cv(mod_var,    min_t = 60L, fc_horizon = 4L,
                      silent = 2L)
```

```{r lfo_load, include = FALSE}
lfo_spline <- read_cache("lfo_spline")
lfo_ar     <- read_cache("lfo_ar")
lfo_var    <- read_cache("lfo_var")
```

`lfo_cv()` returns an `mvgam_lfo` object whose `refits_at` slot
records the timepoints at which Pareto-k forced an actual refit:

```{r refits}
list(
  spline = lfo_spline$refits_at,
  ar     = lfo_ar$refits_at,
  var    = lfo_var$refits_at
)
```

All three refit at the same four timepoints here, so the
latent dynamics are not what drives the gate on this dataset;
the window is simply short enough that each model's weights
degrade at much the same places. The hybrid PSIS step still
skips the remaining timepoints, which is what keeps
rolling-origin evaluation affordable.

`loo_compare()` works on `mvgam_lfo` objects directly:

```{r lfo-compare}
loo_compare(lfo_spline, lfo_ar, lfo_var)
```

```{r lfo-numbers, include = FALSE}
lfo_tab <- loo_compare(lfo_spline, lfo_ar, lfo_var)
lfo_worst <- round(abs(min(lfo_tab[, "elpd_diff"])), 1)
lfo_worst_se <- round(lfo_tab[which.min(lfo_tab[, "elpd_diff"]),
                              "se_diff"], 1)
```

The spline heads this table, which is the reverse of the LOO
ordering. Before reading anything into it, compare each
difference with its own standard error: the largest gap is
`r lfo_worst` units against a standard error of
`r lfo_worst_se`. Every difference here sits inside roughly one
standard error of zero, so this evaluation does not separate the
three models at all. Starting at `min_t = 60` on 68 training
timepoints leaves only a handful of origins, and a four-step
horizon asks a much easier question than the twelve-month
holdout scored above.

So the conclusion is not that the spline forecasts best. It is
that nothing separates these models four steps ahead, where the
twelve-month energy score separated them cleanly. The horizon
changed the answer, not the ranking rule.

`compare_elpds()` is the LFO companion of `compare_scores()`. It
stacks per-step ELPDs from several `mvgam_lfo` objects into a
tidy frame keyed by `model` and `eval_time`.

```{r lfo-trajectory, fig.asp = 0.55, out.width = "85%"}
lfo_df <- compare_elpds(lfo_spline, lfo_ar, lfo_var)
plot(lfo_df)
```

Toggle `cumulative = TRUE` to see how each model pulls ahead or
falls behind across the rolling-origin window:

```{r lfo-cumulative, fig.asp = 0.55, out.width = "85%"}
plot(lfo_df, cumulative = TRUE)
```

The LFO ranking and the LOO ranking disagree here, and LFO is
the one to trust for a forecasting decision, because its
evaluation set is the *future* relative to each training window.
That is the set the production question is asked on. Where LFO
cannot separate the candidates, as here, the honest reading is
that the rolling window was too short to decide, and a longer
holdout scored jointly is the tie-breaker.

## Ensemble forecasting

`loo_model_weights()` on a set of `mvgam_lfo` objects offers two
methods (Yao et al. 2018):

- `method = "pseudo-BMA"` (default): softmax of each model's
  total LFO ELPD. Cheap, requires only the per-step ELPDs that
  every `mvgam_lfo` carries.
- `method = "stacking"`: convex optimisation on the pointwise
  log-density matrix at evaluation points. Gives softer weights
  than pseudo-BMA and is generally preferred. Needs the
  `$log_lik` matrix that `lfo_cv()` only populates when called
  with `save_log_lik = TRUE`.

```{r lfo-weights}
lfo_weights <- loo_model_weights(lfo_spline, lfo_ar, lfo_var)
lfo_weights
stack_weights <- loo_model_weights(
  lfo_spline, lfo_ar, lfo_var,
  method = "stacking"
)
stack_weights
```

The two methods disagree. Pseudo-BMA puts a weight of one on the spline and nothing
anywhere else. It is a softmax of total ELPD, so it exponentiates
differences that the section above showed to be within a standard
error of zero, and a gap that small still swamps its rivals once
raised through `exp()`. The confidence is an artefact of the
transform, not evidence from the data.

Stacking sees the same evaluation points and splits the weight
between the spline and the AR model, giving almost nothing to the
VAR. It optimises the log score of the *mixture* rather than
ranking the candidates, so it can notice that the two leaders
miss on different evaluation points and that averaging them beats
either alone. That is the behaviour to expect whenever the
candidates run close, and it is why stacking is the better
default (Yao et al. 2018, §3.3).

`ensemble()` builds a combined forecast object by resampling
posterior draws from each component in proportion to the weights:

```{r ensemble-build}
fc_ens <- ensemble(
  fc_spline, fc_ar, fc_var,
  weights = as.numeric(stack_weights)
)
fc_ens
```

Score the ensemble alongside the individual models on the same
holdout, using CRPS and then energy. `summary()` on a
`mvgam_compare_scores` frame pivots it to one row per evaluation
cell with a column per model, so the candidates can be read
across on the same data point:

```{r ensemble-crps}
crps_ens <- compare_scores(fc_spline, fc_ar, fc_var, fc_ens,
                             score = "crps",
                             model_names = c("spline", "ar",
                                              "var", "ensemble"))
summary(crps_ens)
```

```{r ensemble-energy}
energy_ens <- compare_scores(fc_spline, fc_ar, fc_var, fc_ens,
                               score = "energy",
                               model_names = c("spline", "ar",
                                                "var", "ensemble"))
summary(energy_ens)
```

The ensemble wins nowhere outright and loses nowhere badly. That
is the property worth having when the candidates disagree about
which species they handle well, as these do: the spline is the
better forecast on PB and PP, the state-space models on DM and
DO, and nothing in the training data tells you which species
will matter most next year. Mixing them buys insurance against
that choice rather than a better score on any one cell.

## Take-aways

Three evaluations of the same three fits gave three different
winners. PSIS-LOO put the VAR first and the spline last, on an
approximation its own Pareto-k values flagged as unreliable, and
for a question about training fit rather than forecasting.
Rolling-origin evaluation at four steps put the spline first by
a margin smaller than its standard error, which is another way
of saying it could not tell them apart. The twelve-month energy
score separated them cleanly and put the AR model first.

None of those contradicts the others, because each asks a
different question. The score and the cross-validation scheme
are not neutral steps that follow the choice of model; they are
the choice. Settle the horizon you care about and whether the
decision is per species or joint, and let that pick the
evaluation. Where the candidates disagree by species, as they do
here, an `ensemble()` weighted by stacking spreads the bet
rather than forcing one.

None of this settles which model to publish. The VAR forecasts
no better than the AR and recovers a correlation between the two
pocket mice that the AR cannot represent; if that structure is
the reason for the study, it outranks a score it was never
competing on.

## Further reading

The natural next functions to explore on the same fit are:

- `pp_check()` for posterior-predictive checks per outcome
  before any scoring runs.
- `conditional_effects()` and `marginaleffects::plot_predictions()`
  for the covariate surface that drives the forecast.
- `posterior_average()` and `pp_average()` for weighted
  posterior summaries across the candidate models the
  ensemble already mixes.
- `irf()`, `fevd()` and `stability()` for impulse response,
  forecast-error variance decomposition and stationarity
  diagnostics when the latent dynamics are a VAR.
- `methods_md()` for a citation-ready methods paragraph naming
  the architecture pieces the fit actually used.

Sibling articles in this series carry the workflow further:

- The [VAR article](https://nicholasjclark.github.io/mvgam/articles/var.html)
  takes the `VAR(cor = TRUE)` fit used above apart cell by cell,
  with impulse responses, variance decompositions and stability
  metrics.
- The [multi-response article](https://nicholasjclark.github.io/mvgam/articles/mvbf.html)
  shows how to share a single latent state across several
  outcomes with different observation families.
- The [JSDM article](https://nicholasjclark.github.io/mvgam/articles/jsdgam.html)
  applies the latent factor variant to multispecies count data
  with informed loadings via `jsdgam()`.
- The [overview vignette](https://nicholasjclark.github.io/mvgam/articles/mvgam_overview.html)
  introduces the trend constructors the three candidates here
  draw on.
- For grouped trend dynamics (one transition matrix and
  innovation covariance per group, shrunk toward a global mean),
  see the `gr` and `subgr` arguments in `?VAR` and the
  [hierarchical VAR article](https://nicholasjclark.github.io/mvgam/articles/hierarchical_var.html).

## Acknowledgements

The Portal Project rodent capture data ship with mvgam via
`data(portal_data)` and are maintained by the Weecology group
at <https://github.com/weecology/PortalData>. The forecast
evaluation scheme comes from Bürkner, Gabry & Vehtari
([2020](https://doi.org/10.1080/00949655.2020.1783262)). The
PSIS approximation is from Vehtari, Gelman & Gabry
([2017](https://doi.org/10.1007/s11222-016-9696-4)). Stacking
weights are from Yao et al.
([2018](https://doi.org/10.1214/17-BA1091)). The univariate
scoring framework is from Gneiting & Raftery
([2007](https://doi.org/10.1198/016214506000001437)). The
energy score is from Gneiting et al.
([2008](https://doi.org/10.1007/s11749-008-0114-x)). The
variogram score is from Scheuerer & Hamill
([2015](https://doi.org/10.1175/MWR-D-14-00269.1)). The
multispecies forecasting framing follows Clark et al.
([2025](https://doi.org/10.7717/peerj.18929)).

## References

Bürkner P-C, Gabry J and Vehtari A (2020). Approximate
leave-future-out cross-validation for Bayesian time series
models. *Journal of Statistical Computation and Simulation*
90(14), 2499-2523.
[doi:10.1080/00949655.2020.1783262](https://doi.org/10.1080/00949655.2020.1783262)

Clark NJ, Ernest SKM, Senyondo H, Simonis J, White EP, Yenni
GM and Karunarathna KANK (2025). Beyond single-species models:
leveraging multispecies forecasts to navigate the dynamics <!-- prose-lint-disable-line -->
of ecological predictability. *PeerJ* 13, e18929.
[doi:10.7717/peerj.18929](https://doi.org/10.7717/peerj.18929)

Gneiting T and Raftery AE (2007). Strictly proper scoring rules,
prediction, and estimation. *Journal of the American
Statistical Association* 102(477), 359-378.
[doi:10.1198/016214506000001437](https://doi.org/10.1198/016214506000001437)

Gneiting T, Stanberry LI, Grimit EP, Held L and Johnson NA
(2008). Assessing probabilistic forecasts of multivariate
quantities, with an application to ensemble predictions of
surface winds. *TEST* 17(2), 211-235.
[doi:10.1007/s11749-008-0114-x](https://doi.org/10.1007/s11749-008-0114-x)

Scheuerer M and Hamill TM (2015). Variogram-based proper scoring
rules for probabilistic forecasts of multivariate quantities.
*Monthly Weather Review* 143(4), 1321-1334.
[doi:10.1175/MWR-D-14-00269.1](https://doi.org/10.1175/MWR-D-14-00269.1)

Vehtari A, Gelman A and Gabry J (2017). Practical Bayesian model
evaluation using leave-one-out cross-validation and WAIC.
*Statistics and Computing* 27(5), 1413-1432.
[doi:10.1007/s11222-016-9696-4](https://doi.org/10.1007/s11222-016-9696-4)

Yao Y, Vehtari A, Simpson D and Gelman A (2018). Using stacking
to average Bayesian predictive distributions. *Bayesian
Analysis* 13(3), 917-1003.
[doi:10.1214/17-BA1091](https://doi.org/10.1214/17-BA1091)
