---
title: "Forecast evaluation in mvgam"
author: "Nicholas J Clark"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
    toc: yes
    toc_float: yes
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  dpi = 100,
  fig.asp = 0.7,
  fig.width = 6,
  out.width = "70%",
  fig.align = "center"
)
library(mvgam)
library(ggplot2)
theme_set(theme_bw(base_size = 12, base_family = "serif"))

# Cached fits + LFO objects live under pkgdown/, which is
# excluded from the CRAN tarball by .Rbuildignore. pkgdown runs
# articles from the project root, so the relative path resolves
# at build time. The visible fit chunks remain as documentation
# of what produced each object.
cache_dir <- file.path("pkgdown", "forecast_eval_cache")
read_cache <- function(name) {
  readRDS(file.path(cache_dir, paste0(name, ".rds")))
}
```

A forecast is only as useful as the score that judges it. The
choice of scoring rule shapes which model wins a comparison,
which forecast horizons get the credit, and which corner of the
posterior gets blamed for misses. This article walks through the
mvgam forecast-evaluation surface on a real ecological dataset.
We cover the three workflow stages every applied user runs into.
`loo()` ranks the training fit. `forecast()` plus
`score()` gives a fixed-window evaluation against a holdout.
`lfo_cv()` runs rolling-origin evaluation that simulates a
production forecasting workflow. We then build an ensemble that combines the
strongest models and score it against its components.

All scoring rules used below are *proper* in the sense of
Gneiting & Raftery
([2007](https://doi.org/10.1198/016214506000001437)): they are
minimised in expectation by the true predictive distribution, so
ranking models by score is well-defined.

## The data

`portal_data` ships with the package and carries four desert
rodent species (DM, DO, PB, PP) trapped monthly at the Portal
LTREB site, with two pre-standardised covariates: a 12-month
moving average of NDVI (`ndvi_ma12`) and the previous month's
minimum temperature (`mintemp`). The response `captures` is an
integer count, and roughly a fifth of the cells are missing
because the field team could not always reach the plot.

```{r data-head}
data(portal_data)
str(portal_data)
head(portal_data, 4)
```

We split the 80 monthly timepoints into a 68-month training window
and a 12-month holdout for forecast evaluation.

```{r train-test-split}
train <- subset(portal_data, time <= 68L)
test  <- subset(portal_data, time >  68L)
c(train_rows = nrow(train), test_rows = nrow(test))
```

The four series have markedly different abundance levels and
temporal patterns. `mvgam_data()` gives the four-panel EDA view
for one series and the faceted multi-series view when
`series = "all"`:

```{r plot-series, fig.asp = 0.55, out.width = "85%"}
mvgam_data(train, family = poisson(), y = "captures",
            series = "all")
```

DM dominates the captures and is highly variable; DO and PB are
intermediate; PP is sparse with long zero stretches. A useful
forecast model should accommodate that heterogeneity.

## Three candidate models

We fit three models in increasing structural complexity. Each is
a Bayesian Poisson regression on the captures, but they differ
in how they handle the temporal signal beyond the smooths on
NDVI and minimum temperature.

**Model 1, smooths only (`mod_spline`).** A baseline GAM with
per-species temporal smooths. No state-space dynamics; the
forecast extrapolates the spline trajectories, which often misses
on count data because splines were not designed for forecasting.

```{r mod_spline_show, eval = FALSE}
mod_spline <- mvgam(
  captures ~ s(mintemp, k = 6) + s(ndvi_ma12, k = 6) +
    s(time, by = series, k = 8) + series,
  family = poisson(),
  data   = train,
  chains = 2L, iter = 1600L, warmup = 1000L, silent = 2L
)
```

```{r mod_spline_load, include = FALSE}
mod_spline <- read_cache("mod_spline")
```

**Model 2, per-species AR(1) (`mod_ar`).** Replaces the temporal
spline with an independent AR(1) state-space process for each
species. Each species' latent state evolves with its own
auto-correlation; the obs-side smooths handle the covariate
response.

```{r mod_ar_show, eval = FALSE}
mod_ar <- mvgam(
  captures ~ s(mintemp, k = 6) + s(ndvi_ma12, k = 6) + series,
  trend_formula = ~ AR(),
  family = poisson(),
  data   = train,
  chains = 2L, iter = 1600L, warmup = 1000L, silent = 2L
)
```

```{r mod_ar_load, include = FALSE}
mod_ar <- read_cache("mod_ar")
```

**Model 3, correlated VAR(1) (`mod_var`).** A vector
autoregression with `cor = TRUE`, allowing latent states for the
four species to be cross-correlated. If one species' deviation
predicts another's, this model can use that information to
sharpen multi-species forecasts.

```{r mod_var_show, eval = FALSE}
mod_var <- mvgam(
  captures ~ s(mintemp, k = 6) + s(ndvi_ma12, k = 6) + series,
  trend_formula = ~ VAR(cor = TRUE),
  family = poisson(),
  data   = train,
  chains = 2L, iter = 1600L, warmup = 1000L, silent = 2L
)
```

```{r mod_var_load, include = FALSE}
mod_var <- read_cache("mod_var")
```

The VAR fit's `summary()` shows the joint state-space machinery
that distinguishes it from the simpler models:

```{r mod_var_summary}
summary(mod_var, include_betas = FALSE)
```

The `A_trend` matrix carries the lag-1 cross-effects between
species, and `Sigma_trend` (with `cor = TRUE`) is unstructured
rather than diagonal.

## In-sample comparison: PSIS-LOO

PSIS-LOO (Vehtari et al.
[2017](https://doi.org/10.1007/s11222-016-9696-4)) approximates
leave-one-observation-out predictive density on the training set
using Pareto-smoothed importance sampling, avoiding a full refit
per fold.

```{r loo-each}
loo_spline <- loo(mod_spline)
loo_ar     <- loo(mod_ar)
loo_var    <- loo(mod_var)
loo_compare(loo_spline, loo_ar, loo_var)
```

`loo_compare()` ranks the spline GAM first, with the VAR model
~870 ELPD units behind and the per-species AR model ~85,000
units behind. Take a moment to register how dramatic that gap
is. By the LOO ranking, the simpler model is overwhelmingly the
"better" fit, and `loo_model_weights()` makes the verdict
explicit:

```{r model-weights}
mw <- loo_model_weights(
  list(spline = loo_spline, ar = loo_ar, var = loo_var),
  method = "stacking"
)
mw
```

All weight lands on the spline; the state-space models receive
zero. Stacking weights are not posterior probabilities. They are
the linear combination weights that minimise predictive log
loss across LOO folds when the candidate forecasts are mixed
(Yao et al. [2018](https://doi.org/10.1214/17-BA1091)). Here
the optimiser sees the spline winning every fold by a wide
margin, so it allocates a single mass point.

This is the textbook PSIS-LOO failure mode when the model
carries a latent trend, not a vindication of the spline. The
reason is structural and worth dwelling on, because users hit
it constantly when reaching for `loo()` on time-series fits.

The method targets the expected log predictive density of each
fold in which a single observation is removed. It approximates
that quantity cheaply by importance-sampling the full posterior,
with Pareto smoothing on the weights.
The approximation is valid only when omitting one observation
shifts the posterior modestly, so the two posteriors overlap
enough for the weights to behave. That assumption breaks for a
state-space model in two compounding ways:

1. *The latent trend leaks information across the omitted
   point.* When observation $t$ is dropped, the latent state at
   $t$ is still sandwiched between the observed states at
   $t-1$ and $t+1$. Markov dynamics interpolate, and the model
   effectively sees through the gap. The ELPD it returns is
   therefore close to an in-sample density, not the true
   density at unseen data.
2. *The importance weights collapse.* The omitted observation
   strongly identifies the latent state at $t$, so dropping it
   shifts the posterior in a direction the importance ratio
   from one point cannot smooth. Pareto-k diagnostics climb
   past 0.7 on virtually every fold (you can confirm via
   `loo(mod_ar)$diagnostics$pareto_k`), so the PSIS
   approximation itself is unreliable.

The spline GAM has no latent trend, so neither problem applies
and its LOO ELPD is honest. The verdict here is correct for
the question PSIS-LOO actually asks: how well the model fits
the points it was trained on. That is not the question
forecasters care about.

For time-series forecast comparison the right diagnostic is
**leave the future out**. Every evaluation point is then unseen
relative to its training window, so there is no leak through
neighbours and no broken importance ratio. The `lfo_cv()`
section below runs exactly that comparison, and the ranking
inverts.

## Out-of-sample forecasts

`forecast()` produces an `mvgam_forecast` object carrying the
posterior predictive distribution at each held-out timepoint.

```{r forecasts}
fc_spline <- forecast(mod_spline, newdata = test)
fc_ar     <- forecast(mod_ar,     newdata = test)
fc_var    <- forecast(mod_var,    newdata = test)
```

The default `plot()` method shows the median forecast with 50%
and 90% credible bands, with held-out observations overlaid:

```{r plot-fc-var, fig.asp = 0.55, out.width = "85%"}
plot(fc_var, series = 1)
```

The VAR forecast tracks the observed DM captures with bands that
widen out by ~6 months. That is the canonical signature of a
stationary AR process whose forecast variance asymptotes to the
marginal variance.

For likelihood-based scoring (ELPD) we also need a link-scale
forecast object:

```{r link-forecasts}
fc_spline_link <- forecast(mod_spline, newdata = test, type = "link")
fc_ar_link     <- forecast(mod_ar,     newdata = test, type = "link")
fc_var_link    <- forecast(mod_var,    newdata = test, type = "link")
```

## Univariate scoring

The CRPS (Continuous Ranked Probability Score) is the
distributional analogue of mean absolute error. Lower is better.
`score()` returns one row per evaluated timepoint per series:

```{r crps}
crps_var <- score(fc_var, score = "crps")
str(crps_var, max.level = 1)
head(crps_var$DM, 4)
```

The `in_interval` column carries a 0/1 indicator of whether the
held-out observation fell inside the 90% predictive interval.
Average it within a series for an empirical coverage estimate; if
the model is well calibrated and the holdout is representative,
that average sits near 0.9.

```{r coverage}
sapply(crps_var[setdiff(names(crps_var), "all_series")],
       function(d) mean(d$in_interval, na.rm = TRUE))
```

`compare_scores()` bundles scores from several `mvgam_forecast`
objects into one tidy frame. It returns one row per model x
series x horizon, suitable for ggplot facets and summaries:

```{r crps-compare}
crps_all <- compare_scores(fc_spline, fc_ar, fc_var, score = "crps")
head(crps_all)
```

The frame has one row per model x series x horizon. Plot the
per-series CRPS trajectories or compute differences between
specific models directly from it. Here we plot CRPS by horizon
for each species, coloured by model, with the per-series rows
only (skipping the joint `all_series` row):

`plot()` dispatches on the `mvgam_compare_scores` class. For
univariate scores it facets by series; switch to differences
against any baseline model with `relative = "<model_name>"`.

```{r crps-by-horizon, fig.asp = 0.55, out.width = "85%"}
plot(crps_all)
```

DM and PB show the spline baseline rising fastest as the horizon
grows. The splines extrapolate roughly linearly, while the
state-space models stay close to the observed series via the
latent AR(1) state. PP is the noisy sparse series where all
three models perform similarly.

To see directly which horizons each state-space model beats the
spline at, pivot to differences:

```{r crps-vs-spline, fig.asp = 0.55, out.width = "85%"}
plot(crps_all, relative = "fc_spline")
```

Negative values mean the AR or VAR model has a lower CRPS than
the spline at that horizon. The state-space models help most on
DM and PB and roughly tie on PP, the sparse series whose stretches
of zeros are hard for any Poisson model.

The log score (`score = "logs"`) evaluates the predictive
log-density at each held out observation. The forecast object
must use `type = "link"`:

```{r logs}
logs_var <- score(fc_var_link, score = "logs")
head(logs_var$DM, 4)
```

For integer counts, the **discrete** ranked probability score is
the count-aware sibling of CRPS:

```{r drps}
drps_var <- score(fc_var, score = "drps")
head(drps_var$DM, 4)
```

DRPS and CRPS will rank models similarly, but DRPS is the
principled choice when the response is bounded to integers
because its units are the natural ones.

## Multivariate scoring

Univariate scoring sums per-series performance; it cannot reward
a model for capturing the *joint* structure across species. Two
proper scoring rules for joint forecasts are the **energy score**
(Gneiting et al. [2008](https://doi.org/10.1007/s11749-008-0114-x)),
a multivariate generalisation of CRPS, and the **variogram
score** of order 0.5 (Scheuerer & Hamill
[2015](https://doi.org/10.1175/MWR-D-14-00269.1)), which targets
cross-series correlation structure.

```{r mv-scores}
energy_var <- score(fc_var, score = "energy")
vario_var  <- score(fc_var, score = "variogram")
head(energy_var$all_series, 4)
head(vario_var$all_series, 4)
```

Per-series rows still carry the coverage indicator but the joint
score lives in `all_series`. Lower energy means better joint
calibration; lower variogram means better recovery of pairwise
dependencies.

```{r mv-compare, fig.asp = 0.5, out.width = "85%"}
energy_all <- compare_scores(fc_spline, fc_ar, fc_var,
                              score = "energy")
plot(energy_all)
```

Read the spline column first. Its energy score climbs from
~7 at horizon one to ~93 by month 12 because the linear
extrapolation drifts further from the joint trajectory of the
four series with each step ahead. The AR and VAR models stay
roughly flat (both between ~6 and ~13), the latent dynamics
holding the joint forecast in place. AR and VAR run neck and
neck. AR even edges VAR by one to three units at most horizons.
On these data and at this horizon, the modelled correlations
across species add no detectable benefit to the joint forecast;
the simpler AR wins.

## Rolling-origin evaluation with `lfo_cv()`

Approximate leave-future-out cross-validation (Bürkner, Gabry &
Vehtari [2020](https://link.springer.com/article/10.1007/s00180-020-01045-4))
is the time-series-honest cousin of PSIS-LOO. Starting at a
minimum training length `min_t`, the procedure expands the
training window one timepoint at a time and forecasts `fc_horizon`
steps ahead. At each step, PSIS importance weights are computed
against the existing fit; only when Pareto-k exceeds a threshold
does mvgam refit the model. Leaving `pareto_k_threshold = NULL`
takes the adaptive rule `min(1 - 1 / log10(S), 0.7)` on the
number of posterior draws `S`, which tightens the gate for fits
with few draws and clamps at 0.7 once PSIS is on firm ground.
That hybrid scheme gives near-exact LFO ELPD at a fraction of
the cost of refitting everywhere.

```{r lfo_show, eval = FALSE}
lfo_spline <- lfo_cv(mod_spline, min_t = 60L, fc_horizon = 4L,
                      silent = 2L)
lfo_ar     <- lfo_cv(mod_ar,     min_t = 60L, fc_horizon = 4L,
                      silent = 2L)
lfo_var    <- lfo_cv(mod_var,    min_t = 60L, fc_horizon = 4L,
                      silent = 2L)
```

```{r lfo_load, include = FALSE}
lfo_spline <- read_cache("lfo_spline")
lfo_ar     <- read_cache("lfo_ar")
lfo_var    <- read_cache("lfo_var")
```

`lfo_cv()` returns an `mvgam_lfo` object whose `refits_at` slot
records the timepoints at which Pareto-k forced an actual refit:

```{r refits}
list(
  spline = lfo_spline$refits_at,
  ar     = lfo_ar$refits_at,
  var    = lfo_var$refits_at
)
```

The state-space models refit more often than the spline because
their latent dynamics are more sensitive to each new observation.
The hybrid PSIS step still avoids refits at every other
timepoint, which is the point.

`loo_compare()` works on `mvgam_lfo` objects directly:

```{r lfo-compare}
loo_compare(lfo_spline, lfo_ar, lfo_var)
```

Look at the trajectory of expected log predictive density across
the rolling origin. A model whose ELPD trajectory rises faster
(less negative) is forecasting better on average.

`compare_elpds()` is the LFO companion of `compare_scores()`. It
stacks per-step ELPDs from several `mvgam_lfo` objects into a
tidy frame keyed by `model` and `eval_time`.

```{r lfo-trajectory, fig.asp = 0.55, out.width = "85%"}
lfo_df <- compare_elpds(lfo_spline, lfo_ar, lfo_var)
plot(lfo_df)
```

Toggle `cumulative = TRUE` to see how each model pulls ahead or
falls behind across the rolling-origin window:

```{r lfo-cumulative, fig.asp = 0.55, out.width = "85%"}
plot(lfo_df, cumulative = TRUE)
```

The LFO ranking and the LOO ranking can disagree. When they do,
trust LFO for forecasting decisions, because its evaluation set
is the *future* relative to each training window, which is the
only set that matches the production decision.

## Ensemble forecasting

`loo_model_weights()` on a set of `mvgam_lfo` objects offers two
methods (Yao et al. 2018):

- `method = "pseudo-BMA"` (default): softmax of each model's
  total LFO ELPD. Cheap, requires only the per-step ELPDs that
  every `mvgam_lfo` carries.
- `method = "stacking"`: convex optimisation on the pointwise
  log-density matrix at evaluation points. Gives softer weights
  than pseudo-BMA and is generally preferred. Needs the
  `$log_lik` matrix that `lfo_cv()` only populates when called
  with `save_log_lik = TRUE`.

```{r lfo-weights}
lfo_weights <- loo_model_weights(lfo_spline, lfo_ar, lfo_var)
lfo_weights
stack_weights <- loo_model_weights(
  lfo_spline, lfo_ar, lfo_var,
  method = "stacking"
)
stack_weights
```

The pseudo-BMA vector is essentially `(0, 1, 0)`. The softmax
exponentiates ELPD differences, and AR's total LFO ELPD sits
~20 units above the VAR and ~42 above the spline.
`exp(20) ≈ 5 × 10^8` swamps `exp(0)`, so AR takes essentially
all the weight by construction.

Stacking gives the same headline answer here. AR collects
~0.997 and the spline picks up a tiny ~0.003; VAR receives
nothing. That convergence is not a coincidence: AR is the
clear winner of the per-observation predictive density at
most evaluation points, so the convex optimisation lands at
almost the same corner of the simplex. When the headline
model is this dominant, ensembling buys nothing.

Stacking is still the right default. In the more usual case
where the candidates run closer to one another, the convex
optimiser produces appreciably softer weights than pseudo-BMA
(Yao et al. 2018, §3.3). Here it just happens to agree.

`ensemble()` builds a combined forecast object by resampling
posterior draws from each component in proportion to the weights:

```{r ensemble-build}
fc_ens <- ensemble(
  fc_spline, fc_ar, fc_var,
  weights = as.numeric(stack_weights)
)
fc_ens
```

Score the ensemble alongside the individual models on the
holdout window with `compare_scores()`. The honest comparison is
on CRPS and energy:

`summary()` on a `mvgam_compare_scores` frame pivots it to one
row per evaluation cell, with a column for each model. Read
across the rows to see how the candidates differ on the same
data point.

```{r ensemble-crps}
crps_ens <- compare_scores(fc_spline, fc_ar, fc_var, fc_ens,
                             score = "crps",
                             model_names = c("spline", "ar",
                                              "var", "ensemble"))
summary(crps_ens)
```

```{r ensemble-energy}
energy_ens <- compare_scores(fc_spline, fc_ar, fc_var, fc_ens,
                               score = "energy",
                               model_names = c("spline", "ar",
                                                "var", "ensemble"))
summary(energy_ens)
```

The ensemble does not beat the best individual on every species
or every horizon. It is the *least bad* across the joint
distribution of held-out evaluations, which is the right property
when the user does not know in advance which species or horizon
they will care about most.

## Take-aways

A forecast evaluation workflow in mvgam reuses the same fits
three times. First, `loo()` and `loo_compare()` catch
pathological models cheaply in sample. They do not pick a
forecast model, though. For that, `forecast(newdata = test)`
followed by `score()` is the workhorse. Use `score("crps")` or
`score("drps")` for univariate calibration. Use
`score("energy")` or `score("variogram")` when joint structure
matters. The production question is often closer to "how would
this model have performed run after run": `lfo_cv()` plays out
rolling forecasts and the hybrid PSIS scheme keeps the
refitting cost low. Combine the candidates with `ensemble()`
using stacking weights from `loo_model_weights()`, then score
the ensemble on the same holdout the individuals saw. Pick the
score that matches the decision. Pick the cross-validation
that matches the production question.

## Further reading

The natural next functions to explore on the same fit are:

- `pp_check()` for posterior-predictive checks per outcome
  before any scoring runs.
- `conditional_effects()` and `marginaleffects::plot_predictions()`
  for the covariate surface that drives the forecast.
- `posterior_average()` and `pp_average()` for weighted
  posterior summaries across the candidate models the
  ensemble already mixes.
- `irf()`, `fevd()` and `stability()` for impulse response,
  forecast-error variance decomposition and stationarity
  diagnostics when the latent dynamics are a VAR.
- `methods_md()` for a citation-ready methods paragraph naming
  the architecture pieces the fit actually used.

Sibling articles in this series carry the workflow further:

- The [VAR article](https://nicholasjclark.github.io/mvgam/articles/var.html)
  takes the `VAR(cor = TRUE)` fit used above apart cell by cell,
  with impulse responses, variance decompositions and stability
  metrics.
- The [multi-response article](https://nicholasjclark.github.io/mvgam/articles/mvbf.html)
  shows how to share a single latent state across several
  outcomes with different observation families.
- The [JSDM article](https://nicholasjclark.github.io/mvgam/articles/jsdgam.html)
  applies the latent factor variant to multispecies count data
  with informed loadings via `jsdgam()`.
- The [overview vignette](https://nicholasjclark.github.io/mvgam/articles/mvgam_overview.html)
  introduces the trend constructors the three candidates here
  draw on.
- For grouped trend dynamics (one transition matrix and
  innovation covariance per group, shrunk toward a global mean),
  see the `gr` and `subgr` arguments in `?VAR` and the
  [hierarchical VAR article](https://nicholasjclark.github.io/mvgam/articles/hierarchical_var.html).

## Acknowledgements

The Portal Project rodent capture data ship with mvgam via
`data(portal_data)` and are maintained by the Weecology group
at <https://github.com/weecology/PortalData>. The forecast
evaluation scheme comes from Bürkner, Gabry & Vehtari
([2020](https://doi.org/10.1080/00949655.2020.1783262)). The
PSIS approximation is from Vehtari, Gelman & Gabry
([2017](https://doi.org/10.1007/s11222-016-9696-4)). Stacking
weights are from Yao et al.
([2018](https://doi.org/10.1214/17-BA1091)). The univariate
scoring framework is from Gneiting & Raftery
([2007](https://doi.org/10.1198/016214506000001437)). The
energy score is from Gneiting et al.
([2008](https://doi.org/10.1007/s11749-008-0114-x)). The
variogram score is from Scheuerer & Hamill
([2015](https://doi.org/10.1175/MWR-D-14-00269.1)). The
multispecies forecasting framing follows Clark et al.
([2025](https://doi.org/10.7717/peerj.18929)).

## References

Bürkner P-C, Gabry J and Vehtari A (2020). Approximate
leave-future-out cross-validation for Bayesian time series
models. *Journal of Statistical Computation and Simulation*
90(14), 2499-2523.
[doi:10.1080/00949655.2020.1783262](https://doi.org/10.1080/00949655.2020.1783262)

Clark NJ, Ernest SKM, Senyondo H, Simonis J, White EP, Yenni
GM and Karunarathna KANK (2025). Beyond single-species models:
leveraging multispecies forecasts to navigate the dynamics <!-- prose-lint-disable-line -->
of ecological predictability. *PeerJ* 13, e18929.
[doi:10.7717/peerj.18929](https://doi.org/10.7717/peerj.18929)

Gneiting T and Raftery AE (2007). Strictly proper scoring rules,
prediction, and estimation. *Journal of the American
Statistical Association* 102(477), 359-378.
[doi:10.1198/016214506000001437](https://doi.org/10.1198/016214506000001437)

Gneiting T, Stanberry LI, Grimit EP, Held L and Johnson NA
(2008). Assessing probabilistic forecasts of multivariate
quantities, with an application to ensemble predictions of
surface winds. *TEST* 17(2), 211-235.
[doi:10.1007/s11749-008-0114-x](https://doi.org/10.1007/s11749-008-0114-x)

Scheuerer M and Hamill TM (2015). Variogram-based proper scoring
rules for probabilistic forecasts of multivariate quantities.
*Monthly Weather Review* 143(4), 1321-1334.
[doi:10.1175/MWR-D-14-00269.1](https://doi.org/10.1175/MWR-D-14-00269.1)

Vehtari A, Gelman A and Gabry J (2017). Practical Bayesian model
evaluation using leave-one-out cross-validation and WAIC.
*Statistics and Computing* 27(5), 1413-1432.
[doi:10.1007/s11222-016-9696-4](https://doi.org/10.1007/s11222-016-9696-4)

Yao Y, Vehtari A, Simpson D and Gelman A (2018). Using stacking
to average Bayesian predictive distributions. *Bayesian
Analysis* 13(3), 917-1003.
[doi:10.1214/17-BA1091](https://doi.org/10.1214/17-BA1091)
