---
title: "Hierarchical vector autoregressions with mvgam"
author: "Nicholas J Clark"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
    toc: yes
    toc_float: yes
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE, dpi = 100, fig.asp = 0.62, fig.width = 7,
  out.width = "85%", fig.align = "center"
)
library(mvgam)
library(brms)
library(dplyr)
library(tidyr)
library(ggplot2)
theme_set(mvgam_theme(base_size = 12, base_family = "serif"))
```

When several units are each observed over time on the same set
of outcomes, which VAR should you fit? Pooling every panel into
one shared transition matrix and one shared innovation
covariance squeezes out real heterogeneity between units.
Estimating a separate VAR for each panel throws away the shared
signal that similar units behave similarly, and on short panels
the individual fits collapse into noise. Partial pooling is the
usual answer. Each panel gets its own $(A, \Sigma)$. Each of
those objects is then shrunk toward a global mean the sampler
learns from the whole collection (Savage
[2016](https://rpubs.com/jimsavage/hierarchical_var); Gelman &
Hill [2007](https://doi.org/10.1017/CBO9780511790942)). This
article replicates Jim Savage's 2016 tutorial in mvgam on fresh
World Development Indicators macro data for eight economies. It
then pushes past his summaries into per-country impulse
responses, forecast error variance decompositions and system
stability metrics, all of which fall straight out of the same
fitted object.

## What Savage set out to answer

There are really only three ways to fit a VAR to a group of
economies observed over time. You can pool the data and estimate
a single VAR that applies to every country. You can fit each
country's VAR by itself. Or you can partially pool: keep a
transition matrix and innovation covariance per country, and
shrink each of them toward a global mean the sampler learns
alongside them. The pooled fit throws away all the interesting
differences between countries; the independent fits waste the
common structure that the panels share, and struggle to pin down
anything with only 30 to 50 annual observations each. Savage
([2016](https://rpubs.com/jimsavage/hierarchical_var)) argued
that the third option is the sensible one in practice, and
worked through it in Stan. His formulation kept each country's
$(A_c, \Sigma_c)$ but mixed its innovation correlation with the
global one through a scalar `rho`, using the LKJ-mixture
construction Ben Goodrich contributed to the original post.

Savage's own numbers (WDI data up to about 2015, roughly 313
observations after log-differencing and dropping incomplete
rows) reported a posterior mean global correlation of

$$
\bar\Omega \;\approx\;
\begin{bmatrix}
1.00 & 0.89 & 0.93 \\
0.89 & 1.00 & 0.74 \\
0.93 & 0.74 & 1.00
\end{bmatrix},
$$

a partial pooling weight $\mathrm{E}[\rho] = 0.82$ with a 90%
credible interval of $[0.75,\ 0.90]$, and effective sample sizes
in the low hundreds at $\hat R \approx 1.01$. Those are the
targets a modern mvgam refit should recover on the same data.
The sections below check whether it does. Then they push past
the summaries Savage stopped at into per-country IRFs, FEVDs,
stability metrics and forecasts, all of which fall out of the
same fitted object.

## The data

We use annual growth rates of three real macroeconomic series
(GDP, consumption, gross fixed capital formation) from the World
Development Indicators (World Bank
[2024](https://data.worldbank.org/)) for the eight economies
Savage used (Australia, Canada, Chile, Ireland, New Zealand,
South Africa, the United Kingdom and the United States). The
WDI package pulls the raw levels; a log-difference gives growth
rates.

```{r data-fetch}
raw <- WDI::WDI(
  indicator = c("NY.GDP.MKTP.KN", "NE.CON.TOTL.KN", "NE.GDI.FTOT.KN"),
  start = 1970
)
```

Keep the years where all three indicators are populated, take the
annual log-difference within each country and restrict the panel
to Savage's eight economies.

```{r growth-rates}
wide <- raw |>
  dplyr::filter(stats::complete.cases(raw)) |>
  dplyr::rename(
    GDP  = NY.GDP.MKTP.KN,
    CONS = NE.CON.TOTL.KN,
    GFCF = NE.GDI.FTOT.KN
  ) |>
  dplyr::filter(GDP > 0, CONS > 0, GFCF > 0) |>
  dplyr::group_by(country) |>
  dplyr::arrange(year) |>
  dplyr::mutate(
    dl_gdp  = c(NA, diff(log(GDP))),
    dl_cons = c(NA, diff(log(CONS))),
    dl_gfcf = c(NA, diff(log(GFCF)))
  ) |>
  dplyr::ungroup() |>
  dplyr::filter(stats::complete.cases(dplyr::across(
    c(dl_gdp, dl_cons, dl_gfcf)
  ))) |>
  dplyr::filter(country %in% c(
    "United States", "United Kingdom", "Australia",
    "New Zealand", "Chile", "Canada", "Ireland", "South Africa"
  )) |>
  dplyr::group_by(country) |>
  dplyr::arrange(year) |>
  dplyr::mutate(time = seq_len(dplyr::n())) |>
  dplyr::ungroup()
```

Growth rates then get standardised per outcome. The three
outcomes sit on quite different scales: investment growth has
roughly three times the standard deviation of consumption or GDP
growth. Standardising puts them onto a common variance so a
single prior on AR coefficients and innovation correlations is
equally informative for each outcome. Countries and outcomes
become factors, and every `(country, outcome)` pair gets a
unique `series` label that mvgam uses to identify the 24
outcome-series inside the panel.

```{r reshape}
long <- wide |>
  tidyr::pivot_longer(
    cols = c(dl_gdp, dl_cons, dl_gfcf),
    names_to = "outcome",
    values_to = "y"
  ) |>
  dplyr::mutate(
    country = factor(country),
    outcome = factor(outcome),
    series  = factor(paste(country, outcome, sep = "."))
  )

scales <- long |>
  dplyr::group_by(outcome) |>
  dplyr::summarise(
    mean = mean(y, na.rm = TRUE),
    sd   = stats::sd(y, na.rm = TRUE),
    .groups = "drop"
  )
long <- long |>
  dplyr::left_join(scales, by = "outcome") |>
  dplyr::mutate(y = (y - mean) / sd) |>
  dplyr::select(-mean, -sd)
```

### Padding ragged tails

Panel data in the wild rarely line up cleanly. The WDI pull ends
in different years for different countries: two of the eight in
this panel (New Zealand and the United States) stop one year
earlier than the other six. mvgam requires a **rectangular
panel** on the shared union of time points, so every
`(series, time)` cell must exist as a row with `y = NA` at any
cell where the observation is missing. The per-response NA
machinery then handles the missing rows internally so the
sampler treats them correctly. This padding step is the most
common data-prep gotcha for hierarchical panel fits. Doing it
with `tidyr::expand_grid()` and `dplyr::left_join()` makes the
intent explicit.

```{r padding-example}
full_grid <- tidyr::expand_grid(
  country = levels(long$country),
  outcome = levels(long$outcome),
  time    = seq_len(max(long$time))
) |>
  dplyr::mutate(
    country = factor(country, levels = levels(long$country)),
    outcome = factor(outcome, levels = levels(long$outcome)),
    series  = factor(
      paste(country, outcome, sep = "."),
      levels = levels(long$series)
    )
  )
long <- dplyr::left_join(
  full_grid, long,
  by = c("country", "outcome", "time", "series")
)
```

After padding, `long` has 55 time points × 24 series = 1320 rows,
with `y = NA` at the six cells (two countries × three outcomes)
where WDI stopped early. mvgam validates this rectangle at fit
time; if you skip the pad you will see a data-shape error naming
the mismatched counts.

```{r data-glimpse}
head(long)
data.frame(
  rows      = nrow(long),
  countries = nlevels(long$country),
  outcomes  = nlevels(long$outcome),
  series    = nlevels(long$series)
)
```

## Building the hierarchical VAR

The target model gives each country its own transition matrix
$A_c$ and its own innovation covariance $\Sigma_c$, and shrinks
both toward a global mean the sampler learns simultaneously.
mvgam encodes that structure through three arguments to `VAR()`:
`gr = country` specifies the grouping factor (one matrix per
country), `subgr = outcome` specifies the matrix dimension (3
outcomes per country), and `cor = TRUE` turns on the correlated
innovations that carry the LKJ-mixture pooling on the
innovation correlations. Together the three arguments map to
the multi-panel prior in Savage's formulation:

$$
\begin{aligned}
y_{c, o, t} \;&=\; z_{c, o, t} \\
\mathbf{z}_{c, t} \;&=\; A_c \, \mathbf{z}_{c, t-1} + \boldsymbol{\varepsilon}_{c, t}, \quad
  \boldsymbol{\varepsilon}_{c, t} \sim \mathrm{MVN}\!\left(\mathbf{0},\; \Sigma_c\right), \\
\mathrm{vec}(A_c) \;&\sim\; \mathrm{Normal}\!\left(\mathrm{vec}(\bar A),\; \sigma_A\right), \\
\Sigma_c \;&=\; \mathrm{diag}(\tau_c)\, \Omega_c\, \mathrm{diag}(\tau_c), \\
\Omega_c \;&\sim\; \mathrm{LKJ\text{-}mix}\!\left(\Omega_{\text{global}},\; \alpha_{\text{cor}}\right).
\end{aligned}
$$

Country $c$ has its own transition matrix $A_c$ and innovation
correlation $\Omega_c$; both are pulled toward global objects
$\bar A$ and $\Omega_{\text{global}}$ the sampler learns
simultaneously. The scalar $\alpha_{\text{cor}} \in (0, 1)$
controls how tightly $\Omega_c$ is pulled toward the global
correlation (Savage's `rho`); when
$\alpha_{\text{cor}} \to 1$ every country's correlation matrix
sits at the global mean, when $\alpha_{\text{cor}} \to 0$ each
country's correlation is estimated freely with almost no
pooling.

We replace three priors the mvgam defaults leave diffuse: the AR
coefficient location (`Amu_trend`, tighter around zero than the
default `normal(0, sqrt(0.455))`), the global correlation Cholesky
(`L_Omega_global_trend`, mildly shrunk toward the identity with
`lkj(2)`), and the residual observation SD (already small on
standardised data). The rest of the priors are the mvgam
defaults.

```{r fit-code}
custom_priors <- c(
  prior(normal(0, 0.15),         class = "Amu_trend"),
  prior(lkj_corr_cholesky(2),    class = "L_Omega_global_trend"),
  prior(student_t(3, 0, 0.5),    class = "sigma")
)

mod_hier <- mvgam(
  formula       = y ~ 0,
  trend_formula = ~ VAR(p = 1, gr = country, subgr = outcome,
                          cor = TRUE),
  data          = long,
  family        = gaussian(),
  prior         = custom_priors,
  chains        = 4,
  cores         = 4,
  iter          = 700,
  warmup        = 500,
  control       = list(adapt_delta = 0.95, max_treedepth = 15),
  silent        = 2
)
```

The Heaps ([2022](https://doi.org/10.1080/10618600.2022.2079648))
stationarity transform runs once per country per gradient
evaluation, so a hierarchical VAR is roughly $C$ times slower
per iteration than a single-panel fit of the same dimension.
The sampler needs `adapt_delta = 0.95` and
`max_treedepth = 15` to keep divergence counts low on this
panel; the tight `Amu_trend` prior at `normal(0, 0.15)` reflects
that macro AR(1) coefficients on log-difference growth rates
concentrate near zero.

## Model summary

`summary()` on a hierarchical VAR fit reports the shared
observation parameters (a single residual SD here), the trend
metadata, and the global trend hyperparameters. Each country
carries its own $A_c$ and $\Sigma_c$, and those arrays run to
hundreds of cells on this panel, so `include_betas = FALSE`
keeps the printout focused on the shrinkage targets `Amu_trend`,
`Aomega_trend`, `L_Omega_global_trend`, the pooling weight
`alpha_cor_trend`, and the innovation SDs `sigma_group_trend`
for each country and outcome. Individual $A_c$ and $\Sigma_c$
blocks are pulled out with `posterior_transition_matrix()`,
`residual_cor()`, and `mcmc_plot()` in the sections below.

```{r summary}
summary(mod_hier, include_betas = FALSE)
```

## Diagnostics: convergence and identification

The `summary()` above runs mvgam's internal convergence advisor,
which filters out latent state parameters (the trajectories
$\mathbf{z}_{c,t}$) before checking $\hat R$ and effective
sample size on the structural parameters that carry the story.
The advisor emits warnings for parameters with $\hat R > 1.05$
or bulk ESS below $100 \times$ chains, so any surviving message
tells you which block needs attention.

For this fit the country transition matrices ($A_c$), the
country innovation covariances ($\Sigma_c$), the global
shrinkage targets (`Amu_trend`, `Aomega_trend`,
`L_Omega_global_trend`) and the partial pooling weight
(`alpha_cor_trend`) all converge cleanly at $\hat R \leq 1.01$.
The one weak link is `sigma`, the residual observation SD: a
Gaussian state space model on continuous data cannot fully
separate observation noise from process innovation noise
(Auger-Méthé et al.
[2021](https://doi.org/10.1002/ecm.1470)). Both scales enter the
likelihood only through the sum of their variances at any given
timepoint, so the sampler traces a ridge along which their
posteriors trade off.

The `summary()` output above reports `sigma` with a small
posterior mean (near 0.05 on the standardised scale) but very
large `Rhat` and small effective sample size. Practically this
says the observation SD hugs zero while the process innovation
SDs (`sigma_group_trend[c, k]`, further down the summary) carry
almost all of the variance. Treat `sigma` as poorly identified
here and interpret only the structural blocks below.

## Transition matrices per country

How much does the partial-pooling prior shrink each country's
AR persistence toward the global mean? Plotting the medians and
95% intervals for the self-persistence diagonal by country
answers that directly. `posterior_transition_matrix(mod_hier,
groups = <name>)` returns the `[ndraws, K, K]` array of posterior
draws for one country's $A_c$, `groups = "all"` returns the set
of them classed for a faceted plot, and the default `groups =
NULL` returns the shrinkage target built from the
hyperparameters `Amu_trend[1, 1]` (diagonal mean) and
`Amu_trend[2, 1]` (off-diagonal mean).

```{r plot-A-country-diag, fig.cap = "Posterior medians and 95% intervals for the three self-persistence coefficients (diagonal of A_c) by country. Points close together across countries indicate that the partial-pooling prior has pulled the country-specific estimates toward the same global value."}
A_all <- posterior_transition_matrix(mod_hier, groups = "all")
plot(A_all, type = "diagonal")
```

## The global innovation correlation

The global correlation matrix $\Omega_{\text{global}}$ tells you
which outcomes tend to share contemporaneous shocks across
countries once the country-specific piece has been factored out.
`residual_cor(mod_hier)` reads the LKJ Cholesky
`L_Omega_global_trend` internally and returns a classed
`mvgam_residcor` object. That object carries, per cell, the
posterior mean, standard error, credible interval, effective
sample size, and probability of a nonzero correlation. The `cor`
slot is the correlation matrix a reader will usually want to
print.

```{r sigma-cor-global}
rc_global <- residual_cor(mod_hier)
round(rc_global$cor, 2)
```

## The partial-pooling weight

The scalar `alpha_cor_trend` mixes each country's innovation
correlation between its own data-driven estimate and the global
$\Omega_{\text{global}}$. Values close to one indicate strong
pooling (country correlations sit near the global mean); values
close to zero indicate little pooling (each country's correlation
is estimated almost freely).

```{r plot-alpha-cor, fig.cap = "Posterior distribution of the partial-pooling weight alpha_cor_trend. Values near one mean country-specific innovation correlations are pulled tightly toward the global correlation; values near zero mean each country is estimated with little pooling."}
mcmc_plot(mod_hier,
          variable = "alpha_cor_trend",
          type = "hist")
```

## Country-specific innovation correlations

The pooling weight `alpha_cor_trend` sets how tightly each
country's innovation correlation is pulled toward the global
mean; heatmaps of the per-country posterior median correlations
show how that pull plays out. If the weight is high, the country
heatmaps should look close to the global heatmap above. If it is
low, some countries should show idiosyncratic structure.

```{r plot-country-cor, fig.width = 8, fig.height = 5, fig.cap = "Posterior mean innovation correlation matrix for each country. Each cell is the correlation between two outcomes (GDP, consumption, investment) once each country's transition matrix has absorbed the lagged structure. Similarity across countries reflects the pull of the alpha_cor_trend partial-pooling weight."}
plot(
  residual_cor(mod_hier, by_group = TRUE),
  include_global = FALSE
)
```

## Impulse responses (per country)

`irf()` traces how a unit shock to one outcome propagates
through the VAR over the next $h$ years. On the hierarchical
fit the routine operates on the full block-diagonal
`A_trend` (dimension $24 \times 24$), so the default plot
would fan out to 24 response panels per shock; `responses`
narrows the display to the panels of interest. The three
Australian processes are packed contiguously in slots 1-3, so
selecting `series = 2` (Australian GDP) with
`responses = 1:3` shows Australia's within-country IRF.

```{r irf, fig.cap = "Generalized impulse response functions for Australia over an eight-year horizon. Each panel shows how one Australian outcome responds to a positive shock in Australian GDP."}
irfs <- irf(mod_hier, h = 8)
plot(irfs, series = 2, responses = 1:3)
```

The diagonal panel (`Process_2 -> Process_2`) traces own-lag
persistence in GDP. The off-diagonal panels trace spillover to
consumption and investment. Both spillovers are positive and
substantial at short horizons, matching the strong global
correlations we already saw in $\Omega_{\text{global}}$.

## Forecast error variance decomposition

`fevd()` turns the same lagged machinery around: at horizon $h$,
what fraction of the forecast variance for outcome $i$ was
generated by shocks in each source outcome? On a $24$-dimensional
hierarchical VAR the raw output has 24 target panels each
decomposed over 24 sources, so we pass `series = 1:3` and
`contributing = 1:3` to keep the display within Australia. The
retained shares get renormalised per (target, horizon) so each
bar still reads as a proper decomposition.

```{r fevd, fig.cap = "Forecast error variance decomposition for Australia. Each panel is an Australian target outcome; each bar breaks that outcome's forecast variance at a given horizon into contributions from the three Australian source outcomes."}
fevds <- fevd(mod_hier, h = 8)
plot(fevds, series = 1:3, contributing = 1:3)
```

At horizon one, the share attributed to an outcome's own past
dominates: the one-step forecast error is almost entirely that
outcome's own innovation. Longer horizons let the VAR integrate
shocks through the transition matrix, so the cross-outcome
slices grow.

## System-wide stability metrics

`stability()` returns posterior draws of three interpretable
system dynamics metrics on the fitted transition matrix (Ives
et al.
[2003](https://doi.org/10.1890/0012-9615(2003)073[0301:ECSAEI]2.0.CO;2);
Neubert & Caswell
[1997](https://doi.org/10.1890/0012-9658(1997)078[0653:ATRFMT]2.0.CO;2)):
reactivity, mean return rate and variance return rate. On a
hierarchical fit the routine operates on the assembled
$A_{\text{trend}}$ of the whole eight-country pool. The
histograms below therefore describe the system as a whole,
not any single country. Two accompanying summaries mask on
the structural zeros the block diagonal enforces:
`prop_int_offdiag` isolates the share of interaction strength
that lies within a country's own block, and
`prop_cov_offdiag` does the same for the innovation
correlations. Both quantities let you check how much of the
system's dynamics live within countries and how much crosses
between them.

```{r stability, fig.width = 8, fig.asp = 0.45, fig.cap = "Posterior distributions of three system dynamics metrics for the assembled 24-dimensional VAR. The dashed line marks zero, the reactivity threshold above which small shocks would grow immediately after the shock before decaying."}
stab <- stability(mod_hier)
plot(stab)
```

The posterior median mean return rate lands near 0.5. A
typical shock therefore decays to roughly one-half of its
magnitude after a single year, and to a negligible size within
a handful of years thereafter, reflecting the mild persistence
implied by the AR coefficients in $\bar A$. The reactivity
distribution straddles zero. Some draws show a brief transient
amplification of a shock before decay sets in; others do not.

## Take-aways

Once fitted, a small set of helpers exposes the hierarchical
VAR at any level:

- `summary(mod, include_betas = FALSE)` keeps the printout
  focused on the hyperparameters.
- `posterior_transition_matrix(mod, groups = <name>)` returns
  each country's $A_c$ as a `[ndraws, K, K]` array. Leave
  `groups` at its default for the shrinkage target built from
  `Amu_trend`.
- `residual_cor(mod)` returns the global innovation correlation
  as an `mvgam_residcor` object with print, summary and plot
  methods. Pass `by_group = TRUE` to receive the same object per
  country plus a `_global` reference matrix.
- `mcmc_plot(mod, variable = "alpha_cor_trend")` reads off how
  much pooling the data support on the innovation correlations.
- `irf(mod)` and `fevd(mod)` operate on the assembled
  $24 \times 24$ transition matrix. Their `plot` methods take
  `series`, `responses` and `contributing` selectors, so a
  hierarchical fit can be inspected one country at a time.

Partial pooling on the correlation structure
(`alpha_cor_trend`) balances information sharing with
country-level adaptation. Each economy keeps its own dynamics
and its own shocks, but every panel borrows strength from the
whole pool. That middle ground matters. Pooled VARs cannot
reach it, because forcing one shared $(A, \Sigma)$ on eight
economies that differ in volatility and persistence throws
away the very heterogeneity a reader would look at these
countries for. Fully unpooled VARs abandon it too, because 30
to 50 annual observations per country are far too short to pin
down a 3 × 3 transition matrix and a 3 × 3 innovation
covariance in isolation. Only the hierarchical fit recovers
Savage's global correlation (diagonal 1, off-diagonals near 0.9
and 0.65 on this WDI vintage) while still letting Chile carry
three times the innovation scale of Australia.

## Further reading

Complementary in-package tools:

- `?VAR` for the trend-side constructor; the `gr` and `subgr`
  arguments are what turn on the hierarchical panel structure
  used here.
- `?forecast.mvgam` and `?score` still work on hierarchical VAR
  fits and forecast each country's outcomes forward using its own
  $A_c$ and $\Sigma_c$.
- `?how_to_cite` for a methods paragraph you can drop into a
  manuscript. It describes the Heaps stationary prior, the LKJ
  mixture pooling on the correlation matrices and the sampler
  configuration.

Sibling articles in this series:

- The
  [VAR article](https://nicholasjclark.github.io/mvgam/articles/var.html)
  is the single-panel version of the same machinery, with a full
  treatment of impulse responses, forecast error variance
  decomposition and community stability metrics.
- The
  [multi-response article](https://nicholasjclark.github.io/mvgam/articles/mvbf.html)
  covers a different generalisation: several outcomes on
  different observation families sharing one latent process.

## Acknowledgements

The World Development Indicators dataset is maintained by the
World Bank ([2024](https://data.worldbank.org/)); the accompanying
`WDI` R package (Arel-Bundock
[2013](https://doi.org/10.32614/RJ-2013-023)) provides the long
panel pull used here. The hierarchical VAR framing follows
Jim Savage's tutorial post
([2016](https://rpubs.com/jimsavage/hierarchical_var)) and the
partial-pooling perspective in Gelman & Hill
([2007](https://doi.org/10.1017/CBO9780511790942)). The
stationary Heaps prior on the VAR coefficients follows Heaps
([2022](https://doi.org/10.1080/10618600.2022.2079648)); the
LKJ-mixture prior on the correlation matrices follows the
suggestion Ben Goodrich contributed to Savage's original post.
Formulas rely on mgcv smooths (Wood
[2017](https://doi.org/10.1201/9781315370279)) and brms's
multilevel apparatus (Bürkner
[2017](https://doi.org/10.18637/jss.v080.i01)); sampling is
performed with Stan (Carpenter et al.
[2017](https://doi.org/10.18637/jss.v076.i01)). The state-space
DGAM framing follows Clark & Wells
([2023](https://doi.org/10.1111/2041-210X.13974)).

## References

Arel-Bundock V (2013). WDI: World Development Indicators and
other World Bank data. *The R Journal* 5(2), 190-192.
[doi:10.32614/RJ-2013-023](https://doi.org/10.32614/RJ-2013-023)

Bürkner PC (2017). brms: An R package for Bayesian multilevel
models using Stan. *Journal of Statistical Software* 80(1),
1-28.
[doi:10.18637/jss.v080.i01](https://doi.org/10.18637/jss.v080.i01)

Carpenter B, Gelman A, Hoffman MD, Lee D, Goodrich B, Betancourt
M, Brubaker M, Guo J, Li P and Riddell A (2017). Stan: A
probabilistic programming language. *Journal of Statistical
Software* 76(1), 1-32.
[doi:10.18637/jss.v076.i01](https://doi.org/10.18637/jss.v076.i01)

Clark NJ and Wells K (2023). Dynamic Generalized Additive Models
(DGAMs) for forecasting discrete ecological time series.
*Methods in Ecology and Evolution* 14, 771-784.
[doi:10.1111/2041-210X.13974](https://doi.org/10.1111/2041-210X.13974)

Gelman A and Hill J (2007). *Data Analysis Using Regression and
Multilevel/Hierarchical Models*. Cambridge University Press.
[doi:10.1017/CBO9780511790942](https://doi.org/10.1017/CBO9780511790942)

Heaps SE (2023). Enforcing stationarity through the prior in
vector autoregressions. *Journal of Computational and Graphical
Statistics* 32(1), 74-83.
[doi:10.1080/10618600.2022.2079648](https://doi.org/10.1080/10618600.2022.2079648)

Savage J (2016). Hierarchical Vector Autoregression.
RPubs, 27 November 2016.
[https://rpubs.com/jimsavage/hierarchical_var](https://rpubs.com/jimsavage/hierarchical_var)

Wood SN (2017). *Generalized Additive Models: An Introduction
with R* (2nd edition). Chapman and Hall/CRC.
[doi:10.1201/9781315370279](https://doi.org/10.1201/9781315370279)

World Bank (2024). *World Development Indicators*. World Bank
Group.
[https://data.worldbank.org/](https://data.worldbank.org/)
