---
title: "Integrated distribution models with mvgam"
author: "Nicholas J Clark"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
    toc: yes
    toc_float: yes
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE, dpi = 100, fig.asp = 0.62, fig.width = 7,
  out.width = "85%", fig.align = "center"
)
library(mvgam)
library(brms)
library(dplyr)
library(ggplot2)
library(patchwork)
theme_set(theme_bw(base_size = 12, base_family = "serif"))
# The article's plots reuse the same red single-hue scheme every
# other mvgam plotting method locks in via `set_color_scheme_local`;
# fixing it globally in setup keeps hand-rolled overlays and
# `bayesplot::mcmc_plot()` panels visually consistent with
# `plot.mvgam_forecast()`, `plot.mvgam_conditional_smooths()` etc.
bayesplot::color_scheme_set("red")
red_pal <- bayesplot::color_scheme_get("red")

cache_dir <- if (dir.exists("pkgdown")) {
  file.path("pkgdown", "idm_cache")
} else {
  file.path("..", "..", "pkgdown", "idm_cache")
}
read_cache <- function(name) {
  readRDS(file.path(cache_dir, paste0(name, ".rds")))
}
```

Ecologists studying a single species rarely see it through a
single lens. A long-running roadside count route delivers
structured counts. A regional occupancy protocol adds
detection / non-detection. Citizen-science portals stream in
opportunistic sightings whose sampling effort is largely
unmodelled. Analysing each stream in isolation throws away the
shared signal every stream is a noisy view of, and no single
stream is sharp enough to recover a nonlinear environmental
response on its own. The integrated distribution model tradition
(Fithian et al.
[2015](https://doi.org/10.1111/2041-210X.12242); Isaac et al.
[2020](https://doi.org/10.1016/j.tree.2019.08.006); Miller et
al. [2019](https://doi.org/10.1111/2041-210X.13110)) resolves
this by fitting one likelihood per data stream and tying every
likelihood to the same underlying intensity. mvgam expresses
that pattern natively via `mvbf()`: one `bf()` outcome per data
stream, each with its own observation family, all sharing a
single latent process on the log-intensity scale. This article
walks through a simulated three-stream monitoring scenario where
the truth is known, so the joint fit's inferences can be checked
directly.

## The data

We simulate one species observed monthly for eight years (96
time points). The latent log-intensity carries a smooth
environmental response and a sparse seasonal autoregression:

$$
\log \lambda_t \;=\; \alpha \;+\; f(\text{env}_t) \;+\; z_t,
\qquad z_t \sim \mathrm{AR}\big(\phi_1 z_{t-1} + \phi_{12} z_{t-12}, \sigma^2\big)
$$

with $f$ a bell-shaped optimum, $\phi_1 = 0.55$, $\phi_{12} = 0.30$
and $\sigma = 0.35$. An autoregression with lag-1 momentum and a
lag-12 seasonal recurrence fits naturally with
`AR(p = c(1, 12))`. Auger-Méthé et al.
([2021](https://doi.org/10.1002/ecm.1470)) review the wider state
space family. Three observation processes deliver a partial view
of $\lambda_t$:

- **Counts.** A structured roadside route delivers Poisson counts
  at effort $E_t$: $\text{count}_t \sim \text{Poisson}(\lambda_t E_t)$,
  modelled with `family = poisson()` and `offset(log_effort)`.
- **Presence-absence.** An occupancy protocol surveys a fixed area
  $A$ per visit. The cloglog link makes
  $p_t = 1 - \exp(-\lambda_t A)$, so detection probability ties
  directly to the log intensity (McCullagh & Nelder
  [1989](https://doi.org/10.1201/9780203753736), §4.2.5). The
  matching mvgam call is `family = bernoulli(link = "cloglog")`.
- **Presence-only.** Opportunistic sightings arrive under an
  unmodelled observer effort. Treating them as a thinned Poisson
  process with its own intercept absorbs the mean log thinning
  constant and leaves the shape of $f(\text{env})$ identified
  (Fithian et al.
  [2015](https://doi.org/10.1111/2041-210X.12242); Renner et al.
  [2015](https://doi.org/10.1111/2041-210X.12352)). The matching
  call is `family = poisson()` with no offset and a free
  intercept.

The simulator is deliberately compact so the whole workflow
stays inspectable. Every step is shown here in full:

```{r data-sim}
set.seed(2026)
n_t <- 96L

# Environmental covariate: a smoothly varying index across time.
env <- as.numeric(scale(cumsum(rnorm(n_t, 0, 0.3))))

# True environmental response: bell-shaped optimum with peak
# near env = 0.4.
f_env <- function(x) 1.5 * exp(-0.5 * (x - 0.4)^2) - 0.4

# Sparse AR(1, 12) latent state with monthly momentum and an
# annual seasonal recurrence.
phi_1 <- 0.55; phi_12 <- 0.30; sigma_z <- 0.35
z <- numeric(n_t)
for (t in seq_len(n_t)) {
  eta <- rnorm(1, 0, sigma_z)
  z[t] <- eta +
    (if (t > 1L)  phi_1  * z[t - 1L]  else 0) +
    (if (t > 12L) phi_12 * z[t - 12L] else 0)
}

# Shared log-intensity that every measurement family will
# link back to.
alpha_shared <- 0.5
log_lambda   <- alpha_shared + f_env(env) + z
lambda       <- exp(log_lambda)

# Three observation streams.
E_c   <- runif(n_t, 0.5, 2)                    # per-time effort
count <- rpois(n_t, lambda = lambda * E_c)

A_pa  <- 0.25                                  # occupancy area
p_pa  <- 1 - exp(-lambda * A_pa)               # cloglog p
pa    <- rbinom(n_t, size = 1L, prob = p_pa)

alpha_po_offset <- -0.8
log_bias <- alpha_po_offset + rnorm(n_t, 0, 0.15)
po       <- rpois(n_t, lambda = lambda * exp(log_bias))

# The mvbf fit expects a single frame where each response column
# carries one measurement. A `series` factor keeps a single
# biological unit tied together across responses.
ipm_data <- data.frame(
  time       = seq_len(n_t),
  env        = env,
  log_effort = log(E_c),
  count      = count,
  pa         = pa,
  po         = po,
  series     = factor("series1")
)

# Truth vectors used later to score recovery.
truth <- list(env = env, f_env_true = f_env(env),
               z = z, phi_1 = phi_1, phi_12 = phi_12,
               sigma_z = sigma_z)

head(ipm_data)
```

The three streams sit on very different scales but track the
same latent process. `mvgam_data()` builds the package's
standard series display from a long-format frame with one row
per measurement / time pair:

```{r data-view, fig.width = 9, fig.height = 3.4, fig.cap = "Simulated three-stream monitoring data over 96 months."}
long <- data.frame(
  time   = rep(ipm_data$time, 3L),
  y      = c(ipm_data$count, ipm_data$pa, ipm_data$po),
  series = factor(rep(c("count", "pa", "po"), each = n_t),
                    levels = c("count", "pa", "po"))
)
mvgam_data(long, y = "y", family = poisson(), series = "all")
```

Counts and opportunistic sightings both spiked in the middle of
the record where $f(\text{env})$ crossed its optimum. The
occupancy stream carried only one bit per time point.

## The joint fit

The joint model fits all three measurements in one `mvgam()`
call. Three `bf()` outcomes chain with `+`. Each keeps its own
family and its own observation-scale predictors; the trend
formula is written once and applies to all of them. Setting
`trend_map = "shared"` says explicitly that every measurement
reads off the same latent process $\log \lambda_t$, so
`s(env)` and the seasonal autoregression are estimated once
from the pooled information in all three streams rather than
three times from three narrower views of it.

```{r joint-fit, eval = FALSE}
joint_formula <-
  bf(count ~ 1 + offset(log_effort), family = poisson()) +
  bf(pa    ~ 1, family = bernoulli(link = "cloglog")) +
  bf(po    ~ 1, family = poisson())

mod_joint <- mvgam(
  joint_formula,
  trend_formula = ~ s(env, k = 8) + AR(p = c(1, 12)),
  trend_map     = "shared",
  data = ipm_data,
  chains = 2L, samples = 500L,
  control = list(adapt_delta = 0.95),
  silent = 2L
)
```

```{r joint-load, include = FALSE}
mod_joint <- read_cache("mod_joint")
```

`trend_map = "shared"` collapses every measurement onto a
single latent process, which is the whole point of an
integrated distribution model. The alternative `"identity"`
gives each measurement its own trend; a `data.frame(series,
trend)` handles the intermediate case where some measurements
share and others do not.

Each measurement has its own intercept. The count intercept
sits on the log-count scale after removing `log_effort`. The
presence-absence intercept absorbs $\log A$, so under the
cloglog link it acts as the reference detection probability.
The presence-only intercept absorbs the mean log observer
effort. None of these three constants are of scientific interest
on their own. The shape they leave behind in the shared trend
is.

Convergence, per-measurement intercepts and the parameters on
the trend side all arrive together from `summary()`:

```{r joint-summary}
summary(mod_joint, smooth_test = FALSE)
```

The `ar1_trend[1]`, `ar12_trend[1]` and `sigma_trend[1]`
posteriors each cover the simulation truth: $\phi_1 = 0.55$,
$\phi_{12} = 0.30$, $\sigma = 0.35$. `Rhat` sits at 1.00 and
effective sample sizes stay well above the usual 400-per-chain
floor, so the trend parameters are identified from the data
rather than mopped up by whatever else the model can absorb.

## Interpreting the joint fit

The environmental response, the seasonal autoregression and the
shared latent trajectory carry the ecological content of the
fit. Each has a one-line accessor in the package.

### The environmental response

`plot(mod, type = "smooths")` renders the trend-side smooth via
the `mvgam_conditional_smooths` class. Median and 95% band come
from posterior draws; the display matches the house theme every
other `plot.mvgam*` method uses:

```{r smooth-env, fig.width = 6.5, fig.height = 3.4, fig.cap = "Trend-side s(env) recovered by the joint fit. The bell peaks near env = 0.4, matching the simulation truth."}
plot(mod_joint, type = "smooths")
```

The peak sits near env $= 0.4$, and the band tightens where env
values were densely observed. Read as an ecological response,
the joint fit places this species' optimum in the middle of the
observed gradient and prefers a symmetric decline in either
direction. That is precisely the shape `f_env()` produced.

### The seasonal autoregression

The trend-side stochastic parameters tell you how much of the
time series is monthly momentum, annual recurrence and
unstructured shocks. `bayesplot::mcmc_plot()` overlays their
posterior densities:

```{r joint-pars, fig.width = 6.5, fig.height = 3, fig.cap = "Marginal posteriors for the seasonal AR coefficients and the innovation SD from the joint fit."}
mcmc_plot(
  mod_joint,
  variable = c("ar1_trend[1]", "ar12_trend[1]", "sigma_trend[1]"),
  type = "areas"
)
```

Each posterior concentrates on the correct value. The lag-12
posterior in particular is well separated from zero, which
matters: the fit has picked up the annual recurrence rather
than mopping it into the smooth.

### The shared latent trajectory

`hindcast(mod, type = "trend")` returns a `mvgam_forecast`
object over the training window; its plot method draws the
median trajectory with 50 / 80 / 95% credible bands.

```{r trend-hindcast, fig.width = 7, fig.height = 3.4, fig.cap = "Posterior hindcast of the shared latent process from the joint fit. Ribbons are 50 / 80 / 95% credible intervals."}
plot(hindcast(mod_joint, type = "trend")) +
  labs(x = "month", y = "shared trend")
```

Every measurement's likelihood reads off this trajectory, so
its dynamics belong to the model as a whole rather than to any
one data stream. Read here: annual recurrence visible in the
peaks, a monthly-scale wobble on top, band tightest where the
count stream is most informative.

### Posterior predictive checks per measurement

Each observation family also has to reproduce its own marginal.
`pp_check(mod, resp = ...)` scopes to one response, and the
`type` argument picks the display appropriate for that
measurement's support.

```{r pp-count, fig.width = 6.5, fig.height = 3, fig.cap = "Density overlay of yrep versus y for the count measurement. Twenty posterior draws (light lines) sit over the observed histogram (dark line)."}
pp_check(mod_joint, resp = "count",
         type = "dens_overlay", ndraws = 20L)
```

```{r pp-pa, fig.width = 6.5, fig.height = 3, fig.cap = "Bar plot of the 0/1 marginal for the presence-absence measurement."}
pp_check(mod_joint, resp = "pa", type = "bars", ndraws = 200L)
```

```{r pp-po, fig.width = 6.5, fig.height = 3, fig.cap = "ECDF overlay of yrep versus y for the presence-only measurement. The ECDFs bracket the observed step function across the range."}
pp_check(mod_joint, resp = "po",
         type = "ecdf_overlay", ndraws = 50L)
```

Choosing the display matters. Density overlays are natural for
overdispersed counts, bars for a binary response, ECDFs for
right-skewed integers where the density smoothing muddies the
tail. All three checks pass here; if any were failing, that
would be the response whose observation family needed to
change.

## When integration breaks

The joint fit assumes every measurement responds to the
environment through the same functional form of the same
underlying intensity. When that assumption fails, integration
does not help. Two situations are common enough to name.

The first is species mismatch. Suppose the count route surveys
one species and the citizen-science portal collects records
for an easily confused congener. The joint fit treats the
mixture as one species and puts the optimum somewhere between
the two truths. Fithian et al.
([2015](https://doi.org/10.1111/2041-210X.12242)) analyse this
identifiability directly. The practical check is a residual
plot per measurement: correlated residuals across measurements
that both rise or fall together at the same time points suggest
one shared driver; residuals that move oppositely suggest the
measurements are pointing at different biological states.

The second is response mismatch. If the presence-absence
protocol picks up the species in a broader habitat than the
count route (say, edge habitat that the count route misses
because it walks transects only through core habitat), the
smooth of env each protocol *would* fit alone will differ. The
joint fit averages them. That average is nobody's response
function. The remedy is to let the observation side carry
whichever part of the response depends on the protocol: put
a smooth on the `bf()` outcome for the measurement where the
mismatch lives (for example, `bf(pa ~ 1 + s(env), family = bernoulli(link = "cloglog"))`),
and keep only the shared response on the trend side.
Koshkina et al.
([2017](https://doi.org/10.1111/2041-210X.12738)) develop the
formal framework for combining presence-background data with
site occupancy under exactly this assumption.

Both failure modes are diagnostic. A joint `pp_check()` that
sits noticeably worse than any baseline fit on one measurement
alone is the signal that the shared response assumption is
wrong for the data at hand.

## Comparing the joint fit to alternatives

Real analyses do not have the truth to check inferences against.
This section runs the checks that make the joint fit
defensible in that setting: contrast against single-measurement
baselines, contrast against a plausibly misspecified trend, and
comparison via `loo()` and `lfo_cv()`.

### Single-measurement baselines

Each stream can be fit alone with the same trend specification.
That yields one posterior per stream over the environmental
response, ready for direct comparison against the joint fit.

```{r baselines-fit, eval = FALSE}
tf <- ~ s(env, k = 8) + AR(p = c(1, 12))

mod_count <- mvgam(
  bf(count ~ 1 + offset(log_effort), family = poisson()),
  trend_formula = tf,
  data = ipm_data,
  chains = 2L, samples = 500L,
  control = list(adapt_delta = 0.95),
  silent = 2L
)
mod_pa <- mvgam(
  bf(pa ~ 1, family = bernoulli(link = "cloglog")),
  trend_formula = tf,
  data = ipm_data,
  chains = 2L, samples = 500L,
  control = list(adapt_delta = 0.95),
  silent = 2L
)
mod_po <- mvgam(
  bf(po ~ 1, family = poisson()),
  trend_formula = tf,
  data = ipm_data,
  chains = 2L, samples = 500L,
  control = list(adapt_delta = 0.95),
  silent = 2L
)
```

```{r baselines-load, include = FALSE}
mod_count <- read_cache("mod_count")
mod_pa    <- read_cache("mod_pa")
mod_po    <- read_cache("mod_po")
```

To compare recovered smooths on a common footing, we evaluate
each fit on the same env grid and centre draws at the median
grid value so the per-measurement intercept drops out of the
shape comparison.

```{r compare-helpers}
grid <- seq(min(ipm_data$env), max(ipm_data$env),
             length.out = 60L)
f_true_grid <- 1.5 * exp(-0.5 * (grid - 0.4)^2) - 0.4
ref_env     <- median(grid)
f_true_ref  <- 1.5 * exp(-0.5 * (ref_env - 0.4)^2) - 0.4

build_nd <- function() {
  nd <- ipm_data[rep(1L, length(grid)), , drop = FALSE]
  nd$env        <- grid
  nd$log_effort <- median(ipm_data$log_effort)
  nd$count      <- NA_integer_
  nd$pa         <- NA_integer_
  nd$po         <- NA_integer_
  nd
}

env_summary <- function(mod, label, resp = NULL) {
  nd    <- build_nd()
  lpred <- if (is.null(resp)) {
    posterior_linpred(mod, newdata = nd)
  } else {
    posterior_linpred(mod, newdata = nd, resp = resp)
  }
  lpred <- as.matrix(lpred)
  ref_j <- which.min(abs(grid - ref_env))
  centered <- sweep(lpred, 1L, lpred[, ref_j], "-")
  q <- apply(centered, 2L, quantile,
              probs = c(0.025, 0.5, 0.975), na.rm = TRUE)
  data.frame(
    model = label, env = grid,
    med = q["50%", ], lo = q["2.5%", ], hi = q["97.5%", ]
  )
}

truth_df <- data.frame(env = grid,
                        y_true = f_true_grid - f_true_ref)
```

```{r compare-sharpening, fig.width = 10, fig.height = 3.6, fig.cap = "Recovery of the environmental response by each single-measurement baseline and by the joint fit. Solid black is truth; the red ribbon is the 95% credible interval for the log-intensity shift relative to the median env."}
sharp_summ <- rbind(
  env_summary(mod_joint, "joint", resp = "count"),
  env_summary(mod_count, "count only"),
  env_summary(mod_pa,    "pa only"),
  env_summary(mod_po,    "po only")
)
sharp_summ$model <- factor(sharp_summ$model,
  levels = c("joint", "count only", "pa only", "po only")
)
ggplot(sharp_summ) +
  geom_ribbon(aes(env, ymin = lo, ymax = hi),
               fill = red_pal$light, alpha = 0.6) +
  geom_line(aes(env, med), colour = red_pal$dark_highlight,
             linewidth = 0.7) +
  geom_line(data = truth_df, aes(env, y = y_true),
             colour = "black", linewidth = 0.7) +
  facet_wrap(~ model, nrow = 1L) +
  labs(x = "env", y = "log-intensity shift")
```

The count-only and joint panels both concentrate around the
correct bell, with the joint fit reporting a slightly narrower
band. The presence-only baseline recovers the shape but with a
softer peak; the presence-absence baseline lands on a monotone
smooth. An ecologist reading only the occupancy fit would
conclude the species prefers the driest end of the gradient.
The species does not.

Recovery numbers make the same point in one table:

```{r compare-metric}
metric <- sharp_summ %>%
  group_by(model) %>%
  summarise(
    rmse          = sqrt(mean((med - truth_df$y_true)^2)),
    mean_ci_width = mean(hi - lo),
    coverage_95   = mean(truth_df$y_true >= lo &
                          truth_df$y_true <= hi)
  )
metric_out <- as.data.frame(metric)
metric_out[, -1L] <- round(metric_out[, -1L], 3)
metric_out
```

The joint fit's band is the narrowest and its 95% coverage
sits close to nominal. Presence-absence has the widest band
yet undercovers, because its posterior is confidently the
wrong shape. Integrated distribution models exist for this
pattern: pool likelihoods that see the same underlying process
through different sampling channels, so the shared functionals
of that process tighten around the same answer.

### A misspecified trend

Real fits are chosen from a small set of candidates. Dropping
the lag-12 term is a common enough shortcut that we can measure
what it costs. `AR(p = 1)` in place of `AR(p = c(1, 12))`
leaves everything else untouched:

```{r wrong-trend-fit, eval = FALSE}
mod_wrong_trend <- mvgam(
  bf(count ~ 1 + offset(log_effort), family = poisson()) +
    bf(pa ~ 1, family = bernoulli(link = "cloglog")) +
    bf(po ~ 1, family = poisson()),
  trend_formula = ~ s(env, k = 8) + AR(p = 1),
  trend_map     = "shared",
  data = ipm_data,
  chains = 2L, samples = 500L,
  control = list(adapt_delta = 0.95),
  silent = 2L
)
```

```{r wrong-trend-load, include = FALSE}
mod_wrong_trend <- read_cache("mod_wrong_trend")
```

The annual recurrence has to go somewhere. With no lag-12
term to absorb it, the smooth over-fits the segments of the
record where env happens to align with the recurring peak, and
the recovered response function is visibly flatter than the
truth:

```{r wrong-trend-plot, fig.width = 8, fig.height = 3.6, fig.cap = "Recovered env response: joint fit versus the AR(1)-only trend. The wrong-trend fit flattens the bell as its smooth absorbs unmodelled seasonal variance."}
mis_summ <- rbind(
  env_summary(mod_joint,       "joint",         resp = "count"),
  env_summary(mod_wrong_trend, "trend: AR(1)",  resp = "count")
)
mis_summ$model <- factor(mis_summ$model,
  levels = c("joint", "trend: AR(1)")
)
ggplot(mis_summ) +
  geom_ribbon(aes(env, ymin = lo, ymax = hi),
               fill = red_pal$light, alpha = 0.6) +
  geom_line(aes(env, med), colour = red_pal$dark_highlight,
             linewidth = 0.7) +
  geom_line(data = truth_df, aes(env, y = y_true),
             colour = "black", linewidth = 0.7) +
  facet_wrap(~ model, nrow = 1L) +
  labs(x = "env", y = "log-intensity shift")
```

Read as a habitat quality summary, the AR(1)-only panel would
report a weaker env dependence than the species actually shows.

### `loo()` on a state-space fit

`loo::loo()` reports the leave-one-out log density at every
observation. On plain regression it ranks models well. On
state-space fits it does not. Bürkner et al.
([2020](https://doi.org/10.1080/00949655.2020.1783262)) trace
the reason to the shared latent state between neighbouring time
points: the pointwise density is not really out of sample. On
the two candidates here:

```{r loo-compare, cache = TRUE}
loo::loo_compare(
  loo::loo(mod_joint),
  loo::loo(mod_wrong_trend)
)
```

The wrong-trend fit tops the table. Its `s(env)` smooth absorbs
the annual recurrence the lag-12 term should carry, and the
extra flexibility earns higher pointwise log density even
though the causal story is worse.

### `lfo_cv()` and choosing the forecast horizon

The honest comparator refits on the first $T$ time points and
scores at time $T+1$, rolling forward.
`lfo_cv(fc_horizon = h)` scores a full $h$-step forecast at
each rolling refit; the choice of $h$ matters. For a
misspecification whose signature is a seasonal recurrence, the
forecast has to reach far enough for that recurrence to enter
the score. At $h = 1$ the AR(1) branch of the misspecified fit can
approximate the correct model's next-step prediction, and LFO
cannot separate them. At $h = 12$, a full seasonal period, the
missing lag-12 term is exactly what the score measures.

```{r lfo-fit, eval = FALSE}
lfo_j_h1  <- lfo_cv(mod_joint,       min_t = 60L, fc_horizon = 1L)
lfo_wt_h1 <- lfo_cv(mod_wrong_trend, min_t = 60L, fc_horizon = 1L)

lfo_j_h6  <- lfo_cv(mod_joint,       min_t = 60L, fc_horizon = 6L)
lfo_wt_h6 <- lfo_cv(mod_wrong_trend, min_t = 60L, fc_horizon = 6L)

lfo_j_h12  <- lfo_cv(mod_joint,       min_t = 60L, fc_horizon = 12L)
lfo_wt_h12 <- lfo_cv(mod_wrong_trend, min_t = 60L, fc_horizon = 12L)
```

```{r lfo-load, include = FALSE}
lfo_j_h1   <- read_cache("lfo_joint_h1")
lfo_wt_h1  <- read_cache("lfo_wrong_trend_h1")
lfo_j_h6   <- read_cache("lfo_joint_h6")
lfo_wt_h6  <- read_cache("lfo_wrong_trend_h6")
lfo_j_h12  <- read_cache("lfo_joint_h12")
lfo_wt_h12 <- read_cache("lfo_wrong_trend_h12")
```

Rolling from time 60 onwards over 96 monthly observations, the
three comparisons look like this:

```{r lfo-compare-table}
lfo_row <- function(lj, lw, h) {
  cmp <- loo::loo_compare(lj, lw,
                            model_names = c("joint", "wrong_trend"))
  winner <- rownames(cmp)[1L]
  data.frame(
    horizon    = h,
    n_evals    = length(lj$elpds),
    joint_elpd = round(lj$sum_ELPD, 1),
    wrong_elpd = round(lw$sum_ELPD, 1),
    winner     = winner,
    elpd_diff  = round(cmp[2L, "elpd_diff"], 1),
    se_diff    = round(cmp[2L, "se_diff"],  1)
  )
}
lfo_tab <- rbind(
  lfo_row(lfo_j_h1,  lfo_wt_h1,  1L),
  lfo_row(lfo_j_h6,  lfo_wt_h6,  6L),
  lfo_row(lfo_j_h12, lfo_wt_h12, 12L)
)
lfo_tab
```

At $h = 1$ the two fits were indistinguishable. At $h = 6$ the
misspecified fit was slightly ahead, because a horizon covering
half a seasonal cycle let its AR(1) branch absorb the
recurrence into its own coefficient. At $h = 12$ the joint fit
won by roughly the same amount it had lost at $h = 6$. On this
single 96-month realisation the $h = 12$ margin is not
overwhelming (a $Z$-ratio near 1), but the sign and the
direction of change across horizons are the interpretable
signal: choose the LFO horizon to reach at least one full
period of the recurrence you are testing.

The per-step ELPD trajectory at $h = 12$ shows the same story
step by step:

```{r lfo-plot, fig.width = 7, fig.height = 3.4, fig.cap = "Per-step LFO log-densities at fc_horizon = 12 across a rolling window from month 61 to month 85. The joint fit tracks the wrong-trend fit at most steps but pulls ahead at those where the annual recurrence hits."}
lfo_long <- rbind(
  data.frame(t = lfo_j_h12$eval_timepoints,
              elpd = lfo_j_h12$elpds,
              model = "joint"),
  data.frame(t = lfo_wt_h12$eval_timepoints,
              elpd = lfo_wt_h12$elpds,
              model = "wrong_trend")
)
ggplot(lfo_long, aes(t, elpd, colour = model)) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 1.4) +
  scale_colour_manual(
    values = c(joint = red_pal$dark_highlight,
                wrong_trend = "black")
  ) +
  labs(x = "eval time", y = "LFO ELPD (fc_horizon = 12)")
```

### A practical checklist

Real analyses do not need the simulation truth to make this
choice. The order that carries the joint fit here is:

1. `pp_check(mod, resp = ...)` per response, with a display
   type suited to each family's support. Fix a bad marginal
   before doing anything else.
2. `plot(mod, type = "smooths")` on each single-measurement
   baseline plus the joint fit, side by side. A stream whose
   smooth wildly disagrees with the others is a candidate for
   its own response function on the observation side rather
   than for the shared trend.
3. `lfo_cv()` on the surviving candidate joint fits at a
   forecast horizon that reaches the timescale of the
   misspecification you are worried about. Reserve `loo()`
   for a sanity check; do not rank state-space fits on it.

## Take-aways

An integrated distribution model is one `mvgam()` call. Each
measurement contributes its own `bf()` outcome, its own family,
its own predictors on the observation side; a shared latent
process on the trend side ties them together. Every per-stream
intercept absorbs the constant its measurement carries so the
environmental response can be interpreted once rather than
three times.

In this example the joint fit put the env optimum near
env $= 0.4$ with a 95% interval of about $\pm 0.3$, matching
the simulation truth. The occupancy baseline placed the
optimum at the high end of the observed env range because a
monotone smooth was the best it could do with one detection
bit per time point. An ecologist reading only the occupancy
fit would report the wrong niche.

Comparisons carry the model choice when the truth is unknown.
Running `pp_check(mod, resp = ...)` on each response catches
gross mismatches in the observation family. Single-measurement
smooths overlaid on the joint smooth reveal streams whose
response is not shared. `lfo_cv()` at a horizon matching the
timescale of the misspecification you are worried about ranks
the surviving candidates. `loo()` on state-space fits will lie
to you (Bürkner et al.
[2020](https://doi.org/10.1080/00949655.2020.1783262)); do not
trust it to break a tie.

## Further reading

- The `mvbf` article (`vignette("mvbf",
  package = "mvgam")`) demonstrates the same mvbf architecture
  on generic count / biomass / camera measurements without the
  ecological framing.
- The `dfm` in-package vignette
  (`vignette("dfm", package = "mvgam")`) shows how
  `trend_map` shares fewer latent processes than observed
  measurements when there is more than one biological signal
  to disentangle.
- `?trend_constructors` documents the sparse-lag
  `AR(p = c(...))` and every alternative you might substitute
  for the trend side of a joint fit.
- Isaac et al.
  ([2020](https://doi.org/10.1016/j.tree.2019.08.006)) survey
  the wider integrated distribution model literature. Miller
  et al.
  ([2019](https://doi.org/10.1111/2041-210X.13110)) review the
  same field from the methodological side.

## Acknowledgements

The identifiability arguments behind the presence-only
intercept and the cloglog occupancy link follow Fithian et al.
([2015](https://doi.org/10.1111/2041-210X.12242)) and Renner
et al. ([2015](https://doi.org/10.1111/2041-210X.12352)). The
joint fit here specialises the combined presence-background and
site occupancy formulation developed by Koshkina et al.
([2017](https://doi.org/10.1111/2041-210X.12738)). The state
space framing follows Auger-Méthé et al.
([2021](https://doi.org/10.1002/ecm.1470)) and the dynamic
generalised additive model perspective in Clark & Wells
([2023](https://doi.org/10.1111/2041-210X.13974)). The cloglog
link on binary responses is treated in McCullagh & Nelder
([1989](https://doi.org/10.1201/9780203753736)). Formulas rely
on mgcv smooths (Wood
[2017](https://doi.org/10.1201/9781315370279)) and brms's
multilevel apparatus (Bürkner
[2017](https://doi.org/10.18637/jss.v080.i01)); sampling is
performed with Stan (Carpenter et al.
[2017](https://doi.org/10.18637/jss.v076.i01)). The LFO
comparator follows Bürkner et al.
([2020](https://doi.org/10.1080/00949655.2020.1783262)).

## References

Auger-Méthé M, Newman K, Cole D, Empacher F, Gryba R, King AA,
Leos-Barajas V, Mills Flemming J, Nielsen A, Petris G and Thomas
L (2021). A guide to state-space modeling of ecological time
series. *Ecological Monographs* 91(4), e01470.
[doi:10.1002/ecm.1470](https://doi.org/10.1002/ecm.1470)

Bürkner PC (2017). brms: An R package for Bayesian multilevel
models using Stan. *Journal of Statistical Software* 80(1),
1-28.
[doi:10.18637/jss.v080.i01](https://doi.org/10.18637/jss.v080.i01)

Bürkner PC, Gabry J and Vehtari A (2020). Approximate
leave-future-out cross-validation for Bayesian time series
models. *Journal of Statistical Computation and Simulation*
90(14), 2499-2523.
[doi:10.1080/00949655.2020.1783262](https://doi.org/10.1080/00949655.2020.1783262)

Carpenter B, Gelman A, Hoffman MD, Lee D, Goodrich B, Betancourt
M, Brubaker M, Guo J, Li P and Riddell A (2017). Stan: A
probabilistic programming language. *Journal of Statistical
Software* 76(1), 1-32.
[doi:10.18637/jss.v076.i01](https://doi.org/10.18637/jss.v076.i01)

Clark NJ and Wells K (2023). Dynamic Generalized Additive Models
(DGAMs) for forecasting discrete ecological time series.
*Methods in Ecology and Evolution* 14, 771-784.
[doi:10.1111/2041-210X.13974](https://doi.org/10.1111/2041-210X.13974)

Fithian W, Elith J, Hastie T and Keith DA (2015). Bias
correction in species distribution models: pooling survey and
collection data for multiple species. *Methods in Ecology and
Evolution* 6(4), 424-438.
[doi:10.1111/2041-210X.12242](https://doi.org/10.1111/2041-210X.12242)

Isaac NJB, Jarzyna MA, Keil P, Dambly LI, Boersch-Supan PH,
Browning E, Freeman SN, Golding N, Guillera-Arroita G, Henrys
PA, Jarvis S, Lahoz-Monfort J, Pagel J, Pescott OL, Schmucki R,
Simmonds EG and O'Hara RB (2020). Data integration for
large-scale models of species distributions. *Trends in Ecology
& Evolution* 35(1), 56-67.
[doi:10.1016/j.tree.2019.08.006](https://doi.org/10.1016/j.tree.2019.08.006)

Koshkina V, Wang Y, Gordon A, Dorazio RM, White M and Stone L
(2017). Integrated species distribution models: combining
presence-background data and site-occupancy data with
imperfect detection. *Methods in Ecology and Evolution* 8(4),
420-430.
[doi:10.1111/2041-210X.12738](https://doi.org/10.1111/2041-210X.12738)

McCullagh P and Nelder JA (1989). *Generalized Linear Models*
(2nd edition). Chapman and Hall/CRC.
[doi:10.1201/9780203753736](https://doi.org/10.1201/9780203753736)

Miller DAW, Pacifici K, Sanderlin JS and Reich BJ (2019). The
recent past and promising future for data integration methods
to estimate species' distributions. *Methods in Ecology and
Evolution* 10(1), 22-37.
[doi:10.1111/2041-210X.13110](https://doi.org/10.1111/2041-210X.13110)

Renner IW, Elith J, Baddeley A, Fithian W, Hastie T, Phillips
SJ, Popovic G and Warton DI (2015). Point process models for
presence-only analysis. *Methods in Ecology and Evolution*
6(4), 366-379.
[doi:10.1111/2041-210X.12352](https://doi.org/10.1111/2041-210X.12352)

Wood SN (2017). *Generalized Additive Models: An Introduction
with R* (2nd edition). Chapman and Hall/CRC.
[doi:10.1201/9781315370279](https://doi.org/10.1201/9781315370279)
