---
title: "Joint species distribution GAMs with informed loadings"
author: "Nicholas J Clark"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
    toc: yes
    toc_float: yes
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  dpi = 100,
  fig.asp = 0.7,
  fig.width = 6,
  out.width = "70%",
  fig.align = "center"
)
library(mvgam)
library(mvabund)
library(ggplot2)
library(patchwork)
theme_set(mvgam_theme(base_size = 12, base_family = "serif"))

# Cached jsdgam fits live under pkgdown/, which is excluded from
# the CRAN tarball by .Rbuildignore. pkgdown runs articles from
# the project root, so the relative path resolves at build time.
# The visible fit chunks remain as documentation of what produced
# each object.
cache_dir <- file.path("pkgdown", "jsdgam_cache")
read_cache <- function(name) {
  readRDS(file.path(cache_dir, paste0(name, ".rds")))
}
```

A community of species rarely behaves as a collection of
independent units. After accounting for measured environmental
gradients, the residual co-occurrence between species still
carries signal. Some of it is unmeasured climate or sampling
artefact; some of it is shared phylogeny or biotic
interaction. Joint Species Distribution Models (JSDMs)
compress that residual covariance into a low-rank decomposition
`Sigma = Z Zᵀ`, where `Z` is an `n_species x n_lv` loading
matrix mapping each species to its position in a small latent
space (Warton et al.
[2015](https://doi.org/10.1016/j.tree.2015.09.007); Ovaskainen
et al. [2017](https://doi.org/10.1111/ele.12757)). The latent
positions double as an ordination, and the implied `Sigma`
is the residual species correlation matrix.

mvgam's `jsdgam()` function fits this class of model using the
Heaps & Jermyn ([2024](https://doi.org/10.1007/s11222-024-10454-0))
identification scheme, which applies a QR decomposition to the
sampled loadings inside generated quantities. The default prior
treats the columns of `Z` as a priori exchangeable, but two
extensions let domain knowledge enter directly. Species traits
or a phylogeny can shape the prior on `Z` through an automatic
relevance determination (ARD) kernel; functionally or
phylogenetically similar species are pulled toward similar
loadings before the data speaks. As an alternative to picking
`n_lv` directly, a column shrinkage prior built from a
multiplicative gamma process lets the data choose how many
factors are active. We walk through both on the Australian ant
trait dataset (Warton et al.
[2015](https://doi.org/10.1016/j.tree.2015.09.007)).

## The data

`antTraits` ships with the `mvabund` package. It carries 30
sites x 41 ant species capture counts together with five
environmental covariates (`Bare.ground`, `Canopy.cover`,
`Shrub.cover`, `Volume.lying.CWD`, `Feral.mammal.dung`) and a
five-column species trait table. The traits cover morphology
(`Femur.length` and `Webers.length` for body size), defence
(`No.spines`), pilosity and worker-caste polymorphism.

```{r data-head}
data(antTraits, package = "mvabund")
str(antTraits, max.level = 1)
head(antTraits$abund[, 1:5])
head(antTraits$env)
head(antTraits$traits, 3)
```

Two of the trait columns (`Pilosity` and `Polymorphism`) ship
as ordered factors with numeric levels. We coerce them to
numeric so the ARD kernel reads them as ordered continuous
traits rather than dummy-encoded categories.

```{r traits-coerce}
abund  <- antTraits$abund
env    <- antTraits$env
traits <- antTraits$traits
traits$Pilosity     <- as.numeric(as.character(traits$Pilosity))
traits$Polymorphism <- as.numeric(as.character(traits$Polymorphism))
```

`jsdgam()` expects long-form data with one row per site x
species cell. We reshape, repeat the environmental covariates
across species rows, and standardise the covariates so the
linear coefficients sit on a comparable scale.

```{r reshape}
n_sites   <- nrow(abund)
n_species <- ncol(abund)
species_names <- colnames(abund)

ant_long <- data.frame(
  site    = factor(rep(seq_len(n_sites), times = n_species)),
  species = factor(rep(species_names, each = n_sites),
                   levels = species_names),
  time    = rep(seq_len(n_sites), times = n_species),
  counts  = as.integer(unlist(abund))
)
ant_long <- cbind(ant_long, env[as.integer(ant_long$site), ])
for (col in colnames(env)) {
  ant_long[[col]] <- as.numeric(scale(ant_long[[col]]))
}
head(ant_long)
```

The capture counts span 0 to 20. Plotting variance against the
mean for each species shows the variance climbing well above
the 1:1 Poisson line, so a Poisson observation model would
systematically underestimate predictive uncertainty.

```{r mean-variance, fig.asp = 0.7, out.width = "65%"}
red <- bayesplot::color_scheme_get("red")$dark
per_sp <- do.call(rbind, tapply(
  ant_long$counts, ant_long$species,
  function(x) c(mean = mean(x), var = var(x))
))
per_sp <- as.data.frame(per_sp)
ggplot(per_sp, aes(x = mean, y = var)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed",
              colour = "grey50") +
  geom_point(size = 2.2, alpha = 0.85, colour = red) +
  scale_x_continuous(trans = "log1p") +
  scale_y_continuous(trans = "log1p") +
  labs(x = "Mean count per species (log1p)",
       y = "Variance of count per species (log1p)")
```

A negative binomial observation model accommodates that
dispersion through its species-level `shape` parameter.

## A baseline JSDM: uninformed loadings

A useful first move is to fit the model with no trait
information at all. The loadings have to find their structure
from the data alone, which gives a clean reference point for
asking what an informed prior adds afterwards. We keep the
default loading prior: each entry of `Z` is independent under
a weak `student_t(3, 0, 0.5)`. The latent space is
two-dimensional (`n_lv = 2L`) and `factor_formula = ~ -1`
asks for no smooths on the factors themselves. Counts get
the `negbinomial()` family to absorb dispersion beyond a
Poisson mean.

```{r mod_baseline_show, eval = FALSE}
mod_baseline <- jsdgam(
  formula = counts ~ Bare.ground + Canopy.cover + Shrub.cover +
    Volume.lying.CWD + Feral.mammal.dung,
  factor_formula = ~ -1,
  data    = ant_long,
  family  = negbinomial(),
  n_lv    = 2L,
  unit    = time,
  species = species,
  chains  = 2L, iter = 1600L, warmup = 1000L, silent = 2L
)
```

```{r mod_baseline_load, include = FALSE}
mod_baseline <- read_cache("mod_baseline")
```

The summary prints the observation-model intercept, the
dispersion parameter `shape` and the identified species
loadings `Z_tilde[i, k]`. Per-coefficient detail for the five
environmental terms lives in `posterior_summary()`.

```{r mod_baseline_summary}
summary(mod_baseline)
```

```{r mod_baseline_env_coefs}
posterior_summary(mod_baseline)[paste0("b_", c(
  "Bare.ground", "Canopy.cover", "Shrub.cover",
  "Volume.lying.CWD", "Feral.mammal.dung"
)), c("Estimate", "Q2.5", "Q97.5")]
```

All five environmental coefficients have credible intervals
straddling zero on the log scale of the linear predictor.
The mean component is doing very little work in this baseline
fit; the bulk of the between-site structure lives in the
residual factor space, which the biplot below visualises.

Calling `ordinate()` returns a biplot. Site scores and species
loadings are taken at the posterior median and the axes are
rotated to a singular-value basis, so the first axis carries
the dominant ordination gradient.

```{r baseline-biplot, fig.asp = 0.85, out.width = "75%"}
ordinate(mod_baseline) +
  ggtitle("Baseline: uninformed loadings")
```

Two species sit at the extremes of LV2 (`Aphaenogaster.longiceps`
high, `Cardiocondyla.nuda.atalanta` low) with the rest of the
community clustered near the origin. Nothing in the layout
references trait similarity yet, because the prior had no
reason to encode any. Before reading further into the
ordination it pays to confirm the observation layer fits the
counts. Calling `pp_check(mod_baseline)` dispatches to
`pp_check.mvgam()` for that diagnostic: a misfit on the zero
counts or on the upper tail means the family is wrong, and
fixing that takes priority over any ecological reading of the
biplot.

## Heaps-architecture trait-informed loadings

The Heaps & Jermyn
([2024](https://doi.org/10.1007/s11222-024-10454-0))
architecture replaces the iid loadings prior with a structured
one. Each column of `Z` is drawn from
`multi_normal_cholesky(0, L_Φ · √Ψ_diag[k])`, where the
covariance `Φ` is built by an ARD kernel on a feature matrix
supplied by the user. The kernel estimates one inverse scale
per feature column. Features that predict co-occurrence pull
the relevant species toward similar loadings; features that do
not predict it shrink to large scales and contribute little.
For species traits the result is a prior that says "ants whose
body size and defensive traits are similar start out near each
other in latent space, before seeing the counts."

Passing the trait table through the `traits` argument is the
only change to the call.

```{r mod_traits_show, eval = FALSE}
mod_traits <- jsdgam(
  formula = counts ~ Bare.ground + Canopy.cover + Shrub.cover +
    Volume.lying.CWD + Feral.mammal.dung,
  factor_formula = ~ -1,
  data    = ant_long,
  family  = negbinomial(),
  n_lv    = 2L,
  traits  = traits,
  unit    = time,
  species = species,
  chains  = 2L, iter = 1600L, warmup = 1000L,
  adapt_delta = 0.95, silent = 2L
)
```

```{r mod_traits_load, include = FALSE}
mod_traits <- read_cache("mod_traits")
```

We bump `adapt_delta` from the Stan default of 0.8 to 0.95.
The ARD geometry on 41 species and 5 trait columns has a
funnel where the kernel scales hand off prior mass to the
loadings; the tighter leapfrog step keeps the sampler from
skipping over it.

The summary now carries an extra `loadings_prior` block. It
lists `theta_features[k]`, the ARD kernel length scale for each
trait, alongside `Psi_diag[k]`, the prior scale on the kth
loading column.

```{r mod_traits_summary}
summary(mod_traits)
```

Reading `theta_features[k]` follows the standard ARD
convention. The kernel takes the form
`exp(-||(x_i - x_j) / theta||^2)`. A posterior
`theta_features[k]` that is small relative to the spread of
trait `k` says the kernel decays sharply along that
dimension, which pulls trait-different species apart in the
latent space. A posterior length scale much larger than the
trait spread says the kernel barely changes with that trait,
so the data has effectively dropped it from the loadings
prior. The Heaps & Jermyn
([2024](https://doi.org/10.1007/s11222-024-10454-0)) prior
expresses the ARD covariance across species through the
matrix `Phi` and ties it to the observable shared variation
`Delta = Z Zᵀ` via the identity `E(Delta) = tr(Psi^2) * Phi`.
The function `shared_variation()` returns posterior summaries
of `Delta` directly when those magnitudes matter.

The biplot, this time with `traits = "auto"` to overlay each
trait's regression onto the rotated loadings, makes the
restructuring visible.

```{r traits-biplot, fig.asp = 0.85, out.width = "75%"}
ordinate(mod_traits, rotation = "varimax", traits = "auto") +
  ggtitle("Trait-informed loadings + trait arrows")
```

Four of the trait arrows point into the upper portion of the
biplot (`Pilosity`, `Webers.length`, `Polymorphism` and
`Femur.length`), where the larger-bodied `Camponotus` and
`Iridomyrmex purpureus` cluster sits. `No.spines` is the only
arrow pointing downward, into the region occupied by
`Cardiocondyla.nuda.atalanta`, `Solenopsis.sp..A`,
`Meranoplus.sp..A` and other smaller species. The latent
space has acquired an ecological gradient that the baseline
fit did not have: body size and defensive armature pulling in
opposite directions along LV2.

`compare_loadings()` makes the effect on individual species
explicit. The two fits sit in independent QR bases. The
function applies an orthogonal Procrustes rotation that brings
the second fit into the basis of the first, then draws a
segment from each species' baseline position to its position
under the trait prior.

```{r compare-loadings, fig.asp = 0.8, out.width = "85%"}
compare_loadings(
  mod_baseline, mod_traits,
  labels = c("uninformed", "trait-informed"),
  label_species = "shift"
)
```

Species labelled in the plot are the ten with the largest
displacement after alignment. The trait-informed cluster is
visibly tighter than the uninformed one. The kernel has pulled
species with shared body and spine traits closer together, at
the price of mild shrinkage on the two outliers
(`Aphaenogaster.longiceps`, `Cardiocondyla.nuda.atalanta`)
that drove the uninformed biplot.

## Identification: rotations of the latent basis

The factor model is identified only up to rotation of the
latent space. Any orthogonal rotation of the columns of `Z`,
paired with the inverse rotation of the latent factor scores,
produces the same likelihood. The Stan model samples
unconstrained `Z` and applies a thin QR decomposition in
generated quantities. The result is a `Z_tilde` that is
lower-triangular with positive diagonal entries (Heaps & Jermyn
[2024](https://doi.org/10.1007/s11222-024-10454-0)). That is
what `ordinate()` reads by default. Different rotations
applied after fitting re-express the same posterior on axes
that are easier to interpret for different questions.

```{r rotations, fig.asp = 0.5, out.width = "100%"}
p_svd <- ordinate(mod_traits, rotation = "svd") +
  ggtitle("svd (default)")
p_vmx <- ordinate(mod_traits, rotation = "varimax") +
  ggtitle("varimax")
p_pro <- ordinate(mod_traits, rotation = "promax") +
  ggtitle("promax")
p_non <- ordinate(mod_traits, rotation = "none") +
  ggtitle("none (raw QR)")
(p_svd | p_vmx) / (p_pro | p_non)
```

The `svd` rotation reorders axes by singular-value variance
and matches the BORAL convention (Hui
[2016](https://doi.org/10.1111/2041-210X.12514)). The
`varimax` and `promax` rotations target loading sparsity. Each
makes per-axis interpretation cleaner when individual species
tend to load on one factor more than the others. The `none`
option plots the raw identified loadings. All four panels show
the same fit. The choice between them is about which axis
interpretation suits the question at hand. See
`?ordinate.jsdgam` for the full set of plotting arguments
(including `alpha` for the BORAL site / loading variance
split, and `label_sites` for switching between text and point
markers). `plot(fit, type = "factors")` draws the identified
latent processes themselves against the unit index, annotated
with the share of series-level variance each one explains.

## Residual species correlations

The factor decomposition implies a residual species covariance
`Sigma = Z Zᵀ` after accounting for the measured environment.
`residual_cor()` returns posterior summaries of this matrix.
The heatmap carries the median correlation as fill and the
weight of evidence as opacity, so a large correlation the data
barely resolve is drawn faintly rather than erased. The
`cluster = TRUE` argument reorders rows and columns by an
approximate Robinson ordering so block structure shows up on
the diagonal. The companion accessor `shared_variation()`
returns the full posterior over the covariance matrix itself,
useful when entry magnitudes matter for downstream analysis;
both surface entry-level posterior probabilities via
`summary()` on the returned object.

```{r resid-cor, fig.asp = 0.95, out.width = "85%"}
rcor <- residual_cor(mod_traits, summary = TRUE)
plot(rcor, cluster = TRUE)
```

```{r rufoniger, include = FALSE}
foc <- "Iridomyrmex.rufoniger"
foc_r <- rcor$cor[foc, ]
foc_r <- foc_r[names(foc_r) != foc]
foc_top <- sort(foc_r)[1:5]
foc_lo <- rcor$cor_lower[foc, names(foc_top)]
foc_hi <- rcor$cor_upper[foc, names(foc_top)]
foc_names <- paste0("`", names(foc_top), "`", collapse = ", ")
foc_worst <- sprintf("%.2f", min(foc_top))
foc_excl <- sum(sign(foc_lo) == sign(foc_hi))
```

Positive blocks line up on the diagonal: groups of species that
tend to co-occur once the environment is accounted for. A band
of negative correlation cuts across the heatmap on
`Iridomyrmex.rufoniger`. Its five strongest negative partners
are `r foc_names`, reaching `r foc_worst`, and `r foc_excl` of
the five have an interval clear of zero. Its loadings sit in a
region of the latent space that the rest of the community
avoids, which is consistent with competitive dominance or with
an unmeasured environmental contrast that we have not
conditioned on.

Residual correlation is *evidence consistent with* shared
unmeasured drivers, not proof of any specific mechanism. A
positive entry can reflect either a shared unmeasured niche
or a simultaneous detection bias on the two species; a
negative entry can reflect competition or an environmental
gradient the model did not condition on. The factor model
gives us a coherent posterior over the matrix but does not
adjudicate between these explanations on its own.

## How many factors? Data-driven dimensionality with MGP

The fits so far pinned `n_lv = 2L`. A more principled choice is
to set a ceiling and let the posterior decide. The
multiplicative gamma process (MGP) prior of Bhattacharya &
Dunson ([2011](https://doi.org/10.1093/biomet/asr013)) puts
increasing shrinkage on later columns of `Z`. Only the leading
few columns carry mass under the prior, and the data shifts
that envelope as needed. Legramanti, Durante & Dunson
([2020](https://doi.org/10.1093/biomet/asaa008)) propose a
cumulative shrinkage criterion for declaring a factor active.

Passing the MGP spec through `loadings_prior` composes cleanly
with the trait kernel. Traits supply the ARD covariance across
species within each column; MGP supplies the shrinkage across
columns.

```{r mod_mgp_show, eval = FALSE}
mod_mgp <- jsdgam(
  formula = counts ~ Bare.ground + Canopy.cover + Shrub.cover +
    Volume.lying.CWD + Feral.mammal.dung,
  factor_formula = ~ -1,
  data    = ant_long,
  family  = negbinomial(),
  n_lv    = 10L,
  loadings_prior = list(column_shrinkage = "mgp", features = traits),
  unit    = time,
  species = species,
  chains  = 2L, iter = 1600L, warmup = 1000L, silent = 2L
)
```

```{r mod_mgp_load, include = FALSE}
mod_mgp <- read_cache("mod_mgp")
```

`active_factors()` summarises the posterior distribution of how
many factors clear the activity threshold. The default
threshold marks factor `k` as active when the posterior
probability that its squared column-norm exceeds 1% of the
leading column's norm is at least 0.5.

```{r active-factors}
af <- active_factors(mod_mgp)
af
```

Plotting the same object gives the per-factor probability bars
with the activity cutoff shown as a dashed line, plus the
posterior median and 95% CI of the active count in the
subtitle.

```{r active-factors-plot, fig.asp = 0.55, out.width = "75%"}
plot(af)
```

Factors 1-3 sit at probability 1, factors 4-6 cross the
cutoff with diminishing posterior support, factor 7 sits right
at the cutoff, and factors 8-10 are essentially shrunk to
zero. The posterior median active count is 6, with a 95%
interval of 4 to 9. Where the right `n_lv` is not known a
priori, this is the specification to reach for first: set a
generous ceiling, let the shrinkage prior decide and read the
answer off `active_factors()` instead of refitting at a
sequence of candidate dimensions. The `fraction` and
`prob_threshold` arguments move the activity cutoff when the
defaults, which follow Legramanti et al., prove too permissive
for the question at hand.

## Take-aways

A negative binomial JSDM with Heaps-identified loadings gives
a single coherent posterior over the environmental responses,
the latent factor scores, the species loadings and the implied
residual species correlation matrix. The observation family
chooses how counts disperse: `family = negbinomial()` was the
right pick here. The loadings prior carries the domain
knowledge: passing `traits =` (or `phylo =`) routes that
knowledge through an ARD kernel, so functionally similar
species cluster a priori before the data speaks. And the
factor count is rarely known up front, so reach for
`loadings_prior = list(column_shrinkage = "mgp", ...)` to let
the posterior shrink later columns of `Z` and surface the
active count via `active_factors()`.

## Further reading

Three extensions sit outside the scope of this article and
deserve their own write-ups. The `phylo =` argument accepts an
`ape::phylo` object or a pairwise distance matrix and builds
the ARD kernel from phylogenetic distance rather than from
trait similarity; that route is useful when the available
signal is evolutionary rather than morphological. The
`trait_slopes =` argument implements the trait-mediated
regression of Pollock et al.
([2012](https://doi.org/10.1111/j.1600-0587.2011.07085.x)) and
the Hmsc framework of Tikhonov et al.
([2020](https://doi.org/10.1111/2041-210X.13345)): each
species' environmental response is itself regressed on traits.
That complements the loadings story by modelling how trait
similarity shapes responses to measured environment, not just
residual co-occurrence. Finally, the `occ()` family
marginalises imperfect detection at the closure-unit level for
joint occupancy fits; `?jsdgam` shows a worked example.

The rest of the workflow is the standard mvgam machinery
applied to a jsdgam fit. To confirm the count layer absorbs
the dispersion, `pp_check.mvgam()` runs the usual posterior
predictive battery. To compare loadings priors on the same
data, reach for `loo.mvgam()` as the cheap first pass and
`kfold.mvgam()` when LOO Pareto-k goes red. To interrogate
how an environmental gradient drives abundance once the
factors have absorbed the residual structure between species,
`conditional_effects()` plots the marginal means on the
response scale. And when writing the methods section,
`how_to_cite(mod_traits)` returns the citation paragraph that
names the architecture pieces the fit actually used. The
`?jsdgam` page covers the remaining options, including
`lv_axis()` for per-factor smooths inside `factor_formula`.

Sibling articles in this series:

- The
  [dynamic factor vignette](https://nicholasjclark.github.io/mvgam/articles/dfm.html)
  is the temporal counterpart of the same factor architecture,
  where `trend_map` fixes which series share a latent path
  instead of leaving the loadings to a prior.
- The
  [N-mixture article](https://nicholasjclark.github.io/mvgam/articles/nmix.html)
  covers the closure-unit families that the `occ()` route above
  belongs to, including the goodness-of-fit checks they need.
- The
  [overview vignette](https://nicholasjclark.github.io/mvgam/articles/mvgam_overview.html)
  places the factor model inside the wider state-space surface
  the two fitters share.

## Acknowledgements

The `antTraits` data are released with the `mvabund` package
(Wang et al. [2012](https://doi.org/10.1111/j.2041-210X.2012.00190.x)).
The factor model architecture follows Heaps & Jermyn
([2024](https://doi.org/10.1007/s11222-024-10454-0)). The MGP
prior is Bhattacharya & Dunson
([2011](https://doi.org/10.1093/biomet/asr013)) with the
active-count criterion of Legramanti, Durante & Dunson
([2020](https://doi.org/10.1093/biomet/asaa008)).

## References

Bhattacharya A and Dunson DB (2011). Sparse Bayesian infinite
factor models. *Biometrika* 98(2), 291-306.
[doi:10.1093/biomet/asr013](https://doi.org/10.1093/biomet/asr013)

Heaps SE and Jermyn IH (2024). Structured prior distributions
for the covariance matrix in latent factor models. *Statistics
and Computing* 34, 143.
[doi:10.1007/s11222-024-10454-0](https://doi.org/10.1007/s11222-024-10454-0)

Hui FKC (2016). boral - Bayesian Ordination and Regression
Analysis of Multivariate Abundance Data in R. *Methods in
Ecology and Evolution* 7(6), 744-750.
[doi:10.1111/2041-210X.12514](https://doi.org/10.1111/2041-210X.12514)

Legramanti S, Durante D and Dunson DB (2020). Bayesian
cumulative shrinkage for infinite factorizations.
*Biometrika* 107(3), 745-752.
[doi:10.1093/biomet/asaa008](https://doi.org/10.1093/biomet/asaa008)

Ovaskainen O, Tikhonov G, Norberg A, Blanchet FG, Duan L,
Dunson D, Roslin T and Abrego N (2017). How to make more out
of community data? A conceptual framework and its
implementation as models and software. *Ecology Letters*
20(5), 561-576.
[doi:10.1111/ele.12757](https://doi.org/10.1111/ele.12757)

Pollock LJ, Morris WK and Vesk PA (2012). The role of
functional traits in species distributions revealed through a
hierarchical model. *Ecography* 35(8), 716-725.
[doi:10.1111/j.1600-0587.2011.07085.x](https://doi.org/10.1111/j.1600-0587.2011.07085.x)

Tikhonov G, Opedal OH, Abrego N, Lehikoinen A, de Jonge MMJ,
Oksanen J and Ovaskainen O (2020). Joint species distribution
modelling with the R-package Hmsc. *Methods in Ecology and
Evolution* 11(3), 442-447.
[doi:10.1111/2041-210X.13345](https://doi.org/10.1111/2041-210X.13345)

Wang Y, Naumann U, Wright ST and Warton DI (2012). mvabund -
an R package for model-based analysis of multivariate
abundance data. *Methods in Ecology and Evolution* 3(3),
471-474.
[doi:10.1111/j.2041-210X.2012.00190.x](https://doi.org/10.1111/j.2041-210X.2012.00190.x)

Warton DI, Blanchet FG, O'Hara RB, Ovaskainen O, Taskinen S,
Walker SC and Hui FKC (2015). So many variables: joint
modeling in community ecology. *Trends in Ecology and
Evolution* 30(12), 766-779.
[doi:10.1016/j.tree.2015.09.007](https://doi.org/10.1016/j.tree.2015.09.007)
