---
title: "One shared process, many sensors: multi-response mvgam"
author: "Nicholas J Clark"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
    toc: yes
    toc_float: yes
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE, dpi = 100, fig.asp = 0.62, fig.width = 7,
  out.width = "85%", fig.align = "center"
)
library(mvgam)
library(brms)
library(ggplot2)
theme_set(theme_bw(base_size = 12, base_family = "serif"))

cache_dir <- file.path("pkgdown", "multivariate_cache")
read_cache <- function(name) {
  readRDS(file.path(cache_dir, paste0(name, ".rds")))
}
```

Field monitoring rarely produces a single response variable. An
ecologist counting animals along a transect typically also
operates a baited trap and runs a camera array. Each survey
delivers a noisy, partial view of the same underlying ecological
quantity. Standard methods for time series data either fit each
response separately (throwing away the information sitting in
the other methods' columns) or force every column into the same
Gaussian observation model, which is
wrong for counts, for biomass and for detection records. This
article walks through the mvgam alternative: **one shared latent
dynamic process linked to many response variables, each with its
own observation family and per-arm covariates**. The latent state
is informed jointly by every response through the likelihood.

That is the package's value proposition for multi-response work.
No alternative R package combines cross-series latent dynamics,
mixed observation families, smooths/GPs/REs per response, full
Bayesian posterior, and proper joint forecasts. We demonstrate
this on a simulated three-method monitoring scenario where we
know the truth and can verify recovery directly.

## The data

We simulate a single hypothetical animal population observed over
80 monthly time points by three parallel survey methods. The
underlying latent state `x[t]` follows a stationary AR(1) with
`phi = 0.7` and `sigma = 0.4`. Each method observes that state
through a different measurement process:

- `count`: visual transect counts (Poisson, log link). Survey
  effort is logged each month and enters the linear predictor as
  an offset.
- `biomass`: trap-collected biomass per unit night (Gamma, log
  link). A cyclic seasonal smooth captures the natural seasonal
  rhythm in biomass independent of survey effort.
- `camera`: camera-trap presence/absence (Bernoulli, logit link).
  The number of deployment days enters as a linear slope.

The camera array is operated on a sparser schedule than the other
two: only 60 of the 80 months carry camera observations. mvgam
keeps the camera arm on its own non-NA rows so the shared latent
state is still informed during the 20 months without a camera by
counts and biomass alone.

```{r data-load}
mv_data <- read_cache("mv_data")
truth   <- read_cache("truth")
str(mv_data, give.attr = FALSE)
```

```{r truth-summary}
cat("True phi      =", truth$phi,   "\n")
cat("True sigma    =", truth$sigma, "\n")
cat("True nu       =", truth$nu,    "(Gamma shape)\n")
cat("camera obs    =", sum(!is.na(mv_data$camera)), "/", nrow(mv_data), "\n")
```

A scatter of the three response columns by time shows what each
method records and where the camera arm is missing.

```{r plot-data, fig.asp = 0.7, fig.width = 8, out.width = "95%"}
long <- with(mv_data, rbind(
  data.frame(time = time, value = count,   resp = "count"),
  data.frame(time = time, value = biomass, resp = "biomass"),
  data.frame(time = time, value = camera,  resp = "camera")
))
ggplot(long, aes(x = time, y = value)) +
  geom_line(colour = "gray30", alpha = 0.6) +
  geom_point(size = 1.2, na.rm = TRUE) +
  facet_wrap(~ resp, ncol = 1L, scales = "free_y") +
  labs(x = "Time (month)", y = "Observed value")
```

## Three independent univariate baselines

To see how much the shared-process model gains over the
default that fits each method alone, we first run three
independent fits with the same AR(1) latent dynamics.

```{r uni-fits, eval = FALSE}
mod_uni_count <- mvgam(
  count ~ 1 + offset(log_effort),
  trend_formula = ~ AR(p = 1),
  family = poisson(), data = mv_data,
  chains = 2L, samples = 1000L, silent = 2L
)
mod_uni_biomass <- mvgam(
  biomass ~ 1 + s(month, bs = "cc", k = 8),
  trend_formula = ~ AR(p = 1),
  family = Gamma(link = "log"), data = mv_data,
  knots = list(month = c(0.5, 12.5)),
  chains = 2L, samples = 1000L, silent = 2L
)
mod_uni_camera <- mvgam(
  camera ~ 1 + deploy_days,
  trend_formula = ~ AR(p = 1),
  family = bernoulli(link = "logit"), data = mv_data,
  chains = 2L, samples = 1000L, silent = 2L
)
```

```{r uni-load, include = FALSE}
mod_uni_count   <- read_cache("mod_uni_count")
mod_uni_biomass <- read_cache("mod_uni_biomass")
mod_uni_camera  <- read_cache("mod_uni_camera")
```

Each fit converges (max R-hat below 1.05) but the camera arm has
very little information by itself. The Bernoulli observations
above provide only one bit per time point, so its AR(1) estimate
of `x[t]` is correspondingly diffuse.

## The joint multi-response fit

The joint fit lives in one `mvgam()` call. We chain three `bf()`
arms with the `+` operator (a brms idiom that mvgam re-exports).
Each arm gets its own family and its own predictors. Every arm
shares the same `trend_formula = ~ AR(p = 1)`: one latent state
feeding three observation models.

```{r joint-fit, eval = FALSE}
joint_formula <-
  bf(count   ~ 1 + offset(log_effort), family = poisson()) +
  bf(biomass ~ 1 + s(month, bs = "cc", k = 8),
     family = Gamma(link = "log")) +
  bf(camera  ~ 1 + deploy_days,
     family = bernoulli(link = "logit"))

mod_joint <- mvgam(
  joint_formula,
  trend_formula = ~ AR(p = 1),
  data  = mv_data,
  knots = list(month = c(0.5, 12.5)),
  chains = 2L, samples = 1000L, silent = 2L
)
```

```{r joint-load, include = FALSE}
mod_joint <- read_cache("mod_joint")
```

mvgam handles the bookkeeping. brms's default `mvbf` would
drop those 20 rows from every arm via listwise deletion. mvgam
instead expands the standata so each arm keeps its own non-NA
rows. The Stan model then evaluates the count likelihood at all
80 time points, the biomass likelihood at all 80 and the camera
likelihood at the 60 observed months, all contributing to a
single shared AR(1) state.

The shared dynamics show up clearly in the fixed-effect summary
and parameter recovery.

```{r joint-summary}
s <- summary(mod_joint, smooth_test = FALSE)
print(s$spec)
```

```{r joint-pars}
ps <- posterior_summary(mod_joint)
keep <- grep("Intercept|deploy_days|shape|ar1|sigma_trend",
              rownames(ps))
round(ps[keep, c("Estimate", "Q2.5", "Q97.5")], 3)
```

Recovery against the truth (`phi = 0.7`, `sigma = 0.4`,
`a_count = 1.5`, `a_biomass = 0.5`, `a_camera = -0.5`,
`b_deploy = 0.05`, `nu = 2.0`) is clean: every intercept, the AR
coefficient, the innovation SD and the Gamma shape sit close to
their generating values with 95% intervals that cover.

## Posterior-predictive checks per arm

Each arm gets its own posterior-predictive check via the standard
`pp_check()` interface, scoped with the `resp` argument. For the
Bernoulli camera arm the `"bars"` type is more informative than
the default density overlay (which is built for continuous
responses).

```{r pp-count, fig.asp = 0.55}
pp_check(mod_joint, resp = "count", ndraws = 50L) +
  labs(title = "PPC: visual transect counts")
```

```{r pp-biomass, fig.asp = 0.55}
pp_check(mod_joint, resp = "biomass", ndraws = 50L) +
  labs(title = "PPC: trap biomass")
```

```{r pp-camera, fig.asp = 0.55, fig.width = 5}
pp_check(mod_joint, resp = "camera", type = "bars", ndraws = 50L) +
  labs(title = "PPC: camera detections")
```

The replicate distributions cover the observed marginals for
every arm. Calibration is sound on all three.

## Per-arm covariate effects

`conditional_effects()` returns one list per response, faceted
internally by predictor. The plot for each arm now carries the
correct response on its y-axis.

```{r ce-biomass, fig.asp = 0.55}
ce <- mvgam::conditional_effects(mod_joint)
plot(ce[["biomass"]])[[1L]]
```

The biomass arm picks up the seasonal sinusoid that drove the
simulation: a clear peak in early months and a trough in late
summer. The cyclic smooth penalisation is doing its job, and the
posterior median follows the truth even though we never told the
model that month was seasonal.

```{r ce-camera, fig.asp = 0.55}
plot(ce[["camera"]])[[2L]]
```

The camera arm's detection probability rises with deploy days,
consistent with the simulated effect of 0.05 log-odds per day.
The CI is wide because Bernoulli observations carry little
information per row.

## Per-arm residual diagnostics

Quantile residuals for each arm are also scoped via `resp`. The
dispatch correctly looks up the per-arm family (Poisson for
counts, Gamma for biomass, Bernoulli for camera) and returns the
matching analytic or empirical PIT residuals.

```{r resid-summary}
res_count   <- residuals(mod_joint, resp = "count")
res_biomass <- residuals(mod_joint, resp = "biomass")
res_camera  <- residuals(mod_joint, resp = "camera")
sapply(list(count = res_count, biomass = res_biomass,
             camera = res_camera), dim)
```

## State recovery: joint vs each baseline

The point of the joint fit is the latent state itself.
`compare_latent_states()` extracts the posterior `trend[t]` from a
named list of fits, aligns each fit's sign to the supplied truth
vector when one is given, and returns either an overlay of the
posterior trajectories or a comparison of the per-time posterior
SD across fits.

```{r compare-trajectory, fig.asp = 0.62, fig.width = 9, out.width = "95%"}
fits <- list(
  "Joint (3 families)" = mod_joint,
  "Counts only"        = mod_uni_count,
  "Biomass only"       = mod_uni_biomass,
  "Camera only"        = mod_uni_camera
)
compare_latent_states(fits, truth = truth$x, type = "trajectory")
```

```{r compare-sd, fig.asp = 0.55, fig.width = 9, out.width = "95%"}
compare_latent_states(fits, truth = truth$x, type = "sd")
```

The joint fit's posterior median tracks the truth more closely
than any of the univariate baselines, and its per-time posterior
SD sits below the other three at every time point. The camera-only
line is the diffuse outlier: a single Bernoulli observation per
time point carries little information. The joint fit recovers
strength wherever any of the methods reported.

## Cross-arm posterior forecasts

The same `forecast()` interface that handles univariate fits now
works on a multi-response fit: it fans out per response and
returns one `mvgam_forecast` per arm. We extend the time grid by
5 months and let mvgam propagate the shared latent state forward,
then map back through each arm's observation family.

```{r forecast-multi, eval = TRUE}
T_ <- max(mv_data$time)
new <- data.frame(
  time = (T_ + 1L):(T_ + 5L),
  series = factor(rep("pop1", 5L), levels = levels(mv_data$series)),
  month = ((T_ + 0L:4L) %% 12L) + 1L,
  log_effort = log(rep(1, 5L)),
  deploy_days = rep(15, 5L),
  count = NA_integer_, biomass = NA_real_, camera = NA_integer_
)
fc <- forecast(mod_joint, newdata = new)
sapply(fc, function(f) length(f$forecasts))
```

One forecast object per arm, ready to feed into `score()` (see
the [forecast-evaluation article](https://nicholasjclark.github.io/mvgam/articles/forecast_evaluation.html))
or into per-arm plotting.

## Take-aways

A multi-response mvgam fit lets you keep each measurement model
in its proper family while sharing the dynamics that drive every
observation. The recipe is small:

- write one `bf()` arm per response with its own family and per-arm
  predictors,
- combine them with `+`,
- pass the combined formula to `mvgam()` with a single shared
  `trend_formula = ~ ...`.

Each post-fit method (`summary()`, `pp_check()`,
`conditional_effects()`, `residuals()`, `posterior_predict()`,
`forecast()`, `loo()`) either operates directly on the joint
fit or accepts a `resp` argument that scopes the call to a single
arm. Arms with NA values keep their own non-NA rows; the shared
latent state is still informed at every time point any method was
observed.

## Further reading

- The [forecast-evaluation article](https://nicholasjclark.github.io/mvgam/articles/forecast_evaluation.html)
  walks through `forecast()`, `score()`, `lfo_cv()` and
  `ensemble()` on a worked count-data example.
- The [JSDM article](https://nicholasjclark.github.io/mvgam/articles/jsdgam.html)
  shows the joint-species variant where many response columns
  share a latent factor structure with informed priors.
- For hierarchical / grouped trend dynamics (one shared trend per
  group with hyperpriors across groups), see `?gr` and the
  hierarchical-trends section of `vignette("mvgam_overview")`.
