---
title: "One shared process, many sensors: multi-response mvgam"
author: "Nicholas J Clark"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
    toc: yes
    toc_float: yes
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE, dpi = 100, fig.asp = 0.62, fig.width = 7,
  out.width = "85%", fig.align = "center"
)
library(mvgam)
library(brms)
library(ggplot2)
library(patchwork)
theme_set(mvgam_theme(base_size = 12, base_family = "serif"))

cache_dir <- file.path("pkgdown", "mvbf_cache")
read_cache <- function(name) {
  readRDS(file.path(cache_dir, paste0(name, ".rds")))
}
```

Field monitoring rarely produces a single response variable. An
ecologist counting animals along a transect typically also
operates a baited trap and runs a camera array. Each survey
delivers a noisy, partial view of the same underlying ecological
quantity. Standard methods for time series data either fit each
response separately, throwing away the information sitting in
the other methods' columns, or force every column into the same
Gaussian observation model, which is wrong for counts, for
biomass and for detection records. This article walks through
the mvgam alternative: one shared latent dynamic process linked
to many response variables, each with its own observation family
and per-outcome covariates. The latent state is informed jointly
by every response through the likelihood. We demonstrate this on
a simulated three-method monitoring scenario where we know the
truth and can verify recovery directly.

## The data

We simulate a single hypothetical animal population observed over
80 monthly time points by three parallel survey methods. The
underlying latent state `x[t]` follows a stationary AR(1) with
`phi = 0.7` and `sigma = 0.4`. Each method observes that state
through a different measurement process:

- `count`: visual transect counts (Poisson, log link). Survey
  effort is logged each month and enters the linear predictor as
  an offset.
- `biomass`: trap-collected biomass per unit night (Gamma, log
  link). A cyclic seasonal smooth captures the natural seasonal
  rhythm in biomass independent of survey effort.
- `camera`: camera-trap presence/absence (Bernoulli, logit link).
  The number of deployment days enters as a linear slope.

The camera array is operated on a sparser schedule than the other
two: only 60 of the 80 months carry camera observations. mvgam
keeps the camera outcome on its own non-NA rows so the shared
latent state is still informed during the 20 months without a
camera by counts and biomass alone.

```{r data-load}
mv_data <- read_cache("mv_data")
truth   <- read_cache("truth")
str(mv_data, give.attr = FALSE)
```

```{r truth-summary}
cat("True phi      =", truth$phi,   "\n")
cat("True sigma    =", truth$sigma, "\n")
cat("True nu       =", truth$nu,    "(Gamma shape)\n")
cat("camera obs    =", sum(!is.na(mv_data$camera)), "/", nrow(mv_data), "\n")
```

A scatter of the three response columns by time shows what each
method records and where the camera outcome is missing.

```{r plot-data, fig.asp = 0.7, fig.width = 8, out.width = "95%"}
long <- with(mv_data, rbind(
  data.frame(time = time, value = count,   resp = "count"),
  data.frame(time = time, value = biomass, resp = "biomass"),
  data.frame(time = time, value = camera,  resp = "camera")
))
ggplot(long, aes(x = time, y = value)) +
  geom_line(colour = bayesplot::color_scheme_get()$dark, alpha = 0.6) +
  geom_point(size = 1.2, na.rm = TRUE) +
  facet_wrap(~ resp, ncol = 1L, scales = "free_y") +
  labs(x = "Time (month)", y = "Observed value")
```

## Three univariate baselines

To see how much the shared-process model gains over the
default that fits each method alone, we first run three
independent fits with the same AR(1) latent dynamics.

```{r uni-fits, eval = FALSE}
mod_uni_count <- mvgam(
  count ~ 1 + offset(log_effort),
  trend_formula = ~ AR(p = 1),
  family = poisson(), data = mv_data,
  chains = 2L, iter = 2000L, warmup = 1000L, silent = 2L
)
mod_uni_biomass <- mvgam(
  biomass ~ 1 + s(month, bs = "cc", k = 8),
  trend_formula = ~ AR(p = 1),
  family = Gamma(link = "log"), data = mv_data,
  knots = list(month = c(0.5, 12.5)),
  chains = 2L, iter = 2000L, warmup = 1000L, silent = 2L
)
mod_uni_camera <- mvgam(
  camera ~ 1 + deploy_days,
  trend_formula = ~ AR(p = 1),
  family = bernoulli(link = "logit"), data = mv_data,
  chains = 2L, iter = 2000L, warmup = 1000L, silent = 2L
)
```

```{r uni-load, include = FALSE}
mod_uni_count   <- read_cache("mod_uni_count")
mod_uni_biomass <- read_cache("mod_uni_biomass")
mod_uni_camera  <- read_cache("mod_uni_camera")
```

Each fit converges (max R-hat below 1.05) but the camera outcome
carries very little information by itself. The Bernoulli observations
above provide only one bit per time point, so its AR(1) estimate
of `x[t]` is correspondingly diffuse.

## A joint fit: one process, three observation families

The joint fit lives in one `mvgam()` call. We chain three `bf()`
outcomes with the `+` operator (a brms idiom that mvgam re-exports).
Each outcome gets its own family and its own predictors. Every
outcome shares the same `trend_formula = ~ AR(p = 1)`: one latent
state feeding three observation models.

```{r joint-fit, eval = FALSE}
joint_formula <-
  bf(count   ~ 1 + offset(log_effort), family = poisson()) +
  bf(biomass ~ 1 + s(month, bs = "cc", k = 8),
     family = Gamma(link = "log")) +
  bf(camera  ~ 1 + deploy_days,
     family = bernoulli(link = "logit"))

mod_joint <- mvgam(
  joint_formula,
  trend_formula = ~ AR(p = 1),
  data  = mv_data,
  knots = list(month = c(0.5, 12.5)),
  chains = 2L, iter = 2000L, warmup = 1000L, silent = 2L
)
```

```{r joint-load, include = FALSE}
mod_joint <- read_cache("mod_joint")
```

mvgam handles the bookkeeping. brms's default `mvbf` would
drop those 20 rows from every outcome via listwise deletion.
mvgam instead expands the standata so each outcome keeps its own
non-NA rows. The Stan model then evaluates the count likelihood at all
80 time points, the biomass likelihood at all 80 and the camera
likelihood at the 60 observed months, all contributing to a
single shared AR(1) state.

`summary()` returns a classed object whose `spec` slot holds the
distributional parameters each family contributes, so the Gamma
shape belonging to the biomass arm can be read on its own.
`posterior_summary()` then pulls the intercepts, the covariate
slope and the trend parameters into one table for the recovery
check.

```{r joint-summary}
s <- summary(mod_joint)
print(s$spec)
```

```{r joint-pars}
ps <- posterior_summary(mod_joint)
keep <- grep("Intercept|deploy_days|shape|ar1|sigma_trend",
              rownames(ps))
round(ps[keep, c("Estimate", "Q2.5", "Q97.5")], 3)
```

Recovery against the truth (`phi = 0.7`, `sigma = 0.4`,
`a_count = 1.5`, `a_biomass = 0.5`, `a_camera = -0.5`,
`b_deploy = 0.05`, `nu = 2.0`) is clean: every intercept, the AR
coefficient, the innovation SD and the Gamma shape sit close to
their generating values with 95% intervals that cover.

## Posterior-predictive checks

Each outcome gets its own posterior-predictive check via the
standard `pp_check()` interface, scoped with the `resp` argument.
For the Bernoulli camera outcome the `"bars"` type is more
informative than the default density overlay (which is built for
continuous responses).

```{r pp-count, fig.asp = 0.55}
pp_check(mod_joint, resp = "count", ndraws = 50L) +
  labs(title = "PPC: visual transect counts")
```

```{r pp-biomass, fig.asp = 0.55}
pp_check(mod_joint, resp = "biomass", ndraws = 50L) +
  labs(title = "PPC: trap biomass")
```

```{r pp-camera, fig.asp = 0.55, fig.width = 5}
pp_check(mod_joint, resp = "camera", type = "bars", ndraws = 50L) +
  labs(title = "PPC: camera detections")
```

The replicate distributions cover the observed marginals for
every outcome. Calibration is sound on all three.

## Conditional effects

`conditional_effects()` returns one list per response, faceted
internally by predictor. The plot for each outcome now carries the
correct response on its y-axis.

```{r ce-biomass, fig.asp = 0.55}
ce <- mvgam::conditional_effects(mod_joint)
plot(ce[["biomass"]])[[1L]]
```

The biomass outcome picks up the seasonal sinusoid that drove the
simulation: a clear peak in early months and a trough in late
summer. The cyclic smooth penalisation is doing its job, and the
posterior median follows the truth even though we never told the
model that month was seasonal.

```{r ce-camera, fig.asp = 0.55}
plot(ce[["camera"]])[[2L]]
```

The camera outcome's detection probability rises with deploy days,
consistent with the simulated effect of 0.05 log-odds per day.
The CI is wide because Bernoulli observations carry little
information per row.

## Residual diagnostics

`plot(mod, type = "residuals", resp = X)` returns a four panel
diagnostic of the Dunn-Smyth residuals for one outcome. The
panels show residuals against fitted values, a Normal quantile
plot, ACF and partial ACF. When the model is doing its job the
residuals will sit horizontally around zero with no trend across
the fitted range, will fall close to the quantile line and will
show no autocorrelation lags reaching beyond the dashed bands.

```{r resid-count, fig.asp = 0.7, fig.width = 8}
plot(mod_joint, type = "residuals", resp = "count", ndraws = 100L)
```

```{r resid-biomass, fig.asp = 0.7, fig.width = 8}
plot(mod_joint, type = "residuals", resp = "biomass", ndraws = 100L)
```

```{r resid-camera, fig.asp = 0.7, fig.width = 8}
plot(mod_joint, type = "residuals", resp = "camera", ndraws = 100L)
```

The count and camera quantile plots track the diagonal almost
exactly. The biomass plot shows a mild deviation in the upper
tail, marking one or two large biomass values the Gamma is
slightly undershooting. None of the three outcomes show
residual autocorrelation outside the lag-1 confidence band,
which is the diagnostic the shared AR(1) latent state was put
in place to absorb.

## State recovery: joint vs baselines

The point of the joint fit is the latent state itself. We pull
the per-draw latent trajectory out of each fit with
`hindcast(fit, type = "trend")`. The resulting `mvgam_forecast`
plots with the package's standard ribbon and median style. A
`geom_point()` layer adds the simulated truth on top so recovery
is read off directly.

```{r recovery-plot, fig.asp = 0.62, fig.width = 9, out.width = "95%"}
fits <- list(
  "Joint (3 families)" = mod_joint,
  "Counts only"        = mod_uni_count,
  "Biomass only"       = mod_uni_biomass,
  "Camera only"        = mod_uni_camera
)
truth_pts <- data.frame(
  time = seq_along(truth$x), trend = truth$x
)

plot_recovery <- function(fit, label) {
  # `hindcast(fit, type = "trend")` returns an `mvgam_forecast`
  # object whose `plot()` method already skips the response-scale
  # observation overlay for trend hindcasts. We only need to
  # overlay the simulated truth vector as ×-marks.
  plot(hindcast(fit, type = "trend")) +
    geom_point(
      data = truth_pts,
      aes(x = time, y = trend),
      shape = 4, size = 1.2, colour = "black"
    ) +
    labs(title = label, y = "Latent state")
}

patchwork::wrap_plots(
  lapply(names(fits), function(n) plot_recovery(fits[[n]], n)),
  ncol = 2L
)
```

Two simple numbers back this up: the root mean squared error
between the posterior median and the simulated truth at each
time point, and the average posterior SD across time points.
The first captures how close the median is to the truth; the
second captures how tight the ribbon is.

```{r recovery-table}
recovery_summary <- function(fit) {
  hc <- hindcast(fit, type = "trend")
  # Multi-response fits return the fan-out wrapper; every arm
  # carries the same shared trend, so peel any one for the
  # posterior matrix.
  if (isTRUE(attr(hc, "mv_wrapper"))) hc <- hc[[1L]]
  m <- hc$hindcasts[[1L]]
  med <- apply(m, 2L, median)
  # Latent factor models identify the trend only up to sign; flip
  # the posterior to the sign that maximises correlation with the
  # truth before computing the recovery metrics.
  sgn <- sign(cor(med, truth$x))
  m <- sgn * m
  med <- sgn * med
  c(RMSE     = sqrt(mean((med - truth$x)^2)),
    `Mean SD` = mean(apply(m, 2L, sd)))
}
recovery <- t(sapply(fits, recovery_summary))
round(recovery, 3)
```

The joint fit posts the smallest RMSE and the tightest mean SD;
the camera-only fit is the outlier on both, because a single
Bernoulli observation per time point carries little information.
The joint fit borrows strength wherever any of the methods
reported.

## Per-outcome posterior forecasts

The same `forecast()` interface that handles univariate fits now
works on a multi-response fit: it fans out per response and
returns one `mvgam_forecast` per outcome. We extend the time grid
by 5 months and let mvgam propagate the shared latent state
forward, then map back through each outcome's observation family.

```{r forecast-multi, fig.asp = 1.0, fig.width = 7, out.width = "85%"}
T_ <- max(mv_data$time)
new <- data.frame(
  time = (T_ + 1L):(T_ + 5L),
  series = factor(rep("pop1", 5L), levels = levels(mv_data$series)),
  month = ((T_ + 0L:4L) %% 12L) + 1L,
  log_effort = log(rep(1, 5L)),
  deploy_days = rep(15, 5L),
  count = NA_integer_, biomass = NA_real_, camera = NA_integer_
)
fc <- forecast(mod_joint, newdata = new)
patchwork::wrap_plots(
  lapply(names(fc), function(n) plot(fc[[n]]) + labs(title = n)),
  ncol = 1L
)
```

Each panel shows that outcome's posterior-predictive ribbon
through the 80 training months and the 5 forecast months ahead.
The same forecast objects feed straight into `score()`; see the
[forecast evaluation article](https://nicholasjclark.github.io/mvgam/articles/forecast_evaluation.html)
for the scoring rules and rolling-origin tools that pair with
`forecast()` here.

## Take-aways

A multi-response mvgam fit lets each measurement model keep its
proper observation family while sharing the dynamics that drive
every series. The recipe is small: write one `bf()` block per
response with its own family and predictors, combine the blocks
with `+`, and pass the combined formula to `mvgam()` with a
single shared `trend_formula = ~ ...`. Every post-fit method
(`summary()`, `pp_check()`, `conditional_effects()`,
`residuals()`, `posterior_predict()`, `forecast()`, `loo()`)
either operates directly on the joint fit or accepts a `resp`
argument that scopes the call to a single outcome. Outcomes
with NA values keep their own non-NA rows, so the shared latent
state is still informed at every time point any method actually
reported.

This is the data integration idea applied to time series rather
than to population vital rates: several noisy measurement
processes pool into a single shared latent state, and every
post-fit method downstream dispatches on that state with the
`resp` argument to scope back to a single outcome.

## Further reading

The natural next functions to explore on the same fit are:

- `score()` and `ensemble()` for proper scoring rules and
  stacked weights on per-outcome forecasts.
- `lfo_cv()` for rolling-origin Bayesian leave-future-out
  cross-validation when the question is "how would this model
  have performed in production".
- `loo()`, `kfold.mvgam()` and `posterior_average()` for
  in-sample model comparison and posterior-weighted predictions.
- `irf()`, `fevd()` and `stability()` for impulse response,
  forecast-error variance decomposition and stationarity
  diagnostics when the latent dynamics are a VAR.
- `methods_md()` for a citation-ready methods paragraph naming
  the architecture pieces the fit actually used.

Sibling articles in this series carry the workflow further:

- The [forecast evaluation article](https://nicholasjclark.github.io/mvgam/articles/forecast_evaluation.html)
  walks through `forecast()`, `score()`, `lfo_cv()` and
  `ensemble()` on a single count outcome and the multivariate
  scoring rules that pair with them.
- The [integrated distribution article](https://nicholasjclark.github.io/mvgam/articles/idm.html)
  takes the same architecture into species distribution
  modelling, where the streams are survey protocols and the
  shared state is a log-intensity.
- The [JSDM article](https://nicholasjclark.github.io/mvgam/articles/jsdgam.html)
  shows the joint-species variant where many response columns
  share a latent factor structure with informed priors via
  `jsdgam()`.
- For grouped trend dynamics (one transition matrix and
  innovation covariance per group, shrunk toward a global mean),
  see the `gr` and `subgr` arguments in `?VAR` and the
  [hierarchical VAR article](https://nicholasjclark.github.io/mvgam/articles/hierarchical_var.html).

## Acknowledgements

The data integration framing draws on Schaub & Abadi
([2011](https://doi.org/10.1007/s10336-010-0632-7)). The
latent dynamic perspective follows the review by Auger-Méthé and
colleagues ([2021](https://doi.org/10.1002/ecm.1470)). The
probabilistic backend is Stan (Carpenter et al.
[2017](https://doi.org/10.18637/jss.v076.i01)) accessed through
brms (Bürkner
[2017](https://doi.org/10.18637/jss.v080.i01)).

## References

Auger-Méthé M, Newman K, Cole D, Empacher F, Gryba R, King AA,
Leos-Barajas V, Mills Flemming J, Nielsen A, Petris G and Thomas
L (2021). A guide to state-space modeling of ecological time
series. *Ecological Monographs* 91(4), e01470.
[doi:10.1002/ecm.1470](https://doi.org/10.1002/ecm.1470)

Bürkner P-C (2017). brms: An R Package for Bayesian Multilevel
Models Using Stan. *Journal of Statistical Software* 80(1),
1-28.
[doi:10.18637/jss.v080.i01](https://doi.org/10.18637/jss.v080.i01)

Carpenter B, Gelman A, Hoffman MD, Lee D, Goodrich B, Betancourt
M, Brubaker M, Guo J, Li P and Riddell A (2017). Stan: A
Probabilistic Programming Language. *Journal of Statistical
Software* 76(1), 1-32.
[doi:10.18637/jss.v076.i01](https://doi.org/10.18637/jss.v076.i01)

Schaub M and Abadi F (2011). Integrated population models: a
novel analysis framework for deeper insights into population
dynamics. *Journal of Ornithology* 152(Suppl 1), S227-S237.
[doi:10.1007/s10336-010-0632-7](https://doi.org/10.1007/s10336-010-0632-7)
