---
title: "N-mixture models with mvgam"
author: "Nicholas J Clark"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
    toc: yes
    toc_float: yes
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  dpi = 100,
  fig.asp = 0.7,
  fig.width = 6,
  out.width = "70%",
  fig.align = "center"
)
library(mvgam)
library(ggplot2)
theme_set(theme_bw(base_size = 12, base_family = "serif"))
```

When we count animals at a site we rarely see all of them, so the
count we write down mixes the true population size with our ability
to detect it. If we only survey once there is no way to tell a large
population that was surveyed poorly from a small one that was
surveyed well. N-mixture models solve this by using repeated surveys
(Royle [2004](https://doi.org/10.1111/j.0006-341X.2004.00142.x)). If
we visit a site several times within a period short enough that the
population cannot change, the repeated counts tell us about both
quantities: their average reflects abundance, while how much they
vary reflects detection.

In this article we fit an N-mixture model in `mvgam` to two
simulated species. Because we simulate the data ourselves we know
the true abundances and the true detection probabilities, so we can
check the estimates against them.

## N-mixture models

For a count $y_{i,r}$ at site $i$ on visit $r$, the model factors
into a state layer and an observation layer:

\begin{align*}
y_{i,r} & \sim \text{Binomial}(N_i, p_r) \\
N_i     & \sim \text{Poisson}(\lambda_i).
\end{align*}

Covariates enter through linear predictors for the abundance mean
(log link) and the detection probability (logit link):

\begin{align*}
\log(\lambda_i) & = \beta_0 + \beta^\top \mathbf{x}_i \\
\text{logit}(p_r) & = \gamma_0 + \gamma^\top \mathbf{q}_r.
\end{align*}

This works because the binomial mean $N_i p_r$ and the binomial
variance $N_i p_r (1 - p_r)$ respond differently as $p$ changes. A
single visit only gives us the mean, which is why one survey can
never separate abundance from detection, but repeated visits also
give us the variance, and that is enough to estimate both.

The latent abundance $N_i$ is an integer, and Hamiltonian Monte
Carlo cannot sample discrete parameters. `mvgam` deals with this the
same way the Stan documentation deals with any discrete latent
variable, by summing the likelihood over a finite range of possible
$N$ values so that only continuous parameters are actually sampled.
The upper limit of that sum comes from a `cap` variable that we
supply. If you would like more detail on marginalising discrete
parameters, the Stan User's Guide has a chapter on
[latent discrete parameters](https://mc-stan.org/docs/stan-users-guide/latent-discrete.html),
and McElreath's
[Algebra and the Missing Oxen](https://elevanth.org/blog/2018/01/29/algebra-and-missingness/)
is a good informal introduction.

## Closure-unit data shape

A closure unit is a set of visits over which the population is
assumed closed. `mvgam` needs one row per visit, with `series`
naming the population, `time` identifying the closure unit, and
`cap` giving the upper bound for the marginalisation:

```{r sim-data}
set.seed(999)
truth <- list(sp_1 = c(28, 26, 23, 16, 14, 14),
              sp_2 = c(4, 7, 15, 16, 19, 18))
detection <- c(sp_1 = 0.70, sp_2 = 0.45)

dat <- do.call(rbind, lapply(names(truth), function(sp) {
  data.frame(
    series = sp,
    time = sort(rep(1:6, 5)),
    visit = rep(1:5, 6),
    truth = rep(truth[[sp]], each = 5),
    y = unlist(lapply(truth[[sp]], function(N) {
      rbinom(5, N, detection[[sp]])
    })),
    cap = if (sp == "sp_1") 100L else 50L
  )
}))
dat$series <- factor(dat$series, levels = names(truth))
dat <- dat[order(dat$time, dat$series, dat$visit), ]
head(dat, 10)
```

Here we simulate two species surveyed over six years, with five
replicate visits in each year. Species 1 declines from 28 individuals
to 14 and has a detection probability of 0.7, so we see most of the
individuals that are there. Species 2 increases from 4 to 18 but has
a detection probability of 0.45, so on average we miss more than half
of them. We keep a `truth` column so we can compare our estimates
against the simulated values later, but the model never uses it.

Note that we also add the `cap` variable, which does not need to be
static, to define the maximum possible value that we think the latent
abundance could be. This simply needs to be large enough that we get
a reasonable idea of which latent $N$ values are most likely, without
adding too much computational cost, because every extra unit of `cap`
adds another term to the sum for every closure unit at every
iteration.

## Fitting

The two linear predictors go into one `bf()` call. The first arm is
abundance, the second is detection. Replicate visits sharing a
`(series, time)` cell are one closure unit by construction, so
nothing has to be declared to link them:

```{r fit, results = "hide"}
mod <- mvgam(
  formula = bf(y ~ s(time, k = 4, by = series) + series,
               p ~ series),
  data = dat,
  family = nmix(),
  chains = 4L,
  iter = 4000L,
  warmup = 2000L,
  silent = 2L
)
```

We give abundance a smooth of time for each species so that each
population can follow its own trajectory, and we give detection only
a species intercept, since detection was simulated as constant within
a species.

```{r summary}
summary(mod, include_betas = FALSE)
```

```{r diagnostics, echo = FALSE}
rhat_max <- round(max(posterior::summarise_draws(
  as_draws_df(mod), "rhat")$rhat, na.rm = TRUE), 3)
ess_min <- round(min(posterior::summarise_draws(
  as_draws_df(mod), "ess_bulk")$ess_bulk, na.rm = TRUE))
```

The worst R-hat is `r rhat_max` and the smallest bulk effective
sample size is `r ess_min`, so the chains have mixed and there are
enough independent draws to summarise.

## Recovering detection

Detection probability is the quantity we cannot estimate from a
single survey, so it is worth checking first. The species intercepts
are on the logit scale, so we convert them with `plogis()`:

```{r detection-recovery}
post <- as.matrix(mod$fit)
p_sp1 <- plogis(post[, "Intercept_p"])
p_sp2 <- plogis(post[, "Intercept_p"] + post[, "b_p[1]"])

data.frame(
  species = c("sp_1", "sp_2"),
  truth = as.numeric(detection),
  median = round(c(median(p_sp1), median(p_sp2)), 2),
  lower = round(c(quantile(p_sp1, 0.025), quantile(p_sp2, 0.025)), 2),
  upper = round(c(quantile(p_sp1, 0.975), quantile(p_sp2, 0.975)), 2)
)
```

The model puts species 1 as the more detectable of the two, which is
how we simulated it. Both intervals are wide, because 30 visits per
species is not much information to estimate a probability from. If
detection matters to you, this is the number to look at before
trusting anything downstream of it.

## Latent abundance

The point of the model is $N$. `hindcast(type = "latent_state")`
returns its posterior at every closure unit:

```{r latent-state}
hc <- hindcast(mod, type = "latent_state")
hc
```

Calling `plot()` on that object returns a `ggplot` object, drawn as
steps because $N$ can only take integer values. Since it is a plain
`ggplot`, we can overlay the simulated truth and the raw counts
without writing the plotting code by hand:

```{r latent-plot, fig.asp = 0.45, out.width = "90%"}
truth_long <- unique(dat[, c("series", "time", "truth")])

plot(hc) +
  geom_jitter(data = dat, aes(x = time, y = y),
              width = 0.08, shape = 21, fill = "darkred",
              colour = "white", size = 2.2) +
  geom_line(data = truth_long, aes(x = time, y = truth),
            linewidth = 0.9) +
  geom_point(data = truth_long, aes(x = time, y = truth),
             shape = 21, fill = "black", colour = "white",
             size = 2.6) +
  labs(x = "Year", y = "Latent abundance (N)",
       title = "Counts (red), true abundance (black), posterior")
```

The red points are the counts we actually recorded, and they sit
below the true abundance everywhere, further below it for species 2
because its detection probability is lower. The posterior sits above
the counts rather than following them, which is the whole reason for
fitting an N-mixture: a model of the counts alone would report both
populations as smaller than they are.

```{r coverage-check, echo = FALSE}
state <- as.data.frame(hc)
seen_max <- stats::aggregate(y ~ series + time, data = dat, FUN = max)
checked <- merge(state, seen_max, by = c("series", "time"))
n_below <- sum(checked$median < checked$y)
```

One quick check: a population can never be smaller than the number
of individuals we counted in it. Across all `r nrow(checked)` closure
units, `r n_below` have a posterior median below the largest count
recorded there.

The intervals are wide, and wider for species 2 throughout. The data
mostly tell us about the product $N p$, so separating abundance from
detection costs us precision on both.

## Alternative observation processes

The assumption we have used so far, that detection is an independent
Bernoulli trial for each of the $N$ individuals present, is the most
common N-mixture formulation but it is not the only one. `nmix()`
gives us two further variants through its `type` argument. Latent $N$
is marginalised in the same way for each of them, and everything we
have done above works unchanged.

| Variant | Constructor | Detection scale | When to prefer |
|---------|-------------|-----------------|----------------|
| Poisson-binomial | `nmix()` | logit $p$ per visit | Default. Direct Royle (2004) parameterisation; cleanest when individuals are independently detectable. |
| Royle-Nichols | `nmix("royle_nichols")` | log $r$ per individual | Repeated presence-absence rather than counts. Heterogeneous detection collapses into a per-individual hazard $r$, and the binary observation arises from $1 - (1 - r)^N$. |
| Poisson-Poisson | `nmix("poisson_poisson")` | log $\rho$ per visit | Encounter-rate data, where the count is a noisy Poisson realisation of an underlying encounter rate with no detection ceiling. |

Switching between them only requires changing the `family` argument;
the data and the formula stay the same.

```{r variant-switch, eval = FALSE}
mod_rn <- mvgam(
  formula = bf(y ~ s(time, k = 4, by = series) + series, p ~ series),
  data = dat, family = nmix("royle_nichols"),
  chains = 4L, iter = 4000L, warmup = 2000L, silent = 2L
)
```

The Royle-Nichols variant follows
[Royle and Nichols (2003)](https://doi.org/10.1890/0012-9658\(2003\)084%5B0777:EAFRPA%5D2.0.CO;2);
the Poisson-Poisson variant is the encounter-rate parameterisation
adopted for `mvgam`. See `?nmix` for the full argument list,
including the `multi_season` opt-in for fits with a real seasonal
axis.

## Where to next

- `?nmix` for the constructor's family-specific dpars,
  identification caveats, and the per-variant kernel references.
- `?occ` for the single-season occupancy family, which shares the
  closure-unit data shape with `nmix`.
- `pp_check(mod, type = "fit_stat")` for the closure-unit goodness of
  fit check of MacKenzie and Bailey (2004), which compares the
  observed within-unit discrepancy against its posterior predictive
  distribution.
- The [integrated distribution article](https://nicholasjclark.github.io/mvgam/articles/idm.html)
  for the next step up, when detection-corrected counts are one of
  several survey streams watching the same population.
- The [data formatting vignette](https://nicholasjclark.github.io/mvgam/articles/data.html)
  for the closure-unit column layout in isolation.

## Further reading

Knape, J. and Korner-Nievergelt, F. (2015). Estimates from
non-replicated population surveys rely on critical assumptions.
*Methods in Ecology and Evolution* 6, 902-906.
[doi:10.1111/2041-210X.12377](https://doi.org/10.1111/2041-210X.12377)

MacKenzie, D. I. and Bailey, L. L. (2004). Assessing the fit of
site-occupancy models. *Journal of Agricultural, Biological, and
Environmental Statistics* 9, 300-318.
[doi:10.1198/108571104X3361](https://doi.org/10.1198/108571104X3361)

Royle, J. A. (2004). N-mixture models for estimating population
size from spatially replicated counts. *Biometrics* 60, 108-115.
[doi:10.1111/j.0006-341X.2004.00142.x](https://doi.org/10.1111/j.0006-341X.2004.00142.x)

Royle, J. A. and Nichols, J. D. (2003). Estimating abundance from
repeated presence-absence data or point counts. *Ecology* 84,
777-790.
[doi:10.1890/0012-9658(2003)084](https://doi.org/10.1890/0012-9658\(2003\)084%5B0777:EAFRPA%5D2.0.CO;2)
