---
title: "State-space vector autoregressions with mvgam"
author: "Nicholas J Clark"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
    toc: yes
    toc_float: yes
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE, dpi = 100, fig.asp = 0.62, fig.width = 7,
  out.width = "85%", fig.align = "center"
)
library(mvgam)
library(brms)
library(dplyr)
library(ggplot2)
theme_set(theme_bw(base_size = 12, base_family = "serif"))

cache_dir <- if (dir.exists("pkgdown")) {
  file.path("pkgdown", "var_cache")
} else {
  file.path("..", "..", "pkgdown", "var_cache")
}
read_cache <- function(name) {
  readRDS(file.path(cache_dir, paste0(name, ".rds")))
}
```

Vector autoregressions describe how a small collection of time
series move together over time by letting each series depend on
its own recent past and on the recent past of every other series
(Lütkepohl [2005](https://doi.org/10.1007/978-3-540-27752-1)).
The version implemented in mvgam adds a state-space layer on
top. The observed counts sit above a latent vector autoregression
whose innovations may be correlated across series (Heaps
[2022](https://doi.org/10.1080/10618600.2022.2079648); Auger-Méthé
et al. [2021](https://doi.org/10.1002/ecm.1470)). Once fitted, the
posterior over the transition matrix $A$ and the innovation
covariance $\Sigma$ can be interrogated for lagged interactions
(impulse response functions), for the share of forecast
uncertainty each series contributes (forecast error variance
decompositions), and for community-level stability metrics
(reactivity, return rates). This article walks through those
four post-fit diagnostics on an annual kestrel abundance
dataset shipped with the MARSS package.

## The data

American kestrel (*Falco sparverius*) abundance was surveyed by
the North American Breeding Bird Survey across three adjacent
Canadian regions (British Columbia, Alberta, Saskatchewan) from
1969 to 2008. Detection-adjusted annual counts are distributed
with the MARSS package (Holmes et al.
[2012](https://doi.org/10.32614/RJ-2012-002)); we invert the log
transform inside the raw file so the response sits on the natural
count scale, then reshape to long form with a `series` factor
and a `time` column.

```{r data}
load(url("https://github.com/atsa-es/MARSS/raw/master/data/kestrel.rda"))
regions <- c("BC", "Alb", "Sask")
model_data <- do.call(rbind, lapply(seq_along(regions), function(x) {
  data.frame(
    year = kestrel[, 1],
    adj_count = exp(kestrel[, 1 + x]),
    region = regions[x]
  )
})) |>
  dplyr::mutate(series = as.factor(region), time = year) |>
  as.data.frame()
```

A quick look at the observed series with `mvgam_data()`, holding
the last five years out for forecast scoring:

```{r train-test}
train <- model_data[model_data$year < 2004, ]
test  <- model_data[model_data$year >= 2004, ]
```

```{r plot-observed, fig.cap = "Annual detection-adjusted kestrel counts across the three regions. The last five years are held out."}
mvgam_data(train, y = "adj_count", family = Gamma(), newdata = test)
```

Counts sit in a similar range across regions and tick along
without a strong long-run trend. They also share the kind of
annual excursions that a vector autoregression is designed to
describe (a good year in British Columbia tends to line up with
a reasonable year in Alberta, and so on).

## Building a state-space VAR(1)

The observed part of the model is a Gamma likelihood; because
the response is stacked with `series` as a factor and one
family covers all rows, mvgam fits a single shared shape
parameter by default. The latent part is a first-order vector
autoregression with correlated innovations, `VAR(cor = TRUE)`,
driven by a region intercept:

$$
\begin{aligned}
y_{r, t}  \;&\sim\; \mathrm{Gamma}\!\left(\mathrm{shape},\; \mu_{r, t}\right) \\
1 / \mu_{r, t} \;&=\; z_{r, t} \\
\mathbf{z}_t \;&\sim\; \mathrm{MVN}\!\left(\alpha_{\mathrm{region}} + A\,\mathbf{z}_{t-1},\; \Sigma\right).
\end{aligned}
$$

Priors for the region intercepts and the innovation SD are pulled
in tighter than the defaults so the sampler does not have to
learn scale from three dozen observations per region. The
sampler uses `adapt_delta = 0.95` to handle the tight
`sigma_trend` prior; the fit is cached to avoid re-running at
render time.

```{r fit-code, eval = FALSE}
varmod <- mvgam(
  formula       = adj_count ~ -1,
  trend_formula = ~ region + VAR(cor = TRUE),
  priors        = c(
    prior(std_normal(), class = b_trend),
    prior(exponential(2.5), class = sigma_trend)
  ),
  data          = train,
  newdata       = test,
  family        = Gamma(),
  adapt_delta   = 0.95,
  chains        = 2,
  samples       = 500,
  burnin        = 1000,
  silent        = 2
)
```

If instead you want each region to carry its own shape
parameter, express that through the brms distributional
formula. `mvgam_formula()` accepts a `brmsformula` on the
observation side, so `bf(adj_count ~ 0, shape ~ series - 1)`
would give one shape per region, at the cost of tying up more
parameters on a modest dataset.

```{r fit-load, include = FALSE}
varmod <- read_cache("kestrel_fit")$varmod
train  <- read_cache("kestrel_fit")$train
test   <- read_cache("kestrel_fit")$test
```

## Model summary

`summary()` on a VAR fit splits the parameter table between the
observation model (here a single shared shape) and the trend
model (the region intercepts, the VAR coefficient array
`A_raw_trend`, and the innovation Cholesky `L_Omega_trend`). We
suppress the spline coefficient table with `include_betas =
FALSE` because there are no smooth terms on the observation
side.

```{r summary}
summary(varmod, include_betas = FALSE)
```

The three region intercepts sit close to zero on the
inverse link scale, and the VAR self-coefficients on the diagonal
of $A$ (rows `A_raw_trend[1,i,i]`) are all positive and bounded
well below one, so the Heaps
([2022](https://doi.org/10.1080/10618600.2022.2079648)) prior
that guarantees stationarity is not pushing against the data.

## Inspecting the transition matrix

The lagged dependence structure lives in the $3 \times 3$
matrix $A$. `mcmc_plot()` with `variable = "A_trend"` and
`type = "hist"` shows the posterior for every cell. Passing
`facet_args = list(dir = "v")` walks the facets column-first, so
row $i$, column $j$ (the effect of process $j$'s state at time
$t-1$ on process $i$ at time $t$) sits where you would draw it
by hand.

```{r plot-A, fig.cap = "Posterior histograms for each cell of the VAR transition matrix A. Diagonals capture self-lag; off-diagonals capture lagged cross-region effects."}
mcmc_plot(
  varmod,
  variable    = "A_trend",
  regex       = TRUE,
  type        = "hist",
  facet_args  = list(dir = "v")
)
```

The clearest signal is on the diagonal (`A_trend[1,1,1]`,
`[1,2,2]`, `[1,3,3]`): each region has substantial positive
self-persistence, so a good year raises the expected count in
the following year within the same region. The off-diagonals
are more ambiguous, with `A_trend[1,2,1]` (BC → Alberta)
showing the most concentrated positive posterior, consistent
with the two adjacent regions moving together.

## Inspecting the innovation covariance

The innovation covariance $\Sigma$ tells you which regions share
contemporaneous shocks, the kind of common weather or
detection-effort year that would nudge all three surveys in the
same direction at once. The mvgam Stan model saves `Sigma_trend`
directly, so we can compute the posterior median correlation
matrix by applying `cov2cor()` to each draw:

```{r sigma-cor}
Sig_post <- as.matrix(varmod, variable = "^Sigma_trend\\[", regex = TRUE)
cor_draws <- t(apply(Sig_post, 1, function(v) {
  as.numeric(cov2cor(matrix(v, 3, 3)))
}))
med_cor <- matrix(apply(cor_draws, 2, median), 3, 3)
rownames(med_cor) <- colnames(med_cor) <- c("BC", "Alb", "Sask")
round(med_cor, 2)
```

The two neighbouring pairs (BC and Alberta, r ≈ 0.32; Alberta
and Saskatchewan, r ≈ 0.29) share moderate positive innovation
correlations, while the pair at either end of the range (BC and
Saskatchewan, r ≈ -0.03) share essentially none. Geography falls
out of the fit without being told.

## Forecasts

Passing `newdata = test` at fit time stored the held-out slice
on the fitted object, so `forecast()` extends the VAR five years
into the future without any extra arguments. The plot draws the
hindcast (grey) up to the dashed vertical, then the forecast
posterior (red) over the five held-out years.

```{r plot-forecast, fig.cap = "Hindcast (grey) and five-year posterior forecast (red) from the correlated-innovation VAR. The mean count trajectory stays close to the historical average, with the ribbon widening as the horizon lengthens."}
plot(forecast(varmod, newdata = test))
```

Forecasts fan out gradually as the VAR is integrated forward.
Because the diagonal of $A$ is safely below one, the mean
converges to the region intercept while the ribbon spreads
according to the stationary covariance implied by
$(A, \Sigma)$.

## Impulse response functions

An impulse response function asks the following counterfactual:
if a shock of one standard deviation hit region $j$ at time
zero, how would the other regions be expected to move over the
next $h$ years? (Lütkepohl
[2005](https://doi.org/10.1007/978-3-540-27752-1)). `irf()`
computes the generalized IRF for every source and target pair;
`plot(irfs, series = j)` fixes the shock at region $j$ and draws
one panel per target region.

```{r irf-fit}
irfs <- irf(varmod, h = 12)
```

```{r plot-irf1, fig.cap = "Generalized impulse responses to a positive shock in region 1 (British Columbia). Each panel shows how the process for the target region is expected to respond over the next 12 years."}
plot(irfs, series = 1)
```

A positive shock in British Columbia (region 1) decays within
its own process (left panel) over roughly ten years, moves
Alberta's process up by about 0.03 units on the inverse-link
scale (middle panel), and pushes Saskatchewan's process slightly
positive with a delayed peak (right panel).

```{r plot-irf2, fig.cap = "Impulse responses to a shock in region 2 (Alberta). The self-response decays smoothly; British Columbia responds in the same direction as Alberta, and Saskatchewan shows a delayed positive response consistent with the adjacent-region correlation seen in Sigma."}
plot(irfs, series = 2)
```

Shocks originating in Alberta propagate outwards to both
neighbours, with the delayed Saskatchewan peak echoing the
delayed cross-lag effect the transition matrix hinted at.

```{r plot-irf3, fig.cap = "Impulse responses to a shock in region 3 (Saskatchewan). The self-response is the sharpest of the three, and the effect on Alberta persists longer than the effect on British Columbia."}
plot(irfs, series = 3)
```

Saskatchewan shocks have the tightest self-response and touch
Alberta more than they touch British Columbia, consistent with
the geography embedded in the innovation correlations.

## Forecast error variance decomposition

The forecast error variance decomposition (FEVD) turns the same
underlying quantities around and asks the complementary
question: at horizon $h$, how much of the forecast uncertainty
for region $i$ was generated by shocks that originated in each
region? `fevd()` returns the posterior draws;
`plot(fevds)` renders one stacked-bar panel per target region,
with the bar heights showing the median share attributed to each
source over increasing horizons.

```{r fevd-fit}
fevds <- fevd(varmod, h = 12)
```

```{r plot-fevd, fig.cap = "Forecast error variance decomposition. Each panel is a target region; each bar breaks down that region's forecast uncertainty at a given horizon into contributions from own shocks and shocks originating in the other two regions."}
plot(fevds)
```

Own-region shocks dominate at short horizons for every region
(the darker slice on the bottom of each bar), consistent with
the strong diagonal in $A$. As the horizon lengthens, the
cross-region shares grow: by year 12 Alberta contributes about
25% of the forecast variance in British Columbia and about 20%
of the forecast variance in Saskatchewan. The two regions at
either end of the range contribute least to each other's
forecast uncertainty, which is the flip side of the negligible
BC and Saskatchewan correlation in $\Sigma$.

## Community stability metrics

The transition matrix and innovation covariance also determine
whether the system as a whole would amplify or absorb small
shocks around the stationary distribution. `stability()`
computes posterior draws of three interpretable dynamic-stability
metrics (Ives et al.
[2003](https://doi.org/10.1890/0012-9615(2003)073[0301:ECSAEI]2.0.CO;2);
Neubert & Caswell
[1997](https://doi.org/10.1890/0012-9658(1997)078[0653:ATRFMT]2.0.CO;2)):

- **Reactivity**: the log of the largest singular value of $A$.
  Positive reactivity means small perturbations grow immediately
  after the shock before eventually decaying; negative
  reactivity means the perturbation shrinks straight away.
- **Mean return rate**: the modulus of the largest eigenvalue
  of $A$. Values closer to one imply a slower return of the
  mean to its stationary distribution after a shock.
- **Variance return rate**: the analogous quantity for $A
  \otimes A$, describing how quickly forecast variance decays
  back to its stationary level.

`plot()` on the returned `mvgam_stability` object renders each
metric as a posterior histogram with a dashed reference at zero
(the reactivity threshold).

```{r plot-stability, fig.width = 8, fig.asp = 0.45, fig.cap = "Posterior distributions of three community-stability metrics for the fitted VAR. The dashed line marks zero, the reactivity threshold above which small shocks would be amplified rather than absorbed."}
stab <- stability(varmod)
plot(stab)
```

Reactivity is centred near zero (posterior median $\approx$
-0.03, 90% CI $\approx$ [-0.39, 0.69]) with 45% of the posterior
mass sitting on the positive side of the dashed line. The data
leave genuine uncertainty about whether the system is reactive
or non-reactive, but any amplification would be small. Mean and
variance return rates sit around 0.84 and 0.70 respectively,
consistent with a stable system that takes a handful of years
to relax back to its stationary distribution after a shock.

## Take-aways

Once a VAR is fitted in mvgam, the same four post-fit
diagnostics work end to end for any observation family and any
correlated innovation structure:

- `mcmc_plot(mod, variable = "A_trend", ...)` shows the lagged
  dependence structure per cell of the transition matrix.
- `as.matrix(mod, variable = "Sigma_trend", ...)` + `cov2cor()`
  recovers the posterior distribution of the innovation
  correlation matrix, which tells you which series share
  contemporaneous shocks.
- `irf()` + `plot(irfs, series = j)` traces how a shock at a
  focal series is expected to move each other series over a
  chosen horizon.
- `fevd()` + `plot(fevds)` decomposes the forecast variance for
  each series into contributions from each source of shocks.
- `stability()` + `plot(stab)` summarises the dynamic-stability
  posterior in the units ecologists use (reactivity, return
  rates).

The transition matrix, innovation covariance, IRFs, FEVDs and
stability metrics all rely on the same posterior draws of
$(A, \Sigma)$; the four post-fit diagnostics just ask different
questions of that joint object.

## Further reading

Complementary in-package tools:

- `?VAR` for the trend-side constructor. Hierarchical panel VAR
  support is available via the `gr` and `subgr` arguments.
- `?forecast.mvgam`, `?score` and `?ensemble` for the forecast
  scoring and averaging surface layered on top of any fitted
  mvgam VAR.
- `?how_to_cite` and `?methods_md` for citation-ready methods
  paragraphs that describe the Heaps stationary prior and the
  sampler configuration in place.

Sibling articles in this series:

- The
  [forecast evaluation article](https://nicholasjclark.github.io/mvgam/articles/forecast_evaluation.html)
  runs the same forecast / score machinery on a longer Portal
  rodent time series.
- The
  [multivariate mvgam article](https://nicholasjclark.github.io/mvgam/articles/mvbf.html)
  covers the different case where several outcomes on different
  families share one latent process via a multivariate
  observation formula.
- The `dfm_trend_map` in-package vignette covers dynamic factor
  models when the number of latent processes is deliberately
  smaller than the number of observed series.

## Acknowledgements

The kestrel dataset was compiled from the North American
Breeding Bird Survey and is redistributed with MARSS (Holmes et
al. [2012](https://doi.org/10.32614/RJ-2012-002)). The stationary
Heaps prior on the VAR coefficients follows Heaps
([2022](https://doi.org/10.1080/10618600.2022.2079648)). The
state-space framing follows the review by Auger-Méthé and
colleagues ([2021](https://doi.org/10.1002/ecm.1470)) and the
dynamic generalised additive model perspective in Clark & Wells
([2023](https://doi.org/10.1111/2041-210X.13974)). Community
stability metrics follow Ives et al.
([2003](https://doi.org/10.1890/0012-9615(2003)073[0301:ECSAEI]2.0.CO;2))
and Neubert & Caswell
([1997](https://doi.org/10.1890/0012-9658(1997)078[0653:ATRFMT]2.0.CO;2)).
Impulse response function and forecast error variance
decomposition definitions follow Lütkepohl
([2005](https://doi.org/10.1007/978-3-540-27752-1)). Formulas
rely on mgcv smooths (Wood
[2017](https://doi.org/10.1201/9781315370279)) and brms's
multilevel apparatus (Bürkner
[2017](https://doi.org/10.18637/jss.v080.i01)); sampling is
performed with Stan (Carpenter et al.
[2017](https://doi.org/10.18637/jss.v076.i01)).

## References

Auger-Méthé M, Newman K, Cole D, Empacher F, Gryba R, King AA,
Leos-Barajas V, Mills Flemming J, Nielsen A, Petris G and Thomas
L (2021). A guide to state-space modeling of ecological time
series. *Ecological Monographs* 91(4), e01470.
[doi:10.1002/ecm.1470](https://doi.org/10.1002/ecm.1470)

Bürkner PC (2017). brms: An R package for Bayesian multilevel
models using Stan. *Journal of Statistical Software* 80(1),
1-28.
[doi:10.18637/jss.v080.i01](https://doi.org/10.18637/jss.v080.i01)

Carpenter B, Gelman A, Hoffman MD, Lee D, Goodrich B, Betancourt
M, Brubaker M, Guo J, Li P and Riddell A (2017). Stan: A
probabilistic programming language. *Journal of Statistical
Software* 76(1), 1-32.
[doi:10.18637/jss.v076.i01](https://doi.org/10.18637/jss.v076.i01)

Clark NJ and Wells K (2023). Dynamic Generalized Additive Models
(DGAMs) for forecasting discrete ecological time series.
*Methods in Ecology and Evolution* 14, 771-784.
[doi:10.1111/2041-210X.13974](https://doi.org/10.1111/2041-210X.13974)

Heaps SE (2022). Enforcing stationarity through the prior in
vector autoregressions. *Journal of Computational and Graphical
Statistics* 32(1), 74-83.
[doi:10.1080/10618600.2022.2079648](https://doi.org/10.1080/10618600.2022.2079648)

Holmes EE, Ward EJ and Wills K (2012). MARSS: Multivariate
Autoregressive State-space Models for Analyzing Time-series
Data. *The R Journal* 4(1), 11-19.
[doi:10.32614/RJ-2012-002](https://doi.org/10.32614/RJ-2012-002)

Ives AR, Dennis B, Cottingham KL and Carpenter SR (2003).
Estimating community stability and ecological interactions from
time-series data. *Ecological Monographs* 73(2), 301-330.
[doi:10.1890/0012-9615(2003)073[0301:ECSAEI]2.0.CO;2](https://doi.org/10.1890/0012-9615(2003)073[0301:ECSAEI]2.0.CO;2)

Lütkepohl H (2005). *New Introduction to Multiple Time Series
Analysis*. Springer, Berlin.
[doi:10.1007/978-3-540-27752-1](https://doi.org/10.1007/978-3-540-27752-1)

Neubert MG and Caswell H (1997). Alternatives to resilience for
measuring the responses of ecological systems to perturbations.
*Ecology* 78(3), 653-665.
[doi:10.1890/0012-9658(1997)078[0653:ATRFMT]2.0.CO;2](https://doi.org/10.1890/0012-9658(1997)078[0653:ATRFMT]2.0.CO;2)

Wood SN (2017). *Generalized Additive Models: An Introduction
with R* (2nd edition). Chapman and Hall/CRC.
[doi:10.1201/9781315370279](https://doi.org/10.1201/9781315370279)
