---
title: "State-space vector autoregressions with mvgam"
author: "Nicholas J Clark"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
    toc: yes
    toc_float: yes
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE, dpi = 100, fig.asp = 0.62, fig.width = 7,
  out.width = "85%", fig.align = "center"
)
library(mvgam)
library(brms)
library(dplyr)
library(ggplot2)
theme_set(mvgam_theme(base_size = 12, base_family = "serif"))

cache_dir <- if (dir.exists("pkgdown")) {
  file.path("pkgdown", "var_cache")
} else {
  file.path("..", "..", "pkgdown", "var_cache")
}
read_cache <- function(name) {
  readRDS(file.path(cache_dir, paste0(name, ".rds")))
}
```

Vector autoregressions describe how a small collection of time
series move together over time by letting each series depend on
its own recent past and on the recent past of every other series
(Lütkepohl [2005](https://doi.org/10.1007/978-3-540-27752-1)).
The version implemented in mvgam adds a state-space layer on
top. The observed counts sit above a latent vector autoregression
whose innovations may be correlated across series (Heaps
[2022](https://doi.org/10.1080/10618600.2022.2079648); Auger-Méthé
et al. [2021](https://doi.org/10.1002/ecm.1470)). The payoff for
that extra machinery is what the posterior over the transition
matrix $A$ and the innovation covariance $\Sigma$ then supports.
Lagged interactions can be read cell by cell, impulse responses
trace how a shock at one series moves the others, forecast error
variance decompositions apportion predictive uncertainty across
its sources, and community stability metrics such as reactivity
and return rate say whether the system amplifies or absorbs
perturbations. This article works through each of them on an
annual kestrel abundance dataset shipped with the MARSS package.

## The data

American kestrel (*Falco sparverius*) abundance was surveyed by
the North American Breeding Bird Survey across three adjacent
Canadian regions (British Columbia, Alberta, Saskatchewan) from
1969 to 2008. Detection-adjusted annual counts are distributed
with the MARSS package (Holmes et al.
[2012](https://doi.org/10.32614/RJ-2012-002)); we invert the log
transform inside the raw file so the response sits on the natural
count scale, then reshape to long form with a `series` factor
and a `time` column.

```{r data}
load(url("https://github.com/atsa-es/MARSS/raw/master/data/kestrel.rda"))
regions <- c("BC", "Alb", "Sask")
model_data <- do.call(rbind, lapply(seq_along(regions), function(x) {
  data.frame(
    year = kestrel[, 1],
    adj_count = exp(kestrel[, 1 + x]),
    region = regions[x]
  )
})) |>
  dplyr::mutate(series = as.factor(region), time = year) |>
  as.data.frame()
```

A quick look at the observed series with `mvgam_data()`, holding
the last five years out for forecast scoring:

```{r train-test}
train <- model_data[model_data$year < 2004, ]
test  <- model_data[model_data$year >= 2004, ]
```

```{r plot-observed, fig.cap = "Annual detection-adjusted kestrel counts across the three regions. The last five years are held out."}
mvgam_data(train, y = "adj_count", family = Gamma(), newdata = test)
```

Counts sit in a similar range across regions and tick along
without a strong long-run trend. They also share the kind of
annual excursions that a vector autoregression is designed to
describe (a good year in British Columbia tends to line up with
a reasonable year in Alberta, and so on).

## Building a state-space VAR(1)

The observed part of the model is a Gamma likelihood; because
the response is stacked with `series` as a factor and one
family covers all rows, mvgam fits a single shared shape
parameter by default. The latent part is a first-order vector
autoregression with correlated innovations, `VAR(cor = TRUE)`,
driven by a region intercept:

$$
\begin{aligned}
y_{r, t}  \;&\sim\; \mathrm{Gamma}\!\left(\mathrm{shape},\; \mu_{r, t}\right) \\
1 / \mu_{r, t} \;&=\; z_{r, t} \\
\mathbf{z}_t \;&\sim\; \mathrm{MVN}\!\left(\alpha_{\mathrm{region}} + A\,\mathbf{z}_{t-1},\; \Sigma\right).
\end{aligned}
$$

Priors for the region intercepts and the innovation SD are pulled
in tighter than the defaults so the sampler does not have to
learn scale from three dozen observations per region. A latent
state per region per year is a lot of correlated parameters for
a series this short, so the fit runs four chains at
`adapt_delta = 0.99` rather than the defaults; at two chains and
a looser target the individual states mix too poorly to support
the intervals reported below. The fit is cached to avoid
re-running it at render time.

```{r fit-code, eval = FALSE}
varmod <- mvgam(
  formula       = adj_count ~ -1,
  trend_formula = ~ region + VAR(cor = TRUE),
  priors        = c(
    prior(std_normal(), class = b_trend),
    prior(exponential(2.5), class = sigma_trend)
  ),
  data          = train,
  newdata       = test,
  family        = Gamma(),
  adapt_delta   = 0.99,
  chains        = 4,
  iter          = 3000,
  warmup        = 1500,
  silent        = 2
)
```

If instead you want each region to carry its own shape
parameter, express that through the brms distributional
formula. `mvgam_formula()` accepts a `brmsformula` on the
observation side, so `bf(adj_count ~ 0, shape ~ series - 1)`
would give one shape per region, at the cost of tying up more
parameters on a modest dataset.

```{r fit-load, include = FALSE}
varmod <- read_cache("kestrel_fit")$varmod
train  <- read_cache("kestrel_fit")$train
test   <- read_cache("kestrel_fit")$test
# The latent processes are ordered by the levels of the series
# factor, which is alphabetical and not the order the regions were
# assembled in. Everything downstream labels itself from this.
region_order <- levels(train$series)
n_reg <- length(region_order)
```

## Model summary

`summary()` on a VAR fit splits the parameter table between the
observation model (here a single shared shape) and the trend
model (the region intercepts, the VAR coefficient array
`A_raw_trend`, and the innovation Cholesky `L_Omega_trend`). We
suppress the spline coefficient table with `include_betas =
FALSE` because there are no smooth terms on the observation
side.

```{r summary}
summary(varmod, include_betas = FALSE)
```

The three region intercepts sit close to zero on the
inverse link scale, and the VAR self-coefficients on the diagonal
of $A$ (rows `A_raw_trend[1,i,i]`) are all positive and bounded
well below one, so the Heaps
([2022](https://doi.org/10.1080/10618600.2022.2079648)) prior
that guarantees stationarity is not pushing against the data.

## Inspecting the transition matrix

The lagged dependence structure lives in the $3 \times 3$
matrix $A$. `mcmc_plot()` with `variable = "A_trend"` and
`type = "hist"` shows the posterior for every cell. Passing
`facet_args = list(dir = "v")` walks the facets column-first, so
row $i$, column $j$ (the effect of process $j$'s state at time
$t-1$ on process $i$ at time $t$) sits where you would draw it
by hand.

```{r plot-A, fig.cap = "Posterior histograms for each cell of the VAR transition matrix A. Diagonals capture self-lag; off-diagonals capture lagged cross-region effects."}
mcmc_plot(
  varmod,
  variable    = "A_trend",
  regex       = TRUE,
  type        = "hist",
  facet_args  = list(dir = "v")
)
```

```{r a-numbers, include = FALSE}
A_post <- as.matrix(varmod, variable = "^A_trend\\[", regex = TRUE)
a_cell <- function(i, j) A_post[, sprintf("A_trend[1,%d,%d]", i, j)]
a_diag_min <- min(vapply(seq_len(n_reg),
                         function(i) mean(a_cell(i, i) > 0), numeric(1)))
off <- expand.grid(i = seq_len(n_reg), j = seq_len(n_reg))
off <- off[off$i != off$j, ]
off$p_pos <- apply(off, 1L, function(r) mean(a_cell(r[["i"]], r[["j"]]) > 0))
best <- off[which.max(off$p_pos), ]
fe_a_diag <- sprintf("%.2f", a_diag_min)
fe_a_best <- sprintf("%s on %s", region_order[best$j], region_order[best$i])
fe_a_best_p <- sprintf("%.2f", best$p_pos)
```

The diagonal carries the clearest signal. Each region's own state
at $t-1$ raises its state at $t$, with at least
`r fe_a_diag` of the posterior above zero for all three, so a
good year tends to be followed by another. The off-diagonals are
weaker and none of them separates from zero. The strongest is the
effect of `r fe_a_best`, at `r fe_a_best_p` of the posterior
above zero, which is suggestive and no more. Note the index
order: `A_trend[1,i,j]` is the effect of process $j$ on process
$i$, and the processes follow the levels of the series factor, so
`region_order` above is what names them.

## Inspecting the innovation covariance

The innovation covariance $\Sigma$ says which regions take the
same shock in the same year, the common weather or
detection-effort excursion that nudges several surveys at once.
`Sigma_trend` is saved directly, so `cov2cor()` on each draw puts
it on a correlation scale, labelled by `region_order` rather than
by hand.

```{r sigma-cor}
Sig_post <- as.matrix(varmod, variable = "^Sigma_trend\\[", regex = TRUE)
cor_draws <- t(apply(Sig_post, 1, function(v) {
  as.numeric(cov2cor(matrix(v, n_reg, n_reg)))
}))
med_cor <- matrix(apply(cor_draws, 2, median), n_reg, n_reg)
dimnames(med_cor) <- list(region_order, region_order)
round(med_cor, 2)
```

```{r sigma-numbers, include = FALSE}
inn <- function(a, b) {
  i <- match(a, region_order)
  j <- match(b, region_order)
  z <- cor_draws[, (j - 1) * n_reg + i]
  sprintf("%.2f (%.2f to %.2f)", median(z),
          quantile(z, 0.05), quantile(z, 0.95))
}
fe_alb_bc <- inn("Alb", "BC")
fe_alb_sask <- inn("Alb", "Sask")
fe_bc_sask <- inn("BC", "Sask")
```

Read those against the map before reading anything into them.
British Columbia and Saskatchewan sit at either end of the
transect with Alberta between them, so adjacency would predict
the two neighbouring pairs correlating and the far pair not.
That is not the pattern. Alberta and British Columbia sit at
`r fe_alb_bc`, the two non-adjacent provinces at `r fe_bc_sask`,
and the other adjacent pair, Alberta and Saskatchewan, at
`r fe_alb_sask`. Every one of those intervals covers zero, so the
honest summary is that 35 years of annual counts do not resolve
the contemporaneous correlations at all, and geography is not
waiting inside them to be found.

The correlation between the *states*, rather than between the
shocks, is a different question and a better answered one.
`residual_cor()` reports it, reading the stationary covariance
the model derives from $A$ and $\Sigma$ together, so it describes
where the latent processes settle rather than how far a single
year moves them:

```{r resid-cor}
rc <- residual_cor(varmod)
round(rc$cor, 2)
```

```{r resid-cor-numbers, include = FALSE}
sta <- function(a, b) {
  i <- match(a, rownames(rc$cor))
  j <- match(b, colnames(rc$cor))
  sprintf("%.2f (%.2f to %.2f)", rc$cor[i, j],
          rc$cor_lower[i, j], rc$cor_upper[i, j])
}
fe_s_alb_bc <- sta("Alb", "BC")
fe_s_bc_sask <- sta("BC", "Sask")
```

These are larger and two of the three exclude zero: Alberta with
British Columbia at `r fe_s_alb_bc`, and British Columbia with
Saskatchewan at `r fe_s_bc_sask`. Accumulating shocks through
$A$ year after year is what builds the shared variation; any one
year's shocks barely share anything. Which of the two matrices
answers your question is worth settling before reading either.
Neither restores the geography.

## Forecasts## Forecasts

Passing `newdata = test` at fit time stored the held-out slice
on the fitted object, so `forecast()` extends the VAR five years
into the future without any extra arguments. The plot draws the
hindcast (grey) up to the dashed vertical, then the forecast
posterior (red) over the five held-out years.

```{r plot-forecast, fig.cap = "Hindcast (grey) and five-year posterior forecast (red) from the correlated-innovation VAR. The mean count trajectory stays close to the historical average, with the ribbon widening as the horizon lengthens."}
plot(forecast(varmod, newdata = test))
```

Forecasts fan out gradually as the VAR is integrated forward.
Because the diagonal of $A$ is safely below one, the mean
converges to the region intercept while the ribbon spreads
according to the stationary covariance implied by
$(A, \Sigma)$.

## Impulse response functions

An impulse response function asks the following counterfactual:
if a shock of one standard deviation hit region $j$ at time
zero, how would the other regions be expected to move over the
next $h$ years? (Lütkepohl
[2005](https://doi.org/10.1007/978-3-540-27752-1)). `irf()`
computes the generalized IRF for every source and target pair;
`plot(irfs, series = j)` fixes the shock at region $j$ and draws
one panel per target region.

```{r irf-fit}
irfs <- irf(varmod, h = 12)
```

`plot()` takes the shocked process by position, so `series = 1`
is the first level of `region_order` rather than the first region
named when the data were assembled.

```{r irf-numbers, include = FALSE}
ir <- as.data.frame(irfs)
ir$src <- region_order[as.integer(sub(".*Process_(\\d+) ->.*", "\\1",
                                       ir$shock))]
ir$tgt <- region_order[as.integer(sub(".*-> Process_(\\d+)", "\\1",
                                       ir$shock))]
self_pk <- max(abs(ir$irfQ50[ir$src == ir$tgt]))
cross_pk <- max(abs(ir$irfQ50[ir$src != ir$tgt]))
cross <- ir[ir$src != ir$tgt, ]
cross_resolved <- sum(sign(cross$irfQ2.5) == sign(cross$irfQ97.5))
fe_self_pk <- sprintf("%.2f", self_pk)
fe_cross_pk <- sprintf("%.2f", cross_pk)
```

```{r plot-irf1, fig.cap = "Generalized impulse responses to a positive shock in region 1 (British Columbia). Each panel shows how the process for the target region is expected to respond over the next 12 years."}
plot(irfs, series = 1)
```

The self-response is the one that shows. A shock to
`r region_order[1]` lifts its own process by
`r fe_self_pk` on the inverse-link scale in the first year and
decays from there; the responses in the other two regions peak in
year two at a fraction of that.

```{r plot-irf2, fig.cap = "Impulse responses to a shock in region 2 (Alberta). The self-response decays smoothly; British Columbia responds in the same direction as Alberta, and Saskatchewan shows a delayed positive response consistent with the adjacent-region correlation seen in Sigma."}
plot(irfs, series = 2)
```

The same shape repeats for a shock to `r region_order[2]`: a
first-year self-response, then small positive movements
elsewhere that peak a year later.

```{r plot-irf3, fig.cap = "Impulse responses to a shock in region 3 (Saskatchewan). The self-response is the sharpest of the three, and the effect on Alberta persists longer than the effect on British Columbia."}
plot(irfs, series = 3)
```

Across all three panels the largest cross-region response is
`r fe_cross_pk`, roughly a third of the largest self-response,
and `r cross_resolved` of the cross-region responses have a
credible interval excluding zero at any horizon. The medians
trace a tidy story of shocks spreading between provinces; the
intervals do not support telling it. What the figures establish
is that each region responds to its own shocks and that any
spillover is too small for this series to pin down.

## Forecast error variance decomposition

The forecast error variance decomposition (FEVD) turns the same
quantities around. Rather than tracing one shock outward, it asks
how much of the forecast uncertainty for region $i$ at horizon
$h$ was generated by shocks originating in each region. `fevd()`
reports the posterior median share and an interval for each
source, keeping the draws behind `summary = FALSE`;
`plot(fevds)` renders one stacked-bar panel per target region,
the bar heights showing the median share attributed to each
source over increasing horizons.

```{r fevd-fit}
fevds <- fevd(varmod, h = 12)
```

```{r plot-fevd, fig.cap = "Forecast error variance decomposition. Each panel is a target region; each bar breaks down that region's forecast uncertainty at a given horizon into contributions from own shocks and shocks originating in the other two regions."}
plot(fevds)
```

```{r fevd-numbers, include = FALSE}
fv <- as.data.frame(fevds)
fv$src <- region_order[as.integer(sub(".*Process_(\\d+) ->.*", "\\1",
                                       fv$shock))]
fv$tgt <- region_order[as.integer(sub(".*-> Process_(\\d+)", "\\1",
                                       fv$shock))]
f12 <- fv[fv$horizon == 12, ]
sh <- function(a, b) sprintf("%.0f%%", 100 *
  f12$fevdQ50[f12$src == a & f12$tgt == b])
fe_ab <- sh("Alb", "BC")
fe_as <- sh("Alb", "Sask")
fe_own_min <- sprintf("%.0f%%", 100 * min(
  f12$fevdQ50[f12$src == f12$tgt]))
```

Own-region shocks dominate at short horizons for every region,
the darker slice at the bottom of each bar, which is what the
strong diagonal in $A$ implies. The cross-region shares grow with
the horizon without ever taking over: by year 12 Alberta accounts
for `r fe_ab` of the forecast variance in British Columbia and
`r fe_as` of it in Saskatchewan, while every region still owns at
least `r fe_own_min` of its own. Read this beside the impulse
responses rather than instead of them. A variance share is a
point estimate of a ratio and carries no statement about whether
the underlying effect is distinguishable from zero, and the
intervals above say these ones are not.

## Community stability metrics

The transition matrix and innovation covariance also determine
whether the system as a whole would amplify or absorb small
shocks around the stationary distribution. `stability()`
reports three interpretable dynamic-stability metrics, each
summarised to a median and an interval with the draws available
under `summary = FALSE` (Ives et al.
[2003](https://doi.org/10.1890/0012-9615(2003)073[0301:ECSAEI]2.0.CO;2);
Neubert & Caswell
[1997](https://doi.org/10.1890/0012-9658(1997)078[0653:ATRFMT]2.0.CO;2)):

- Reactivity, the log of the largest singular value of $A$.
  Positive reactivity means small perturbations grow immediately
  after the shock before eventually decaying; negative
  reactivity means the perturbation shrinks straight away.
- Mean return rate, the modulus of the largest eigenvalue
  of $A$. Values closer to one imply a slower return of the
  mean to its stationary distribution after a shock.
- Variance return rate, the analogous quantity for $A
  \otimes A$, describing how quickly forecast variance decays
  back to its stationary level.

`plot()` on the returned `mvgam_stability` object renders each
metric as a posterior histogram with a dashed reference at zero
(the reactivity threshold).

```{r plot-stability, fig.width = 8, fig.asp = 0.45, fig.cap = "Posterior distributions of three community-stability metrics for the fitted VAR. The dashed line marks zero, the reactivity threshold above which small shocks would be amplified rather than absorbed."}
stab <- stability(varmod)
plot(stab)
```

```{r stability-numbers, include = FALSE}
st_tab <- as.data.frame(stab)
st_val <- function(m) st_tab$Estimate[st_tab$metric == m]
st_draws <- stability(varmod, summary = FALSE)
react <- if (is.list(st_draws)) st_draws$reactivity else
  st_draws[, "reactivity"]
fe_react <- sprintf("%.2f", stats::median(react))
fe_react_lo <- sprintf("%.2f", stats::quantile(react, 0.05))
fe_react_hi <- sprintf("%.2f", stats::quantile(react, 0.95))
fe_react_pos <- sprintf("%.0f%%", 100 * mean(react > 0))
fe_mrr <- sprintf("%.2f", st_val("mean_return_rate"))
fe_vrr <- sprintf("%.2f", st_val("var_return_rate"))
```

Reactivity is centred near zero, with a median of `r fe_react`
and a 90% interval from `r fe_react_lo` to `r fe_react_hi`, and
`r fe_react_pos` of the posterior sitting above the dashed line.
The data do not settle whether this system amplifies a small
perturbation before absorbing it, though they do bound how much
it could amplify by. Mean and variance return rates of
`r fe_mrr` and `r fe_vrr` describe a stable system that takes
several years to relax back to its stationary distribution.

## Take-aways

Once a VAR is fitted in mvgam, the same handful of accessors
work end to end for any observation family and any correlated
innovation structure:

- `mcmc_plot(mod, variable = "A_trend", ...)` shows the lagged
  dependence structure per cell of the transition matrix.
- `residual_cor(mod)` turns the innovation covariance into
  correlations with intervals and evidence, telling you which
  series share contemporaneous shocks.
- `irf()` + `plot(irfs, series = j)` traces how a shock at a
  focal series is expected to move each other series over a
  chosen horizon.
- `fevd()` + `plot(fevds)` decomposes the forecast variance for
  each series into contributions from each source of shocks.
- `stability()` + `plot(stab)` summarises the dynamic-stability
  posterior in the units ecologists use (reactivity, return
  rates).

Every one of these views draws on the same posterior for
$(A, \Sigma)$, differing only in which question they put to that
joint object. Each returns a summary by default and the draws
under `summary = FALSE`, so a figure costs a call rather than a
refit.

## Further reading

Complementary in-package tools:

- `?VAR` for the trend-side constructor. Hierarchical panel VAR
  support is available via the `gr` and `subgr` arguments.
- `?forecast.mvgam`, `?score` and `?ensemble` for the forecast
  scoring and averaging surface layered on top of any fitted
  mvgam VAR.
- `?how_to_cite` and `?methods_md` for citation-ready methods
  paragraphs that describe the Heaps stationary prior and the
  sampler configuration in place.

Sibling articles in this series:

- The
  [forecast evaluation article](https://nicholasjclark.github.io/mvgam/articles/forecast_evaluation.html)
  runs the same forecast / score machinery on a longer Portal
  rodent time series.
- The
  [hierarchical VAR article](https://nicholasjclark.github.io/mvgam/articles/hierarchical_var.html)
  scales the same model to a panel of units, giving each its own
  $(A, \Sigma)$ and shrinking both toward a global mean.
- The
  [multi-response article](https://nicholasjclark.github.io/mvgam/articles/mvbf.html)
  covers the different case where several outcomes on different
  families share one latent process via a multivariate
  observation formula.
- The
  [dynamic factor vignette](https://nicholasjclark.github.io/mvgam/articles/dfm.html)
  covers the case where the number of latent processes is
  deliberately smaller than the number of observed series.

## Acknowledgements

The kestrel dataset was compiled from the North American
Breeding Bird Survey and is redistributed with MARSS (Holmes et
al. [2012](https://doi.org/10.32614/RJ-2012-002)). The stationary
Heaps prior on the VAR coefficients follows Heaps
([2022](https://doi.org/10.1080/10618600.2022.2079648)). The
state-space framing follows the review by Auger-Méthé and
colleagues ([2021](https://doi.org/10.1002/ecm.1470)) and the
dynamic generalised additive model perspective in Clark & Wells
([2023](https://doi.org/10.1111/2041-210X.13974)). Community
stability metrics follow Ives et al.
([2003](https://doi.org/10.1890/0012-9615(2003)073[0301:ECSAEI]2.0.CO;2))
and Neubert & Caswell
([1997](https://doi.org/10.1890/0012-9658(1997)078[0653:ATRFMT]2.0.CO;2)).
Impulse response function and forecast error variance
decomposition definitions follow Lütkepohl
([2005](https://doi.org/10.1007/978-3-540-27752-1)). Formulas
rely on mgcv smooths (Wood
[2017](https://doi.org/10.1201/9781315370279)) and brms's
multilevel apparatus (Bürkner
[2017](https://doi.org/10.18637/jss.v080.i01)); sampling is
performed with Stan (Carpenter et al.
[2017](https://doi.org/10.18637/jss.v076.i01)).

## References

Auger-Méthé M, Newman K, Cole D, Empacher F, Gryba R, King AA,
Leos-Barajas V, Mills Flemming J, Nielsen A, Petris G and Thomas
L (2021). A guide to state-space modeling of ecological time
series. *Ecological Monographs* 91(4), e01470.
[doi:10.1002/ecm.1470](https://doi.org/10.1002/ecm.1470)

Bürkner PC (2017). brms: An R package for Bayesian multilevel
models using Stan. *Journal of Statistical Software* 80(1),
1-28.
[doi:10.18637/jss.v080.i01](https://doi.org/10.18637/jss.v080.i01)

Carpenter B, Gelman A, Hoffman MD, Lee D, Goodrich B, Betancourt
M, Brubaker M, Guo J, Li P and Riddell A (2017). Stan: A
probabilistic programming language. *Journal of Statistical
Software* 76(1), 1-32.
[doi:10.18637/jss.v076.i01](https://doi.org/10.18637/jss.v076.i01)

Clark NJ and Wells K (2023). Dynamic Generalized Additive Models
(DGAMs) for forecasting discrete ecological time series.
*Methods in Ecology and Evolution* 14, 771-784.
[doi:10.1111/2041-210X.13974](https://doi.org/10.1111/2041-210X.13974)

Heaps SE (2023). Enforcing stationarity through the prior in
vector autoregressions. *Journal of Computational and Graphical
Statistics* 32(1), 74-83.
[doi:10.1080/10618600.2022.2079648](https://doi.org/10.1080/10618600.2022.2079648)

Holmes EE, Ward EJ and Wills K (2012). MARSS: Multivariate
Autoregressive State-space Models for Analyzing Time-series
Data. *The R Journal* 4(1), 11-19.
[doi:10.32614/RJ-2012-002](https://doi.org/10.32614/RJ-2012-002)

Ives AR, Dennis B, Cottingham KL and Carpenter SR (2003).
Estimating community stability and ecological interactions from
time-series data. *Ecological Monographs* 73(2), 301-330.
[doi:10.1890/0012-9615(2003)073[0301:ECSAEI]2.0.CO;2](https://doi.org/10.1890/0012-9615(2003)073[0301:ECSAEI]2.0.CO;2)

Lütkepohl H (2005). *New Introduction to Multiple Time Series
Analysis*. Springer, Berlin.
[doi:10.1007/978-3-540-27752-1](https://doi.org/10.1007/978-3-540-27752-1)

Neubert MG and Caswell H (1997). Alternatives to resilience for
measuring the responses of ecological systems to perturbations.
*Ecology* 78(3), 653-665.
[doi:10.1890/0012-9658(1997)078[0653:ATRFMT]2.0.CO;2](https://doi.org/10.1890/0012-9658(1997)078[0653:ATRFMT]2.0.CO;2)

Wood SN (2017). *Generalized Additive Models: An Introduction
with R* (2nd edition). Chapman and Hall/CRC.
[doi:10.1201/9781315370279](https://doi.org/10.1201/9781315370279)
